### 2020-07-24 generated/tables/table_time.tex
restructure

\begin{table}

\caption{\label{tab:}Average runtime in hours, for evaluating one grid and one optimization run of MBO.}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{linear} & \multicolumn{3}{c}{paper} & \multicolumn{3}{c}{paper2} & \multicolumn{3}{c}{sigmoid} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
 & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000\\
\midrule
grid & 54.8 & 60.9 & 60.1 & 62.7 & 62.9 & 66.2 & 59.3 & 64 & 66.6 & 54.3 & 52.8 & 60.5\\
mbo & 3.7 & 3.9 & 3.7 & 4.8 & 4.5 & 4.3 & 3.9 & 4 & 3.7 & 3.4 & 3.1 & 3.9\\
\bottomrule
\end{tabular}
\end{table}

### 2020-07-24 main.tex
restructure

\documentclass[bimj,fleqn]{w-art}
\usepackage{times}
\usepackage{w-thm}
\usepackage[authoryear]{natbib}
\setlength{\bibsep}{2pt}
\setlength{\bibhang}{2em}
\newcommand{\J}{J\"{o}reskog}
\newcommand{\So}{S\"{o}rbom}
\newcommand{\bcx}{{\bf X}}
\newcommand{\bcy}{{\bf Y}}
\newcommand{\bcz}{{\bf Z}}
\newcommand{\bcu}{{\bf U}}
\newcommand{\bcv}{{\bf V}}
\newcommand{\bcw}{{\bf W}}
\newcommand{\bci}{{\bf I}}
\newcommand{\bch}{{\bf H}}
\newcommand{\bcb}{{\bf B}}
\newcommand{\bcr}{{\bf R}}
\newcommand{\bcm}{{\bf M}}
\newcommand{\bcf}{{\bf F}}
\newcommand{\bcg}{{\bf G}}
\newcommand{\bcs}{{\bf S}}
\newcommand{\bca}{{\bf A}}
\newcommand{\bcd}{{\bf D}}
\newcommand{\bcc}{{\bf C}}
\newcommand{\bce}{{\bf E}}
\newcommand{\ba}{{\bf a}}
\newcommand{\bb}{{\bf b}}
\newcommand{\bc}{{\bf c}}
\newcommand{\bd}{{\bf d}}
\newcommand{\bx}{{\bf x}}
\newcommand{\by}{{\bf y}}
\newcommand{\bz}{{\bf z}}
\newcommand{\bu}{{\bf u}}
\newcommand{\bv}{{\bf v}}
\newcommand{\bh}{{\bf h}}
\newcommand{\bl}{{\bf l}}
\newcommand{\be}{{\bf e}}
\newcommand{\br}{{\bf r}}
\newcommand{\bw}{{\bf w}}
\newcommand{\de}{\stackrel{D}{=}}
\newcommand{\bt}{\bigtriangleup}
\newcommand{\bfequiv}{\mbox{\boldmath $\equiv$}}
\newcommand{\bmu}{\mbox{\boldmath $\mu$}}
\newcommand{\bnu}{\mbox{\boldmath $\nu$}}
\newcommand{\bxi}{\mbox{\boldmath $\xi$}}
\newcommand{\btau}{\mbox{\boldmath $\tau$}}
\newcommand{\bgamma}{\mbox{\boldmath $\Gamma$}}
\newcommand{\bphi}{\mbox{\boldmath $\Phi$}}
\newcommand{\bfphi}{\mbox{\boldmath $\varphi$}}
\newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
\newcommand{\bpi}{\mbox{\boldmath $\Pi$}}
\newcommand{\bequiv}{\mbox{\boldmath $\equiv$}}
\newcommand{\bvarepsilon}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\btriangle}{\mbox{\boldmath $\triangle$}}
\newcommand{\bdelta}{\mbox{\boldmath $\Delta$}}
\newcommand{\beps}{\mbox{\boldmath $\epsilon$}}
\newcommand{\btheta}{\mbox{\boldmath $\theta$}}
\newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
\newcommand{\bsphi}{\mbox{\boldmath $\varphi$}}
\newcommand{\bsig}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfpsi}{\mbox{\boldmath $\psi$}}
\newcommand{\bfdelta}{\mbox{\boldmath $\delta$}}
\newcommand{\bsigma}{{\bf \Sigma}}
\newcommand{\bzero}{{\bf 0}}
\newcommand{\bpsi}{\mbox{\boldmath $\Psi$}}
\newcommand{\bep}{\mbox{\boldmath $\epsilon$}}
\newcommand{\bomega}{\mbox{\boldmath $\Omega$}}
\newcommand{\bfomega}{\mbox{\boldmath $\omega$}}
\newcommand{\blambda}{\mbox{\boldmath $\Lambda$}}
\newcommand{\bflambda}{\mbox{\boldmath $\lambda$}}
\newcommand{\bfsigma}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfpi}{{\mbox{\boldmath $\pi$}}}
\newcommand{\bupsilon}{\mbox{\boldmath $\upsilon$}}
\newcommand{\obs}{{\rm obs}}
\newcommand{\mis}{{\rm mis}}
\theoremstyle{plain}
\newtheorem{criterion}{Criterion}
\theoremstyle{definition}
\newtheorem{condition}[theorem]{Condition}
\usepackage[]{graphicx}
\chardef\bslash=`\\ % p. 424, TeXbook
\newcommand{\ntt}{\normalfont\ttfamily}
\newcommand{\cn}[1]{{\protect\ntt\bslash#1}}
\newcommand{\pkg}[1]{{\protect\ntt#1}}
\let\fn\pkg
\let\env\pkg
\let\opt\pkg
\hfuzz1pc % Don't bother to report overfull boxes if overage is < 1pc
\newcommand{\envert}[1]{\left\lvert#1\right\rvert}
\let\abs=\envert

% own packages

% kableExtra
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2020}
\pagespan{1}{}
\keywords{Model-Based Optimization; Design of Experiments;\\ %FIXME: name 5 %Up to five keywords are allowed and should be given in alphabetical order. Please capitalize the key
\noindent \hspace*{-4pc}\\
\hspace*{-4pc} {\small\it words)}\\[1pc]
\noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not
applicable)
}  %%% semicolon and fullpoint added here for keyword style

\title[Model-based Optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian Optimization} %(please use sentence case, e.g. Score tests for exploring complex models)
%% Information for the first author.
\author[Jakob Richter {\it{et al.}}]{Jakob Richter\footnote{Corresponding author: {\sf{e-mail: richter@statistik.tu-dortmund.de}}}\inst{,1}} 
\address[\inst{1}]{Fakultät Statistik, Technische Universität Dortmund, 44221 Dortmund}
%%%%    Information for the second author
\author[dd]{Tim Friede\inst{2}}
\address[\inst{2}]{Second address} %please include the department and the postal address
%%%%    Information for the third author
\author[]{Jörg Rahnenführer\inst{1}} %please provide full author names. Middle names should be indicated by initials only, i.e. Henry J. James
%%%%    \dedicatory{This is a dedicatory.}
\Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}

% max 250 words
\begin{abstract}
Planning the patient group sizes and test methids for clinical trials in such a way that the test power is as high as possible for given effect sizes under the null hypothesis and for a fixed number of treatments is a crucial task.
From a wide set of possible set ups we want to chose the most promising plan in short time to drive fast decisions.
While it is possible to simulate the power for all various set ups, the space of possible set ups soon becomes too big.
Either you need big computational resources or you have to wait for an exhausive exploration of the possible set ups.
In this paper we show that you can successfully apply Model-based Optimization to find a clinical trial set-up that yields an high power from a wide set of possible set-ups.
We optimize the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhausitve search to show that our optimization approach finds a promising set-up in a fraction of the time.
\end{abstract}

%% maketitle must follow the abstract.
\maketitle                   % Produces the title.

%% If there is not enough space inside the running head
%% for all authors including the title you may provide
%% the leftmark in one of the following three forms:

%% \renewcommand{\leftmark}
%% {First Author: A Short Title}

%% \renewcommand{\leftmark}
%% {First Author and Second Author: A Short Title}

%% \renewcommand{\leftmark}
%% {First Author et al.: A Short Title}

%% \tableofcontents  % Produces the table of contents.

\section{Introduction}

Clinical research and in particular drug development is typically structured into phases, e.g. phases I to IV in drug development; similar approaches for e.g. complex interventions (MRC framework). 
The delepment phases include elements of learning and comfirmation. 
For instance, Sheiner (1997) described the drug development process as two learning and confirming cycles. 
First learning / confirming cycle (Phase I-IIa): Learning about tolerated dose (Phase I). 
Then confirming of efficacy of selected dose in selected group of patients (Phase IIa). 
Second learning / confirming cycle (Phase IIb-III): Learning about optimal use in respresentative patients (Phase IIb); Then confirming of acceptable benefit / risk ratio (Phase III). 
Traditionally, separate studies for learning and confirming.

Changing landscape in clinical research: Dissolving boundaries between development phases; Master protocols: Broader developments focusing not only on one single treatment or one single indication / disease; Adaptive designs: Study designs become more flexible; More principled approach to decision making using statistical modelling and integrated analyses across studies

More complex designs and analyses of clinical trials. 
Examples: adaptive design; evidence synthesis.

With increasing complexity of both designs and analyses of  clinical trials the planning of such studies often relies on computer simulations, since no closed form solutions even for usually fairly straightforward tasks as sample size calculations are available. 
Frameworks and guidance have been developed over the past years on how to planning, execution and interpretation of such simulation studies. 
Method comparison studies (references), references on simulation studies, in particular Clinical Scenario Evaluation (CSE) \\

Depending on the setting and the purpose of the simulation study suitable metrics are chosen to compare alternative design or analysis strategies. 
While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios. 
Here we propose to use a formal approach to optimization. As the opitimization will typically be on multiple dimensions which makes the exercise computationally demanding, we demonstrate here the used of the so-called Model-based Optimization (MBO) approach (reference).

The remainder of the manuscript is structured as follows. \dots

%\subsection{Related work}
%subsub Clinical 
% TODO: Tim

%subsub MBO
% TODO: Jakob

\section{Motivating Example}
% TODO: Jacob 
COPD trial
% Friede et al (2020) Biom J, Section 5.1 Clinical trial in COPD with treatment selection; other references: Barnes et al. (2011) and Donohue et al. (2010); Cuffe, Lawrence, Stone,&Vandemeulebroecke, 2014.

\section{Adaptive Seamless Designs}
% TODO: Tim
Intro 3- 4 Sätze

\subsection{Two-stage design with treatment selection}
Multi-arm trial; two-stages; treatment selection in interim analysis; final analysis based on combination test approach with closed test principle

% first stage = interrim analysis
% second stage = final analysys
% stick to first second stage througout text but mention interrim here?

\subsection{Selection rules}
\paragraph{Epsilion Rule}

\paragraph{Threshold Rule}

\subsection{Clinical scenario evaluation for ASD}
Friede et al (2010) DIJ, Figure 2

\subsection{Simulating ASD}
R package asd by Nick Parsons
Test statistics are simulated, not individual participant data

\section{Model-based Optimization}

Sequential \textbf{Model-based Optimization (MBO)}~\citep{jones_taxonomy_2001} (also known as Bayesian optimization) is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simmulation with certain parameters, here denoted by $\theta$) is very time consuming, as it becomes infeasible to evaluate the black-box for thousands of configurations.
MBO solves the optimization problem within a bounded search space $\Theta$:
\[
\theta^\ast := \operatorname{argmin}_{\theta \in \Theta} f(\theta),
\]
where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
To reduce the number of evaluations on $f$ the key idea of MBO is to only evaluate values of $\theta$ that are expected to lead to a small value of $f(\theta)$.
The estimate $\hat{f}(\theta)$ is generated by a so called \emph{surrogate model}.
Typically, this is a regression model that predicts the outcome of $f$ based on previous evaluations of $f$.
First, an initial design of already evaluated configurations is needed.
Then, iteratively, the MBO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
The steps are repeated until a budget is exhausted.

A so called infill criterion guides the proposal of new configurations $\theta$ based on $\hat{f}$.
It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation, i.e.\ the search in regions that promise best outcomes.
We use the augmented expected improvement~\citep{huang_global_2006} which is well suited for noisy functions.
For non-noisy optimization we would choose the configuration $\theta^\ast$ that has led to the best outcome of $f$ to be returned as the optimization result.
If the function outcome is noisy the best observed outcome is likely overly optimistic and not located at the true posterior mean.
Therefore, we employ the surrogate to estimate the posterior mean for each evaluated configuration to cancel out the noise.
The configuration for which the surrogate estimates the best outcome is then returned as the optimization result $\theta^\ast$.
%To obtain a final outcome $\nu^\ast = f(\theta^\ast)$ that is independent of the optimization process, $f(\theta^\ast)$ is calculated again.
Using the stochastic outcome $\nu^\ast = f(\theta^\ast)$ observed during the optimization process would still potentially lead to an overly optimistic result
Therefore an independent calculation of $f(\theta^\ast)$ should be conducted to obtain a fair estimation of the reached best outcome.

We apply Kriging (also called Gaussian process regression) to fit the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$.
We use the implementation in the R-package \emph{DiceKriging}~\citep{roustant_dicekriging_2012} and configure it to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
We use one-hot encoding to handle categorical parameters.
The hierarchical structure of the search space introduces inactive values.
If a value is inactive it is set to a value two times as high as the maximum in its active range.

\section{Simulation Study}
% TODO : Jakob

The goal of this simulation study is to show that Model-based optimization can effectively find the trail design that yields the highest statistical power for the given \emph{number of total treatments} $n_{\text{treat}}$, \emph{effect sizes} for all arms and both stages and the \emph{correlation between early and final outcomes}.
The effect sizes we consider in this study are given in Table~\ref{tab:table_effect_names}.
\input{generated/tables/table_effect_names.tex}
As this table shows, we restrict ourselves to studys with five arms.
The space of possible trial design setups we consider is given in Table~\ref{tab:search_space}.
\begin{table}[h]
  \caption{Search Space of possible trial design setups.}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter          & Range \\
  \hline
  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thr} \}$ \\
  Ratio ($r$)        & $(0,1)$ \\
  Epsilion           & $[0,4]$ \\
  Threshold          & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}
The different selection strategies are explained in Table~\ref{tab:selection_strategies}.
\begin{table}[h]
  \caption{How are the arms for the second stage selected?}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Selection strategy &  \\
  \hline
  $all$              & Select all arms  \\
  $\{1,2,3\}$        & Select 1, 2 or 3 arms with the maximal test statistic \\
  eps                & Select all arms that are within \emph{Epsilon} of the maximal test statistic \\
  thr                & Select all arms which have a test statistic above the \emph{Threshold} \\
  \hline
  \end{tabular}
\end{table}
This table also explains the meaning of the \emph{Epsilon} and the \emph{Threshold} parameter which are only active for the respective selection strategy.
The \emph{Ratio} ($r$) parameter defines how many treatments are conducted in the first and in the second stage respectively.

As mentioned before, one restriction is that the \emph{number of total treatments} ($n_{\text{treat}}$) should be constant.
The number depends on the number of arms including the control group in stage 1 ($k_1$) and stage 2 ($k_2$) is calculated as follows:
\begin{equation}
  \label{eq:ntreat}
  n_{\text{treat}} = (k_1) \cdot n_{\text{stage1}} + (k_2) \cdot n_{\text{stage2}},
\end{equation}
whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ denote the number of treatments per arm in the first and second stage.
In our scenario $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$.
Within one stage, all arms obtain the same number of treatments.

To keep the number of total treatments constant we optimize the ratio ($r \in (0,1)$) of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ instead of optimizing these values directly.
Therefore we define
\begin{align}
  \label{eq:stagec}
  \begin{split}
  n_{\text{stage1}} &= r \cdot c \\
  n_{\text{stage2}} &= (1-r) \cdot c \ ,
  \end{split}
\end{align}
whereas $c$ is a constant that depends on $n_{\text{treat}}$, $k_1$ and $k_2$.
For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
For a general solution, we can insert the variables from (\ref{eq:stagec}) into (\ref{eq:ntreat}) and obtain
\begin{align}
  \label{eq:ntreatc}
  n_{\text{treat}} &= k_1 \cdot r \cdot c + k_2 \cdot (1-r) \cdot c \Leftrightarrow \\
  c &= \frac{n_{\text{treat}}}{k_1 \cdot r + k_2 \cdot (1-r)}.
\end{align}
Placing $c$ into (\ref{eq:stagec}) lets us calculate $n_{\text{stage1}}$ and $n_{\text{stage2}}$ for the given constraints and a given ratio $r$.

This formula gives us the treatments per arm and per stage if $k_2$ is known. 
Unfortunately, this is only the case for \emph{selection strategy} $\in \{\text{all},1,2,3\}$.
Hence, it is necessary to run a calibration step for the selection strategies \emph{thr} and \emph{eps} to adjust the number of $n_{\text{stage1}}$ so that it matches the desired ratio $r$.
Therefore, another simulation is run that returns the expected number of arms at stage 2 $k_2$ for given values of $n_{\text{stage1}}$ and  \emph{thr} or \emph{eps} respectively.
TODO: Explain in more detail.

%FIXME Frage: Warum reicht es nicht wenn der Suchraum select=0,1,2,3,4,5 ist?

The outcome of the simulation and therefore our optimization objective is stochastic.
Also our search space is hierarchical because \emph{Epsilon} and \emph{Threshold} are only active for the respective selection strategy.

Challenges
  \begin{itemize}
    \item Power is determined stochastically through simulation
    \item Number of treatments in stage 2 depends on \emph{selection strategy} ($k_2$): \\
    $n_{\text{treat}} = (k_1+1) \cdot n_{\text{stage1}} + (k_2+1) \cdot n_{\text{stage2}}$ \\
    Remember: $n_{\text{treat}}$, $k_1$ fixed.
    \begin{itemize}
      \item $k_2 = 1,2,3$ fixed if we \emph{select} 1/2/3 best arms.
      \item $k_2$ only as expected value through \underline{simulation} for \emph{epsilon} and \emph{threshold} strategy.
    \end{itemize}
    Optimize \emph{ratio} $r$ instead of $n_{\text{stage1}}$ and $n_{\text{stage2}}$.
    \begin{align}
      c &= \frac{n_{\text{treat}}}{(k_1 + 1) \cdot r + (k_2+1) \cdot (1-r)} \\
      \Rightarrow n_{\text{stage1}} &= r \cdot c \ \land\\
       n_{\text{stage2}} &= (1-r) \cdot c 
    \end{align}
    \item Problem: \underline{Simulation} depends on $n_{\text{stage1}}$ \\
      Solution: Adapt $n_{\text{stage1}}$ so that $\hat{n}_{\text{treat}}$ is close to given $n_{\text{treat}}$ and $r$.
  \end{itemize}

\begin{itemize}
  \item Discretize search space and evaluate grid (possible with enough computational resources)
  \item Apply MBO and compare results
\end{itemize}

\subsection{Design}

The underlying algorithms in this study are implemented in R, for the model-based optimization the R-package mlrMBO~\citep{bischl_mlrmbo_2017} is used, and the simulation of the clinical trials is performed using the R-package asd~\citep{parsons_software_2011}.

\subsection{Results}

To obtain an unbiased result, we do not simply report the best observed power of both optimization approaches.
Instead we use the best found parameters to run the simulation again independently, repeated times, similar to a cross-validation in machine learning applications.
This eliminates the risk of only reporting outcomes that are optimal by chance.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{generated/figures/plot_allbest.pdf}
\caption{Power of the different selection rules. For the epsilon and threshold rule only the curve for the epsilon and threshold value that reached the best power is dispayed.}
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
\caption{Best found $x$-values from MBO runs.}
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
\caption{%
  Optimization curves of each optimization run and the mean of all runs drawn as a black line.
  }
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Validated performance of both approaches.
  }
\end{center}
\end{figure}

\input{generated/tables/table_best.tex}

\input{generated/tables/table_time.tex}

\subsection{Second level heading}

This is the body text. Please note that cross-references in the body text should be shown as follows:
(Miller, 1900), (Miller and Baker, 1900) or if three or more authors (Miller {\it{et al}}., 1900)
\vspace*{12pt}

\noindent Bullet lists are not allowed. Always use (i), (ii), etc.
\vspace*{12pt}

\noindent Sentences should never start with a symbol.
\vspace*{12pt}

\noindent Names of software packages and website addresses should be written in {\tt{Courier new, i.e. Stata, the R package
MASS, http://www.biometrical-journal.com.}}

\section{Discussion}

% subsection Summary

% subsection Outlook

% \begin{table}[htb]
% \begin{center}
% \caption{The caption of a table.}
% \begin{tabular}{lll}
% \hline
% Description 1 & Description 2 & Description 3\\
% \hline
% Row 1, Col 1 & Row 1, Col 2 & Row 1, Col 3\\
% Row 2, Col 1 & Row 2, Col 2 & Row 2, Col 3\\
% \hline
% \end{tabular}
% \end{center}
% \end{table}
% \begin{equation}
% \left({\theta^{0}_{i}}\atop{\theta^{1}_{i}}\right) \sim N(\theta,\Sigma),\quad {\mathrm{with}}\ 
% {{\theta}} = \left({\theta_{0}}\atop{\theta_{1}}\right)\ {\mathrm{and}}\ \Sigma =
% \left(\begin{array}{cc}
% \sigma^{2}_{0} & \rho\sigma_{0}\sigma_{1}\\
% \rho\sigma_{0}\sigma_{2} & \sigma^{2}_{1}
% \end{array}\right).
% \end{equation}

\noindent This is the body text. Only number equations which are referred to in the text body. If equations
are numbered, these should be numbered continuously throughout the text. Not section wise! Please
carefully follow the rules for mathematical expressions in the ``Instructions to Authors''.

\begin{acknowledgement}
This work was partly supported by Deutsche Forschungsgemeinschaft (DFG) within the Collaborative Research Center SFB 876, A3.
\end{acknowledgement}
\vspace*{1pc}

\noindent {\bf{Conflict of Interest}}

\noindent {\it{The authors have declared no conflict of interest.}} %(or please state any conflicts of interest)

%\section*{Appendix {\it(please insert here, if applicable)}}

%\subsection*{A.1.\enspace Second level heading}

%Please insert appendices before the references.

% \begin{thebibliography}{10}
% \bibitem[Bauer and Bauer(1994)Bauer, P. and Bauer, M.M.]{bib1}Bauer, P. and  Bauer, M. M. (1994). Testing equivalence simultaneously for location and  dispersion of two normally distributed populations.  \textit{Biometrical  Journal} \textbf{36}, 643--660.
% \bibitem[Farrington, C. P. and Andrews, N. (2003)]{bib2}Farrington, C.P. and Andrews, N. (2003). Outbreak detection:
% Application to infectious disease surveillance. In: Monitoring the Health of Populations (eds. R. Brookmeyer and D. F. Stroup), Oxford University Press, Oxford,\break 203--231.
% \bibitem[Rencher(1998)Rencher, A.C.]{bib3}Rencher, A. C. (1998).  \textit{Multivariate Statistical Inference and Applications}. Wiley, New  York. 
% \end{thebibliography}

\bibliographystyle{biometrical}
\bibliography{../literature.bib}

\newpage
\phantom{aaaa}
\end{document}

### 2020-07-24 benchmark/batchtools_grid.R

@@ -1,0 +1,285 @@
+library(batchtools)
+library(smoof)
+library(asd.mod)
+library(data.table)
+library(mlr3misc)
+
+TESTMODE = Sys.info()[["nodename"]] %in% c("JayBook2")
+
+if (TESTMODE) {
+  #try out params
+  NSIM = 50
+  MBOITERS = 10
+  GRIDRES = 4
+  REPLS = 1
+  NCASES = c(1000,2000)
+  EFFECTS = list(
+    paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20))
+  )
+} else {
+  #real params
+  NSIM = 1000
+  MBOITERS = 100
+  GRIDRES = 25
+  REPLS = 10
+  NCASES = c(500,1000,2000)
+  EFFECTS = list(
+    paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20)),
+    linear = {x = c(0,2,4,6,8)/10; list(early = x, final = x/4)},
+    #threashold = {x = c(0,0,0,4,8)/10; list(early = x, final = x/4)},
+    #saturation = {x = c(0,5,7,7.8,8)/10; list(early = x, final = x/4)},
+    sigmoid = {x = c(0,1,2,7,8)/10; list(early = x, final = x/4)},
+    #paper_mod = list(early = c(0,0.68,0.82,0.93,0.93), final = c(0,0.13,0.17,0.23,0.20)),
+    paper2 = list(early = c(0,0.68,0.82,0.95,0.91), final = 2*c(0,0.13,0.17,0.23,0.20))# ,
+    #paper_mod2 = list(early = c(0,0.68,0.82,0.93,0.93), final = 2*c(0,0.13,0.17,0.23,0.20))
+  )
+}
+
+## Define Problem
+
+fn = function(
+  x, 
+  n_cases = 1000, 
+  effect = list(early=c(0,0.68,0.82,0.95,0.91),final=c(0,0.13,0.17,0.23,0.20)), 
+  corr = 0.4, 
+  nsim = 100) {
+
+  if (is.character(effect)) {
+    effect = dictionaries$effects[[effect]]
+  }
+  k_plus_1 = length(effect$early)
+  p = x$stage_ratio
+
+  calculate_n = function(n_select) { # calculate n for each arm in stage1 and stage2 given the number of arrms in stage2, and the ratio p
[…]

### 2020-07-24 generator/main.R

@@ -1,0 +1,309 @@
+## ----setup, include=FALSE----------------------------------------------------- 
+library(data.table)
+library(ggplot2)
+library(scales)
+library(stringi)
+library(knitr)
+library(kableExtra)
+library(dplyr)
+library(mlr3misc)
+
+plot_wrapper = function(name, fig.height, fig.width, expr) {
+  path = paste0("../generated/figures/", name, ".pdf")
+  pdf(file = path, width = fig.width, height = fig.height, onefile = TRUE)
+  print(eval(expr))
+  dev.off()
+  #browser()
+  knitr::plot_crop(path)
+}
+
+kable_to_text = function(text, name) {
+  writeLines(text, con = paste0("../generated/tables/", name, ".tex"))
+}
+
+own_theme = function(...) {
+  theme_light(...) + 
+  theme(legend.margin=margin(0,0,0,0))
+}
+theme_set(own_theme())
+
+
+
+FIG_WIDTH = 8
+FIG_HEIGHT = 6
+FNT_SMALL = 6
+
+
+options(
+  knitr.table.format = "latex",
+  knitr.kable.NA = '', 
+  knitr.kable.NaN = ''
+)
+
+
+## ----helpers------------------------------------------------------------------
+# paper parameters
+EFFECTS = list(
+  paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20)),
+  linear = {x = c(0,2,4,6,8)/10; list(early = x, final = x/4)}, #yes
+  # threshold = {x = c(0,0,0,4,8)/10; list(early = x, final = x/4)},
+  # saturation = {x = c(0,5,7,7.8,8)/10; list(early = x, final = x/4)},
+  sigmoid = {x = c(0,1,2,7,8)/10; list(early = x, final = x/4)}, #yes
+  # paper_mod = list(early = c(0,0.68,0.82,0.93,0.93), final = c(0,0.13,0.17,0.23,0.20)), 
+  paper2 = list(early = c(0,0.68,0.82,0.95,0.91), final = 2*c(0,0.13,0.17,0.23,0.20)) #yes
+  # paper_mod2 = list(early = c(0,0.68,0.82,0.93,0.93), final = 2*c(0,0.13,0.17,0.23,0.20))
+)
+effect_names = setNames(names(EFFECTS), names(EFFECTS)) #to be populated with better names
+
+
+## ----table_effect_names-------------------------------------------------------
+tmp = lapply(EFFECTS, do.call, what = rbind)
[…]

### 2020-07-24 generator/main.R
fix labels

@@ -1,3 +1,5 @@
+set.seed(1) #actually nothing stochastic happens here
+
 ## ----setup, include=FALSE----------------------------------------------------- 
 library(data.table)
 library(ggplot2)
@@ -18,6 +20,7 @@ plot_wrapper = function(name, fig.height, fig.width, expr) {
 }
 
 kable_to_text = function(text, name) {
+  text = stri_replace_all_fixed(text, pattern = "\\label{tab:}", sprintf("\\label{tab:%s}", name))
   writeLines(text, con = paste0("../generated/tables/", name, ".tex"))
 }
 

### 2020-07-24 main.tex
Updates from Overleaf

@@
 \begin{abstract}
-Planning the patient group sizes and test methids for clinical trials in such a way that the test power is as high as possible for given effect sizes under the null hypothesis and for a fixed number of treatments is a crucial task.
+Planning the patient group sizes and test methods for clinical trials in such a way that the test power is as high as possible for given effect sizes under the null hypothesis and for a fixed number of treatments is a crucial task.
 From a wide set of possible set ups we want to chose the most promising plan in short time to drive fast decisions.
 While it is possible to simulate the power for all various set ups, the space of possible set ups soon becomes too big.
-Either you need big computational resources or you have to wait for an exhausive exploration of the possible set ups.
+Either you need big computational resources or you have to wait for an exhaustive exploration of the possible set ups.
 In this paper we show that you can successfully apply Model-based Optimization to find a clinical trial set-up that yields an high power from a wide set of possible set-ups.
-We optimize the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhausitve search to show that our optimization approach finds a promising set-up in a fraction of the time.
+We optimize the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhaustive search to show that our optimization approach finds a promising set-up in a fraction of the time.
 \end{abstract}
@@
 Clinical research and in particular drug development is typically structured into phases, e.g. phases I to IV in drug development; similar approaches for e.g. complex interventions (MRC framework). 
-The delepment phases include elements of learning and comfirmation. 
-For instance, Sheiner (1997) described the drug development process as two learning and confirming cycles. 
+The delepment phases include elements of learning and confirmation. 
+For instance, \cite{sheiner_learning_1997} described the drug development process as two learning and confirming cycles. 
 First learning / confirming cycle (Phase I-IIa): Learning about tolerated dose (Phase I). 
@@
 COPD trial
-% Friede et al (2020) Biom J, Section 5.1 Clinical trial in COPD with treatment selection; other references: Barnes et al.
-(2011) and Donohue et al.
-(2010); Cuffe, Lawrence, Stone,&Vandemeulebroecke, 2014.
+% Friede et al (2020) Biom J, Section 5.1 Clinical trial in COPD with treatment selection;
+\cite{friede_adaptive_2020}
+
+%other references: Barnes et al. (2011) and Donohue et al. (2010); 
+\cite{cuffe_when_2014}
 
@@
 
-% \begin{thebibliography}{10}
-% \bibitem[Bauer and Bauer(1994)Bauer, P.
-and Bauer, M.M.]{bib1}Bauer, P.
-and  Bauer, M.
-M.
-(1994).
-Testing equivalence simultaneously for location and  dispersion of two normally distributed populations.
-\textit{Biometrical  Journal} \textbf{36}, 643--660.
-% \bibitem[Farrington, C. P. and Andrews, N. (2003)]{bib2}Farrington, C.P. and Andrews, N. (2003). Outbreak detection:
-% Application to infectious disease surveillance. In: Monitoring the Health of Populations (eds. R. Brookmeyer and D. F. Stroup), Oxford University Press, Oxford,\break 203--231.
-% \bibitem[Rencher(1998)Rencher, A.C.]{bib3}Rencher, A. C. (1998).  \textit{Multivariate Statistical Inference and Applications}. Wiley, New  York. 
-% \end{thebibliography}
-
 \bibliographystyle{biometrical}
-\bibliography{../literature.bib}
+\bibliography{literature.bib}
 

### 2020-07-24 main.tex

@@
 
+Assurance: \cite{stallard_optimal_2009}
+
 \section{Adaptive Seamless Designs}
@@
 
-Challenges
-  \begin{itemize}
-    \item Power is determined stochastically through simulation
-    \item Number of treatments in stage 2 depends on \emph{selection strategy} ($k_2$): \\
-    $n_{\text{treat}} = (k_1+1) \cdot n_{\text{stage1}} + (k_2+1) \cdot n_{\text{stage2}}$ \\
-    Remember: $n_{\text{treat}}$, $k_1$ fixed.
-    \begin{itemize}
-      \item $k_2 = 1,2,3$ fixed if we \emph{select} 1/2/3 best arms.
-      \item $k_2$ only as expected value through \underline{simulation} for \emph{epsilon} and \emph{threshold} strategy.
-    \end{itemize}
-    Optimize \emph{ratio} $r$ instead of $n_{\text{stage1}}$ and $n_{\text{stage2}}$.
-    \begin{align}
-      c &= \frac{n_{\text{treat}}}{(k_1 + 1) \cdot r + (k_2+1) \cdot (1-r)} \\
-      \Rightarrow n_{\text{stage1}} &= r \cdot c \ \land\\
-       n_{\text{stage2}} &= (1-r) \cdot c 
-    \end{align}
-    \item Problem: \underline{Simulation} depends on $n_{\text{stage1}}$ \\
-      Solution: Adapt $n_{\text{stage1}}$ so that $\hat{n}_{\text{treat}}$ is close to given $n_{\text{treat}}$ and $r$.
-  \end{itemize}
-
 \begin{itemize}
@@
 \bibliographystyle{biometrical}
-\bibliography{literature.bib}
+\bibliography{literature,literature_zotero}
 

### 2020-08-03 benchmark/batchtools_grid.R
add nsim 5000

@@ -212,6 +212,10 @@ addAlgorithm("mbo", fun = function(job, data, instance, corr, nsim, n_cases, mbo
 })
 
 const_design = expand.grid(n_cases = NCASES, nsim = NSIM, corr = 0.4, effect = names(EFFECTS), stringsAsFactors = FALSE)
+# add 5times sim steps for one experiment
+new_line = subset(const_design, n_cases == 2000 & effect == "paper")
+new_line$nsim = 5 * new_line$nsim
+const_design = rbind(const_design, new_line)
 eval_design = generateGridDesign(getParamSet(fun), GRIDRES)
 eval_design = subset(eval_design, stage_ratio != 1 & stage_ratio != 0)
 eval_design = merge.data.frame(eval_design, const_design)

### 2020-08-04 main.tex

@@
 \author[dd]{Tim Friede\inst{2}}
-\address[\inst{2}]{Second address} %please include the department and the postal address
+\address[\inst{2}]{Institut für Medizinische Statistik, Universitätsmedizin Göttingen, 37073 Göttingen}
 %%%%    Information for the third author
@@
 Either you need big computational resources or you have to wait for an exhaustive exploration of the possible set ups.
-In this paper we show that you can successfully apply Model-based Optimization to find a clinical trial set-up that yields an high power from a wide set of possible set-ups.
-We optimize the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhaustive search to show that our optimization approach finds a promising set-up in a fraction of the time.
+In this paper we show that you can successfully apply Model-based Optimization to find a clinical trial set-up that yields a high power from a wide set of possible set-ups.
+We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhaustive search to show that our optimization approach finds a promising set up in a fraction of the time.
 \end{abstract}
@@
 Clinical research and in particular drug development is typically structured into phases, e.g. phases I to IV in drug development; similar approaches for e.g. complex interventions (MRC framework). 
-The delepment phases include elements of learning and confirmation. 
+The development phases include elements of learning and confirmation. 
 For instance, \cite{sheiner_learning_1997} described the drug development process as two learning and confirming cycles. 
@@
 Then confirming of efficacy of selected dose in selected group of patients (Phase IIa). 
-Second learning / confirming cycle (Phase IIb-III): Learning about optimal use in respresentative patients (Phase IIb); Then confirming of acceptable benefit / risk ratio (Phase III). 
+Second learning / confirming cycle (Phase IIb-III): Learning about optimal use in representative patients (Phase IIb); Then confirming of acceptable benefit / risk ratio (Phase III). 
 Traditionally, separate studies for learning and confirming.
 
-Changing landscape in clinical research: Dissolving boundaries between development phases; Master protocols: Broader developments focusing not only on one single treatment or one single indication / disease; Adaptive designs: Study designs become more flexible; More principled approach to decision making using statistical modelling and integrated analyses across studies
+Changing landscape in clinical research: Dissolving boundaries between development phases; Master protocols: Broader developments focusing not only on one single treatment or one single indication / disease; Adaptive designs: Study designs become more flexible; More principled approach to decision making using statistical modeling and integrated analyses across studies
 
@@
 Frameworks and guidance have been developed over the past years on how to planning, execution and interpretation of such simulation studies. 
-Method comparison studies (references), references on simulation studies, in particular Clinical Scenario Evaluation (CSE) \\
+Method comparison studies (references), references on simulation studies, in particular Clinical Scenario Evaluation (CSE).
 
@@
 Here we propose to use a formal approach to optimization. 
-As the opitimization will typically be on multiple dimensions which makes the exercise computationally demanding, we demonstrate here the used of the so-called Model-based Optimization (MBO) approach (reference).
+The set of possible scenarios can be described as a multi dimensional search space, including real-valued and categorical parameters.
+This vast space of possible set ups makes the optimization computationally demanding.
+We demonstrate here, that the use of the so-called Model-based Optimization (MBO) approach~\cite{jones_taxonomy_2001} can be applied successfully to maximize the statistical power of a clinical trial design for given effect sizes and number of treatments.
 
-The remainder of the manuscript is structured as follows. \dots
+The remainder of the manuscript is structured as follows:
 
@@
 
-%subsub MBO
-% TODO: Jakob
+Model-based optimization, or often used interchangeably Bayesian optimization has been successfully applied to optimize expensive black-boxes in many scenarios.
+Popular applications include hyper-parameter tuning of machine learning methods~\cite{snoek_practical_2012}, general algorithm configuration~\cite{hutter_sequential_2011}, and many more.
+Various implementations of Model-based optimization methods are used widely, such as SMAC~\cite{hutter_sequential_2011}, Spearmint~\cite{snoek_practical_2012} and BoTorch~\cite{balandat_botorch_2020}.
+For this work we use the implementation in the R-package mlrMBO~\cite{bischl_mlrmbo_2017}.
+This implementations has been successfully applied to optimize hyper-parameters of machine learning methods on various tasks~\cite{bischl_mlrmbo_2017, wozniak_candle_2018}, and to optimize model-weights~\cite{richter_modelbased_2019,browaeys_nichenet_2020}.
 
 \section{Motivating Example}
-% TODO: Jacob 
+% TODO: Jakob 
 COPD trial
@@
 
-%other references: Barnes et al. (2011) and Donohue et al. (2010); 
+\cite{barnes_integrating_2010}
+\cite{donohue_oncedaily_2010}
 \cite{cuffe_when_2014}
@@
 Unfortunately, this is only the case for \emph{selection strategy} $\in \{\text{all},1,2,3\}$.
-Hence, it is necessary to run a calibration step for the selection strategies \emph{thr} and \emph{eps} to adjust the number of $n_{\text{stage1}}$ so that it matches the desired ratio $r$.
-Therefore, another simulation is run that returns the expected number of arms at stage 2 $k_2$ for given values of $n_{\text{stage1}}$ and  \emph{thr} or \emph{eps} respectively.
-TODO: Explain in more detail.
+For the selection strategies \emph{thr} and \emph{eps}, the number of arms in the second stage $k_2$ depends on the choice of the \emph{Threshold} or \emph{Epsilon} parameter.
+However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 $\hat{k}_2$, by running a simulation of the first stage.
+This simulation is run multiple times for different values of $n_{\text{stage1}}$ in a calibration step until the desired number of $n_{\text{treat}}$ is reached approximately.
+In fact, this calibration step formulates an independent optimization problem:
+We minimize 
+\begin{align}
+  \label{eq:targettreat}
+  n_{\text{stage1}}^\ast &= \operatorname{argmin}_{n_{\text{stage1}} \in [l,u]} h(n_{\text{stage1}}) \,  \text{, with} \\
+  h(n_{\text{stage1}}) &:= \left( (k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot \underbrace{\frac{1-r}{r} \cdot n_{\text{stage1}}}_{=n_{\text{stage2}}}) - n_{\text{treat}} \right)^2 
+\end{align}
 

### 2020-08-07 benchmark/batchtools_grid.R
nsim fix

@@ -119,7 +119,7 @@ fn = function(
         mean_select = getOptPathEl(res$opt.path, getOptPathBestIndex(res$opt.path))$extra$mean_select)
       return(n)
     }
-    calib_res = calibrate_n(x)
+      = calibrate_n(x)
     select2 = calib_res$mean_select
   }
   n = calculate_n(select2)

### 2020-08-07 generator/main.R

@@ -63,7 +63,7 @@ effect_names = setNames(names(EFFECTS), names(EFFECTS)) #to be populated with be
 tmp = lapply(EFFECTS, do.call, what = rbind)
 tmp = lapply(tmp, function(x) cbind.data.frame(stage = rownames(x), x))
 tmp = rbindlist(tmp, idcol = TRUE)
-setnames(tmp, ".id", "setup")
+setnames(tmp, ".id", "effect set")
 kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation") %>% 
   collapse_rows(columns = 1:2, latex_hline = "major") %>% 
   pack_rows("Set 1", 2, ncol(tmp), indent = FALSE) %>% 
@@ -160,7 +160,7 @@ res_mbo[, time.running := time.running - as.difftime(sum(.SD$opt.path[[1]][dob =
 ## ----plot_allbest, fig.height = 1.5 * FIG_HEIGHT------------------------------
 # find best performing solution for each select methods and across all epsilon and threshold choices
 plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
-  res_ave = res_eval[, list(mean_y = mean(y)), by = c(algo.par.names)]
+  res_ave = res_eval[nsim == 1000, list(mean_y = mean(y)), by = c(algo.par.names)]
   res_best = res_ave[, .SD[order(-mean_y)[1], ], by = c("select", algo.par.names.meta)]
   res_best = res_best[, !c("mean_y", "stage_ratio")]
   # merge so the plot only contains the best curves (applies for epsilon and threshold rule)
@@ -169,7 +169,7 @@ plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
   g = ggplot(df, aes(x = (stage_1_arms * stage_1_n), y = y, color = select, group = paste(select, epsilon, thresh)))
   g = g + geom_line(data = dfmean)
   g = g + geom_point(alpha = 0.1)
-  g = g + geom_point(data = res_mbo, size = 3)
+  g = g + geom_point(data = res_mbo[nsim == 1000,], size = 3)
   g = g + scale_color_manual(values = select_labels_colors)
   g = g + facet_grid(effect~n_cases, scales = "free", labeller = label_both)
   g = g + theme(legend.position = "bottom")
[…]

### 2020-08-14 main.tex

@@
 Planning the patient group sizes and test methods for clinical trials in such a way that the test power is as high as possible for given effect sizes under the null hypothesis and for a fixed number of treatments is a crucial task.
-From a wide set of possible set ups we want to chose the most promising plan in short time to drive fast decisions.
-While it is possible to simulate the power for all various set ups, the space of possible set ups soon becomes too big.
-Either you need big computational resources or you have to wait for an exhaustive exploration of the possible set ups.
+From a wide set of possible setups we want to chose the most promising plan in short time to drive fast decisions.
+While it is possible to simulate the power for all various setups, the space of possible setups soon becomes too big.
+Either you need big computational resources or you have to wait for an exhaustive exploration of the possible setups.
 In this paper we show that you can successfully apply Model-based Optimization to find a clinical trial set-up that yields a high power from a wide set of possible set-ups.
-We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhaustive search to show that our optimization approach finds a promising set up in a fraction of the time.
+We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhaustive search to show that our optimization approach finds a promising setup in a fraction of the time.
 \end{abstract}
@@
 The development phases include elements of learning and confirmation. 
-For instance, \cite{sheiner_learning_1997} described the drug development process as two learning and confirming cycles. 
+For instance, \citet{sheiner_learning_1997} described the drug development process as two learning and confirming cycles. 
 First learning / confirming cycle (Phase I-IIa): Learning about tolerated dose (Phase I). 
@@
 The set of possible scenarios can be described as a multi dimensional search space, including real-valued and categorical parameters.
-This vast space of possible set ups makes the optimization computationally demanding.
-We demonstrate here, that the use of the so-called Model-based Optimization (MBO) approach~\cite{jones_taxonomy_2001} can be applied successfully to maximize the statistical power of a clinical trial design for given effect sizes and number of treatments.
+This vast space of possible setups makes the optimization computationally demanding.
+We demonstrate here, that the use of the so-called Model-based Optimization (MBO) approach~\citep{jones_taxonomy_2001} can be applied successfully to maximize the statistical power of a clinical trial design for given effect sizes and number of treatments.
 
@@
 Model-based optimization, or often used interchangeably Bayesian optimization has been successfully applied to optimize expensive black-boxes in many scenarios.
-Popular applications include hyper-parameter tuning of machine learning methods~\cite{snoek_practical_2012}, general algorithm configuration~\cite{hutter_sequential_2011}, and many more.
-Various implementations of Model-based optimization methods are used widely, such as SMAC~\cite{hutter_sequential_2011}, Spearmint~\cite{snoek_practical_2012} and BoTorch~\cite{balandat_botorch_2020}.
-For this work we use the implementation in the R-package mlrMBO~\cite{bischl_mlrmbo_2017}.
-This implementations has been successfully applied to optimize hyper-parameters of machine learning methods on various tasks~\cite{bischl_mlrmbo_2017, wozniak_candle_2018}, and to optimize model-weights~\cite{richter_modelbased_2019,browaeys_nichenet_2020}.
+Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
+Various implementations of Model-based optimization methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012} and BoTorch~\citep{balandat_botorch_2020}.
+For this work we use the implementation in the R-package mlrMBO~\citep{bischl_mlrmbo_2017}.
+This implementations has been successfully applied to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
 
@@
 Sequential \textbf{Model-based Optimization (MBO)}~\citep{jones_taxonomy_2001} (also known as Bayesian optimization) is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
-In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simmulation with certain parameters, here denoted by $\theta$) is very time consuming, as it becomes infeasible to evaluate the black-box for thousands of configurations.
+In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming, and it becomes infeasible to evaluate the black-box for thousands of configurations.
 MBO solves the optimization problem within a bounded search space $\Theta$:
 \[
-\theta^\ast := \operatorname{argmin}_{\theta \in \Theta} f(\theta),
+\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta),
 \]
 where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
-To reduce the number of evaluations on $f$ the key idea of MBO is to only evaluate values of $\theta$ that are expected to lead to a small value of $f(\theta)$.
-The estimate $\hat{f}(\theta)$ is generated by a so called \emph{surrogate model}.
+To reduce the number of evaluations on $f$ the key idea of MBO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
+The estimate is generated by a so called \emph{surrogate model}.
 Typically, this is a regression model that predicts the outcome of $f$ based on previous evaluations of $f$.
@@
 
-A so called infill criterion guides the proposal of new configurations $\theta$ based on $\hat{f}$.
-It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation, i.e.\ the search in regions that promise best outcomes.
-We use the augmented expected improvement~\citep{huang_global_2006} which is well suited for noisy functions.
-For non-noisy optimization we would choose the configuration $\theta^\ast$ that has led to the best outcome of $f$ to be returned as the optimization result.
-If the function outcome is noisy the best observed outcome is likely overly optimistic and not located at the true posterior mean.
-Therefore, we employ the surrogate to estimate the posterior mean for each evaluated configuration to cancel out the noise.
-The configuration for which the surrogate estimates the best outcome is then returned as the optimization result $\theta^\ast$.
+The proposal of a new configuration $\theta$ is obtained by maximizing a so called acquisition function.
+The acquisition function guides the search to promising new regions in the search space $\Theta$ by facilitating the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty prediction $\hat{s}^2(\theta)$ of the surrogate.
+It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation.
+The acquisition function achieves exploration by giving a high value to areas where the surrogate predicts a high uncertainty.
+Exploitation is achieved by giving high values to values of $\theta$ where the surrogate predicts a high outcome of $f$.
+For deterministic functions the expected improvement~\citep{jones_efficient_1998} is arguably the most popular acquisition function.
+
+\paragraph{Acquisition function for noisy outcomes}
+Since this work focuses on optimization of a stochastic simulation, we assume that our non-deterministic optimization problem can be formulated as follows:
+\begin{equation}
+\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon \, \text{,with}  \varepsilon \sim N(0, \sigma^2_{n}) \text{iid.}
+\end{equation}
+In this setting the expected improvement is not feasible as shown in \citet{huang_global_2006}.
+They propose to use the augmented expected improvement instead.
+Therefore, in a first step, we calculate the \emph{effective best solution}:
+\begin{equation}
+  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot \hat{s}(\theta),
+\end{equation}
+with $c$ as a tuning parameter that is usually set to 1 and $\mathcal{D}$ as the design that contains all previously evaluated values of $\theta$.
+The effective best point is the pessimistic estimate of the best observed outcome so far.
+In the final step, we calculate the augmented expected improvement~\citep{huang_global_2006} as follows:
+\begin{multline}
+  \label{eq:aei}
+  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{\hat{s}(\theta)} \biggr) \\
+   + \hat{s}(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{\hat{s}(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma_{n}}{\sqrt{\sigma_{n}^2 + \hat{s}^2(\theta)}}}_{\text{correction}}\biggr),
+\end{multline}
+with $\Phi$ and $\phi$ as the distribution and density function of the standard normal distribution, and
+with $\sigma_{n}^2$ denoting the random error (nugget effect) in the Kriging model.
+The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
+This assumption is also reflected by using a Gaussian process regression as a surrogate.
+However, the intuition in this formula is that the first summand favors exploitation while the second summand favors exploration.
+The correction term is necessary to avoid exploration in areas where the model uncertainty ($\hat{s}^2(\theta)$) equals the random error ($\sigma_{n}^2$), because in this case further evaluations will not decrease the model uncertainty.
+
+\paragraph{Choosing the best point from noisy outcomes}
+If we chose the configuration ($\theta$) from all previously evaluated configurations that led to the best outcome of $f$ as the optimization result, we would be overly optimistic.
+It is likely that a single best outcome can be attributed to the random error ($\sigma_{n}^2$).
+Instead we are interested in the optimum of the true posterior mean.
+Therefore, we employ the surrogates estimate of the posterior mean for each evaluated configuration to cancel out the noise.
+The configuration for which the surrogate estimates the best outcome is then returned as the optimization result $\hat{\theta}^\ast$:
+\begin{equation}
+  \hat{\theta}^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta),
+\end{equation}
 %To obtain a final outcome $\nu^\ast = f(\theta^\ast)$ that is independent of the optimization process, $f(\theta^\ast)$ is calculated again.
-Using the stochastic outcome $\nu^\ast = f(\theta^\ast)$ observed during the optimization process would still potentially lead to an overly optimistic result
-Therefore an independent calculation of $f(\theta^\ast)$ should be conducted to obtain a fair estimation of the reached best outcome.
-
-We apply Kriging (also called Gaussian process regression) to fit the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$.
-We use the implementation in the R-package \emph{DiceKriging}~\citep{roustant_dicekriging_2012} and configure it to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
-We use one-hot encoding to handle categorical parameters.
-The hierarchical structure of the search space introduces inactive values.
-If a value is inactive it is set to a value two times as high as the maximum in its active range.
+Using the stochastic outcome $\hat{\nu}^\ast = f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
+Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimation of the reached best outcome.
 
@@
 
-The goal of this simulation study is to show that Model-based optimization can effectively find the trail design that yields the highest statistical power for the given \emph{number of total treatments} $n_{\text{treat}}$, \emph{effect sizes} for all arms and both stages and the \emph{correlation between early and final outcomes}.
+The goal of this simulation study is to show that Model-based optimization can effectively find the trial design setup that yields the highest statistical power for the given \emph{number of total treatments} ($n_{\text{treat}}$), \emph{effect sizes} for all arms and both stages.
+
 The effect sizes we consider in this study are given in Table~\ref{tab:table_effect_names}.
 \input{generated/tables/table_effect_names.tex}
-As this table shows, we restrict ourselves to studys with five arms.
+As this table shows, we restrict ourselves to studies with five arms.
+
 The space of possible trial design setups we consider is given in Table~\ref{tab:search_space}.
@@
   Ratio ($r$)        & $(0,1)$ \\
-  Epsilion           & $[0,4]$ \\
+  Epsilon            & $[0,4]$ \\
   Threshold          & $[0,10]$ \\
@@
   \hline
-  $all$              & Select all arms  \\
+  \emph{all}     & Select all arms  \\
   $\{1,2,3\}$    & Select 1, 2 or 3 arms with the maximal test statistic \\
-  eps                & Select all arms that are within \emph{Epsilon} of the maximal test statistic \\
-  thr                & Select all arms which have a test statistic above the \emph{Threshold} \\
+  \emph{eps}     & Select all arms that are within \emph{Epsilon} of the maximal test statistic \\
+  \emph{thr}     & Select all arms which have a test statistic above the \emph{Threshold} \\
   \hline
@@
 
+\paragraph{Definition of the simulation}
+The simulation and calculation of the statistical power of the trial design can be formulated as a black-box as follows:
+\begin{align}
+  \label{eq:bbox}
+  (y,k_2)' &= f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold}).
+\end{align}
+We notice, that we obtain two measures of interest: The simulated statistical power ($y$) and the number of arms in the second stage ($k_2$).
+Our main optimization goal is to maximize $y$.
+$k_2$ is used in a later step.
+
 As mentioned before, one restriction is that the \emph{number of total treatments} ($n_{\text{treat}}$) should be constant.
-The number depends on the number of arms including the control group in stage 1 ($k_1$) and stage 2 ($k_2$) is calculated as follows:
+The number depends on the number of arms including the control group in stage 1 ($k_1$) and stage 2 ($k_2$) and is calculated as follows:
 \begin{equation}
   \label{eq:ntreat}
-  n_{\text{treat}} = (k_1) \cdot n_{\text{stage1}} + (k_2) \cdot n_{\text{stage2}},
+  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}},
 \end{equation}
@@
   n_{\text{treat}} &= k_1 \cdot r \cdot c + k_2 \cdot (1-r) \cdot c \Leftrightarrow \\
+  \label{eq:ntreatcc}
   c &= \frac{n_{\text{treat}}}{k_1 \cdot r + k_2 \cdot (1-r)}.
@@
 Placing $c$ into (\ref{eq:stagec}) lets us calculate $n_{\text{stage1}}$ and $n_{\text{stage2}}$ for the given constraints and a given ratio $r$.
-
 This formula gives us the treatments per arm and per stage if $k_2$ is known. 
 Unfortunately, this is only the case for \emph{selection strategy} $\in \{\text{all},1,2,3\}$.
-For the selection strategies \emph{thr} and \emph{eps}, the number of arms in the second stage $k_2$ depends on the choice of the \emph{Threshold} or \emph{Epsilon} parameter.
-However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 $\hat{k}_2$, by running a simulation of the first stage.
+
+For the selection strategies \emph{thr} and \emph{eps}, the number of arms in the second stage ($k_2$) depends on the choice of the \emph{Threshold} or \emph{Epsilon} parameter.
+However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 ($\hat{k}_2$), by running a simulation of the first stage.
 This simulation is run multiple times for different values of $n_{\text{stage1}}$ in a calibration step until the desired number of $n_{\text{treat}}$ is reached approximately.
@@
   \label{eq:targettreat}
-  n_{\text{stage1}}^\ast &= \operatorname{argmin}_{n_{\text{stage1}} \in [l,u]} h(n_{\text{stage1}}) \,  \text{, with} \\
-  h(n_{\text{stage1}}) &:= \left( (k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot \underbrace{\frac{1-r}{r} \cdot n_{\text{stage1}}}_{=n_{\text{stage2}}}) - n_{\text{treat}} \right)^2 
+  \begin{split}
+  & \operatorname{min}_{n_{\text{stage1}} \in [l,u]} h(n_{\text{stage1}}) \,  \text{, with} \\
+  h(n_{\text{stage1}}) &:= \biggl( (k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot \underbrace{\frac{1-r}{r} \cdot n_{\text{stage1}}}_{=n_{\text{stage2}}}) - n_{\text{treat}} \biggr)^2 \, \text{and} \\
+  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
+  \end{split}
 \end{align}
+and $l = \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil$ and $u = \lceil \frac{n_{\text{treat}}}{k_1} \rceil$. 
+$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
+This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
+The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
+
+With the calibration step that is necessary for selection strategy $\in \{\text{thr}, \text{eps} \}$ we can always set $n_{\text{stage1}}$ and $n_{\text{stage2}}$ according to $n_{\text{treat}}$, $r$.
+This allows us to reparametrize the black-box from equation~\ref{eq:bbox} to directly depend from the parameters of interest:
+\begin{equation}
+  y = \tilde{f}_{n_{\text{treat}}}(r, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold}) \, ,
+\end{equation}
+whereas $n_{\text{treat}}$ is considered to be set constant.
+This constraint is necessary because otherwise bigger number of treatments almost certainly imply a higher statistical power of the study design.
+
+In the following we try to find the clinical design setup, represented by the parameter settings $r$, \emph{effect set}, \emph{Selection strategy}, \emph{Epsilon} and \emph{Thresold} that yield the highest statistical power.
+The statistical power is determined by evaluating the simulator $\tilde{f}$ as explained before.
+
+\paragraph{General simulation settings}
+For the simulation we call the \texttt{treatsel.sim} function from the \texttt{asd}~package~\cite{parsons_software_2011} with the following parameters as fixed:
+number of simulation iterations (\texttt{nsim}~$= 1000$), correlation between early and final outcomes  (\texttt{corr}~$= 0.4$), test level (\texttt{level}~$=0.025$), and (TODO Not sure what that means) (\texttt{ptest}~$=(3,4)$).
 
 %FIXME Frage: Warum reicht es nicht wenn der Suchraum select=0,1,2,3,4,5 ist?
+\paragraph{Exhaustive Search}
+For each effect set (see Table~\ref{tab:table_effect_names}) we conduct an exhaustive grid search with a resolution of 25 points per dimension.
+This results in total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
+Each conf
+Additionally
+Each numeric parameter ($r$, \emph{Epsilon}, \emph{Threshold}) is discretized over 25 equidistant points.
+
+\paragraph{Application of MBO}
+We apply Kriging (also called Gaussian process regression) to fit the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$.
+We use the implementation in the R-package \emph{DiceKriging}~\citep{roustant_dicekriging_2012} and configure it to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
+We use one-hot encoding to handle categorical parameters.
+Also our search space is hierarchical because \emph{Epsilon} and \emph{Threshold} are only active for the respective \emph{selection strategy}.
+The hierarchical structure of introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter of \emph{thr} inactive.
+If a value is inactive it is set to a value two times as high as the maximum in its active range.
 

### 2020-08-14 generator/main.R
add 5000 nsim plot

@@ -244,6 +244,18 @@ plot_wrapper(name = "plot_boxplot_valid_y", fig.height = 1.6 * FIG_HEIGHT, expr 
   g + geom_boxplot()
 })
 
+## ----plot_boxplot_valid_y_5000-----------------------------------------------------
+plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT, expr = {
+  tmp = rbind(res_grid, res_mbo[, colnames(res_grid), with = FALSE])
+  tmp = tmp[n_cases == 2000 & effect == "paper",]
+  g = ggplot(tmp, aes(x = algorithm, y = y, color = algorithm, fill = algorithm))
+  g = g + facet_grid(.~nsim, scales = "free", labeller = label_both)
+  darker_colors = colorspace::darken(algorithm_labels_color, amount = 0.6)
+  names(darker_colors) = names(algorithm_labels_color)
+  g = g + scale_fill_manual(values = algorithm_labels_color) + scale_color_manual(values = darker_colors)
+  g + geom_boxplot()
+})
+
 
 ## ----table_best---------------------------------------------------------------
 #best of grid

### 2020-08-20 main.tex
overleag
mbo 5000

@@
 % own packages
+\usepackage{siunitx}
 
@@
 This vast space of possible setups makes the optimization computationally demanding.
-We demonstrate here, that the use of the so-called Model-based Optimization (MBO) approach~\citep{jones_taxonomy_2001} can be applied successfully to maximize the statistical power of a clinical trial design for given effect sizes and number of treatments.
+We demonstrate here, that the use of the so-called Model-based optimization (MBO) approach~\citep{jones_taxonomy_2001} can be applied successfully to maximize the statistical power of a clinical trial design for given effect sizes and number of treatments.
 
@@
 Various implementations of Model-based optimization methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012} and BoTorch~\citep{balandat_botorch_2020}.
-For this work we use the implementation in the R-package mlrMBO~\citep{bischl_mlrmbo_2017}.
+For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
 This implementations has been successfully applied to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
+Also, MBO is well suited for multi-criteria optimization.
+Different adaptions of MBO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
+Many of them are implemented in \texttt{mlrMBO} as well.
+However, in this work we restrict ourself to single-criteria optimization.
 
 \section{Motivating Example}
+
+This paper introduces the idea of model-based optimization (MBO) for finding a trail design that is optimal in respect to a certain criterion.
+To show the efficiency of this approach, we apply MBO to optimize the statistical power of a trail design.
+We compare the computation time as well as the best obtained power between MBO and an exhaustive grid search.
+The presented optimization approach can be applied to optimize any trail design if the criterion is numeric and the parameters of the trail design are numeric and include none or just few categorical choices.
+
+One trail design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} (What is THE asd paper?) as implemented in the \texttt{asd}~\cite{parsons_software_2011} package.
+In~\citet{friede_adaptive_2020} the authors present an application of \emph{adaptive seamless designs} for chronic obstructive pulmonary disease (COPD).
+We use the COPD study to apply MBO for finding a trail design with the maximal statistical power within the scenarios restrictions.
+
+% taken from Friede et al (2020) Biom J, Section 5.1 Clinical trial in COPD with treatment selection;
+\citet{barnes_integrating_2010} and \citet{donohue_oncedaily_2010} report a seamless adaptive design with dose selection, which was also discussed elsewhere \citep{cuffe_when_2014}. %COPIED!
+Patients were randomized to four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}), active controls, and placebo control. %COPIED!
+For the purpose of illustration, we ignore the active controls in the following and use the observed results to illustrate the design process. %COPIED!
+The primary outcome was the percentage of days of poor control over 26 weeks. %COPIED!
+As recruitment for the entire study took only 6 months, the interim treatment selection could not be based on the primary endpoint. %COPIED!
+Trough forced expiratory volume in 1 s (FEV1) at 15 days was identified as a suitable early outcome to inform the interim analysis. %COPIED!
+From Figure 1 of Barnes et al.
+(2011), difference in trough FEV1 compared to placebo at 15 days is 150 ml, 180 ml, 210 ml, and 200 ml for the indacaterol doses \SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram}, and \SI{600}{\micro\gram}, respectively.%COPIED!
+Reported 95\% confidence intervals suggest two standard errors of treatment difference is approximately 60 ml, so assuming equal stage 1 sample sizes n1 = 110, then the standard deviation of the measurements is approximately 220 ml.
+%COPIED!
+Standardized effect sizes are thus approximately 0.68, 0.82, 0.95, and 0.91 for the indacaterol doses \SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram}, and \SI{600}{\micro\gram}, respectively.
+%COPIED!
+Regarding the final outcome days of poor control (\%) over 26 weeks, we gather from the report of stage 2 results at \texttt{http://clinicaltrials.gov/show/NCT00463567} that the placebo rate was 35\%.
+%COPIED!
+Let us assume rates of 31\%, 30\%, 28\%, and 29\% for the four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}). %COPIED!
+Based on reported standard errors, we estimate the standard deviation to be approximately 30\%.
+Hence, the standardized effect sizes are approximately 0.13 (\SI{75}{\micro\gram}), 0.17 (\SI{150}{\micro\gram}), 0.23 (\SI{300}{\micro\gram}), and 0.20 (\SI{600}{\micro\gram}).
+%COPIED!
+The approximate sample sizes per arm were 100 patients in stage 1 and 300 patients in stage 2.%COPIED!
+Since the aim was to select two doses of indacaterol at the interim analysis to take into stage 2, we consider an overall sample size for the two stage of 5 × 100 + 3 × 300 = 1,400 patients.
+Furthermore, a moderate positive correlation between early and final outcomes of 0.4 is assumed.%COPIED!
+In the following, we will consider three settings: %COPIED!
+(i) continuous early and final outcomes and selecting always two doses for the second stage, as in the original study; %COPIED!
+(ii) continuous early and final outcomes, as in the original study, but for the purpose of illustration, we vary the selection rule using the threshold rule to allow varying numbers of doses being taken forward for confirmatory testing in the second stage; %COPIED!
+(iii) as final outcomes are non-normal and early and final outcomes are on different scales, we assume in another setting that the final is binary and not continuous.%COPIED!
+
+% FIXME Question: 
+% What does exactly specifies 'corr'?
+
 % TODO: Jakob 
-COPD trial
-% Friede et al (2020) Biom J, Section 5.1 Clinical trial in COPD with treatment selection;
-\cite{friede_adaptive_2020}
+% phase 1 idea something fast
+% phase 2
+% Describe the study we look at. / The Data
+%COPD trial
 
-\cite{barnes_integrating_2010}
-\cite{donohue_oncedaily_2010}
-\cite{cuffe_when_2014}
+% What about these three papers?
+%\cite{barnes_integrating_2010}
+%\cite{donohue_oncedaily_2010}
+%\cite{cuffe_when_2014}
 
+% Do we really take in this Assurance?
 Assurance: \cite{stallard_optimal_2009}
@@
 \section{Adaptive Seamless Designs}
+\label{sec:adaptive_seamless_designs}
+
+% TODO: Jakob copy contents from paper
+% Is there a condensed version? In Friede et al (2020) it is only explained in detail?
 % TODO: Tim
@@
 \subsection{Two-stage design with treatment selection}
+\label{ssec:two_stage_design_with_treatment_selection}
+
 Multi-arm trial; two-stages; treatment selection in interim analysis; final analysis based on combination test approach with closed test principle
@@
 \subsection{Selection rules}
+\label{ssec:selection_rules}
+
 \paragraph{Epsilion Rule}
@@
 \subsection{Clinical scenario evaluation for ASD}
-Friede et al (2010) DIJ, Figure 2
+\label{ssec:clinical_scenario_evaluation_for_asd}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{assets/friede_refinement_2010_fig2.png}
+\caption{Application of the refined clinical scenario evalua- tion framework to adap- tive seamless designs in progressive multiple scle- rosis. REPLACE WITH TikZ}
+\label{fig:friede_refinement_2010_fig2}
+\end{figure}
+Friede et al (2010) DIJ, Figure 2 % FIXME: Which paper? https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
 
 \subsection{Simulating ASD}
+\label{ssec:simulating_asd}
 R package asd by Nick Parsons
@@
 \section{Model-based Optimization}
+\label{sec:model_based_optimization}
 
@@
 \section{Simulation Study}
+\label{sec:simulation_study}
+
 % TODO : Jakob
@@
 We apply Kriging (also called Gaussian process regression) to fit the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$.
-We use the implementation in the R-package \emph{DiceKriging}~\citep{roustant_dicekriging_2012} and configure it to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
+We use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012} and configure it to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
 We use one-hot encoding to handle categorical parameters.
@@
 \begin{figure}[htb]
-\begin{center}
+\centering
 \includegraphics[width=\linewidth]{generated/figures/plot_allbest.pdf}
 \caption{Power of the different selection rules. For the epsilon and threshold rule only the curve for the epsilon and threshold value that reached the best power is dispayed.}
-\end{center}
+\label{fig:plot_allbest}
 \end{figure}
@@
 \begin{figure}[htb]
-\begin{center}
+\centering
 \includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
 \caption{Best found $x$-values from MBO runs.}
-\end{center}
+\label{fig:plot_best_x}
 \end{figure}
@@
 \begin{figure}[htb]
-\begin{center}
+\centering
 \includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
@@
   }
-\end{center}
+\label{fig:plot_opt_path} 
 \end{figure}
@@
 \begin{figure}[htb]
-\begin{center}
+\centering
 \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
@@
   }
-\end{center}
+\label{fig:plot_boxplot_valid_y}
 \end{figure}
@@
 
-\subsection{Second level heading}
+\subsection{More simulation rounds}
 
-This is the body text. Please note that cross-references in the body text should be shown as follows:
-(Miller, 1900), (Miller and Baker, 1900) or if three or more authors (Miller {\it{et al}}., 1900)
-\vspace*{12pt}
+To investigate the effect of the function noise variance ($\sigma^2_n$) on the optimization result we increased the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario with the \emph{paper} effect set and  $n_{\text{treat}} = 2000$.
+This is expected to increase the runtime by a factor of 5.
+Also we expect that due to the decreased noise variance the grid search might be less prone to select over optimistic configurations.
+Furthermore, MBO might need less iterations because the signal to noise ratio is higher.
 
-\noindent Bullet lists are not allowed. Always use (i), (ii), etc.
-\vspace*{12pt}
+Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that a single optimization run leads to a results closer to the mean result.
+Also the outcome of the best found point by MBO is closer to the best outcome found by the exhaustive grid search.
+We have to remember that this plot reports the best observed outcome so far. 
+However, as explained in Section~\ref{sec:model_based_optimization} the final point is not determined by the best found outcome but by the surrogates mean prediction of all evaluated point.
+This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for MBO in both cases.
+Interestingly the validation error of the grid search is lower for $n_\text{sim} = 5000$ then for $n_\text{sim} = 1000$.
+We would have expected the opposite, since more noise during optimization leads to overly optimistic choices that result in a bad validation error.
+This does not seem to be the case here but also the results are not drastically different and the illustrated IQR for the grid search becomes smaller for $n_\text{sim} = 5000$.
 
-\noindent Sentences should never start with a symbol.
-\vspace*{12pt}
+\begin{figure}[htb]
+\centering
+  \begin{minipage}{0.642\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
+    \captionof{figure}{Opt Path}
+    \label{fig:plot_opt_path_5000}
+  \end{minipage}%
+  \begin{minipage}{0.1\textwidth}
+  \end{minipage}%
+  \begin{minipage}{0.357\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
+    \captionof{figure}{Boxplot}
+    \label{fig:plot_boxplot_valid_y_5000}
+  \end{minipage}
+\end{figure}
 
-\noindent Names of software packages and website addresses should be written in {\tt{Courier new, i.e. Stata, the R package
-MASS, http://www.biometrical-journal.com.}}
+% This is the body text. Please note that cross-references in the body text should be shown as follows:
+% (Miller, 1900), (Miller and Baker, 1900) or if three or more authors (Miller {\it{et al}}., 1900)
+% \vspace*{12pt}
+
+% \noindent Bullet lists are not allowed. Always use (i), (ii), etc.
+% \vspace*{12pt}
+
+% \noindent Sentences should never start with a symbol.
+% \vspace*{12pt}
+
+% \noindent Names of software packages and website addresses should be written in {\texttt{Courier new, i.e. Stata, the R package
+% MASS, http://www.biometrical-journal.com.}}
 
@@
 % subsection Outlook
+
+Instead of keeping the number of total trials constant, we could apply multi-criteria optimization and include this number as an additional outcome.
+This multi-criteria approach would result in a set of points (Pareto front) that minimize the number of trails as well as maximize the statistical power of the trail design.
+The user could then see for which number of trails the power is above a certain threshold (e.g. $95\%$).
+
+Another way of speeding up the optimization would be to adaptively increase the iterations of the design simulation. 
+If few simulations already indicate a bad outcome it might not be promising to evaluate more simulation runs.
+Such procedure can be guided through statistical testing and is already common in machine learning under the terms racing or multi-fidelity optimization.
 

### 2020-08-20 generator/main.R

@@ -245,7 +245,7 @@ plot_wrapper(name = "plot_boxplot_valid_y", fig.height = 1.6 * FIG_HEIGHT, expr 
 })
 
 ## ----plot_boxplot_valid_y_5000-----------------------------------------------------
-plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT, expr = {
+plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT * 0.35, fig.width = 0.357 * FIG_WIDTH, expr = {
   tmp = rbind(res_grid, res_mbo[, colnames(res_grid), with = FALSE])
   tmp = tmp[n_cases == 2000 & effect == "paper",]
   g = ggplot(tmp, aes(x = algorithm, y = y, color = algorithm, fill = algorithm))
@@ -253,7 +253,41 @@ plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT, 
   darker_colors = colorspace::darken(algorithm_labels_color, amount = 0.6)
   names(darker_colors) = names(algorithm_labels_color)
   g = g + scale_fill_manual(values = algorithm_labels_color) + scale_color_manual(values = darker_colors)
+  g = g + theme(legend.position = "bottom")
   g + geom_boxplot()
+})
+
+## ----plot_opt_path_5000-----------------------------
+
+# calculate y perf of mbo runs
+plot_wrapper(name = "plot_opt_path_5000", fig.height = 1.6 * FIG_HEIGHT * 0.35, fig.width = 0.642 * FIG_WIDTH, expr = {
+  df = res_mbo[n_cases == 2000 & effect == "paper",]
+  common_names = intersect(colnames(df), colnames(df$opt.path[[1]]))
+  data.table::setnames(df, common_names, paste0("opt.", common_names))
+  df = tidyr::unnest(df, "opt.path")
+  setDT(df)
+  df = df[, cummax_y := cummax(y), by = c(algo.par.names.meta, "repl")]
+  
+  # calculate theoretical best y from grid
+  res_ave = res_eval[n_cases == 2000 & effect == "paper", list(mean_y = mean(y)), by = algo.par.names]
+  df_best = res_ave[,.SD[order(-mean_y)[1]], by = algo.par.names.meta][, c(algo.par.names.meta, "mean_y"), with = FALSE]
[…]

### 2020-09-10 main.tex
joerg edits

@@
 \pagespan{1}{}
-\keywords{Model-Based Optimization; Design of Experiments;\\ %FIXME: name 5 %Up to five keywords are allowed and should be given in alphabetical order. Please capitalize the key
-\noindent \hspace*{-4pc}\\
-\hspace*{-4pc} {\small\it words)}\\[1pc]
+\keywords{Adaptive Seamless Designs; Clinical Trials; Model-Based Optimization; Treatment Selection\\\\ %FIXME: name 5 %Up to five keywords are allowed and should be given in alphabetical order.
+Please capitalize the key
+%\noindent \hspace*{-4pc}\\
+%\hspace*{-4pc} 
+%{\small\it words)}\\[1pc]
 \noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not
@@
 \begin{abstract}
-Planning the patient group sizes and test methods for clinical trials in such a way that the test power is as high as possible for given effect sizes under the null hypothesis and for a fixed number of treatments is a crucial task.
-From a wide set of possible setups we want to chose the most promising plan in short time to drive fast decisions.
-While it is possible to simulate the power for all various setups, the space of possible setups soon becomes too big.
-Either you need big computational resources or you have to wait for an exhaustive exploration of the possible setups.
-In this paper we show that you can successfully apply Model-based Optimization to find a clinical trial set-up that yields a high power from a wide set of possible set-ups.
-We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes and compare the result to an exhaustive search to show that our optimization approach finds a promising setup in a fraction of the time.
+In clinical trials, planning patient group sizes and test methods is a crucial task.
+A common goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
+From a wide range of possible alternatives we want to select the best plan in a short period of time, to enable quick decisions.
+The standard approach to simulate the power for each single setup can become too time-consuming, when the number of possible setups becomes very large.
+Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
+Here, propose a new optimization approach. We show how Model-based Optimization can be used to quickly select a clinical trial setup with high power from a large number of setups options.
+We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
+We compare the MBO approach with an exhaustive evaluation of all candidate designs.
+The optimization approach finds competitive setups in a fraction of the time.
 \end{abstract}
@@
 
-Clinical research and in particular drug development is typically structured into phases, e.g. phases I to IV in drug development; similar approaches for e.g. complex interventions (MRC framework). 
+Clinical research and in particular drug development is typically structured into phases, e.g.
+phases I to IV in drug development.
+Similar approaches exist for complex interventions, such as the MRC framework for non-pharmacological interventions (ref)\cite{}.
 The development phases include elements of learning and confirmation.
-For instance, \citet{sheiner_learning_1997} described the drug development process as two learning and confirming cycles. 
-First learning / confirming cycle (Phase I-IIa): Learning about tolerated dose (Phase I). 
-Then confirming of efficacy of selected dose in selected group of patients (Phase IIa). 
-Second learning / confirming cycle (Phase IIb-III): Learning about optimal use in representative patients (Phase IIb); Then confirming of acceptable benefit / risk ratio (Phase III). 
-Traditionally, separate studies for learning and confirming.
+For instance, \citet{sheiner_learning_1997} describe the drug development process as two learning and confirming cycles:
+The first cycle includes learning about the tolerated dose (Phase I) and confirming the efficacy of the selected dose in a selected group of patients (Phase IIa), whereas the second cycle consists of learning about the optimal use in representative patients (Phase IIb) and confirming of an acceptable benefit or risk ratio (Phase III).
 
-Changing landscape in clinical research: Dissolving boundaries between development phases; Master protocols: Broader developments focusing not only on one single treatment or one single indication / disease; Adaptive designs: Study designs become more flexible; More principled approach to decision making using statistical modeling and integrated analyses across studies
+Traditionally, separate studies are performed for learning and confirming.
+However, the landscape in clinical research is changing.
+Boundaries between the development phases are increasingly dissolving in multiple aspects.
+Master protocols are a new concept where multiple subgroups and substudies are integrated (ref).
+Here, simultaneously, more than one treatment and more than one disease type are investigated within the same overall trial structure.
+Adaptive designs are established, but are being further developed.
+In general, study designs become more flexible, and more principled approaches to decision making using statistical modeling and integrated analyses across studies become popular. 
+A related principle on the analysis side is evidence synthesis, which is a way of combining information from several studies that have examined the same question to come to an overall understanding.
 
-More complex designs and analyses of clinical trials. 
-Examples: adaptive design; evidence synthesis.
+The increasing complexity of both the designs and the analyses of clinical trials has the consequence that the planning of such studies often relies also on computer simulations, especially if no closed form solutions for calculations are available.
+This applies even to usually fairly straightforward tasks as sample size calculations. 
+Various frameworks and guidance proposals have been developed over the past years on how to plan, execute and interpret the results of such simulation studies.
+Method comparison studies are used to give clinicians and biostatisticians guidance when an approach improves currently used methods.  
+Depending on the setting and the purpose of the simulation study, suitable metrics must be chosen to compare alternative designs or analysis strategies. 
+Especially relevant are neutral comparison studies, which are characterized as follows: The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way (ref).
+%While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios.
+A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation). ref
+CSE was particularly developed to support the overall process of performing simulations of clinical trials, especially for complex designs and analysis strategies.
+The problem is decomposed into data models, analysis models, and evaluation models that specify the evaluation metrics.
 
-With increasing complexity of both designs and analyses of clinical trials the planning of such studies often relies on computer simulations, since no closed form solutions even for usually fairly straightforward tasks as sample size calculations are available.
-Frameworks and guidance have been developed over the past years on how to planning, execution and interpretation of such simulation studies. 
-Method comparison studies (references), references on simulation studies, in particular Clinical Scenario Evaluation (CSE).
-
-Depending on the setting and the purpose of the simulation study suitable metrics are chosen to compare alternative design or analysis strategies. 
-While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios.
-Here we propose to use a formal approach to optimization. 
-The set of possible scenarios can be described as a multi dimensional search space, including real-valued and categorical parameters.
-This vast space of possible setups makes the optimization computationally demanding.
-We demonstrate here, that the use of the so-called Model-based optimization (MBO) approach~\citep{jones_taxonomy_2001} can be applied successfully to maximize the statistical power of a clinical trial design for given effect sizes and number of treatments.
-
-The remainder of the manuscript is structured as follows:
-
+Here, we propose to use a formal approach to optimization of the clinical trial design, in the situation of a very large number of available options for the parameters specifying the design. 
+This set of candidate designs can be regarded as a multi-dimensional search space, including potentially many real-valued and categorical parameters.
+The vast space of possible designs makes the optimization computationally demanding.
+In the following, we demonstrate that the so-called Model-based optimization (MBO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
+As evaluation metric We consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
+Further, it is crucial that such a design is found in a short time.
 %\subsection{Related work}
@@
 
-Model-based optimization, or often used interchangeably Bayesian optimization has been successfully applied to optimize expensive black-boxes in many scenarios.
+MBO, often also called Bayesian optimization, has been successfully applied to optimize expensive black-boxes in many different scenarios.
 Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
-Various implementations of Model-based optimization methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012} and BoTorch~\citep{balandat_botorch_2020}.
+Various implementations of MBO methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
 For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
-This implementations has been successfully applied to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
+This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
 Also, MBO is well suited for multi-criteria optimization.
@@
 Many of them are implemented in \texttt{mlrMBO} as well.
-However, in this work we restrict ourself to single-criteria optimization.
+However, in this work we restrict ourselves to single-criteria optimization.
+
+The remainder of the manuscript is structured as follows.
+In the next section we present as motivating example an application of adaptive seamless designs for chronic obstructive pulmonary disease (COPD).
+In the sections 3 and 4, we explain the background on adaptive seamless designs (asd) and model-based optimization (MBO), respectively.
+In section 5, we present a simulation study that is closely related to the motivating example and in which the suitability of MBO for optimizing trial designs is demonstrated.
+In the last Section 6, we discuss promising extensions as well as limitations of the approach.
 
 \section{Motivating Example}
+\label{sect.motivation}
 
-This paper introduces the idea of model-based optimization (MBO) for finding a trail design that is optimal in respect to a certain criterion.
-To show the efficiency of this approach, we apply MBO to optimize the statistical power of a trail design.
-We compare the computation time as well as the best obtained power between MBO and an exhaustive grid search.
-The presented optimization approach can be applied to optimize any trail design if the criterion is numeric and the parameters of the trail design are numeric and include none or just few categorical choices.
+The general idea of this paper is to apply MBO for finding a trial design that is optimal with respect to a certain metric.
+As an example, we consider the maximization of the statistical power of a trail design, given a set of parameters to be chosen for the trial design.
+We compare the computation time as well as the maximal obtained power between the results obtained with MBO and with an exhaustive grid search on the space of adjustable parameters.
+The presented optimization approach can also be applied for optimizing other metrics than the power, as long as the metric itself is numeric and the parameters of the trial design are numeric and include none or just few categorical choices.
 
-One trail design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} (What is THE asd paper?) as implemented in the \texttt{asd}~\cite{parsons_software_2011} package.
+One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd}~\cite{parsons_software_2011} package and described in Section~\ref{sec:adaptive_seamless_designs}.
+%(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, ) 
+% https://doi.org/10.1002/bimj.200510232
+% seamless trials (assoziert)
+%  https://doi.org/10.1002/bimj.200510231
 In~\citet{friede_adaptive_2020} the authors present an application of \emph{adaptive seamless designs} for chronic obstructive pulmonary disease (COPD).
-We use the COPD study to apply MBO for finding a trail design with the maximal statistical power within the scenarios restrictions.
+We will use this COPD study to apply MBO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
+%within the given scenarios restrictions.
 
-% taken from Friede et al (2020) Biom J, Section 5.1 Clinical trial in COPD with treatment selection;
-\citet{barnes_integrating_2010} and \citet{donohue_oncedaily_2010} report a seamless adaptive design with dose selection, which was also discussed elsewhere \citep{cuffe_when_2014}. %COPIED!
-Patients were randomized to four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}), active controls, and placebo control. %COPIED!
-For the purpose of illustration, we ignore the active controls in the following and use the observed results to illustrate the design process. %COPIED!
-The primary outcome was the percentage of days of poor control over 26 weeks. %COPIED!
-As recruitment for the entire study took only 6 months, the interim treatment selection could not be based on the primary endpoint. %COPIED!
-Trough forced expiratory volume in 1 s (FEV1) at 15 days was identified as a suitable early outcome to inform the interim analysis. %COPIED!
-From Figure 1 of Barnes et al.
-(2011), difference in trough FEV1 compared to placebo at 15 days is 150 ml, 180 ml, 210 ml, and 200 ml for the indacaterol doses \SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram}, and \SI{600}{\micro\gram}, respectively.%COPIED!
-Reported 95\% confidence intervals suggest two standard errors of treatment difference is approximately 60 ml, so assuming equal stage 1 sample sizes n1 = 110, then the standard deviation of the measurements is approximately 220 ml.
-%COPIED!
-Standardized effect sizes are thus approximately 0.68, 0.82, 0.95, and 0.91 for the indacaterol doses \SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram}, and \SI{600}{\micro\gram}, respectively.
-%COPIED!
-Regarding the final outcome days of poor control (\%) over 26 weeks, we gather from the report of stage 2 results at \texttt{http://clinicaltrials.gov/show/NCT00463567} that the placebo rate was 35\%.
-%COPIED!
-Let us assume rates of 31\%, 30\%, 28\%, and 29\% for the four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}). %COPIED!
-Based on reported standard errors, we estimate the standard deviation to be approximately 30\%.
-Hence, the standardized effect sizes are approximately 0.13 (\SI{75}{\micro\gram}), 0.17 (\SI{150}{\micro\gram}), 0.23 (\SI{300}{\micro\gram}), and 0.20 (\SI{600}{\micro\gram}).
-%COPIED!
-The approximate sample sizes per arm were 100 patients in stage 1 and 300 patients in stage 2.%COPIED!
-Since the aim was to select two doses of indacaterol at the interim analysis to take into stage 2, we consider an overall sample size for the two stage of 5 × 100 + 3 × 300 = 1,400 patients.
-Furthermore, a moderate positive correlation between early and final outcomes of 0.4 is assumed.%COPIED!
-In the following, we will consider three settings: %COPIED!
-(i) continuous early and final outcomes and selecting always two doses for the second stage, as in the original study; %COPIED!
-(ii) continuous early and final outcomes, as in the original study, but for the purpose of illustration, we vary the selection rule using the threshold rule to allow varying numbers of doses being taken forward for confirmatory testing in the second stage; %COPIED!
-(iii) as final outcomes are non-normal and early and final outcomes are on different scales, we assume in another setting that the final is binary and not continuous.%COPIED!
+%phase II III klinische studien je in Phase
+% nicht 100% II = stage 1, 
+%stage 1 lung function test
+%stage 2
+
+For the COPD trial, an interim treatment selection was performed. 
+Indacaterol is a drug administered to COPD patients.
+The patients were randomized to four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}), active controls, and placebo control.
+As in \citet{friede_adaptive_2020}, here we ignore the active controls.
+The primary outcome of the trial was the percentage of days of poor control over 26 weeks of the COPD patients.
+Since the recruitment took only a short time, another outcome was required for treatment selection, in this case forced expiratory volume in 1 s (FEV1) at 15 days.
+In the original study, the difference in FEV1 compared to placebo at 15 days for the different doses and corresponding confidence intervals were estimated.
+From these, standardized effect sizes were estimated as approximately 0.68, 0.82, 0.95, and 0.91 for the four  indacaterol doses \SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram}, and \SI{600}{\micro\gram}, respectively.
+For the final outcome days of poor control, also, from real data the standardized effect sizes were estimated.
+These were approximately 0.13, 0.17, 0.23, and 0.20, for the four doses.
+The approximate sample sizes per arm were 100 patients in stage 1 and 300 patients in stage 2.
+
+In the analysis of \citet{friede_adaptive_2020}, the aim was to select two doses of indacaterol at the interim analysis that are passed to stage 2, resulting in a total sample size $5 \cdot 100 + 3 \cdot 300 = 1,400$ patients.
+In the simulations, a positive correlation between early and final outcomes of 0.4 was assumed.
+Further, different settings with continuous early and continuous or binary final outcomes and different treatment selection strategies were compared.
 
@@
 % What does exactly specifies 'corr'?
+% auf der individuen ebene
 
@@
 
-% Do we really take in this Assurance?
-Assurance: \cite{stallard_optimal_2009}
+% FIXME: Do we really take in this Assurance?
+Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
 
@@
 \includegraphics[width=\linewidth]{assets/friede_refinement_2010_fig2.png}
-\caption{Application of the refined clinical scenario evalua- tion framework to adap- tive seamless designs in progressive multiple scle- rosis. REPLACE WITH TikZ}
+\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis. REPLACE WITH TikZ}
 \label{fig:friede_refinement_2010_fig2}
 \end{figure}
-Friede et al (2010) DIJ, Figure 2 % FIXME: Which paper? https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
+%FIXME Is this the right paper? Convert to tikz! Tim schickt mir das Original!
+%Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
 
@@
 
+\clearpage
+
 \section{Model-based Optimization}
@@
 Sequential \textbf{Model-based Optimization (MBO)}~\citep{jones_taxonomy_2001} (also known as Bayesian optimization) is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
-In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming, and it becomes infeasible to evaluate the black-box for thousands of configurations.
-MBO solves the optimization problem within a bounded search space $\Theta$:
+In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
+Then it becomes infeasible to evaluate the black-box for thousands of configurations.
+
+\subsection{General principle}
+MBO solves an optimization problem within a bounded search space $\Theta$:
 \[
@@
 It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation.
-The acquisition function achieves exploration by giving a high value to areas where the surrogate predicts a high uncertainty.
-Exploitation is achieved by giving high values to values of $\theta$ where the surrogate predicts a high outcome of $f$.
+The acquisition function achieves exploration by assigning high values to areas where the surrogate predicts a high uncertainty.
+Exploitation is achieved by assigning high values to areas of $\theta$ where the surrogate predicts a high outcome of $f$.
 For deterministic functions the expected improvement~\citep{jones_efficient_1998} is arguably the most popular acquisition function.
 
-\paragraph{Acquisition function for noisy outcomes}
+\subsection{Acquisition function for noisy outcomes}
 Since this work focuses on optimization of a stochastic simulation, we assume that our non-deterministic optimization problem can be formulated as follows:
 \begin{equation}
-\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon \, \text{,with}  \varepsilon \sim N(0, \sigma^2_{n}) \text{iid.}
+\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2_{n}) %\text{ i.i.d.}
 \end{equation}
-In this setting the expected improvement is not feasible as shown in \citet{huang_global_2006}.
+In this setting, the expected improvement is not feasible, as shown in \citet{huang_global_2006}.
 They propose to use the augmented expected improvement instead.
@@
 \end{multline}
-with $\Phi$ and $\phi$ as the distribution and density function of the standard normal distribution, and
-with $\sigma_{n}^2$ denoting the random error (nugget effect) in the Kriging model.
+where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma_{n}^2$ denotes the random error (nugget effect) in the Kriging model.
 The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
 This assumption is also reflected by using a Gaussian process regression as a surrogate.
-However, the intuition in this formula is that the first summand favors exploitation while the second summand favors exploration.
-The correction term is necessary to avoid exploration in areas where the model uncertainty ($\hat{s}^2(\theta)$) equals the random error ($\sigma_{n}^2$), because in this case further evaluations will not decrease the model uncertainty.
+The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
+The correction term is necessary to avoid exploration in areas where the model uncertainty $\hat{s}^2(\theta)$ equals the random error $\sigma_{n}^2$, because in this case further evaluations will not decrease the model uncertainty.
 
-\paragraph{Choosing the best point from noisy outcomes}
-If we chose the configuration ($\theta$) from all previously evaluated configurations that led to the best outcome of $f$ as the optimization result, we would be overly optimistic.
-It is likely that a single best outcome can be attributed to the random error ($\sigma_{n}^2$).
-Instead we are interested in the optimum of the true posterior mean.
-Therefore, we employ the surrogates estimate of the posterior mean for each evaluated configuration to cancel out the noise.
+\subsection{Selection of the best point from noisy outcomes}
+If we chose the configuration $\theta$ from all previously evaluated configurations according to the best outcome of $f$ as the optimization result, we would be overly optimistic.
+It is likely that a single best outcome can be partially attributed to the random error $\sigma_{n}^2$.
+Instead, we are interested in the optimum of the true posterior mean.
+Therefore, we employ the surrogate estimate of the posterior mean for each evaluated configuration to cancel out the noise.
 The configuration for which the surrogate estimates the best outcome is then returned as the optimization result $\hat{\theta}^\ast$:
 \begin{equation}
-  \hat{\theta}^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta),
+  \hat{\theta}^{\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta),
 \end{equation}
 %To obtain a final outcome $\nu^\ast = f(\theta^\ast)$ that is independent of the optimization process, $f(\theta^\ast)$ is calculated again.
-Using the stochastic outcome $\hat{\nu}^\ast = f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
-Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimation of the reached best outcome.
+Using the stochastic outcome $\hat{\nu}^\ast := f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
+Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimate of the true value of the best outcome.
+
+\clearpage
 
@@
 
-The goal of this simulation study is to show that Model-based optimization can effectively find the trial design setup that yields the highest statistical power for the given \emph{number of total treatments} ($n_{\text{treat}}$), \emph{effect sizes} for all arms and both stages.
+The goal of this simulation study is to investigate the potential benefit of the MBO approach for effectively finding a clinical trial design that maximizes a metric.
+The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sect.motivation}, with treatment selection in the first of two stages.
+In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
+The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.
 
-The effect sizes we consider in this study are given in Table~\ref{tab:table_effect_names}.
+\subsection{Clinical trial design and parameters}
+For the true effect sizes in the simulation, we consider four \emph{effect sets} whose values are inspired and partially based on the corresponding numbers of the motivating example, see Table~\ref{tab:table_effect_names}.
 \input{generated/tables/table_effect_names.tex}
-As this table shows, we restrict ourselves to studies with five arms.
+For the effect set \emph{paper} we use exactly the same numbers as in~\citet{friede_adaptive_2020}.
+For the effect set \emph{paper2} the effect sizes for the second stage are doubled.
+As further case we consider a \emph{linear} effect set with effect sizes linear increasing from 20\% to 80\% for the first stage and effect sizes divided by 4 for the second stage as well as a \emph{sigmoid} relationship, again with 4 times larger effect sizes for the first stage.
+These sets reflect different realistic situations.
 
-The space of possible trial design setups we consider is given in Table~\ref{tab:search_space}.
+The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
+The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control).
+The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage.
+In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
 \begin{table}[h]
-  \caption{Search Space of possible trial design setups.}
+  \caption{Overview of the selection strategies that determine which trial arms are kept for the second stage}
+  \label{tab:selection_strategies}
+  \centering
+  \begin{tabular}{ll}
+  \hline
+  Notation & Procedure of selection strategy  \\
+  \hline
+  \emph{all}     & Select all arms  \\
+  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
+  \emph{eps}     & Select all arms with test statistic less than \emph{eps} (epsilon) below maximal test statistic \\
+  \emph{thresh}     & Select all arms with test statistic larger than the value \emph{thresh} (threshold) \\
+  \hline
+  \end{tabular}
+\end{table}
+\begin{table}[h]
+  \caption{Parameters for the selection strategies representing the search space of trial designs}
   \label{tab:search_space}
@@
   Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thr} \}$ \\
-  Ratio ($r$)        & $(0,1)$ \\
-  Epsilon            & $[0,4]$ \\
-  Threshold          & $[0,10]$ \\
+  $r$ (ratio)        & $(0,1)$ \\
+  eps (epsilon)            & $[0,4]$ \\
+  thresh (threshold)          & $[0,10]$ \\
   \hline
@@
 \end{table}
-The different selection strategies are explained in Table~\ref{tab:selection_strategies}.
-\begin{table}[h]
-  \caption{How are the arms for the second stage selected?}
-  \label{tab:selection_strategies}
-  \centering
-  \begin{tabular}{ll}
-  \hline
-  Selection strategy &  \\
-  \hline
-  \emph{all}     & Select all arms  \\
-  $\{1,2,3\}$    & Select 1, 2 or 3 arms with the maximal test statistic \\
-  \emph{eps}     & Select all arms that are within \emph{Epsilon} of the maximal test statistic \\
-  \emph{thr}     & Select all arms which have a test statistic above the \emph{Threshold} \\
-  \hline
-  \end{tabular}
-\end{table}
-This table also explains the meaning of the \emph{Epsilon} and the \emph{Threshold} parameter which are only active for the respective selection strategy.
-The \emph{Ratio} ($r$) parameter defines how many treatments are conducted in the first and in the second stage respectively.
 
-\paragraph{Definition of the simulation}
-The simulation and calculation of the statistical power of the trial design can be formulated as a black-box as follows:
-\begin{align}
-  \label{eq:bbox}
-  (y,k_2)' &= f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold}).
-\end{align}
-We notice, that we obtain two measures of interest: The simulated statistical power ($y$) and the number of arms in the second stage ($k_2$).
-Our main optimization goal is to maximize $y$.
-$k_2$ is used in a later step.
+\subsection{Calculation of the statistical power}
 
-As mentioned before, one restriction is that the \emph{number of total treatments} ($n_{\text{treat}}$) should be constant.
-The number depends on the number of arms including the control group in stage 1 ($k_1$) and stage 2 ($k_2$) and is calculated as follows:
+Let $n_{\text{treat}}$ be the number of total treatments (in our case 5, including the control), and let $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively.
+Let $k_1$ and $k_2$ denote the number of treatments in stage 1 and stage 2, respectively.
+Within one stage, the same number of patients are allocated to all different treatments.
+Then $n_{\text{treat}}$ can be calculated as
 \begin{equation}
@@
 \end{equation}
-whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ denote the number of treatments per arm in the first and second stage.
-In our scenario $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$.
-Within one stage, all arms obtain the same number of treatments.
+In our scenario we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1.
 
+Our main optimization goal is to maximize the statistical power $y$ depending on the trial design, where $k_2$ depends on the results of the first stage. 
+The power $y$ can be formulated as a black-box as follows:
+\begin{align}
+  \label{eq:bbox}
+  (y,k_2)' &= f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
+\end{align}
+
+The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
+We denote the proportion of patients per treatment in stage 1, compared to stage 2, with 
+$$ r=\frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}$$
 To keep the number of total treatments constant we optimize the ratio ($r \in (0,1)$) of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ instead of optimizing these values directly.
@@
   \label{eq:stagec}
-  \begin{split}
-  n_{\text{stage1}} &= r \cdot c \\
-  n_{\text{stage2}} &= (1-r) \cdot c \ ,
-  \end{split}
+  %\begin{split}
+  n_{\text{stage1}} &= r \cdot c, \qquad
+  n_{\text{stage2}} = (1-r) \cdot c,
+  %\end{split}
 \end{align}
@@
 \begin{align}
-  \label{eq:ntreatc}
-  n_{\text{treat}} &= k_1 \cdot r \cdot c + k_2 \cdot (1-r) \cdot c \Leftrightarrow \\
+ % \label{eq:ntreatc}
+  n_{\text{treat}} & = k_1 \cdot r \cdot c + k_2 \cdot (1-r) \cdot c \;\; \Leftrightarrow \;\; %\\
   \label{eq:ntreatcc}
-  c &= \frac{n_{\text{treat}}}{k_1 \cdot r + k_2 \cdot (1-r)}.
+  c = \frac{n_{\text{treat}}}{k_1 \cdot r + k_2 \cdot (1-r)}.
 \end{align}
@@
 number of simulation iterations (\texttt{nsim}~$= 1000$), correlation between early and final outcomes  (\texttt{corr}~$= 0.4$), test level (\texttt{level}~$=0.025$), and (TODO Not sure what that means) (\texttt{ptest}~$=(3,4)$).
+%jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
 
-%FIXME Frage: Warum reicht es nicht wenn der Suchraum select=0,1,2,3,4,5 ist?
 \paragraph{Exhaustive Search}
@@
 This does not seem to be the case here but also the results are not drastically different and the illustrated IQR for the grid search becomes smaller for $n_\text{sim} = 5000$.
+
+%FIXME: Add Runtime comparision!
 

### 2020-09-10 main.tex
improve figures

@@
 Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
-Here, propose a new optimization approach. We show how Model-based Optimization can be used to quickly select a clinical trial setup with high power from a large number of setups options.
+In this work we propose to use Model-based Optimization to quickly find a clinical trial setup with high power from a large number of setups options.
 We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
@@
 
-Clinical research and in particular drug development is typically structured into phases, e.g.
-phases I to IV in drug development.
-Similar approaches exist for complex interventions, such as the MRC framework for non-pharmacological interventions (ref)\cite{}.
+Clinical research and in particular drug development is typically structured into phases, e.g. phases I to IV in drug development. 
+Similar approaches exist for complex interventions, such as the MRC framework for non-pharmacological interventions (ref)%FIXME ref
 The development phases include elements of learning and confirmation.
@@
 Boundaries between the development phases are increasingly dissolving in multiple aspects.
-Master protocols are a new concept where multiple subgroups and substudies are integrated (ref).
+Master protocols are a new concept where multiple subgroups and substudies are integrated (ref). %FIXME ref
 Here, simultaneously, more than one treatment and more than one disease type are investigated within the same overall trial structure.
@@
 Depending on the setting and the purpose of the simulation study, suitable metrics must be chosen to compare alternative designs or analysis strategies. 
-Especially relevant are neutral comparison studies, which are characterized as follows: The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way (ref).
+Especially relevant are neutral comparison studies, which are characterized as follows: %FIXME Is there someting missing? (Jakob)
+The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way (ref).
 %While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios.
-A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation). ref
+A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation). ref %FIXME ref
 CSE was particularly developed to support the overall process of performing simulations of clinical trials, especially for complex designs and analysis strategies.
@@
 In the following, we demonstrate that the so-called Model-based optimization (MBO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
-As evaluation metric We consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
+As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
 Further, it is crucial that such a design is found in a short time.
@@
 In the next section we present as motivating example an application of adaptive seamless designs for chronic obstructive pulmonary disease (COPD).
-In the sections 3 and 4, we explain the background on adaptive seamless designs (asd) and model-based optimization (MBO), respectively.
-In section 5, we present a simulation study that is closely related to the motivating example and in which the suitability of MBO for optimizing trial designs is demonstrated.
-In the last Section 6, we discuss promising extensions as well as limitations of the approach.
+In the sections~\ref{sec:adaptive_seamless_designs} and~\ref{sec:model_based_optimization}, we explain the background on adaptive seamless designs (asd) and model-based optimization (MBO), respectively.
+In section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of MBO for optimizing trial designs is demonstrated.
+In the last Section~\ref{sec:discussion}, we discuss promising extensions as well as limitations of the approach.
 
@@
 As an example, we consider the maximization of the statistical power of a trail design, given a set of parameters to be chosen for the trial design.
-We compare the computation time as well as the maximal obtained power between the results obtained with MBO and with an exhaustive grid search on the space of adjustable parameters.
+We compare the computation time as well as the maximal obtained power between the results obtained with MBO and with an exhaustive grid search over the space of adjustable parameters.
 The presented optimization approach can also be applied for optimizing other metrics than the power, as long as the metric itself is numeric and the parameters of the trial design are numeric and include none or just few categorical choices.
@@
 \centering
-  \begin{minipage}{0.642\textwidth}
+  \begin{minipage}{0.64\textwidth}
     \centering
@@
     \label{fig:plot_opt_path_5000}
-  \end{minipage}%
-  \begin{minipage}{0.1\textwidth}
-  \end{minipage}%
-  \begin{minipage}{0.357\textwidth}
+  \end{minipage}\hspace{0.01\textwidth}%
+  \begin{minipage}{0.35\textwidth}
     \centering
@@
 \section{Discussion}
+\label{sec:discussion}
 

### 2020-09-10 generator/main.R

@@ -9,8 +9,15 @@ library(knitr)
 library(kableExtra)
 library(dplyr)
 library(mlr3misc)
+library(memoise)
 
-plot_wrapper = function(name, fig.height, fig.width, expr) {
+
+FIG_WIDTH = 9
+FIG_HEIGHT = 4.5
+FNT_SMALL = 6
+CACHE = cache_filesystem(path = "memoise")
+
+plot_wrapper_uncached = function(name, fig.height = FIG_HEIGHT, fig.width = FIG_WIDTH, expr) {
   path = paste0("../generated/figures/", name, ".pdf")
   pdf(file = path, width = fig.width, height = fig.height, onefile = TRUE)
   print(eval(expr))
@@ -18,6 +25,7 @@ plot_wrapper = function(name, fig.height, fig.width, expr) {
   #browser()
   knitr::plot_crop(path)
 }
+plot_wrapper = memoise(plot_wrapper_uncached, cache = CACHE)
 
 kable_to_text = function(text, name) {
   text = stri_replace_all_fixed(text, pattern = "\\label{tab:}", sprintf("\\label{tab:%s}", name))
@@ -29,12 +37,6 @@ own_theme = function(...) {
   theme(legend.margin=margin(0,0,0,0))
 }
 theme_set(own_theme())
-
-
-
-FIG_WIDTH = 8
-FIG_HEIGHT = 6
-FNT_SMALL = 6
 
 
 options(
@@ -58,6 +60,19 @@ EFFECTS = list(
 )
 effect_names = setNames(names(EFFECTS), names(EFFECTS)) #to be populated with better names
 
+select_labels = c("0" = "all", "1" = "1 best", "2" = "2 best", "3" = "3 best", "4" = "epsilon rule", "5" = "random", "6" = "threshold rule")
+select_labels_colors = #http://paletton.com/#uid=74Z140kqdu7ghF3lowyu1ppwGk7
+  c(
+    "all" = "#A61E96",
+    "1 best" = "#FF9B55",
+    "2 best" = "#CA5B0D",
+    "3 best" = "#A04200",
+    "epsilon rule" = "#1B9786",
+    "random" = "#000000",
+    "threshold rule" = "#C0E62A"
+  )
+algorithm_labels = c("mbo" = "MBO", "grid" = "Grid")
+algorithm_labels_color = c("mbo" = "#CB1C00", "grid" = "#00962D")
 
[…]

### 2020-09-14 main.tex
results

@@
 \subsection{Selection of the best point from noisy outcomes}
+\label{ssec:best_point}
 If we chose the configuration $\theta$ from all previously evaluated configurations according to the best outcome of $f$ as the optimization result, we would be overly optimistic.
@@
 \paragraph{Exhaustive Search}
-For each effect set (see Table~\ref{tab:table_effect_names}) we conduct an exhaustive grid search with a resolution of 25 points per dimension.
+For each effect set (see Table~\ref{tab:table_effect_names}) we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension for each selection strategy.
 This results in total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
-Each conf
-Additionally
-Each numeric parameter ($r$, \emph{Epsilon}, \emph{Threshold}) is discretized over 25 equidistant points.
+Additionally the grid is evaluated with 10 stochastic repliactions to obtain an estimation of the variance of the simulation.
+
+The best point of one grid is determined by selecting the configuration that let to the best outcome.
+If we would report this outcome, we will likely be overoptimistic, as the chances are high that the noise contributed to the good outcome.
+To be able to report an unbiased performance we take the best found configuration and look up the results in the remaining 9 stochastic replications of the grid evaluations, similar to a cross-validation in machine learning applications.
+This eliminates the risk of only reporting outcomes that are optimal by chance.
 
 \paragraph{Application of MBO}
-We apply Kriging (also called Gaussian process regression) to fit the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$.
-We use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012} and configure it to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
-We use one-hot encoding to handle categorical parameters.
-Also our search space is hierarchical because \emph{Epsilon} and \emph{Threshold} are only active for the respective \emph{selection strategy}.
+Before the application of MBO, we have to define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ is generated.
+As the regression model for the surrogate we chose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is well known to work well for fairly low dimensional search spaces.
+We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
+
+Kringing expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
+First, to transform categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
+Furthermore, our search space is hierarchical because \emph{Epsilon} and \emph{Threshold} are only active for the respective \emph{selection strategy}.
 The hierarchical structure of introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter of \emph{thr} inactive.
-If a value is inactive it is set to a value two times as high as the maximum in its active range.
+If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if \emph{thresh} is inactive it will be set to 20).
+To set the inactive value to a high value outside of the original range ensures that within the original range the Kriging-model does not or only marginally gets affected by the inactive values.
+Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance by the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
+With the above steps we ensured that the input for the Kriging is purely numeric.
 
-The outcome of the simulation and therefore our optimization objective is stochastic.
-Also our search space is hierarchical because \emph{Epsilon} and \emph{Threshold} are only active for the respective selection strategy.
+As an acquisition function we chose the $\operatorname{AEI}$ as explained in Section~\ref{sec:model_based_optimization}.
+We start MBO with an inital design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator.
 
-\begin{itemize}
-  \item Discretize search space and evaluate grid (possible with enough computational resources)
-  \item Apply MBO and compare results
-\end{itemize}
+At the end of the optimization procedure, the best point is chosen according to the best prediction obtained on the set of evaluated poins as explained in Section~\ref{ssec:best_point}.
+The best selected point is then evaluated $10$ times to allow us to report an an unbiased estimate of the true performance of the optimization result.
+
+The whole MBO procedure is repeated $10$ times to ensure that a single optimization run was not just lucky in finding the optimum.
 
@@
 
-To obtain an unbiased result, we do not simply report the best observed power of both optimization approaches.
-Instead we use the best found parameters to run the simulation again independently, repeated times, similar to a cross-validation in machine learning applications.
-This eliminates the risk of only reporting outcomes that are optimal by chance.
-
+As explained before, we obtained unbiased result from the exhaustive grid search and the MBO procedure.
+However, in Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behaviour of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio}.
 \begin{figure}[htb]
@@
 \end{figure}
+Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments at the early stage.
+Also, for the \emph{selection strategy} \emph{epsilon rule} and \emph{threshold rule} only the curves are shown of the repsective \emph{eps} and \emph{thresh} value that led to the best outcome within the respective selection strategy.
 
@@
 We have to remember that this plot reports the best observed outcome so far. 
-However, as explained in Section~\ref{sec:model_based_optimization} the final point is not determined by the best found outcome but by the surrogates mean prediction of all evaluated point.
+However, as explained in Section~\ref{ssec:best_point} the best point is not determined by the best found outcome but by the surrogates mean prediction of all evaluated point.
 This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for MBO in both cases.

### 2020-09-14 generator/main.R

@@ -78,10 +78,10 @@ algorithm_labels_color = c("mbo" = "#CB1C00", "grid" = "#00962D")
 tmp = lapply(EFFECTS, do.call, what = rbind)
 tmp = lapply(tmp, function(x) cbind.data.frame(stage = rownames(x), x))
 tmp = rbindlist(tmp, idcol = TRUE)
-setnames(tmp, ".id", "effect set")
+setnames(tmp, ".id", "Effect Set")
+setnames(tmp, "stage", "Stage")
 kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation") %>% 
   collapse_rows(columns = 1:2, latex_hline = "major") %>% 
-  pack_rows("Set 1", 2, ncol(tmp), indent = FALSE) %>% 
   add_header_above(c(" " = 2, "Treatment" = 5)) %>% 
   kable_styling(position = "center") %>% 
   kable_to_text("table_effect_names")

### 2020-09-17 main.tex
joerg changes

@@
 \begin{table}[h]
-  \caption{Overview of the selection strategies that determine which trial arms are kept for the second stage}
+  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
   \label{tab:selection_strategies}
@@
 The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
-We denote the proportion of patients per treatment in stage 1, compared to stage 2, with 
-$$ r=\frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}$$
-To keep the number of total treatments constant we optimize the ratio ($r \in (0,1)$) of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ instead of optimizing these values directly.
-Therefore we define
+We denote the proportion of patients per treatment in stage 1, compared to patients per treatment in both stages, with 
+$$ r=\frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.$$
+Given the number of treatments $k_1$ and $k_2$ and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
 \begin{align}
@@
   %\begin{split}
-  n_{\text{stage1}} &= r \cdot c, \qquad
-  n_{\text{stage2}} = (1-r) \cdot c,
+  n_{\text{stage1}} & = 
+  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
+  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \\
+  & = n_{\text{treat}}\cdot
+  \frac{r}{k_1\cdot
+  r+k_2\cdot (1-r)}, \\
+  n_{\text{stage2}} & =
+  n_{\text{treat}}\cdot
+  \frac{1-r}{k_1\cdot
+  r+k_2\cdot (1-r)}.
   %\end{split}
 \end{align}
-whereas $c$ is a constant that depends on $n_{\text{treat}}$, $k_1$ and $k_2$.
-For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
-For a general solution, we can insert the variables from (\ref{eq:stagec}) into (\ref{eq:ntreat}) and obtain
-\begin{align}
- % \label{eq:ntreatc}
-  n_{\text{treat}} & = k_1 \cdot r \cdot c + k_2 \cdot (1-r) \cdot c \;\; \Leftrightarrow \;\; %\\
-  \label{eq:ntreatcc}
-  c = \frac{n_{\text{treat}}}{k_1 \cdot r + k_2 \cdot (1-r)}.
-\end{align}
-Placing $c$ into (\ref{eq:stagec}) lets us calculate $n_{\text{stage1}}$ and $n_{\text{stage2}}$ for the given constraints and a given ratio $r$.
-This formula gives us the treatments per arm and per stage if $k_2$ is known. 
-Unfortunately, this is only the case for \emph{selection strategy} $\in \{\text{all},1,2,3\}$.
-
-For the selection strategies \emph{thr} and \emph{eps}, the number of arms in the second stage ($k_2$) depends on the choice of the \emph{Threshold} or \emph{Epsilon} parameter.
+%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
+These formulas give the treatments per arm and per stage if $k_2$ is known. 
+However, this is only the case for selection strategies 1, 2, and 3.
+For the selection strategies \emph{threshold} and \emph{epsilon}, the number of arms in the second stage ($k_2$) depends on the choice of the \emph{thresh} or \emph{eps} parameter.
 However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 ($\hat{k}_2$), by running a simulation of the first stage.

### 2020-09-28 main.tex
joerg?

@@
 \subsection{Clinical trial design and parameters}
+\label{subsec:design}
 For the true effect sizes in the simulation, we consider four \emph{effect sets} whose values are inspired and partially based on the corresponding numbers of the motivating example, see Table~\ref{tab:table_effect_names}.
@@
 
-Let $n_{\text{treat}}$ be the number of total treatments (in our case 5, including the control), and let $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively.
+Let $n_{\text{treat}}$ be the total number of treatments in the overall trial, and let $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively.
 Let $k_1$ and $k_2$ denote the number of treatments in stage 1 and stage 2, respectively.
 Within one stage, the same number of patients are allocated to all different treatments.
-Then $n_{\text{treat}}$ can be calculated as
+Then $n_{\text{treat}}$ is given by
 \begin{equation}
   \label{eq:ntreat}
-  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}},
+  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}}.
 \end{equation}
-In our scenario we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1.
+In our scenario (as described in Section \ref{subsec:design}) we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1 and the control treatment is included in both stages.
 
-Our main optimization goal is to maximize the statistical power $y$ depending on the trial design, where $k_2$ depends on the results of the first stage. 
-The power $y$ can be formulated as a black-box as follows:
+Our main optimization goal is to maximize the statistical power $y$ depending on the trial design. The power can be formulated as a black-box as follows:
 \begin{align}
   \label{eq:bbox}
-  (y,k_2)' &= f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
+  y & = f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
 \end{align}
@@
 We denote the proportion of patients per treatment in stage 1, compared to patients per treatment in both stages, with 
-$$ r=\frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.$$
-Given the number of treatments $k_1$ and $k_2$ and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
+\begin{align}
+  \label{eq:r}
+ r & = \frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.
+\end{align}
+Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
 \begin{align}
@@
   n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
-  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \\
+  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
   & = n_{\text{treat}}\cdot
   \frac{r}{k_1\cdot
-  r+k_2\cdot (1-r)}, \\
+  r+k_2\cdot (1-r)}, \nonumber \\
   n_{\text{stage2}} & =
@@
 %For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
-These formulas give the treatments per arm and per stage if $k_2$ is known. 
-However, this is only the case for selection strategies 1, 2, and 3.
-For the selection strategies \emph{threshold} and \emph{epsilon}, the number of arms in the second stage ($k_2$) depends on the choice of the \emph{thresh} or \emph{eps} parameter.
-However, for given values of $n_{\text{stage1}}$ and \emph{Threshold} or \emph{Epsilon} respectively we can estimate the expected number of arms at stage 2 ($\hat{k}_2$), by running a simulation of the first stage.
-This simulation is run multiple times for different values of $n_{\text{stage1}}$ in a calibration step until the desired number of $n_{\text{treat}}$ is reached approximately.
-In fact, this calibration step formulates an independent optimization problem:
-We minimize 
+These formulas give the numbers of patients per treatment arm in the two stages, if $k_2$ is known. 
+This is only the case for selection strategies 1, 2, and 3.
+For the selection strategies \emph{threshold} and \emph{epsilon}, $k_2$ depends on the outcome of stage 1.
+%choice of the \emph{thresh} or \emph{eps} parameter.
+In these cases, for given values of $n_{\text{stage1}}$ and \emph{thresh} or \emph{eps}, respectively, we can  obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
+For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
+We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed values for $n_{\text{stage1}}$.
+Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total patients is closest to $n_{\text{treat}}$.
+More precisely, we minimize
 \begin{align}
@@
   \begin{split}
-  & \operatorname{min}_{n_{\text{stage1}} \in [l,u]} h(n_{\text{stage1}}) \,  \text{, with} \\
-  h(n_{\text{stage1}}) &:= \biggl( (k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot \underbrace{\frac{1-r}{r} \cdot n_{\text{stage1}}}_{=n_{\text{stage2}}}) - n_{\text{treat}} \biggr)^2 \, \text{and} \\
-  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
+  h(n_{\text{stage1}}) &:= \left( \Big(k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot %\underbrace{
+  \frac{1-r}{r} \cdot n_{\text{stage1}}
+  %}_{=n_{\text{stage2}}}
+  \Big) - n_{\text{treat}} \right)^2
+%  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
   \end{split}
 \end{align}
-and $l = \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil$ and $u = \lceil \frac{n_{\text{treat}}}{k_1} \rceil$. 
-$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
-This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
-The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
+with respect to $n_{\text{stage1}}$, where the estimate $\hat{k}_2$ is obtained as described above dependent on $n_{\text{stage1}}$. 
+%$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
+%This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
+%The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
 
-With the calibration step that is necessary for selection strategy $\in \{\text{thr}, \text{eps} \}$ we can always set $n_{\text{stage1}}$ and $n_{\text{stage2}}$ according to $n_{\text{treat}}$, $r$.
-This allows us to reparametrize the black-box from equation~\ref{eq:bbox} to directly depend from the parameters of interest:
+The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{thr}$ and $\text{eps}$.
+Then the black-box from formula (\ref{eq:bbox}) can be reformulated as
 \begin{equation}
-  y = \tilde{f}_{n_{\text{treat}}}(r, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold}) \, ,
+  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
 \end{equation}
-whereas $n_{\text{treat}}$ is considered to be set constant.
-This constraint is necessary because otherwise bigger number of treatments almost certainly imply a higher statistical power of the study design.
+In our simulations, $n_{\text{treat}}$ is considered to be a constant.
 
-In the following we try to find the clinical design setup, represented by the parameter settings $r$, \emph{effect set}, \emph{Selection strategy}, \emph{Epsilon} and \emph{Thresold} that yield the highest statistical power.
-The statistical power is determined by evaluating the simulator $\tilde{f}$ as explained before.
+Summarizing, in the following we aim to find the clinical design setup, represented by the parameter values for $r$, \emph{effect set}, \emph{Selection strategy}, \emph{epsilon} and \emph{threshold} that yield the highest statistical power, when evaluation $\tilde{f}$.
 
-\paragraph{General simulation settings}
-For the simulation we call the \texttt{treatsel.sim} function from the \texttt{asd}~package~\cite{parsons_software_2011} with the following parameters as fixed:
-number of simulation iterations (\texttt{nsim}~$= 1000$), correlation between early and final outcomes  (\texttt{corr}~$= 0.4$), test level (\texttt{level}~$=0.025$), and (TODO Not sure what that means) (\texttt{ptest}~$=(3,4)$).
-%jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
+\subsection{Optimization strategies}
 
-\paragraph{Exhaustive Search}
-For each effect set (see Table~\ref{tab:table_effect_names}) we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension for each selection strategy.
-This results in total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
-Additionally the grid is evaluated with 10 stochastic repliactions to obtain an estimation of the variance of the simulation.
+In the simulation study we compare an exhaustive grid search on the parameter space with MBO (Model-based optimization) with respect to both performance (power) and runtime.
 
-The best point of one grid is determined by selecting the configuration that let to the best outcome.
-If we would report this outcome, we will likely be overoptimistic, as the chances are high that the noise contributed to the good outcome.
-To be able to report an unbiased performance we take the best found configuration and look up the results in the remaining 9 stochastic replications of the grid evaluations, similar to a cross-validation in machine learning applications.
-This eliminates the risk of only reporting outcomes that are optimal by chance.
+First, we describe the \emph{exhaustive grid search}.
+For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.
+one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and \emph{eps} or \emph{thresh}, respectively).
+This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
+Additionally, the grid is evaluated with 10 stochastic replictions to estimate the variance of the estimated power.
 
-\paragraph{Application of MBO}
-Before the application of MBO, we have to define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ is generated.
-As the regression model for the surrogate we chose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is well known to work well for fairly low dimensional search spaces.
+The best point per evaluated grid is determined by selecting the configuration (parameter set) with the best outcome, i.e. the largest power.
+The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the good outcome.
+To account for this optimism and report unbiased performance estimates, for each of the 10 replicate simulations, we identify the best configuration and consider the corresponding 9 values of the other 9 replicate simulations, similar to a cross-validation framework in machine learning applications.
+\\
+%This eliminates the risk of only reporting outcomes that are optimal by chance.
+
+Next, we describe our realization of the search with the \emph{MBO (Model-based optimization)} approach.
+For the application of MBO, we first have to define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter sets) is generated.
+As the regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is well known to work well for fairly low dimensional search spaces.
 We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
 
-Kringing expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
+Kriging expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
 First, to transform categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
-Furthermore, our search space is hierarchical because \emph{Epsilon} and \emph{Threshold} are only active for the respective \emph{selection strategy}.
-The hierarchical structure of introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter of \emph{thr} inactive.
+Furthermore, our search space is hierarchical because the parameters \emph{epsilon} and \emph{threshold} are only active for the respective \emph{selection strategies}.
+The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter of \emph{thresh} is inactive.
 If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if \emph{thresh} is inactive it will be set to 20).
-To set the inactive value to a high value outside of the original range ensures that within the original range the Kriging-model does not or only marginally gets affected by the inactive values.
-Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance by the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
-With the above steps we ensured that the input for the Kriging is purely numeric.
+This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
+Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
+With the above steps we ensure that the input for the Kriging is purely numeric.
 
-As an acquisition function we chose the $\operatorname{AEI}$ as explained in Section~\ref{sec:model_based_optimization}.
+As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:model_based_optimization}.
 We start MBO with an inital design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator.
 
-At the end of the optimization procedure, the best point is chosen according to the best prediction obtained on the set of evaluated poins as explained in Section~\ref{ssec:best_point}.
-The best selected point is then evaluated $10$ times to allow us to report an an unbiased estimate of the true performance of the optimization result.
+At the end of the optimization procedure, the best point is chosen according to the best prediction obtained on the set of evaluated points as explained in Section~\ref{ssec:best_point}.
+Again, overoptimism must be avoided. Thus, the best selected point (the best parameter setting) is then independently evaluated $10$ times again.
 
@@
 
-\subsection{Design}
+\subsection{Implementation}
 
-The underlying algorithms in this study are implemented in R, for the model-based optimization the R-package mlrMBO~\citep{bischl_mlrmbo_2017} is used, and the simulation of the clinical trials is performed using the R-package asd~\citep{parsons_software_2011}.
+The algorithms used in this study are implemented in R.
+For the model-based optimization, the R-package mlrMBO~\citep{bischl_mlrmbo_2017} is used.
+For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package asd~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
+The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???
+
+\begin{table}[h]
+  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
+  \label{tab:par_implement}
+  \centering
+  \begin{tabular}{lll}
+  \hline
+  Meaning of parameter & Parameter & Value \\
+  \hline
+  Number of simulation iterations & \texttt{nsim} & $1000$ \\
+  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
+  Test level & \texttt{level} & $0.025$ \\
+  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
+  \hline
+  \end{tabular}
+\end{table}
+
+% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
 

### 2020-09-28 main.tex
stupid tikz

@@
 \usepackage{siunitx}
+\usepackage{enumitem}
+% tikz
+\usepackage{tikz}
+% we want ER + above/below + left/right
+\usetikzlibrary{er,arrows, chains, fit, positioning, calc, shapes}
 
@@
 \end{figure}
+
+\tikzset{nearnodes/.style={draw, node distance=0cm,}}
+\begin{tikzpicture}[auto,node distance=0.3cm,align=center,text width = 3cm, font=\scriptsize]
+  \node[draw, ellipse, minimum height=1cm, text width=3cm] (ClinicalScenarios) at (0,0) {
+    \textbf{Clinical scenarios}\\
+    Simulation Studies};
+  \node[name=block, rectangle split, rectangle split parts=2, nearnodes, above left = of ClinicalScenarios] (HistoricalDataPoints)
+  { \begin{itemize}[leftmargin=.3cm]
+      \item Correlation between early interim and late disability outcomes.
+    \end{itemize}
+    \nodepart{second} Combination of options
+  };
+  \node[name=block, rectangle split, rectangle split parts=2, nearnodes, left = of HistoricalDataPoints] (UnknownPoints)
+  { \begin{itemize}[leftmargin=.3cm]
+      \item Treatment effect on early interim outcome.
+      \item Treatment effect on late disability outcome
+    \end{itemize}
+    \nodepart{second} Constrained by health care environment/infrastructure
+  };
+  %\node[nearnodes, inner xsep=1em, inner ysep=1em, above left = of ClinicalScenarios, fit = (HistoricalDataPoints) (UnknownPoints)] (boxDiseaseSpecific) {};
+  \node[draw, above = of HistoricalDataPoints, text width = 6cm] (DiseaseSpecific) {\textbf{Disease-specific features}};
+  %\node[draw, above right = of ClinicalScenarios] (DesignOptions) {Design options};
+  \node[name=block, rectangle split, rectangle split parts=2, nearnodes, above right = of ClinicalScenarios] (CombinationPoints)
+  { \begin{itemize}[leftmargin=.3cm]
+      \item Adaptive seamless design time point of interim analysis
+      \item Treatment selection rules at interim Conventional phase 2/3
+      \item split of resources between phases
+      \item end oh phase futility criteria
+    \end{itemize}
+    \nodepart{second} Combination of options
+  };
+  \node[name=block, rectangle split, rectangle split parts=2, nearnodes, left = of CombinationPoints] (ConstrainedPoints)
+  { \begin{itemize}[leftmargin=.3cm]
+      \item Number of test treatments
+      \item Total subjects in trial
+      \item recruitment rate
+    \end{itemize}
+    \nodepart{second} Constrained by health care environment/infrastructure
+  };
+  \node[draw, above = of ConstrainedPoints,text width = 6cm] (DesignOptions) {\textbf{Disease-specific features}};
+  \node[draw, below = of ClinicalScenarios] (DesignPerformance) {
+    \textbf{Design performance} \\ e.g.\ statistical power};
+  \node[draw, right = of DesignPerformance] (Evaluation) {
+    \textbf{Clinical Scenario Evaluation}\\
+    Design performance measures evaluated across a wide range of clinical scenarios};
+  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
+  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
+  \draw[->] (CombinationPoints) -- (ClinicalScenarios);
+  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
+  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
+  \draw[->] (DesignPerformance) -- (Evaluation);
+\end{tikzpicture}
 %FIXME Is this the right paper? Convert to tikz! Tim schickt mir das Original!

### 2020-09-28 main.tex
improve tikz

--- 2020-09-28 main.tex
@@
 \end{figure}
+
+\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\scriptsize, outer sep= 0cm]
+  \node[draw, ellipse, minimum height=1cm, text width=2cm] (ClinicalScenarios) at (0,0) {
+    \textbf{Clinical scenarios}\\
+    Simulation Studies};
+  \node[name=block, rectangle split, rectangle split parts=2, draw, above = of ClinicalScenarios] (UnknownPoints)
+  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
+      \item Treatment effect on early interim outcome.
+      \item Treatment effect on late disability outcome
+    \end{itemize}
+    \nodepart{second} Unknown
+  };
+  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north east] (HistoricalDataPoints) at (UnknownPoints.north west)
+  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
+      \item Correlation between early interim and late disability outcomes.
+    \end{itemize}
+    \nodepart{second} Historical data
+  };
+  %\node[draw, inner xsep=1em, inner ysep=1em, above left = of ClinicalScenarios, fit = (HistoricalDataPoints) (UnknownPoints)] (boxDiseaseSpecific) {};
+  \node[draw, text width = 6.2cm, anchor = south west] (DiseaseSpecific) at (HistoricalDataPoints.north west) {\textbf{Disease-specific features}};
+  %\node[draw, above right = of ClinicalScenarios] (DesignOptions) {Design options};
+  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (ConstrainedPoints) at ([xshift = 4cm]HistoricalDataPoints.north east)
+  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
+      \item Number of test treatments
+      \item Total subjects in trial
+      \item recruitment rate
+    \end{itemize}
+    \nodepart{second} Constrained by health care environment/infrastructure
+  };
+  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (CombinationPoints) at (ConstrainedPoints.north east)
+  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
+      \item Adaptive seamless design time point of interim analysis
+      \item Treatment selection rules at interim Conventional phase 2/3
+      \item split of resources between phases
+      \item end oh phase futility criteria
+    \end{itemize}
+    \nodepart{second} Combination of options
+  };
+  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design Options}};
+  \node[draw, below = 0.5cm of ClinicalScenarios] (DesignPerformance) {
+    \textbf{Design performance} \\ e.g.\ statistical power};
+  \node[draw, right = of DesignPerformance, text width = 4cm, node distance = 2cm] (Evaluation) {
+    \textbf{Clinical Scenario Evaluation}\\
+    Design performance measures evaluated across a wide range of clinical scenarios};
+  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
+  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
+  \draw[->] (CombinationPoints) -- (ClinicalScenarios);
+  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
+  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
+  \draw[->] (DesignPerformance) -- (Evaluation);
+\end{tikzpicture}
 %FIXME Is this the right paper? Convert to tikz! Tim schickt mir das Original!

### 2020-09-28 main.tex
replace fig

@@
 \centering
-\includegraphics[width=\linewidth]{assets/friede_refinement_2010_fig2.png}
-\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis. REPLACE WITH TikZ}
-\label{fig:friede_refinement_2010_fig2}
-\end{figure}
-
 \begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\scriptsize, outer sep= 0cm]
@@
 %Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
+\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis.}
+\label{fig:friede_refinement_2010_fig2}
+\end{figure}
 

### 2020-10-05 main.tex
reread

@@
   \hline
-  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thr} \}$ \\
+  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thresh} \}$ \\
   $r$ (ratio)        & $(0,1)$ \\
@@
 \end{table}
+
+%FIXME: JR: uniform naming: eps, thersh for the values and epsilon and treshold for the strategies
 
@@
 For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
-We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed values for $n_{\text{stage1}}$.
+We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed value for $n_{\text{stage1}}$.
 Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total patients is closest to $n_{\text{treat}}$.
@@
 
-The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{thr}$ and $\text{eps}$.
+The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{threshold}$ and $\text{epsilon}$.
 Then the black-box from formula (\ref{eq:bbox}) can be reformulated as

### 2020-10-05 main.tex
update plots

@@
 \centering
-\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\scriptsize, outer sep= 0cm]
-  \node[draw, ellipse, minimum height=1cm, text width=2cm] (ClinicalScenarios) at (0,0) {
+\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\sffamily \scriptsize, outer sep= 0cm]
+  \node[draw, ellipse, minimum height=1cm, text width=2.5cm] (ClinicalScenarios) at (0,0) {
     \textbf{Clinical scenarios}\\
@@
 However, in Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behaviour of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio}.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
+\caption{%
+  Validated performance of both approaches.
+  }
+\label{fig:plot_boxplot_valid_y}
+\end{figure}
+
+\input{generated/tables/table_time.tex}
+
 \begin{figure}[htb]
@@
 
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
-\caption{%
-  Optimization curves of each optimization run and the mean of all runs drawn as a black line.
-  }
-\label{fig:plot_opt_path} 
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
-\caption{%
-  Validated performance of both approaches.
-  }
-\label{fig:plot_boxplot_valid_y}
-\end{figure}
-
 \input{generated/tables/table_best.tex}
-
-\input{generated/tables/table_time.tex}
 

### 2020-10-05 generator/main.R

@@ -9,6 +9,8 @@ library(knitr)
 library(kableExtra)
 library(dplyr)
 library(mlr3misc)
+library(grid)
+library(gridExtra)
 library(memoise)
 
 
@@ -205,8 +207,19 @@ plot_wrapper(name = "plot_best_x", fig.height = 1.5 * FIG_HEIGHT, expr = {
   g = g + scale_y_continuous(expand = expansion(mult = 0.2))
   g = g + scale_shape_manual(values = c("mbo" = 19, "grid" = 21))
   g = g + scale_color_manual(values = select_labels_colors)
-  g = g + theme(legend.position = "bottom")
-  g
+  g = g + theme(legend.position = "bottom", strip.background = element_blank(), strip.text.x = element_blank())
+  #grid headlines
+  col_heads = paste0("n_cases: ", unique(tmp$n_cases)) %>% lapply(textGrob, gp = gpar(fontsize = 12))
+  row_heads = levels(factor(tmp$effect)) %>% lapply(textGrob, gp = gpar(fontsize = 12), rot=90)
+  layout_mat = matrix(c(
+    NA, 1, 2, 3,
+     4, 8, 8, 8,
+     5, 8, 8, 8,
+     6, 8, 8, 8,
+     7, 8, 8, 8,
+    NA, 8, 8, 8
+  ), byrow = TRUE, ncol = 4)
+  grid.arrange(grobs=c(col_heads, row_heads, list(g)),layout_matrix=layout_mat, widths=c(1,10,10,10), heights=c(1,5,5,5,5,3))
 })
 
 

### 2020-10-06 benchmark/batchtools_grid.R
more repls for nsim high comparision

@@ -4,7 +4,7 @@ library(asd.mod)
 library(data.table)
 library(mlr3misc)
 
-TESTMODE = Sys.info()[["nodename"]] %in% c("JayBook2")
+TESTMODE = Sys.info()[["nodename"]] %in% c("JayBook2", "HERKULES5ARCH")
 
 if (TESTMODE) {
   #try out params
@@ -57,7 +57,7 @@ fn = function(
       stage2 = round((1-p) * const)
     )
   }
-
+  
   calculate_nselect = function(res) { #calculate the nselect from the results after the simulation ran
     if (!is.null(res$error)) {
       return(0)
@@ -119,7 +119,7 @@ fn = function(
         mean_select = getOptPathEl(res$opt.path, getOptPathBestIndex(res$opt.path))$extra$mean_select)
       return(n)
     }
-      = calibrate_n(x)
+    calib_res = calibrate_n(x)
     select2 = calib_res$mean_select
   }
   n = calculate_n(select2)
@@ -211,23 +211,26 @@ addAlgorithm("mbo", fun = function(job, data, instance, corr, nsim, n_cases, mbo
   res
 })
 
-const_design = expand.grid(n_cases = NCASES, nsim = NSIM, corr = 0.4, effect = names(EFFECTS), stringsAsFactors = FALSE)
-# add 5times sim steps for one experiment
-new_line = subset(const_design, n_cases == 2000 & effect == "paper")
-new_line$nsim = 5 * new_line$nsim
-const_design = rbind(const_design, new_line)
-eval_design = generateGridDesign(getParamSet(fun), GRIDRES)
-eval_design = subset(eval_design, stage_ratio != 1 & stage_ratio != 0)
-eval_design = merge.data.frame(eval_design, const_design)
-setDT(eval_design)
-
-mbo_design = cbind(const_design, mbo_iters = MBOITERS)
-ades = list(
-  eval = eval_design,
-  mbo = mbo_design
-)
-
+generate_eval_mbo_design = function(n_cases = NCASES, nsim = NSIM, effect = names(EFFECTS)) {
+  const_design = expand.grid(n_cases = n_cases, nsim = nsim, corr = 0.4, effect = effect, stringsAsFactors = FALSE)
+  eval_design = generateGridDesign(getParamSet(fun), GRIDRES)
+  eval_design = subset(eval_design, stage_ratio != 1 & stage_ratio != 0)
[…]

### 2020-10-06 generator/main.R

@@ -13,6 +13,9 @@ library(grid)
 library(gridExtra)
 library(memoise)
 
+root = rprojroot::find_root(rprojroot::is_git_root)
+setwd(paste0(root,"/generator"))
+
 
 FIG_WIDTH = 9
 FIG_HEIGHT = 4.5
@@ -36,7 +39,7 @@ kable_to_text = function(text, name) {
 
 own_theme = function(...) {
   theme_light(...) + 
-  theme(legend.margin=margin(0,0,0,0))
+  theme(legend.margin=margin(0,0,0,0), strip.background = element_rect(fill = "white", color = "white"), strip.text = element_text(color = "grey10"))
 }
 theme_set(own_theme())
 
@@ -176,7 +179,7 @@ max_dob = tmp$max_dob
 ## ----plot_allbest, fig.height = 1.5 * FIG_HEIGHT------------------------------
 # find best performing solution for each select methods and across all epsilon and threshold choices
 plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
-  res_ave = res_eval[nsim == 1000, list(mean_y = mean(y)), by = c(algo.par.names)]
+  res_ave = res_eval[nsim == 1000 & repl <= 10, list(mean_y = mean(y)), by = c(algo.par.names)]
   res_best = res_ave[, .SD[order(-mean_y)[1], ], by = c("select", algo.par.names.meta)]
   res_best = res_best[, !c("mean_y", "stage_ratio")]
   # merge so the plot only contains the best curves (applies for epsilon and threshold rule)
@@ -185,7 +188,7 @@ plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
   g = ggplot(df, aes(x = (stage_1_arms * stage_1_n), y = y, color = select, group = paste(select, epsilon, thresh)))
   g = g + geom_line(data = dfmean)
   g = g + geom_point(alpha = 0.2, size = 0.5)
-  g = g + geom_point(data = res_mbo[nsim == 1000,], size = 2.5)
+  g = g + geom_point(data = res_mbo[nsim == 1000 & repl <= 10,], size = 2.5)
   g = g + scale_color_manual(values = select_labels_colors)
   g = g + facet_grid(effect~n_cases, scales = "free", labeller = label_both)
[…]

### 2020-10-06 benchmark/batchtools_grid.R
update plots

@@ -252,23 +252,23 @@ if (FALSE) {
   reg$status[ids_expired[is_ready == TRUE, .(job.id)], done := Sys.time()]
 }
 
-jdf = getJobTable(findDone())
-res_eval = reduceResultsDataTable(jdf[algorithm == "eval", ], fun = function(res, ...) {
+jdf_done = getJobTable(findDone())
+res_eval = reduceResultsDataTable(jdf_done[algorithm == "eval", ], fun = function(res, ...) {
   extras = attr(res, "extras")
   extras = lapply(extras, function(x) {attributes(x) = NULL; x}) #clean extras
   c(y = res[[1]], extras)
 })
 res_eval = unwrap(res_eval, "result")
-res_eval = ljoin(jdf[algorithm == "eval", ], res_eval)
+res_eval = ljoin(jdf_done[algorithm == "eval", ], res_eval)
 saveRDS(res_eval, "batchtools_grid_res_eval.rds")
 
-res_mbo = reduceResultsDataTable(jdf[algorithm == "mbo" & map(algo.pars, "effect") %in% names(EFFECTS),], fun = function(res) {
+res_mbo = reduceResultsDataTable(jdf_done[algorithm == "mbo" & map(algo.pars, "effect") %in% names(EFFECTS),], fun = function(res) {
   opdf = as.data.table(res$opt.path)
-  opdf[, c("error.model", "eol") := NULL]
+  opdf[, c("error.model", "eol", "aei", "train.time", "propose.time", "se", "mean", "tau") := NULL]
   best_el = getOptPathEl(res$opt.path, index = res$best.ind)
-  c(res$x, y = res$y, best_el$extra[c("stage_1_arms", "stage_1_n","stage_2_arms", "stage_2_n")], opt.path = list(opdf))
+  c(res$x, y = res$y, best_index = res$best.ind, best_el$extra[c("stage_1_arms", "stage_1_n","stage_2_arms", "stage_2_n")], opt.path = list(opdf))
 })
-res_mbo = ljoin(jdf[algorithm == "mbo", ], res_mbo)
+res_mbo = ljoin(jdf_done[algorithm == "mbo", ], res_mbo)
 saveRDS(res_mbo, "batchtools_grid_res_mbo.rds")
 
 if (FALSE) {

### 2020-10-06 generator/main.R

@@ -127,7 +127,7 @@ read_results_uncached = function(select_labels) {
       valid_dt = res_eval[repl != this_repl, ]
       res = merge(optim_dt[,-"y"], valid_dt[, c(algo.par.names, "y"), with = FALSE], by = algo.par.names)
       # write result into x values dt
-      optim_dt$y = mean(res$y)
+      optim_dt$y_valid = mean(res$y)
       #strangely we have 3 to 4 missing values here, probably batchtools had a hick up
       times = .SD$time.running
       if (any(is.na(times)) && mean(is.na(times)) < 0.004) {
@@ -147,7 +147,10 @@ read_results_uncached = function(select_labels) {
   # determine max dob (probably 101) because it contains outside mbo evals
   max_dob = max(map_int(map(res_mbo$opt.path, "dob"), max))
 
-  res_mbo[, time.running := time.running - as.difftime(sum(.SD$opt.path[[1]][dob == max_dob, ]$exec.time), units = "secs") , by = rownames(res_mbo)]
+  res_mbo[, time.running := time.running - as.difftime(sum(.SD$opt.path[[1]][prop.type == "final_eval", ]$exec.time), units = "secs") , by = rownames(res_mbo)]
+  data.table::setnames(res_mbo, "y", "y_valid")
+  res_mbo[, y := .SD$opt.path[[1]][.SD$best_index, ]$y, by = rownames(res_mbo)]
+  
 
   list(res_mbo = res_mbo, res_grid = res_grid, res_eval = res_eval, algo.par.names = algo.par.names, algo.par.names.meta = algo.par.names.meta, algo.par.names.optim = algo.par.names.optim, max_dob = max_dob)
 }
@@ -191,6 +194,7 @@ plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
   g = g + geom_point(data = res_mbo[nsim == 1000 & repl <= 10,], size = 2.5)
   g = g + scale_color_manual(values = select_labels_colors)
   g = g + facet_grid(effect~n_cases, scales = "free", labeller = label_both)
+  g = g + labs(x = expression(k[1] %.% n[stage1]))
   g = g + theme(legend.position = "bottom")
   g
 })
[…]

### 2020-10-06 generator/main.R
fix boxplot

@@ -303,7 +303,7 @@ plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT *
   g = g + scale_fill_manual(values = algorithm_labels_color) + scale_color_manual(values = darker_colors)
   g = g + theme(legend.position = "bottom")
   g = g + geom_boxplot()
-  g = g + labs(x = expression(n[sim]), y = expression(y[valid]), color = NULL)
+  g = g + labs(x = expression(n[sim]), y = expression(y[valid]), color = NULL, fill = NULL)
   g
 })
 

### 2020-10-07 benchmark/_functions.R
calc grid results for mbo

@@ -1,0 +1,136 @@
+## Define Problem
+
+fn = function(
+  x, 
+  n_cases = 1000, 
+  effect = list(early=c(0,0.68,0.82,0.95,0.91),final=c(0,0.13,0.17,0.23,0.20)), 
+  corr = 0.4, 
+  nsim = 100) {
+  
+  if (is.character(effect)) {
+    effect = dictionaries$effects[[effect]]
+  }
+  k_plus_1 = length(effect$early)
+  p = x$stage_ratio
+  
+  calculate_n = function(n_select) { # calculate n for each arm in stage1 and stage2 given the number of arrms in stage2, and the ratio p
+    const = n_cases/(k_plus_1 * p + (n_select+1) * (1-p))
+    list(
+      stage1 = round(p * const), 
+      stage2 = round((1-p) * const)
+    )
+  }
+  
+  calculate_nselect = function(res) { #calculate the nselect from the results after the simulation ran
+    if (!is.null(res$error)) {
+      return(0)
+    }
+    as.numeric(sum(res$res.5$sim.res$count.total * seq_along(res$res.5$sim.res$count.total)) / res$res.1$simulations["n"])
+  }
+  
+  # wrapper to safely run treatsel.sim (necessary for init design)
+  treatsel_sim_safe = function(n, effect, nsim, corr, seed = sample(.Machine$integer.max, 1), select, level = 0.025, ptest = c(3, 4), epsilon = epsilon, thresh = thresh, ...) {
+    capture.output({
+      res = tryCatch({
+        treatsel.sim(n = n, effect = effect, outcome = list(early = "N", final = "N"), nsim = nsim, corr = corr, seed = seed, select = select, level = level, ptest = ptest, epsilon = epsilon, thresh = thresh, ...)
+        # on error return result that leads to power 0
+      }, error = function(e) list(res.5 = list(sim.res = list(sim.reject = 0), sim.res.n = nsim), error = as.character(e)))
+    })
+    return(res)
+  }
+  if (x$select %in% c(0,1:3,5)) { # all go to second stage
+    select2 = switch(as.character(x$select),
+                     "0" = k_plus_1-1,
+                     "1" = 1,
+                     "2" = 2,
+                     "3" = 3,
+                     "5" = 1) #1 random
+  } else if (x$select %in% c(4,6)) { # select arms according to thresold of p value or epsilon # 4 and 6
[…]

### 2020-10-07 benchmark/_parameters.R

@@ -1,0 +1,36 @@
+library(batchtools)
+library(smoof)
+library(asd.mod)
+library(data.table)
+library(mlr3misc)
+
+TESTMODE = Sys.info()[["nodename"]] %in% c("JayBook2", "HERKULES5ARCH")
+
+if (TESTMODE) {
+  #try out params
+  NSIM = 50
+  MBOITERS = 10
+  GRIDRES = 4
+  REPLS = 1
+  NCASES = c(1000,2000)
+  EFFECTS = list(
+    paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20))
+  )
+} else {
+  #real params
+  NSIM = 1000
+  MBOITERS = 100
+  GRIDRES = 25
+  REPLS = 10
+  NCASES = c(500,1000,2000)
+  EFFECTS = list(
+    paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20)),
+    linear = {x = c(0,2,4,6,8)/10; list(early = x, final = x/4)},
+    #threashold = {x = c(0,0,0,4,8)/10; list(early = x, final = x/4)},
+    #saturation = {x = c(0,5,7,7.8,8)/10; list(early = x, final = x/4)},
+    sigmoid = {x = c(0,1,2,7,8)/10; list(early = x, final = x/4)},
+    #paper_mod = list(early = c(0,0.68,0.82,0.93,0.93), final = c(0,0.13,0.17,0.23,0.20)),
+    paper2 = list(early = c(0,0.68,0.82,0.95,0.91), final = 2*c(0,0.13,0.17,0.23,0.20))# ,
+    #paper_mod2 = list(early = c(0,0.68,0.82,0.93,0.93), final = 2*c(0,0.13,0.17,0.23,0.20))
+  )
+}

### 2020-10-07 benchmark/batchtools_grid.R

@@ -1,152 +1,8 @@
-library(batchtools)
-library(smoof)
-library(asd.mod)
-library(data.table)
-library(mlr3misc)
+source("_parameters.R")
 
-TESTMODE = Sys.info()[["nodename"]] %in% c("JayBook2", "HERKULES5ARCH")
+# source functions
 
-if (TESTMODE) {
-  #try out params
-  NSIM = 50
-  MBOITERS = 10
-  GRIDRES = 4
-  REPLS = 1
-  NCASES = c(1000,2000)
-  EFFECTS = list(
-    paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20))
-  )
-} else {
-  #real params
-  NSIM = 1000
-  MBOITERS = 100
-  GRIDRES = 25
-  REPLS = 10
-  NCASES = c(500,1000,2000)
-  EFFECTS = list(
-    paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20)),
-    linear = {x = c(0,2,4,6,8)/10; list(early = x, final = x/4)},
-    #threashold = {x = c(0,0,0,4,8)/10; list(early = x, final = x/4)},
-    #saturation = {x = c(0,5,7,7.8,8)/10; list(early = x, final = x/4)},
-    sigmoid = {x = c(0,1,2,7,8)/10; list(early = x, final = x/4)},
-    #paper_mod = list(early = c(0,0.68,0.82,0.93,0.93), final = c(0,0.13,0.17,0.23,0.20)),
-    paper2 = list(early = c(0,0.68,0.82,0.95,0.91), final = 2*c(0,0.13,0.17,0.23,0.20))# ,
-    #paper_mod2 = list(early = c(0,0.68,0.82,0.93,0.93), final = 2*c(0,0.13,0.17,0.23,0.20))
-  )
-}
-
-## Define Problem
-
-fn = function(
-  x, 
-  n_cases = 1000, 
-  effect = list(early=c(0,0.68,0.82,0.95,0.91),final=c(0,0.13,0.17,0.23,0.20)), 
-  corr = 0.4, 
-  nsim = 100) {
-
-  if (is.character(effect)) {
-    effect = dictionaries$effects[[effect]]
-  }
-  k_plus_1 = length(effect$early)
-  p = x$stage_ratio
-
[…]

### 2020-10-07 generator/main.R

@@ -24,7 +24,7 @@ CACHE = cache_filesystem(path = "memoise")
 
 plot_wrapper_uncached = function(name, fig.height = FIG_HEIGHT, fig.width = FIG_WIDTH, expr) {
   path = paste0("../generated/figures/", name, ".pdf")
-  pdf(file = path, width = fig.width, height = fig.height, onefile = TRUE)
+  cairo_pdf(filename = path, width = fig.width, height = fig.height, onefile = TRUE)
   print(eval(expr))
   dev.off()
   #browser()
@@ -96,7 +96,7 @@ kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation") %>%
 read_results_uncached = function(select_labels) {
   res_eval = readRDS("../benchmark_results/batchtools_grid_res_eval.rds")
   res_eval = res_eval[,.(job.id, y, stage_1_arms, stage_1_n, stage_2_arms, stage_2_n, repl, problem, prob.pars, algorithm, algo.pars, time.running)]
-  res_eval = res_eval[map(algo.pars, "effect") %in% names(EFFECTS),]
+  res_eval = res_eval[map_chr(algo.pars, "effect") %in% names(EFFECTS),]
 
   algo.par.names = c("algorithm", names(res_eval$algo.pars[[1]]))
   algo.par.names.meta = c("algorithm", "effect", "corr", "nsim", "n_cases")
@@ -107,7 +107,7 @@ read_results_uncached = function(select_labels) {
 
   res_mbo = readRDS("../benchmark_results/batchtools_grid_res_mbo.rds")
   res_mbo = res_mbo[,.(job.id, result, repl, problem, prob.pars, algorithm, algo.pars, time.running)]
-  res_mbo = res_mbo[map(algo.pars, "effect") %in% names(EFFECTS),]
+  res_mbo = res_mbo[map_chr(algo.pars, "effect") %in% names(EFFECTS),]
   res_mbo = unnest(res_mbo, cols = c("result", "prob.pars", "algo.pars"))
   
   res_eval[, select := factor(select, levels = names(select_labels), labels = select_labels)]
@@ -140,6 +140,23 @@ read_results_uncached = function(select_labels) {
     by = c(algo.par.names.meta, "repl"), 
     .SDcols = c("y", "repl", algo.par.names, "time.running")]
   res_grid[, algorithm := "grid"]
+  
+  ## calculate y result on grid
[…]

### 2020-10-07 main.tex
update plots
calc grid results for mbo
improve plots, mbo grid

@@
 \usepackage{enumitem}
+\usepackage{placeins} %floatbarrier
 % tikz
@@
 \Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}
+
+\makeatletter
+\let\newtitle\@title
+\makeatother
 
@@
   $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
-  \emph{eps}     & Select all arms with test statistic less than \emph{eps} (epsilon) below maximal test statistic \\
-  \emph{thresh}     & Select all arms with test statistic larger than the value \emph{thresh} (threshold) \\
+  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
+  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
   \hline
@@
   $r$ (ratio)                      & $(0,1)$ \\
-  eps (epsilon)            & $[0,4]$ \\
-  thresh (threshold)          & $[0,10]$ \\
+  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
+  $\tau$ (onlt for \emph{thresh})  & $[0,10]$ \\
   \hline
@@
 
-\input{generated/tables/table_best.tex}
-
 \subsection{More simulation rounds}
@@
 \phantom{aaaa}
+%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
+\clearpage
+\begin{center}
+\textbf{\large Supplemental Materials: \newtitle}
+\end{center}
+\FloatBarrier
+%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
+%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
+\setcounter{equation}{0}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\setcounter{page}{1}
+\makeatletter
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\bibnumfmt}[1]{[S#1]}
+%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
+
+\input{generated/tables/table_best.tex}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
+\caption{%
+  Optimization curves of each optimization run and the mean of all runs drawn as a black line.
+  }
+\label{fig:plot_opt_path} 
+\end{figure}
+
 \end{document}

### 2020-10-07 generator/main.R

@@ -76,8 +76,8 @@ select_labels_colors = #http://paletton.com/#uid=74Z140kqdu7ghF3lowyu1ppwGk7
     "random" = "#000000",
     "threshold rule" = "#C0E62A"
   )
-algorithm_labels = c("mbo" = "MBO", "grid" = "Grid")
-algorithm_labels_color = c("mbo" = "#CB1C00", "grid" = "#00962D")
+algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "grid" = "Grid")
+algorithm_labels_color = c("mbo" = "#F04A2B", "eval" = "#099438", "mbo grid" = "#690D86", "grid" = "#099438")
 
 ## ----table_effect_names-------------------------------------------------------
 tmp = lapply(EFFECTS, do.call, what = rbind)
@@ -182,6 +182,38 @@ algo.par.names.meta = tmp$algo.par.names.meta
 algo.par.names.optim = tmp$algo.par.names.optim
 max_dob = tmp$max_dob
 
+cases = list(
+  default = quote(nsim == 1000 & repl <= 10),
+  nsim = quote(n_cases == 2000 & effect == "paper")
+)
+
+get_res = function(dt = res_eval, case = "default") {
+  dt[eval(cases[[case]]),]
+}
+
+get_res_grid = function(case = "default") {
+  get_res(res_grid, case)
+}
+
+get_res_mbo = function(case = "default") {
+  get_res(res_mbo, case)
+}
+
+get_res_eval = function(case = "default") {
+  get_res(res_eval, case)
+}
+
+get_res_mbogrid = function(case = "default") {
+  tmp_mbogrid = get_res_mbo(case = case)
+  tmp_mbogrid[, algorithm := "mbo grid"]
+  tmp_mbogrid[, y_valid := NULL]
+  setnames(tmp_mbogrid, "y_grid", "y_valid")
+}
+
+get_res_eval_average = function(case = "default") {
+  get_res_eval(case = case)[, list(mean_y = mean(y)), by = algo.par.names]
+}
+
 ## ----put_ext_eval_mbo_results, eval = FALSE-----------------------------------
 ## res_mbo[, y_tuning := y]
 ## map_dbl(res_mbo$opt.path, function(op) mean(op[dob == max(dob), y])) == res_mbo$y_tuning
@@ -199,7 +231,7 @@ max_dob = tmp$max_dob
 ## ----plot_allbest, fig.height = 1.5 * FIG_HEIGHT------------------------------
[…]

### 2020-10-07 main.tex
some result text

@@
   \label{eq:bbox}
-  y & = f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
+  y & = f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{selection strategy}, \epsilon, \tau).
 \end{align}
@@
 This is only the case for selection strategies 1, 2, and 3.
-For the selection strategies \emph{threshold} and \emph{epsilon}, $k_2$ depends on the outcome of stage 1.
+For the selection strategies \emph{eps} and \emph{thresh}, $k_2$ depends on the outcome of stage 1.
 %choice of the \emph{thresh} or \emph{eps} parameter.
-In these cases, for given values of $n_{\text{stage1}}$ and \emph{thresh} or \emph{eps}, respectively, we can  obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
+In these cases, for given values of $n_{\text{stage1}}$ and $\epsilon$ or $\tau$, respectively, we can  obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
 For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
@@
 \begin{equation}
-  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
+  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \epsilon, \tau).
 \end{equation}
@@
 
-Summarizing, in the following we aim to find the clinical design setup, represented by the parameter values for $r$, \emph{effect set}, \emph{Selection strategy}, \emph{epsilon} and \emph{threshold} that yield the highest statistical power, when evaluation $\tilde{f}$.
+Summarizing, in the following we aim to find the clinical design setup, represented by the parameter values for $r$, \emph{effect set}, \emph{Selection strategy}, $\epsilon$ and $\tau$ that yield the highest statistical power, when evaluating $\tilde{f}$.
 
@@
 For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.
-one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and \emph{eps} or \emph{thresh}, respectively).
+one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and $\epsilon$ or $\tau$, respectively).
 This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
-Additionally, the grid is evaluated with 10 stochastic replictions to estimate the variance of the estimated power.
+Additionally, the grid is evaluated with 10 stochastic replications to estimate the variance of the estimated power.
 
@@
 To account for this optimism and report unbiased performance estimates, for each of the 10 replicate simulations, we identify the best configuration and consider the corresponding 9 values of the other 9 replicate simulations, similar to a cross-validation framework in machine learning applications.
-\\
 %This eliminates the risk of only reporting outcomes that are optimal by chance.
@@
 Next, we describe our realization of the search with the \emph{MBO (Model-based optimization)} approach.
-For the application of MBO, we first have to define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter sets) is generated.
-As the regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is well known to work well for fairly low dimensional search spaces.
+For the application of MBO, we first have to define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
+As the regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
 We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
@@
 First, to transform categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
-Furthermore, our search space is hierarchical because the parameters \emph{epsilon} and \emph{threshold} are only active for the respective \emph{selection strategies}.
-The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter of \emph{thresh} is inactive.
-If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if \emph{thresh} is inactive it will be set to 20).
+Furthermore, our search space is hierarchical because the parameters $\epsilon$ and $\tau$ are only active for the respective selection strategies \emph{eps} and \emph{thresh}.
+The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter $\tau$ of \emph{thresh} is inactive.
+If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if $\tau$ is inactive it will be set to 20).
 This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
@@
 As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:model_based_optimization}.
-We start MBO with an inital design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator.
+We start MBO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
 
-At the end of the optimization procedure, the best point is chosen according to the best prediction obtained on the set of evaluated points as explained in Section~\ref{ssec:best_point}.
-Again, overoptimism must be avoided. Thus, the best selected point (the best parameter setting) is then independently evaluated $10$ times again.
+At the end of the optimization procedure, the best configuration is chosen according to the best prediction obtained on the set of evaluated points as explained in Section~\ref{ssec:best_point}.
+Again, overoptimism must be avoided. 
+Thus, the best selected configuration (the best parameter set) is then independently evaluated $10$ times again.
 
 The whole MBO procedure is repeated $10$ times to ensure that a single optimization run was not just lucky in finding the optimum.
+
+Additionally we want to compare whether the results of \emph{MBO} are different from the \emph{grid search} just because \emph{MBO} can propose configurations on within the whole search space and not only the grid.
+Therefore we introduce \textbf{MBO Grid}.
+The optimization is executed as described for \emph{MBO} but with one difference:
+At the end we conduct the independent evaluation for the configuration within the \emph{grid} that is closest to the configuration that is proposed by \emph{MBO}.
+% Mention that only configurations of the same factor level are looked at?
+
+We expect that \emph{MBO} achieves to find configurations that yield better results than the \emph{grid search} for two reasons.
+First, \emph{MBO} will be able to find an optimal configuration that is not restricted to be within the grid but can be between points from the grid.
+Second, \emph{MBO} will not choose configurations that were observed to perform well only due to noise.
 
@@
 For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package asd~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
-The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???
-
 \begin{table}[h]
@@
 \end{table}
+The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
+Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $\tilde{f}$. 
+The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{MBO} are additional replicates.
 
@@
 
-As explained before, we obtained unbiased result from the exhaustive grid search and the MBO procedure.
-However, in Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behaviour of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio}.
-
+As explained before, we obtained unbiased results from the exhaustive grid search and the MBO procedure.
+Figure~\ref{fig:plot_boxplot_valid_y} shows that \emph{MBO} yields better results for each effect set even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{grid search}.
 \begin{figure}[htb]
@@
 \end{figure}
+The difference in the number of evaluations is also reflected in the average runtime of each optimization methods given in Table~\ref{tab:table_time}.
+\input{generated/tables/table_time.tex}
+The good performance of \emph{MBO Grid} indicates that the resolution of the grid is sufficiently high to find a good configuration.
+Accordingly, the good performance of \emph{MBO} can be contributed mostly to the fact that the best configuration is determined by the surrogates mean estimation es explained in Section~\ref{ssec:best_point}.
+In other words, the \emph{grid search} is not only extremely demanding in terms of necessary black-box evaluations but also does take into account the stochasticity of the problem.
+This seemingly leads to proposals of configurations that were observed to yield a high power but that do not perform well, if evaluated independently.
+We will later study on single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of the \emph{gird search}.
 
-\input{generated/tables/table_time.tex}
+However, in Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behavior of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio}.
 

### 2020-10-12 main.tex
result desc

@@
 
-Additionally we want to compare whether the results of \emph{MBO} are different from the \emph{grid search} just because \emph{MBO} can propose configurations on within the whole search space and not only the grid.
+Additionally we want to investigate whether the results of \emph{MBO} are better then \emph{grid search} because \emph{MBO} can propose configurations within the whole search space and not only limited the grid.
 Therefore we introduce \textbf{MBO Grid}.
@@
 \input{generated/tables/table_time.tex}
+For the chosen grid resoultion and evaluation budget for \emph{MBO}, \emph{MBO} is approxamatly 20-times faster and yields better configurations.
+
 The good performance of \emph{MBO Grid} indicates that the resolution of the grid is sufficiently high to find a good configuration.
 Accordingly, the good performance of \emph{MBO} can be contributed mostly to the fact that the best configuration is determined by the surrogates mean estimation es explained in Section~\ref{ssec:best_point}.
-In other words, the \emph{grid search} is not only extremely demanding in terms of necessary black-box evaluations but also does take into account the stochasticity of the problem.
+In other words, the \emph{grid search} is not only extremely demanding in terms of necessary black-box evaluations but also does not take into account the stochasticity of the problem.
 This seemingly leads to proposals of configurations that were observed to yield a high power but that do not perform well, if evaluated independently.
@@
 
-However, in Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behavior of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio}.
-
+After having only looked at the outcome of our probelms and established the superiority of \emph{MBO} over the \emph{grid search} we want to investigate the problems in more detail.
+In Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behavior of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} $r$.
 \begin{figure}[htb]
@@
 \includegraphics[width=\linewidth]{generated/figures/plot_allbest.pdf}
-\caption{Power of the different selection rules. For the epsilon and threshold rule only the curve for the epsilon and threshold value that reached the best power is dispayed.}
+\caption{%
+  All evaluated configurations and their outcomes from the \emph{grid search}.
+  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power is displayed.
+  For each configuration in the grid we obtained 10 stochastic results.
+  The line connects the mean outcomes for each configuration.
+  Additionally the best found configurations from \emph{MBO} are displayed.
+  }
 \label{fig:plot_allbest}
@@
 Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments at the early stage.
-Also, for the \emph{selection strategy} \emph{epsilon rule} and \emph{threshold rule} only the curves are shown of the repsective \emph{eps} and \emph{thresh} value that led to the best outcome within the respective selection strategy.
+Also, for the \emph{selection strategy} \emph{epsilon rule} and \emph{threshold rule} only the curves are shown of the respective \emph{eps} and \emph{thresh} value that led to the best outcome within the respective selection strategy.
+For the selection strategies \emph{eps} and \emph{thresh}, we only show the curve of the one $\epsilon$ and $\tau$ setting that led to the optimal outcome for the respective selection strategy.
+Instead of showing $r$ on the $x$-Axis we show the resulting value of treatments at stage 1 as this allows for easier comparison between the different selection strategies.
+Also note that for effect set \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
+In accordance with intuition, for most scenarios it is the worst decision to use all available treatments in stage 1. 
+In all scenarios \emph{MBO} was able to identify the peak of all curves in combination.
+Within the 10 stochastic repetitions \emph{MBO} found different configurations that all yield similarly good outcomes.
+That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.
+different configurations yield similarly good outcomes.
+In case multiple optimization runs give different configurations it might be advisable to let a human expert choose the final configuration.
 
+In contrast to Figure~\ref{fig:plot_allbest}, Figure~\ref{fig:plot_best_x} shows the outcome of the independent validation and the ratio $r$ directly.
 \begin{figure}[htb]
@@
 \end{figure}
+The results show that in many cases the grid search returned configurations of $r$ that are far from the solutions that \emph{MBO} found.
 

### 2020-10-12 generator/main.R

@@ -389,7 +389,7 @@ plot_wrapper(name = "plot_opt_path_5000", fig.height = 1.6 * FIG_HEIGHT * 0.35, 
   g = g + facet_grid(.~nsim, scales = "free")
   g = g + geom_hline(yintercept = 0 , color = colorspace::darken(algorithm_labels_color[["grid"]], amount = 0.6))
   g = g + geom_label(data = df_best, aes(label = formatC(best_y, 4)), y = 0, x = 90, vjust = 1.5, show.legend = FALSE)
-  g = g + scale_color_manual(values = algorithm_labels_color)
+  g = g + scale_color_manual(labels = algorithm_labels, values = algorithm_labels_color)
   g = g + theme(legend.position = "bottom")
   g = g + labs(x = "iteration", y = expression(y[iter]*"*" - y[grid]*"*"), color = NULL)
   g = g + guides(colour = guide_legend(override.aes = list(size=2)))

### 2020-10-13 generator/main.R
fix grid eval

@@ -123,7 +123,8 @@ read_results_uncached = function(select_labels) {
     {
       optim_dt = copy(.SD)
       this_repl = optim_dt$repl[1]
-      optim_dt = optim_dt[sample(which(y == max(y)), 1),]
+      optim_dt = optim_dt[y == max(y),]
+      if (nrow(optim_dt) > 1) optim_dt = optim_dt[sample.int(nrow(optim_dt), 1),] # in case of multiple best
       valid_dt = res_eval[repl != this_repl, ]
       res = merge(optim_dt[,-"y"], valid_dt[, c(algo.par.names, "y"), with = FALSE], by = algo.par.names)
       # write result into x values dt
@@ -332,20 +333,22 @@ create_boxplot_df = function(case = "default") {
   )
 }
 ## ----plot_boxplot_valid_y-----------------------------------------------------
-plot_wrapper(name = "plot_boxplot_valid_y", fig.height = FIG_HEIGHT * 0.5, expr = {
+plot_wrapper(name = "plot_boxplot_valid_y", fig.height = FIG_HEIGHT * 0.9, expr = {
   tmp = create_boxplot_df()
-  g = ggplot(tmp, aes(x = as.factor(n_cases), y = y_valid, color = algorithm, fill = algorithm))
+  g = ggplot(tmp, aes(x = algorithm, y = y_valid, color = algorithm, fill = algorithm))
   # mylabels = function(labels) {
   #   do.call(map, args = c(list(paste), labels))
   #   map(paste, labels[[1]], labels[[2]])
   #   paste(value, variable)
   # }
-  g = g + facet_wrap(~effect, scales = "free", ncol = 4, labeller = label_both)
+  g = g + facet_wrap(effect~n_cases, scales = "free_y", ncol = 6, labeller = label_both)
   darker_colors = colorspace::darken(algorithm_labels_color, amount = 0.6)
   names(darker_colors) = names(algorithm_labels_color)
   g = g + scale_fill_manual(labels = algorithm_labels, values = algorithm_labels_color) + scale_color_manual(labels = algorithm_labels, values = darker_colors)
+  g = g + scale_x_discrete(labels = algorithm_labels)
   g = g + geom_boxplot() + theme(legend.position = "right")
-  g = g + labs(x = expression(n[treat]), y = expression(y[valid]), color = NULL, fill = NULL)
[…]

### 2020-10-14 generated/tables/table_time.tex
fix labels
tables names
update plots
grid7

@@
 
-\caption{\label{tab:}Average runtime in hours, for evaluating one grid and one optimization run of MBO.}
+\caption{\label{tab:table_time}Average runtime in hours, for evaluating one grid and one optimization run of MBO.}
 \centering
@@
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{linear} & \multicolumn{3}{c}{paper} & \multicolumn{3}{c}{paper2} & \multicolumn{3}{c}{sigmoid} \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
+\multicolumn{1}{c}{ } & \multicolumn{4.5}{c}{linear} & \multicolumn{4.5}{c}{paper} & \multicolumn{4.5}{c}{paper2} & \multicolumn{4.5}{c}{sigmoid} \\
+\cmidrule(l{3pt}r{3pt}){2-5.5} \cmidrule(l{3pt}r{3pt}){6.5-10} \cmidrule(l{3pt}r{3pt}){11-14.5} \cmidrule(l{3pt}r{3pt}){15.5-19}
  & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000\\
 \midrule
-grid & 54.8 & 60.9 & 60.1 & 62.7 & 62.9 & 66.2 & 59.3 & 64 & 66.6 & 54.3 & 52.8 & 60.5\\
-mbo & 3.7 & 3.9 & 3.7 & 4.8 & 4.5 & 4.3 & 3.9 & 4 & 3.7 & 3.4 & 3.1 & 3.9\\
+Grid & 55.1 & 60.9 & 60.5 & 62.7 & 63.2 & 66.6 & 59.7 & 64.4 & 67.1 & 54.6 & 52.8 & 61.0\\
+Grid 7 & 3.4 & 3.8 & 3.7 & 3.8 & 3.9 & 4.1 & 3.8 & 3.9 & 4.1 & 3.4 &  & 3.7\\
+MBO & 3.6 & 3.4 & 3.6 & 3.4 & 4.0 & 4.1 & 3.7 & 3.3 & 3.2 & 3.3 & 2.6 & 3.1\\
 \bottomrule

### 2020-10-14 benchmark/_functions.R

@@ -121,9 +121,9 @@ addToEnvironment = function(f, ...) {
   return(f)
 }
 
-generate_eval_mbo_design = function(n_cases, nsim, effect) {
+generate_eval_mbo_design = function(n_cases, nsim, effect, par_set = getParamSet(fun), grid_res = GRIDRES) {
   const_design = expand.grid(n_cases = n_cases, nsim = nsim, corr = 0.4, effect = effect, stringsAsFactors = FALSE)
-  eval_design = generateGridDesign(getParamSet(fun), GRIDRES)
+  eval_design = generateGridDesign(par_set, grid_res)
   eval_design = subset(eval_design, stage_ratio != 1 & stage_ratio != 0)
   eval_design = merge.data.frame(eval_design, const_design)
   setDT(eval_design)

### 2020-10-14 generator/main.R

@@ -76,8 +76,8 @@ select_labels_colors = #http://paletton.com/#uid=74Z140kqdu7ghF3lowyu1ppwGk7
     "random" = "#000000",
     "threshold rule" = "#C0E62A"
   )
-algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "grid" = "Grid")
-algorithm_labels_color = c("mbo" = "#F04A2B", "eval" = "#099438", "mbo grid" = "#690D86", "grid" = "#099438")
+algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "grid" = "Grid", "grid7" = "Grid 7")
+algorithm_labels_color = c("mbo" = "#F04A2B", "eval" = "#099438", "mbo grid" = "#690D86", "grid" = "#099438", "grid7" = "#44CA6D") #https://paletton.com/#uid=6060R0krPwadDMilcBuwcpxDti6
 
 ## ----table_effect_names-------------------------------------------------------
 tmp = lapply(EFFECTS, do.call, what = rbind)
@@ -119,28 +119,50 @@ read_results_uncached = function(select_labels) {
   # 2. we obtain the y values of the given x-values of the other 9 repls
   # 3. we average the external y values
   # 4. we sum up the time that the grid needed on one replication
-  res_grid = res_eval[,
-    {
-      optim_dt = copy(.SD)
-      this_repl = optim_dt$repl[1]
-      optim_dt = optim_dt[y == max(y),]
-      if (nrow(optim_dt) > 1) optim_dt = optim_dt[sample.int(nrow(optim_dt), 1),] # in case of multiple best
-      valid_dt = res_eval[repl != this_repl, ]
-      res = merge(optim_dt[,-"y"], valid_dt[, c(algo.par.names, "y"), with = FALSE], by = algo.par.names)
-      # write result into x values dt
-      optim_dt$y_valid = mean(res$y)
-      #strangely we have 3 to 4 missing values here, probably batchtools had a hick up
-      times = .SD$time.running
-      if (any(is.na(times)) && mean(is.na(times)) < 0.004) {
-        times[is.na(times)] = mean(times, na.rm = TRUE)
-      }
-      optim_dt$time.running = sum(times)
-      optim_dt[, c(algo.par.names.meta, "repl") := NULL]
-      optim_dt
-    }, 
+  calc_res_grid_internal = function(.SD) {
[…]

### 2020-10-21 main.tex
anything?

@@
 \centering
-  \begin{minipage}{0.64\textwidth}
+  \begin{minipage}{0.62\textwidth}
     \centering
     \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
-    \captionof{figure}{Opt Path}
+    \captionof{figure}{Optimization progress for \emph{MBO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
     \label{fig:plot_opt_path_5000}
-  \end{minipage}\hspace{0.01\textwidth}%
+  \end{minipage}\hspace{0.02\textwidth}%
   \begin{minipage}{0.35\textwidth}
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
-    \captionof{figure}{Boxplot}
+    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_treat = 2000$ with 100 stochastic repititions and different values for \texttt{nsim}}
     \label{fig:plot_boxplot_valid_y_5000}

### 2020-10-21 main.tex
grid7 ...

@@
 As explained before, we obtained unbiased results from the exhaustive grid search and the MBO procedure.
-Figure~\ref{fig:plot_boxplot_valid_y} shows that \emph{MBO} yields better results for each effect set even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{grid search}.
+Figure~\ref{fig:plot_boxplot_valid_y} shows that \emph{MBO} yields better results for each effect set even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{grid search} (126 by \emph{Grid 7}).
 \begin{figure}[htb]

### 2020-10-21 generator/main.R

@@ -456,7 +456,7 @@ kable(tmp2,
   col.names = c("", as.character(tmp[algorithm == "mbo"]$n_cases)),
   caption = "Average runtime in hours, for evaluating one grid and one optimization run of MBO."
 ) %>% 
-  add_header_above(c(" " = 1, 3) %>% 
+  add_header_above(c(" " = 1, table(tmp$effect)/3)) %>% 
   kable_styling(position = "center") %>% 
   kable_to_text("table_time")
 

### 2020-10-23 main.tex
some results

--- 2020-10-21 main.tex
@@
 First, we describe the \emph{exhaustive grid search}.
-For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.
-one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and $\epsilon$ or $\tau$, respectively).
+For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.\ one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and $\epsilon$ or $\tau$, respectively).
 This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
-Additionally, the grid is evaluated with 10 stochastic replications to estimate the variance of the estimated power.
+Additionally, we will use a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \emph{grid small}.
+The grid is evaluated with 10 stochastic replications to estimate the variance of the estimated power.
 
-The best point per evaluated grid is determined by selecting the configuration (parameter set) with the best outcome, i.e. the largest power.
+The best point per evaluated grid is determined by selecting the configuration (parameter set) with the best outcome, i.e.\ the largest power.
 The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the good outcome.
-To account for this optimism and report unbiased performance estimates, for each of the 10 replicate simulations, we identify the best configuration and consider the corresponding 9 values of the other 9 replicate simulations, similar to a cross-validation framework in machine learning applications.
+To account for this optimism and report unbiased performance estimates, for each of the 10 replicate simulations, we identify the best configuration and consider the corresponding 9 outcomes of the other 9 replicate simulations, similar to a cross-validation framework in machine learning applications.
 %This eliminates the risk of only reporting outcomes that are optimal by chance.
@@
 The optimization is executed as described for \emph{MBO} but with one difference:
-At the end we conduct the independent evaluation for the configuration within the \emph{grid} that is closest to the configuration that is proposed by \emph{MBO}.
+At the end we conduct the independent evaluation for the configuration within the \emph{grid} that is closest to the optimal configuration that is proposed by \emph{MBO}.
 % Mention that only configurations of the same factor level are looked at?
@@
 
-As explained before, we obtained unbiased results from the exhaustive grid search and the MBO procedure.
-Figure~\ref{fig:plot_boxplot_valid_y} shows that \emph{MBO} yields better results for each effect set even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{grid search}.
+As explained before, we obtained unbiased results from the exhaustive grid search, the small grid search and the MBO procedure.
+Figure~\ref{fig:plot_boxplot_valid_y} shows that \emph{MBO} yields comparable results to the exhaustive \emph{Grid} for each effect set.
 \begin{figure}[htb]
@@
 \caption{%
-  Validated performance of both approaches.
+  Performance on independent replicates of the best found solution from each optimizer for different effect sets and different numbers of $n_{\text{treat}}$. %FIXME rename n_cases in plot
   }
@@
 \end{figure}
+Looking at the results in detail, we can see that in some cases ($n_{\text{treat}} = 2000$, \emph{effect: linear}; $n_{\text{treat}} = 1000$, \emph{effect: sigmoid}; $n_{\text{treat}} = 2000$, \emph{effect: sigmoid}) the exhaustive \emph{Grid} has a slight lead.
+In other cases ($n_{\text{treat}} = 1000$, \emph{effect: linear}; $n_{\text{treat}} = 500$, \emph{effect: paper2}; $n_{\text{treat}} = 500$, \emph{effect: sigmoid}) MBO yields slightly better results.
+In total \emph{MBO} and the exhaustive \emph{Grid} give equally good results, even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{Grid}.
 The difference in the number of evaluations is also reflected in the average runtime of each optimization methods given in Table~\ref{tab:table_time}.
 \input{generated/tables/table_time.tex}
-For the chosen grid resoultion and evaluation budget for \emph{MBO}, \emph{MBO} is approxamatly 20-times faster and yields better configurations.
+For the given resolution of 25 points per dimension, the exhaustive \emph{Grid} and the evaluation budget for \emph{MBO}, \emph{MBO} is approximately 20-times faster, while giving equally good results.
 
-The good performance of \emph{MBO Grid} indicates that the resolution of the grid is sufficiently high to find a good configuration.
-Accordingly, the good performance of \emph{MBO} can be contributed mostly to the fact that the best configuration is determined by the surrogates mean estimation es explained in Section~\ref{ssec:best_point}.
-In other words, the \emph{grid search} is not only extremely demanding in terms of necessary black-box evaluations but also does not take into account the stochasticity of the problem.
-This seemingly leads to proposals of configurations that were observed to yield a high power but that do not perform well, if evaluated independently.
-We will later study on single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of the \emph{gird search}.
+To compare \emph{MBO} with a grid search that uses approximately the same time, we will look at the results of \emph{Grid Small}.
+The resolution of 7 points per dimension leads to 126 evaluations which roughly take up the same time as \emph{MBO} (see~\ref{tab:table_time}) with 116 allowed evaluations, giving a small advantage to \emph{Grid Small}.
+Yet, in the majority of cases the results of \emph{Grid Small} are inferior the the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
 
-After having only looked at the outcome of our probelms and established the superiority of \emph{MBO} over the \emph{grid search} we want to investigate the problems in more detail.
-In Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behavior of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} $r$.
+Looking at the results of \emph{MBO Grid} in comparison to \emph{MBO} and the exhaustive \emph{Grid}, we can observe two patterns.
+
+First, if \emph{Grid} achieves equal or better results then \emph{MBO}, the results of \emph{MBO Grid} are similar to \emph{MBO}.
+In case \emph{MBO} was not able to find a better solution then \emph{Grid}, also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
+An increase in performance of \emph{MBO Grid} towards \emph{MBO} could just be by chance.
+
+Second and in contrast to the first pattern: If \emph{MBO} is able to find better configurations then the exhaustive \emph{Grid} (e.g.
+for scenario \ $n_{\text{treat}} = 1000$, \emph{effect: linear}), also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.
+This leads to the conclusion that the resolution of the exhaustive grid is sufficiently fine but the selection of the final configuration within the grid search is faulty in such scenarios.
+The grid search has no notion of the stochasticity of the problem.
+The configuration that yielded the best outcome is "optimistically" taken as the final configuration.
+The independent validation that gives the measure of $y_{\text{valid}}$ shows the unbiased performance of this final configuration. %FIXME: Plot y vs y_valid?
+In conclusion we are able to see that even though \emph{Grid} has "seen" all outcomes on the grid, the optimistic choice of the final configuration is indeed overly optimistic and not as good as the final configuration that is determined by the mean prediction of the surrogate of \emph{MBO} as explained in Section~\ref{ssec:best_point}.
+In other words, the exhaustive \emph{Grid} is not only extremely demanding in terms of necessary black-box evaluations but also does not take into account the stochasticity of the problem.
+We will later study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of the \emph{gird search}.
+%FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)
+
+Note that between the scenarios the absolute differences of the optimization results are of different practical significance, %FIXME "practical significance?"
+i.e.\ for scenario \ $n_{\text{treat}} = 1000$, \emph{effect: sigmoid}), \emph{Grid} is superior to \emph{MBO} but the difference is smaller than $0.25\%$.
+
+After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{MBO} to find equally good configurations as the much more demanding exhaustive \emph{Grid} we want to investigate the problems in more detail.
+In Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behavior of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
 \begin{figure}[htb]
@@
 \caption{%
-  All evaluated configurations and their outcomes from the \emph{grid search}.
-  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power is displayed.
+  All evaluated configurations and their outcomes on the \emph{Grid}.
+  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power ($y$) is displayed.
   For each configuration in the grid we obtained 10 stochastic results.
@@
 \end{figure}
-Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments at the early stage.
-Also, for the \emph{selection strategy} \emph{epsilon rule} and \emph{threshold rule} only the curves are shown of the respective \emph{eps} and \emph{thresh} value that led to the best outcome within the respective selection strategy.
-For the selection strategies \emph{eps} and \emph{thresh}, we only show the curve of the one $\epsilon$ and $\tau$ setting that led to the optimal outcome for the respective selection strategy.
-Instead of showing $r$ on the $x$-Axis we show the resulting value of treatments at stage 1 as this allows for easier comparison between the different selection strategies.
+Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments at the first stage.
+Also, for the \emph{selection strategy} \emph{epsilon rule} and \emph{threshold rule} only the curves are shown for the $\epsilon$ and $\tau$ value that led to the best outcome within the respective selection strategy.
 Also note that for effect set \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
@@
 In case multiple optimization runs give different configurations it might be advisable to let a human expert choose the final configuration.
+%FIXME: Write more about the meaning of the results from a clinical point of view.
 
@@
 \end{figure}
-The results show that in many cases the grid search returned configurations of $r$ that are far from the solutions that \emph{MBO} found.
+The results show that for some scenarios most replications of the \emph{Grid} returned the same optimal configurations (e.g. for scenario \ $n_{\text{treat}} = 1000$, \emph{effect: sigmoid})
+In such cases the signal to noise ratio was high enough to reliably find the best configuration.
+These are also the scenarios where \emph{Grid} was superior the \emph{MBO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
+In other cases \emph{Grid} found varying configurations and \emph{MBO} found more stable results that lead to better outcomes (e.g.\ $n_{\text{treat}} = 1000$, \emph{effect: linear}; $n_{\text{treat}} = 500$, \emph{effect: paper2}; $n_{\text{treat}} = 500$, \emph{effect: sigmoid}).
+A special scenario is $n_{\text{treat}} = 2000$, \emph{effect: paper2}: Here nearly all found configurations yield perfect outcomes.
 

### 2020-10-23 generated/tables/table_time.tex
grid7 ...
fix figures and eval

@@
 \centering
-\begin{tabular}[t]{lrrrrrrrrrrrr}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lrrrrrrrrrrrrr}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{4.5}{c}{linear} & \multicolumn{4.5}{c}{paper} & \multicolumn{4.5}{c}{paper2} & \multicolumn{4.5}{c}{sigmoid} \\
-\cmidrule(l{3pt}r{3pt}){2-5.5} \cmidrule(l{3pt}r{3pt}){6.5-10} \cmidrule(l{3pt}r{3pt}){11-14.5} \cmidrule(l{3pt}r{3pt}){15.5-19}
- & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000\\
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{linear} & \multicolumn{3}{c}{paper} & \multicolumn{3}{c}{paper2} & \multicolumn{3}{c}{sigmoid} & \multicolumn{1}{c}{ } \\
+\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
+ & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000 & 500 & 1000 & 2000 & \emph{evals}\\
 \midrule
-Grid & 55.1 & 60.9 & 60.5 & 62.7 & 63.2 & 66.6 & 59.7 & 64.4 & 67.1 & 54.6 & 52.8 & 61.0\\
-Grid 7 & 3.4 & 3.8 & 3.7 & 3.8 & 3.9 & 4.1 & 3.8 & 3.9 & 4.1 & 3.4 &  & 3.7\\
-MBO & 3.6 & 3.4 & 3.6 & 3.4 & 4.0 & 4.1 & 3.7 & 3.3 & 3.2 & 3.3 & 2.6 & 3.1\\
+Grid & 55.1 & 60.9 & 60.5 & 62.7 & 63.2 & 66.6 & 59.7 & 64.4 & 67.1 & 54.6 & 52.8 & 61.0 & 1350\\
+Grid Small & 4.1 & 4.5 & 4.4 & 4.5 & 4.6 & 4.8 & 4.4 & 4.7 & 4.8 & 4.0 & 3.9 & 4.4 & 126\\
+MBO & 3.6 & 3.4 & 3.6 & 3.4 & 4.0 & 4.1 & 3.7 & 3.3 & 3.2 & 3.3 & 2.6 & 3.1 & 116\\
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}

### 2020-10-23 main.tex
fix figures and eval

@@
 
-First, we describe the \emph{exhaustive grid search}.
+First, we describe the \emph{exhaustive grid search} (\emph{Grid}).
 For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.\ one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and $\epsilon$ or $\tau$, respectively).
 This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
-Additionally, we will use a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \emph{grid small}.
-The grid is evaluated with 10 stochastic replications to estimate the variance of the estimated power.
+Additionally, we will use a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \emph{Grid Small}.
+Both grids are evaluated with 10 stochastic replications to estimate the variance of the estimated power.
 
@@
 \end{figure}
-Looking at the results in detail, we can see that in some cases ($n_{\text{treat}} = 2000$, \emph{effect: linear}; $n_{\text{treat}} = 1000$, \emph{effect: sigmoid}; $n_{\text{treat}} = 2000$, \emph{effect: sigmoid}) the exhaustive \emph{Grid} has a slight lead.
-In other cases ($n_{\text{treat}} = 1000$, \emph{effect: linear}; $n_{\text{treat}} = 500$, \emph{effect: paper2}; $n_{\text{treat}} = 500$, \emph{effect: sigmoid}) MBO yields slightly better results.
+Looking at the results in detail, we can see that in some cases (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$) the exhaustive \emph{Grid} has a slight lead.
+In other cases (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$) MBO yields slightly better results.
 In total \emph{MBO} and the exhaustive \emph{Grid} give equally good results, even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{Grid}.
@@
 For the given resolution of 25 points per dimension, the exhaustive \emph{Grid} and the evaluation budget for \emph{MBO}, \emph{MBO} is approximately 20-times faster, while giving equally good results.
-
-To compare \emph{MBO} with a grid search that uses approximately the same time, we will look at the results of \emph{Grid Small}.
-The resolution of 7 points per dimension leads to 126 evaluations which roughly take up the same time as \emph{MBO} (see~\ref{tab:table_time}) with 116 allowed evaluations, giving a small advantage to \emph{Grid Small}.
-Yet, in the majority of cases the results of \emph{Grid Small} are inferior the the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
 
@@
 Second and in contrast to the first pattern: If \emph{MBO} is able to find better configurations then the exhaustive \emph{Grid} (e.g.
-for scenario \ $n_{\text{treat}} = 1000$, \emph{effect: linear}), also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.
+for scenario \ \emph{effect: linear}, $n_{\text{treat}} = 1000$), also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.
 This leads to the conclusion that the resolution of the exhaustive grid is sufficiently fine but the selection of the final configuration within the grid search is faulty in such scenarios.
@@
 In other words, the exhaustive \emph{Grid} is not only extremely demanding in terms of necessary black-box evaluations but also does not take into account the stochasticity of the problem.
-We will later study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of the \emph{gird search}.
+We will later study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of the \emph{grid search}.
 %FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)
 
+To compare \emph{MBO} with a grid search that uses approximately the same time  (see~\ref{tab:table_time}) and number of evaluations, we will look at the results of \emph{Grid Small}.
+The resolution of 7 points per dimension leads to 126 evaluations which roughly take up the same time as \emph{MBO} (see~\ref{tab:table_time}) with 116 allowed evaluations, giving a small advantage to \emph{Grid Small}.
+Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior the the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
+Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
+For the first scenario the absolute differences are negligible as all methods obtain a high outcome.
+For the second scenario we will later see, that the optimal configuration is close to the border of the search space.
+The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
+These points are not included in the coarse grid.
+In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates.
+
 Note that between the scenarios the absolute differences of the optimization results are of different practical significance, %FIXME "practical significance?"
-i.e.\ for scenario \ $n_{\text{treat}} = 1000$, \emph{effect: sigmoid}), \emph{Grid} is superior to \emph{MBO} but the difference is smaller than $0.25\%$.
+i.e.\ for scenario (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$), \emph{Grid} is superior to \emph{MBO} but the difference is smaller than $0.25\%$.
 
@@
 \end{figure}
-The results show that for some scenarios most replications of the \emph{Grid} returned the same optimal configurations (e.g. for scenario \ $n_{\text{treat}} = 1000$, \emph{effect: sigmoid})
+The results show that for some scenarios most replications of the \emph{Grid} returned the same optimal configurations (e.g.\ for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$)
 In such cases the signal to noise ratio was high enough to reliably find the best configuration.
 These are also the scenarios where \emph{Grid} was superior the \emph{MBO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
-In other cases \emph{Grid} found varying configurations and \emph{MBO} found more stable results that lead to better outcomes (e.g.\ $n_{\text{treat}} = 1000$, \emph{effect: linear}; $n_{\text{treat}} = 500$, \emph{effect: paper2}; $n_{\text{treat}} = 500$, \emph{effect: sigmoid}).
-A special scenario is $n_{\text{treat}} = 2000$, \emph{effect: paper2}: Here nearly all found configurations yield perfect outcomes.
+In other cases \emph{Grid} found varying configurations and \emph{MBO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+A special scenario is \emph{effect: paper2}, $n_{\text{treat}} = 2000$: Here nearly all found configurations yield perfect outcomes.
 
 \subsection{More simulation rounds}
+
+%FIXME TODO by Jakob
 

### 2020-10-23 generator/main.R

@@ -76,8 +76,16 @@ select_labels_colors = #http://paletton.com/#uid=74Z140kqdu7ghF3lowyu1ppwGk7
     "random" = "#000000",
     "threshold rule" = "#C0E62A"
   )
-algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "grid" = "Grid", "grid7" = "Grid 7")
+algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "grid" = "Grid", "grid7" = "Grid Small")
 algorithm_labels_color = c("mbo" = "#F04A2B", "eval" = "#099438", "mbo grid" = "#690D86", "grid" = "#099438", "grid7" = "#44CA6D") #https://paletton.com/#uid=6060R0krPwadDMilcBuwcpxDti6
+
+ncases_labels = c(500,1000,2000)
+ncases_labels = paste0("n[cases] ",ncases_labels)
+names(ncases_labels) = c(500,1000,2000)
+
+names(effect_labels) = effect_labels = c("linear", "paper", "paper2", "sigmoid")
+
+
 
 ## ----table_effect_names-------------------------------------------------------
 tmp = lapply(EFFECTS, do.call, what = rbind)
@@ -91,10 +99,27 @@ kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation") %>%
   kable_styling(position = "center") %>% 
   kable_to_text("table_effect_names")
 
-
 ## ----read_data----------------------------------------------------------------
-read_results_uncached = function(select_labels) {
-  res_eval = readRDS("../benchmark_results/batchtools_grid_res_eval.rds")
+read_results_uncached = function(select_labels, case = "default") {
+  if (case == "default") {
+    case_selector = quote(nsim == 1000 & repl <= 10)
+  } else if (case == "nsim") {
+    case_selector = quote(n_cases == 2000 & effect == "paper")
+  } else {
+    stop("not supported case")
+  }
+  
+  read_res_eval_uncached = function() {
+    readRDS("../benchmark_results/batchtools_grid_res_eval.rds")
+  }
+  read_res_eval = memoise(read_res_eval_uncached)
+  
+  read_res_mbo_uncached = function() {
+    readRDS("../benchmark_results/batchtools_grid_res_mbo.rds")
+  }
+  read_res_mbo = memoise(read_res_mbo_uncached)
+  
[…]

### 2020-10-23 generator/main.R
fix plots ...

@@ -389,7 +389,7 @@ plot_wrapper(name = "plot_boxplot_valid_y", fig.height = FIG_HEIGHT * 0.9, expr 
   #   paste(value, variable)
   # }
   
-  g = g + facet_wrap(effect~n_cases, scales = "free_y", ncol = 6, labeller = label_bquote({atop(n[treat]==.(n_cases),.(effect))}))
+  g = g + facet_wrap(effect~n_cases, scales = "free_y", ncol = 6, labeller = label_bquote({atop(.(effect), n[treat]==.(n_cases))}))
   darker_colors = colorspace::darken(algorithm_labels_color, amount = 0.6)
   names(darker_colors) = names(algorithm_labels_color)
   g = g + scale_fill_manual(labels = algorithm_labels, values = algorithm_labels_color) + scale_color_manual(labels = algorithm_labels, values = darker_colors)
@@ -411,6 +411,7 @@ plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT *
   g = g + theme(legend.position = "bottom")
   g = g + geom_boxplot()
   g = g + labs(x = expression(n[sim]), y = expression(y[valid]), color = NULL, fill = NULL)
+  g = g + guides(fill=guide_legend(nrow=2,byrow=FALSE))
   g
 })
 

### 2020-10-23 main.tex
Updates from Overleaf

@@
 Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior the the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
-Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
+Only for the scenarios (\emph{effect: paper}, $n_{\text{treat}} = 2000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
 For the first scenario the absolute differences are negligible as all methods obtain a high outcome.
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
-    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_treat = 2000$ with 100 stochastic repititions and different values for \texttt{nsim}}
+    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{treat}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
     \label{fig:plot_boxplot_valid_y_5000}

### 2020-11-02 main.tex
edits

@@
 
-\subsection{Results}
+\subsection{Main results}
 
-As explained before, we obtained unbiased results from the exhaustive grid search, the small grid search and the MBO procedure.
-Figure~\ref{fig:plot_boxplot_valid_y} shows that \emph{MBO} yields comparable results to the exhaustive \emph{Grid} for each effect set.
+In this chapter we present the results of the unbiased comparison on independent replicates for the exhaustive grid search, the small grid search and the MBO procedure.
+Figure~\ref{fig:plot_boxplot_valid_y} shows the  estimated power values $y$ for the best solutions of the three approaches, for the four different effect sets and for three different numbers of $n_{\text{treat}}$.
 \begin{figure}[htb]
@@
 \caption{%
-  Performance on independent replicates of the best found solution from each optimizer for different effect sets and different numbers of $n_{\text{treat}}$. %FIXME rename n_cases in plot
+  Boxplots of performance (power values) on independent replicates of the best solutions for the four approaches, the different effect sets, and different numbers of $n_{\text{treat}}$.
+%FIXME rename n_cases in plot
   }
@@
 \end{figure}
-Looking at the results in detail, we can see that in some cases (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$) the exhaustive \emph{Grid} has a slight lead.
-In other cases (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$) MBO yields slightly better results.
-In total \emph{MBO} and the exhaustive \emph{Grid} give equally good results, even though \emph{MBO} is just allowed 116 evaluations in contrast to the 1350 evaluations that are conducted by the \emph{Grid}.
-The difference in the number of evaluations is also reflected in the average runtime of each optimization methods given in Table~\ref{tab:table_time}.
+
+The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets.
+Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
+In other cases \emph{MBO} yields slightly better results (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension.
+The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}. \emph{MBO} is approximately 20 times faster.
 \input{generated/tables/table_time.tex}
-For the given resolution of 25 points per dimension, the exhaustive \emph{Grid} and the evaluation budget for \emph{MBO}, \emph{MBO} is approximately 20-times faster, while giving equally good results.
 
-Looking at the results of \emph{MBO Grid} in comparison to \emph{MBO} and the exhaustive \emph{Grid}, we can observe two patterns.
+For \emph{MBO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
+Comparing the results of \emph{MBO Grid} to those of \emph{MBO} and the exhaustive \emph{Grid}, we can observe two patterns.
+First, if \emph{Grid} achieves equal or better results then \emph{MBO}, the results of \emph{MBO Grid} are similar to \emph{MBO}. 
+If \emph{MBO} was not able to find a better solution then \emph{Grid}, then also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
+An increase in performance of \emph{MBO Grid} compared to \emph{MBO} is purely due to chance.
+Second, and in contrast to the first pattern: 
+If \emph{MBO} is able to find better configurations then the exhaustive \emph{Grid} (e.g.
+for scenario \ \emph{effect: linear}, $n_{\text{treat}} = 1000$), then also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.
 
-First, if \emph{Grid} achieves equal or better results then \emph{MBO}, the results of \emph{MBO Grid} are similar to \emph{MBO}.
-In case \emph{MBO} was not able to find a better solution then \emph{Grid}, also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
-An increase in performance of \emph{MBO Grid} towards \emph{MBO} could just be by chance.
-
-Second and in contrast to the first pattern: If \emph{MBO} is able to find better configurations then the exhaustive \emph{Grid} (e.g.
-for scenario \ \emph{effect: linear}, $n_{\text{treat}} = 1000$), also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.
-This leads to the conclusion that the resolution of the exhaustive grid is sufficiently fine but the selection of the final configuration within the grid search is faulty in such scenarios.
-The grid search has no notion of the stochasticity of the problem.
-The configuration that yielded the best outcome is "optimistically" taken as the final configuration.
-The independent validation that gives the measure of $y_{\text{valid}}$ shows the unbiased performance of this final configuration. %FIXME: Plot y vs y_valid?
-In conclusion we are able to see that even though \emph{Grid} has "seen" all outcomes on the grid, the optimistic choice of the final configuration is indeed overly optimistic and not as good as the final configuration that is determined by the mean prediction of the surrogate of \emph{MBO} as explained in Section~\ref{ssec:best_point}.
-In other words, the exhaustive \emph{Grid} is not only extremely demanding in terms of necessary black-box evaluations but also does not take into account the stochasticity of the problem.
-We will later study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of the \emph{grid search}.
+This leads to the conclusion that the resolution of the exhaustive grid is sufficiently fine but the selection of the final configuration within the grid search is not good enough in such scenarios.
+The grid search has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
+Just the configuration on the grid with the best outcome is "optimistically" selected as final configuration.
+The independent validation then is based on $y_{\text{valid}}$ that represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
+\emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
+%FIXME: Plot y vs y_valid?
+Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
 %FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)
 
-To compare \emph{MBO} with a grid search that uses approximately the same time  (see~\ref{tab:table_time}) and number of evaluations, we will look at the results of \emph{Grid Small}.
-The resolution of 7 points per dimension leads to 126 evaluations which roughly take up the same time as \emph{MBO} (see~\ref{tab:table_time}) with 116 allowed evaluations, giving a small advantage to \emph{Grid Small}.
-Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior the the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
+To compare \emph{MBO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
+Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{MBO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
+Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
 Only for the scenarios (\emph{effect: paper}, $n_{\text{treat}} = 2000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
-For the first scenario the absolute differences are negligible as all methods obtain a high outcome.
-For the second scenario we will later see, that the optimal configuration is close to the border of the search space.
+In the first case, the absolute differences are negligible as all methods lead to a high power.
+For the second scenario we will later show, that the optimal configuration is close to the border of the search space.
 The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
 These points are not included in the coarse grid.
-In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates.
+In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
 
-Note that between the scenarios the absolute differences of the optimization results are of different practical significance, %FIXME "practical significance?"
+Note that between the scenarios the absolute differences of the optimization results are of different practical relevance,
 i.e.\ for scenario (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$), \emph{Grid} is superior to \emph{MBO} but the difference is smaller than $0.25\%$.
 
-After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{MBO} to find equally good configurations as the much more demanding exhaustive \emph{Grid} we want to investigate the problems in more detail.
-In Figure~\ref{fig:plot_allbest} we show the observed outcomes during evaluation of the grid as well as the final configurations found by MBO, to demonstrate the behavior of the simulation in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
+%After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{MBO} to find equally good configurations as the much more demanding exhaustive \emph{Grid}
+We now investigate the results in more detail.
+In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{MBO} are shown, in order to analyze the differences in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
 \begin{figure}[htb]
@@
   For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power ($y$) is displayed.
-  For each configuration in the grid we obtained 10 stochastic results.
-  The line connects the mean outcomes for each configuration.
+  For each configuration in the grid 10 stochastic results are obtained.
+  A line connects the mean outcomes for each configuration.
   Additionally the best found configurations from \emph{MBO} are displayed.
@@
 \end{figure}
-Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments at the first stage.
-Also, for the \emph{selection strategy} \emph{epsilon rule} and \emph{threshold rule} only the curves are shown for the $\epsilon$ and $\tau$ value that led to the best outcome within the respective selection strategy.
-Also note that for effect set \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
-In accordance with intuition, for most scenarios it is the worst decision to use all available treatments in stage 1. 
-In all scenarios \emph{MBO} was able to identify the peak of all curves in combination.
-Within the 10 stochastic repetitions \emph{MBO} found different configurations that all yield similarly good outcomes.
+Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments in the first stage.
+For the \emph{selection strategies} \emph{epsilon rule} and \emph{threshold rule} only curves representing the $\epsilon$ and $\tau$ values with the best outcome, respectively, are shown.
+Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are as expected nearly identical.
+As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. 
+In all scenarios, \emph{MBO} identified the peak of all combined curves.
+Within the 10 stochastic repetitions per scenario, \emph{MBO} found different configurations, but those had similarly good outcomes.
 That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.
 different configurations yield similarly good outcomes.
-In case multiple optimization runs give different configurations it might be advisable to let a human expert choose the final configuration.
+In cases in which multiple optimization runs result in different configurations it might be advisable to let a human expert choose the final configuration, also depending on the medical background.
 %FIXME: Write more about the meaning of the results from a clinical point of view.
 
-In contrast to Figure~\ref{fig:plot_allbest}, Figure~\ref{fig:plot_best_x} shows the outcome of the independent validation and the ratio $r$ directly.
+%In contrast to Figure~\ref{fig:plot_allbest},
+Figure~\ref{fig:plot_best_x} visualizes for the configurations selected by \emph{MBO} the corresponding values for $y_{\text{valid}}$ (independent validation) and the ratio $r$.
 \begin{figure}[htb]
@@
 \includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
-\caption{Best found $x$-values from MBO runs.}
+\caption{Best found $x$-values for \emph{Grid} and \emph{MBO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
 \label{fig:plot_best_x}
 \end{figure}
-The results show that for some scenarios most replications of the \emph{Grid} returned the same optimal configurations (e.g.\ for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$)
+For some scenarios most replications of \emph{Grid} returned the same optimal configurations (e.g.\ for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
 In such cases the signal to noise ratio was high enough to reliably find the best configuration.
@@
 In other cases \emph{Grid} found varying configurations and \emph{MBO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
-A special scenario is \emph{effect: paper2}, $n_{\text{treat}} = 2000$: Here nearly all found configurations yield perfect outcomes.
+A special case is scenario \emph{effect: paper2}, $n_{\text{treat}} = 2000$:
+Here nearly all found configurations yield perfect outcomes.
 
-\subsection{More simulation rounds}
+\subsection{Results for more simulation rounds}
 

### 2020-11-02 main.tex
final results

@@
 Additionally, we will use a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \emph{Grid Small}.
-Both grids are evaluated with 10 stochastic replications to estimate the variance of the estimated power.
+Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.
 
@@
 The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the good outcome.
-To account for this optimism and report unbiased performance estimates, for each of the 10 replicate simulations, we identify the best configuration and consider the corresponding 9 outcomes of the other 9 replicate simulations, similar to a cross-validation framework in machine learning applications.
+To account for this optimism and report unbiased performance estimates, for each of the 20 replicate simulations, we identify the best configuration and consider the corresponding 19 outcomes of the other 19 replicate simulations, similar to a cross-validation framework in machine learning applications.
 %This eliminates the risk of only reporting outcomes that are optimal by chance.
@@
 To investigate the effect of the function noise variance ($\sigma^2_n$) on the optimization result we increased the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario with the \emph{paper} effect set and  $n_{\text{treat}} = 2000$.
-This is expected to increase the runtime by a factor of 5.
-Also we expect that due to the decreased noise variance the grid search might be less prone to select over optimistic configurations.
-Furthermore, MBO might need less iterations because the signal to noise ratio is higher.
+% This is expected to increase the runtime by a factor of 5. % obvious, also not of interest here
+Additionally, we run 100 stochastic replicates instead of the previously 20 to obtain a more precise estimation of the independent validation.
 
-Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that a single optimization run leads to a results closer to the mean result.
+We expect, that increasing $n_\text{sim}$ decreases the noise variance of the outcomes of the simulation that returns the statistical power of the design configuration.
+For the exhaustive \emph{Grid}, we expect that the decreased noise leads to less selections of overly optimistic outcomes as result.
+Also \emph{MBO} should profit from a less noisy objective function as the surrogates predictions will be more precise.
+
+Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
+It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
+So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that oberved overoptimistic outcomes in compaision the the right side.
+Ideally, the choise of the final point should not be influenced by these overoptimistic observations for \emph{MBO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
+
 Also the outcome of the best found point by MBO is closer to the best outcome found by the exhaustive grid search.
-We have to remember that this plot reports the best observed outcome so far. 
+We have to remember that this plot 
 However, as explained in Section~\ref{ssec:best_point} the best point is not determined by the best found outcome but by the surrogates mean prediction of all evaluated point.
-This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for MBO in both cases.
-Interestingly the validation error of the grid search is lower for $n_\text{sim} = 5000$ then for $n_\text{sim} = 1000$.
-We would have expected the opposite, since more noise during optimization leads to overly optimistic choices that result in a bad validation error.
-This does not seem to be the case here but also the results are not drastically different and the illustrated IQR for the grid search becomes smaller for $n_\text{sim} = 5000$.
-
-%FIXME: Add Runtime comparision!
-
+This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{MBO} for the low and the high value of $n_\text{sim}$.
+%FIXME: Add Runtime comparision? Maybe boring because it probably is just 5 times the old values
 \begin{figure}[htb]
@@
 \end{figure}
+For all methods the IQR displayed in the is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
+Also the absolute differences between the methods seem even more negligible.
+This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
+But as an increased value of $n_\text{sim}$ also increases the runtime linearly this is not always a feasible choice if the goal is to find the optimal configuration quickly.
+We conclude with the observation, that \emph{MBO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high numer of simulation.
 

### 2020-11-02 benchmark/_parameters.R

@@ -21,7 +21,7 @@ if (TESTMODE) {
   NSIM = 1000
   MBOITERS = 100
   GRIDRES = 25
-  REPLS = 10
+  REPLS = 20
   NCASES = c(500,1000,2000)
   EFFECTS = list(
     paper = list(early = c(0,0.68,0.82,0.95,0.91), final = c(0,0.13,0.17,0.23,0.20)),

### 2020-11-02 benchmark/batchtools_grid.R

@@ -63,7 +63,7 @@ addExperiments(algo.designs = ades, repls = REPLS)
 
 # add 5times sim steps for one experiment
 ades_nsim_high = generate_eval_mbo_design(n_cases = tail(NCASES,1), nsim = c(NSIM, 5 * NSIM), effect = "paper")
-addExperiments(algo.designs = ades_nsim_high, repls = 10 * REPLS)
+addExperiments(algo.designs = ades_nsim_high, repls = max(4 * REPLS, 100))
 
 if (TESTMODE) {
   stop()

### 2020-11-02 generator/main.R

@@ -80,7 +80,7 @@ algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "g
 algorithm_labels_color = c("mbo" = "#F04A2B", "eval" = "#099438", "mbo grid" = "#690D86", "grid" = "#099438", "grid7" = "#44CA6D") #https://paletton.com/#uid=6060R0krPwadDMilcBuwcpxDti6
 
 ncases_labels = c(500,1000,2000)
-ncases_labels = paste0("n[cases] ",ncases_labels)
+ncases_labels = paste0("n[cases] ", ncases_labels)
 names(ncases_labels) = c(500,1000,2000)
 
 names(effect_labels) = effect_labels = c("linear", "paper", "paper2", "sigmoid")
@@ -102,7 +102,7 @@ kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation") %>%
 ## ----read_data----------------------------------------------------------------
 read_results_uncached = function(select_labels, case = "default") {
   if (case == "default") {
-    case_selector = quote(nsim == 1000 & repl <= 10)
+    case_selector = quote(nsim == 1000 & repl <= 20)
   } else if (case == "nsim") {
     case_selector = quote(n_cases == 2000 & effect == "paper")
   } else {
@@ -143,7 +143,7 @@ read_results_uncached = function(select_labels, case = "default") {
   ## ----estimate res_grid--------------------------------------------------------
   # we calculate res_grid which emulates a grid search with an independent validation of the best found x value
   # 1. for each repl we look for the best y and take the x values (randomly 1 on ties)
-  # 2. we obtain the y values of the given x-values of the other 9 repls
+  # 2. we obtain the y values of the given x-values of the other 19 repls
   # 3. we average the external y values
   # 4. we sum up the time that the grid needed on one replication
   calc_res_grid_internal = function(.SD) {
@@ -283,11 +283,11 @@ plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
   df = merge(get_res_eval(), res_best, all.x = FALSE, all.y = TRUE, by = colnames(res_best))
[…]

### 2020-11-23 benchmark/batchtools_grid.R
fix final evals

@@ -47,7 +47,7 @@ addAlgorithm("mbo", fun = function(job, data, instance, corr, nsim, n_cases, mbo
     min(getOptPathY(opt.path), na.rm = TRUE)
   }
 
-  mbo_ctrl = makeMBOControl(on.surrogate.error = "warn", final.method = "best.predicted", impute.y.fun = impute_y, final.evals = 10)
+  mbo_ctrl = makeMBOControl(on.surrogate.error = "warn", final.method = "best.predicted", impute.y.fun = impute_y, final.evals = 20)
   mbo_ctrl = setMBOControlInfill(mbo_ctrl, crit = makeMBOInfillCritAEI(aei.use.nugget = TRUE))
   mbo_ctrl = setMBOControlTermination(mbo_ctrl, iters = mbo_iters)
 

### 2020-11-23 main.tex
Updates from Overleaf

@@
 \section{Motivating Example}
-\label{sect.motivation}
+\label{sec:motivating_example}
 
@@
 % TODO: Tim
-Intro 3- 4 Sätze
+Traditionally drug development is organized in four phases with individual clinical trials for the separate phases.
+Seamless designs combine elements from different development phases in a single trial thereby offering the promise of speeding up development times.
+For instance, seamless phase II/III designs combine the learning about dose-response relationship or subgroup heterogeneity typical for phase II with confirmation of treatment effects in  phase III.
+If the data of the phase II part are used for confirmatory purposes, then this needs to be accounted for in the testing strategy to maintain the familywise type I error rate in the strong sense.
+This type of designs are referred to as adaptive seamless designs.
+A recent overview can be found in \citet{friede_adaptive_2020}.
 
@@
 \label{ssec:two_stage_design_with_treatment_selection}
+Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection.
+The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control.
+Interest is in assessing the treatment effects $\theta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs.
+control and testing the null hypotheses $H_{0,k}: \theta_k=0$ against the alternative  $H_{1,k}: \theta_k>0$ (larger $\theta_k$ being better).
+Fur this purpose test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments.
+In an interim analysis, treatments that are continued into the second stage are selected; the set of selected treatments is denoted by $\mathcal{S}$.
+The selection rules used here are described in Section \ref{ssec:selection_rules}.
+Based on the data from patients recruited into the second stage, test statistics $Z_{2,k}$ with $k \in \mathcal{S}$ are computed.
 
-Multi-arm trial; two-stages; treatment selection in interim analysis; final analysis based on combination test approach with closed test principle
+final analysis based on combination test approach with closed test principle
 
-% first stage = interrim analysis
-% second stage = final analysys
-% stick to first second stage througout text but mention interrim here?
+% first stage = interim analysis
+% second stage = final analysis
+% stick to first second stage throughout text but mention interim here?
 
@@
 
-\paragraph{Epsilion Rule}
+\paragraph{Epsilon Rule}
 
@@
 
-% TODO : Jakob
-
-The goal of this simulation study is to investigate the potential benefit of the MBO approach for effectively finding a clinical trial design that maximizes a metric.
-The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sect.motivation}, with treatment selection in the first of two stages.
+The goal of this simulation study is to investigate the potential benefit of MBO for effectively finding a clinical trial design that maximizes a metric.
+The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
 In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
@@
   $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
-  $\tau$ (onlt for \emph{thresh})  & $[0,10]$ \\
+  $\tau$ (only for \emph{thresh})  & $[0,10]$ \\
   \hline
@@
 \end{table}
-
-%FIXME: JR: uniform naming: eps, thersh for the values and epsilon and treshold for the strategies
 
@@
 \end{equation}
-In our scenario (as described in Section \ref{subsec:design}) we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1 and the control treatment is included in both stages.
+In our scenario (as described in Section \ref{subsec:design}) we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1.
+The control treatment is included in both stages.
 
-Our main optimization goal is to maximize the statistical power $y$ depending on the trial design. The power can be formulated as a black-box as follows:
+Our main optimization goal is to maximize the statistical power $y$ depending on the trial design. 
+The power can be formulated as a black-box as follows:
 \begin{align}
@@
 
-The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{threshold}$ and $\text{epsilon}$.
+The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
 Then the black-box from formula (\ref{eq:bbox}) can be reformulated as 
@@
 
-In the simulation study we compare an exhaustive grid search on the parameter space with MBO (Model-based optimization) with respect to both performance (power) and runtime.
+In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand  MBO (Model-based optimization).
+For different scenarios we compare the statistical power of the best found clinical trial design.
+We also compare the runtime needed to determine the respective optimization results.
+Next we introduce the two optimization strategies.
 
-First, we describe the \emph{exhaustive grid search} (\emph{Grid}).
-For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.\ one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and $\epsilon$ or $\tau$, respectively).
-This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
-Additionally, we will use a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \emph{Grid Small}.
+\paragraph{Grid}
+We conduct an \emph{exhaustive grid search} with a resolution of 25 points per real-valued dimension and one point for each categorical dimension.
+Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of 25.
+All 6 selection strategies have the numeric parameter $r$.
+The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively. 
+This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
+We call this approach \textbf{Grid}.
+Further, we also consider a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \textbf{Grid Small}.
 Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.
 
-The best point per evaluated grid is determined by selecting the configuration (parameter set) with the best outcome, i.e.\ the largest power.
-The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the good outcome.
+The best point per evaluated grid is determined by selecting the configuration with the best outcome, i.e.\  the parameter set with the largest power.
+The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the quality of the outcome.
 To account for this optimism and report unbiased performance estimates, for each of the 20 replicate simulations, we identify the best configuration and consider the corresponding 19 outcomes of the other 19 replicate simulations, similar to a cross-validation framework in machine learning applications.
@@
 
-Next, we describe our realization of the search with the \emph{MBO (Model-based optimization)} approach.
-For the application of MBO, we first have to define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
-As the regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
+\paragraph{MBO (Model-based optimization)}
+For applying MBO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
+As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
 We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
@@
 Kriging expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
-First, to transform categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
+First, to transform the categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
 Furthermore, our search space is hierarchical because the parameters $\epsilon$ and $\tau$ are only active for the respective selection strategies \emph{eps} and \emph{thresh}.
-The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter $\tau$ of \emph{thresh} is inactive.
+The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if \emph{selection strategy} is set to \emph{eps} then the parameter $\tau$ of \emph{thresh} is inactive.
 If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if $\tau$ is inactive it will be set to 20).
@@
 Again, overoptimism must be avoided. 
-Thus, the best selected configuration (the best parameter set) is then independently evaluated $10$ times again.
+Thus, the best selected configuration (the best parameter set) is then independently evaluated $20$ times again.
 
-The whole MBO procedure is repeated $10$ times to ensure that a single optimization run was not just lucky in finding the optimum.
+The whole MBO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.
 
-Additionally we want to investigate whether the results of \emph{MBO} are better then \emph{grid search} because \emph{MBO} can propose configurations within the whole search space and not only limited the grid.
-Therefore we introduce \textbf{MBO Grid}.
-The optimization is executed as described for \emph{MBO} but with one difference:
-At the end we conduct the independent evaluation for the configuration within the \emph{grid} that is closest to the optimal configuration that is proposed by \emph{MBO}.
-% Mention that only configurations of the same factor level are looked at?
-
-We expect that \emph{MBO} achieves to find configurations that yield better results than the \emph{grid search} for two reasons.
-First, \emph{MBO} will be able to find an optimal configuration that is not restricted to be within the grid but can be between points from the grid.
-Second, \emph{MBO} will not choose configurations that were observed to perform well only due to noise.
+Additionally we investigate how \emph{MBO} performs when it is limited to return configurations that are within the grid.
+Therefore we introduce \textbf{MBO Grid}, which executes the optimization exactly as \emph{MBO}, but the best configuration is mapped to the closest value on the grid. 
+In other words, for \emph{MBO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{MBO}.
+The comparison of the results of \emph{MBO Gird} and \emph{Gird} will allow two possible conclusions:
+First, if \emph{MBO Grid} is worse then \emph{MBO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
+Second, if \emph{MBO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
+Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.
 
@@
 
-In this chapter we present the results of the unbiased comparison on independent replicates for the exhaustive grid search, the small grid search and the MBO procedure.
-Figure~\ref{fig:plot_boxplot_valid_y} shows the  estimated power values $y$ for the best solutions of the three approaches, for the four different effect sets and for three different numbers of $n_{\text{treat}}$.
+In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{MBO} and \emph{MBO Grid},
+We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
+The replicated evaluations give us an unbiased estimate of the performance ($y_{\text{valid}}$) of each optimizer.
+
+For this study we consider 12 scenarios.
+The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers of $n_{\text{treat}}$.
+Each optimizer is applied on each scenario.
+
+Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
 \begin{figure}[htb]
@@
 \caption{%
-  Boxplots of performance (power values) on independent replicates of the best solutions for the four approaches, the different effect sets, and different numbers of $n_{\text{treat}}$.
+  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$.
 %FIXME rename n_cases in plot
@@
 Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
-In other cases \emph{MBO} yields slightly better results (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. %FIXME Make sure that this is mentioned for figure 4!
+In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
 Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension.
-The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}. \emph{MBO} is approximately 20 times faster.
+The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
+\emph{MBO} is approximately 20 times faster.
 \input{generated/tables/table_time.tex}
@@
 For \emph{MBO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
-Comparing the results of \emph{MBO Grid} to those of \emph{MBO} and the exhaustive \emph{Grid}, we can observe two patterns.
-First, if \emph{Grid} achieves equal or better results then \emph{MBO}, the results of \emph{MBO Grid} are similar to \emph{MBO}. 
+Comparing the results of \emph{MBO Grid} to those of \emph{MBO} we mainly see similar performances for both.
+This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{MBO}.
+
+Comparing \emph{MBO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{MBO Grid} achieves slightly better performance then \emph{Grid}.
+The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e.
+in this case the noise on the predicted power.
+Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
+The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
+In these scenarios \emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
+Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
+%FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)
+
 If \emph{MBO} was not able to find a better solution then \emph{Grid}, then also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
 An increase in performance of \emph{MBO Grid} compared to \emph{MBO} is purely due to chance.
-Second, and in contrast to the first pattern: 
-If \emph{MBO} is able to find better configurations then the exhaustive \emph{Grid} (e.g.
-for scenario \ \emph{effect: linear}, $n_{\text{treat}} = 1000$), then also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.
-
-This leads to the conclusion that the resolution of the exhaustive grid is sufficiently fine but the selection of the final configuration within the grid search is not good enough in such scenarios.
-The grid search has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
-Just the configuration on the grid with the best outcome is "optimistically" selected as final configuration.
-The independent validation then is based on $y_{\text{valid}}$ that represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
-\emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
-%FIXME: Plot y vs y_valid?
-Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
-%FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)
 
@@
 Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
-Only for the scenarios (\emph{effect: paper}, $n_{\text{treat}} = 2000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
+Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
 In the first case, the absolute differences are negligible as all methods lead to a high power.
-For the second scenario we will later show, that the optimal configuration is close to the border of the search space.
+For the second case we will later show, that the optimal configuration is close to the border of the search space.
 The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
@@
 In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
+The advantage of \emph{MBO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect:paper2}, $n_{\text{treat}} = 500$.
 
@@
 In all scenarios, \emph{MBO} identified the peak of all combined curves.
-Within the 10 stochastic repetitions per scenario, \emph{MBO} found different configurations, but those had similarly good outcomes.
+Within the 20 stochastic repetitions per scenario, \emph{MBO} found different configurations, but those had similarly good outcomes.
 That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.
@@
 In cases in which multiple optimization runs result in different configurations it might be advisable to let a human expert choose the final configuration, also depending on the medical background.
+The curves also show the potential of optimization for the different scenarios.
+If the allowed treatment number is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less necessary.
+If the allowed treatment number is too small $n_{\text{treat}} = 500$, no configuration yields a practically usable design.
+Here, a fast optimization helps to determine if a desired treatment number can actually yield sufficiently powerful designs.
+Of special interest are the scenarios where the correct choice of the design configuration can make a difference between a design with a usable power or a design that is too weak, e.g.
+for the scenarios (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: linear}, $n_{\text{treat}} = 2000$, \emph{effect: paper}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
+Note, that the selection strategies \emph{thresh} and \emph{all} never gave powerful designs.
+
 %FIXME: Write more about the meaning of the results from a clinical point of view.
@@
 %In contrast to Figure~\ref{fig:plot_allbest},
-Figure~\ref{fig:plot_best_x} visualizes for the configurations selected by \emph{MBO} the corresponding values for $y_{\text{valid}}$ (independent validation) and the ratio $r$.
+Figure~\ref{fig:plot_best_x} visualizes in detail which configurations were selected by \emph{MBO} and \emph{Grid} and the corresponding values for $y_{\text{valid}}$ (result of the independent validation).
 \begin{figure}[htb]
@@
 \end{figure}
-For some scenarios most replications of \emph{Grid} returned the same optimal configurations (e.g.\ for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
+The optimal values for $\epsilon$ and $\theta$ are not shown.
+For some scenarios most replications of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
 In such cases the signal to noise ratio was high enough to reliably find the best configuration.
-These are also the scenarios where \emph{Grid} was superior the \emph{MBO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
+These are also those scenarios where \emph{Grid} was superior the \emph{MBO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
+Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500)$ and \emph{effect: paper}, $n_{\text{treat}} = 500)$, small values of $r$ lead to an optimal outcome.
+Here, \emph{MBO} was not able to find these values close to the border of the search space.
+A common remedy is to log-transform this value.
+
 In other cases \emph{Grid} found varying configurations and \emph{MBO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
@@
 
-%FIXME TODO by Jakob
-
-To investigate the effect of the function noise variance ($\sigma^2_n$) on the optimization result we increased the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario with the \emph{paper} effect set and  $n_{\text{treat}} = 2000$.
+To investigate the effect of the function noise variance ($\sigma^2_n$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario \emph{effect: paper}, $n_{\text{treat}} = 2000$.
 % This is expected to increase the runtime by a factor of 5. % obvious, also not of interest here
@@
 Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
+Furthermore, the average optimization curve indicates that less iterations can be sufficient to obtain an outcome that is comparable to \emph{Grid},
 It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
-So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that oberved overoptimistic outcomes in compaision the the right side.
-Ideally, the choise of the final point should not be influenced by these overoptimistic observations for \emph{MBO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
-
-Also the outcome of the best found point by MBO is closer to the best outcome found by the exhaustive grid search.
-We have to remember that this plot 
-However, as explained in Section~\ref{ssec:best_point} the best point is not determined by the best found outcome but by the surrogates mean prediction of all evaluated point.
+So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
+Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{MBO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
 This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{MBO} for the low and the high value of $n_\text{sim}$.
@@
 \end{figure}
-For all methods the IQR displayed in the is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
+For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
 Also the absolute differences between the methods seem even more negligible.
 This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
-But as an increased value of $n_\text{sim}$ also increases the runtime linearly this is not always a feasible choice if the goal is to find the optimal configuration quickly.
-We conclude with the observation, that \emph{MBO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high numer of simulation.
+But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
+We conclude with the observation, that \emph{MBO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.
 
@@
 Such procedure can be guided through statistical testing and is already common in machine learning under the terms racing or multi-fidelity optimization.
+
+IDEA: Restricting MBO to only evaluate configurations on the grid to avoid evaluating identical / very close configurations.
+IDEA: log transform parameters that have optimal settings close to the border.
 

### 2020-12-09 main.tex
friede

@@
 Clinical research and in particular drug development is typically structured into phases, e.g. phases I to IV in drug development. 
-Similar approaches exist for complex interventions, such as the MRC framework for non-pharmacological interventions (ref)%FIXME ref
+Similar approaches exist for complex interventions, such as the MRC framework for non-pharmacological interventions~\citep{campbell_framework_2000}.
 The development phases include elements of learning and confirmation.
@@
 Boundaries between the development phases are increasingly dissolving in multiple aspects.
-Master protocols are a new concept where multiple subgroups and substudies are integrated (ref). %FIXME ref
+Master protocols are a new concept where multiple subgroups and substudies are integrated~\citep{bogin_master_2020}.
 Here, simultaneously, more than one treatment and more than one disease type are investigated within the same overall trial structure.
@@
 Especially relevant are neutral comparison studies, which are characterized as follows: %FIXME Is there someting missing? (Jakob)
-The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way (ref).
+The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way~\citep{boulesteix_plea_2013}.
 %While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios.
@@
 
+We start by providing some more details on the study design considered (Sections \ref{ssec:two_stage_design_with_treatment_selection} and \ref{ssec:selection_rules}) and then introduce briefly a framework for optimizing such designs, namely clinical scenario evaluation (CSE) in Section \ref{ssec:clinical_scenario_evaluation_for_asd}.
+In Section \ref{ssec:simulating_asd} we make some brief comments on the simulation model.
+
 \subsection{Two-stage design with treatment selection}
@@
 control and testing the null hypotheses $H_{0,k}: \theta_k=0$ against the alternative  $H_{1,k}: \theta_k>0$ (larger $\theta_k$ being better).
-Fur this purpose test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments.
-In an interim analysis, treatments that are continued into the second stage are selected; the set of selected treatments is denoted by $\mathcal{S}$.
+For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments.
+These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis.
+In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe.
+In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$.
 The selection rules used here are described in Section \ref{ssec:selection_rules}.
-Based on the data from patients recruited into the second stage, test statistics $Z_{2,k}$ with $k \in \mathcal{S}$ are computed.
-
-final analysis based on combination test approach with closed test principle
-
-% first stage = interim analysis
-% second stage = final analysis
-% stick to first second stage throughout text but mention interim here?
+In the final analysis, the closed test principle is applied to account for the multiple hypotheses.
+The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function.
+Here we use the inverse normal combination function~\citep{lehmacher99}  % Lehmacher and Wassmer https://doi.org/10.1111/j.0006-341X.1999.01286.x.
+%FIXME Show function
 
@@
 \label{ssec:selection_rules}
+In practice, the totality of the data would be suggested by a data monitoring committee (DMC) to make a recommendation regarding the treatment selection.
+However, it is advisable for the sponsor to define a selection rule, although in comparison fairly simplistic, to provide some guidance for the DMC regarding sponsor's preferences.
+Furthermore, formal selection rules are indispensable to conduct simulation studies, e.g.
+to explore sample size and power.
+
+\paragraph{k-Rule} %best k, evtl. anderer Name?
 
@@
 \paragraph{Threshold Rule}
+
+\subsection{Simulating ASD}
+\label{ssec:simulating_asd}
+R package asd by Nick Parsons
+Test statistics are simulated, not individual participant data
 
@@
 \end{figure}
-
-\subsection{Simulating ASD}
-\label{ssec:simulating_asd}
-R package asd by Nick Parsons
-Test statistics are simulated, not individual participant data
-
-\clearpage
 
@@
 % subsection Summary
+When planning clinical trials, determining patient group sizes and test methods is a crucial task.
+The main goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
+If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
+Either because large computational resources are required for an exhaustive exploration of all different configurations, or because it involves a tedious manual manual try and error process.
+In this work we applied Model-based Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
+We compared out approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
+MBO was able to find competitive setups in a fraction of the time.
 

### 2020-12-09 main.tex
reading 1

@@
 
-\title[Model-based Optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian Optimization} %(please use sentence case, e.g.
+\title[Bayesian Optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian Optimization} %(please use sentence case, e.g.
 Score tests for exploring complex models)
@@
 Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
-In this work we propose to use Model-based Optimization to quickly find a clinical trial setup with high power from a large number of setups options.
+In this work we propose to use Bayesian Optimization to quickly find a clinical trial setup with high power from a large number of setups options.
 We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-We compare the MBO approach with an exhaustive evaluation of all candidate designs.
+We compare the Bayesian Optimization approach with an exhaustive evaluation of all candidate designs.
 The optimization approach finds competitive setups in a fraction of the time.
@@
 
-% TODO: Jakob copy contents from paper
 % Is there a condensed version? In Friede et al (2020) it is only explained in detail?
@@
 The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control.
-Interest is in assessing the treatment effects $\theta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs.
-control and testing the null hypotheses $H_{0,k}: \theta_k=0$ against the alternative  $H_{1,k}: \theta_k>0$ (larger $\theta_k$ being better).
+Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs.
+control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better).
 For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments.
@@
 The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function.
-Here we use the inverse normal combination function~\citep{lehmacher99}  % Lehmacher and Wassmer https://doi.org/10.1111/j.0006-341X.1999.01286.x.
+Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}  % Lehmacher and Wassmer https://doi.org/10.1111/j.0006-341X.1999.01286.x.
 %FIXME Show function
@@
 \end{tikzpicture}
-%FIXME Is this the right paper? Convert to tikz! Tim schickt mir das Original!
 %Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf

### 2020-12-09 main.tex
TODO notes

@@
 \pagespan{1}{}
-\keywords{Adaptive Seamless Designs; Clinical Trials; Model-Based Optimization; Treatment Selection\\\\ %FIXME: name 5 %Up to five keywords are allowed and should be given in alphabetical order.
+\keywords{Adaptive Seamless Designs; Bayesian Optimization; Clinical Trials; Treatment Selection\\\\ %Up to five keywords are allowed and should be given in alphabetical order.
 Please capitalize the key
@@
 Depending on the setting and the purpose of the simulation study, suitable metrics must be chosen to compare alternative designs or analysis strategies. 
-Especially relevant are neutral comparison studies, which are characterized as follows: %FIXME Is there someting missing? (Jakob)
+Especially relevant are neutral comparison studies, which are characterized as follows:
+%FIXME Is there someting missing? (Jakob)
 The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way~\citep{boulesteix_plea_2013}.
@@
 
-% TODO: Jakob 
-% phase 1 idea something fast
-% phase 2
-% Describe the study we look at. / The Data
-%COPD trial
-
 % What about these three papers?
@@
 The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function.
-Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}  % Lehmacher and Wassmer https://doi.org/10.1111/j.0006-341X.1999.01286.x.
+Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}
 %FIXME Show function
@@
   Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$.
-%FIXME rename n_cases in plot
   }
@@
 Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
-Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. %FIXME Make sure that this is mentioned for figure 4!
+Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
 In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
@@
 Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
-%FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)
 
@@
 This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{MBO} for the low and the high value of $n_\text{sim}$.
-%FIXME: Add Runtime comparision? Maybe boring because it probably is just 5 times the old values
 \begin{figure}[htb]

### 2020-12-09 main.tex
reading 2

@@
 The vast space of possible designs makes the optimization computationally demanding.
-In the following, we demonstrate that the so-called Model-based optimization (MBO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
+In the following, we demonstrate that the so-called \emph{Bayesian optimization} (BO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
 As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
@@
 
-MBO, often also called Bayesian optimization, has been successfully applied to optimize expensive black-boxes in many different scenarios.
+BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes in many different scenarios.
 Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
-Various implementations of MBO methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
+Various implementations of BO methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
 For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
 This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
-Also, MBO is well suited for multi-criteria optimization.
-Different adaptions of MBO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
+Also, BO is well suited for multi-criteria optimization.
+Different adaptions of BO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
 Many of them are implemented in \texttt{mlrMBO} as well.
@@
 In the next section we present as motivating example an application of adaptive seamless designs for chronic obstructive pulmonary disease (COPD).
-In the sections~\ref{sec:adaptive_seamless_designs} and~\ref{sec:model_based_optimization}, we explain the background on adaptive seamless designs (asd) and model-based optimization (MBO), respectively.
-In section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of MBO for optimizing trial designs is demonstrated.
+In the sections~\ref{sec:adaptive_seamless_designs} and~\ref{sec:bayesian_optimization}, we explain the background on adaptive seamless designs (asd) and Bayesian optimization (BO), respectively.
+In section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of BO for optimizing trial designs is demonstrated.
 In the last Section~\ref{sec:discussion}, we discuss promising extensions as well as limitations of the approach.
@@
 
-The general idea of this paper is to apply MBO for finding a trial design that is optimal with respect to a certain metric.
+The general idea of this paper is to apply BO for finding a trial design that is optimal with respect to a certain metric.
 As an example, we consider the maximization of the statistical power of a trail design, given a set of parameters to be chosen for the trial design.
-We compare the computation time as well as the maximal obtained power between the results obtained with MBO and with an exhaustive grid search over the space of adjustable parameters.
+We compare the computation time as well as the maximal obtained power between the results obtained with BO and with an exhaustive grid search over the space of adjustable parameters.
 The presented optimization approach can also be applied for optimizing other metrics than the power, as long as the metric itself is numeric and the parameters of the trial design are numeric and include none or just few categorical choices.
@@
 One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd}~\cite{parsons_software_2011} package and described in Section~\ref{sec:adaptive_seamless_designs}.
-%(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, ) 
-% https://doi.org/10.1002/bimj.200510232
+%(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, )  bauer_combining_1999
+% https://doi.org/10.1002/bimj.200510232 bretz_confirmatory_2006 
 % seamless trials (assoziert) 
-%  https://doi.org/10.1002/bimj.200510231
+%  https://doi.org/10.1002/bimj.200510231 schmidli_confirmatory_2006
 In~\citet{friede_adaptive_2020} the authors present an application of \emph{adaptive seamless designs} for chronic obstructive pulmonary disease (COPD).
-We will use this COPD study to apply MBO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
+We will use this COPD study to apply BO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
 %within the given scenarios restrictions.
@@
 In the final analysis, the closed test principle is applied to account for the multiple hypotheses.
-The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function.
-Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}
-%FIXME Show function
+The intersection hypotheses are tested by combining stagewise $p$-values for the respective intersection hypotheses using a prespecified combination function.
+Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}:
+\begin{equation}
+  \frac{1}{\sqrt{K}} \sum_{k=1}^K \Phi^{-1}(1-p_k) \,
+\end{equation}
+where $\Phi^{-1}$ denotes the quantile function of the standard normal distribution.
 
@@
 \label{ssec:simulating_asd}
-R package asd by Nick Parsons
-Test statistics are simulated, not individual participant data
+The R-package \texttt{asd} by~\citet{parsons_software_2011} simulates the test statistics and not individual participant data.
 
@@
   { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
-      \item Treatment effect on early interim outcome.
+      \item Treatment effect on early interim outcome
       \item Treatment effect on late disability outcome
@@
   { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
-      \item Correlation between early interim and late disability outcomes.
+      \item Correlation between early interim and late disability outcomes
     \end{itemize}
@@
       \item Total subjects in trial
-      \item recruitment rate
+      \item Recruitment rate
     \end{itemize}
@@
   \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (CombinationPoints) at (ConstrainedPoints.north east)
-  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
-      \item Adaptive seamless design time point of interim analysis
-      \item Treatment selection rules at interim Conventional phase 2/3
+  { Adaptive seamless design
+    \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
+      \item time point of interim analysis
+      \item treatment selection rules at interim Conventional phase 2/3
       \item split of resources between phases
@@
   \draw[->] (UnknownPoints) -- (ClinicalScenarios);
-  \draw[->] (CombinationPoints) -- (ClinicalScenarios);
+  \draw[->] (CombinationPoints.223) -- (ClinicalScenarios);
   \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
@@
 
-\section{Model-based Optimization}
-\label{sec:model_based_optimization}
+\section{Bayesian Optimization}
+\label{sec:bayesian_optimization}
 
-Sequential \textbf{Model-based Optimization (MBO)}~\citep{jones_taxonomy_2001} (also known as Bayesian optimization) is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
-In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
+Bayesian Optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
+In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
 Then it becomes infeasible to evaluate the black-box for thousands of configurations.
@@
 \subsection{General principle}
-MBO solves an optimization problem within a bounded search space $\Theta$:
+BO solves an optimization problem within a bounded search space $\Theta$:
 \[
@@
 where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
-To reduce the number of evaluations on $f$ the key idea of MBO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
+To reduce the number of evaluations on $f$ the key idea of BO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
 The estimate is generated by a so called \emph{surrogate model}.
@@
 First, an initial design of already evaluated configurations is needed.
-Then, iteratively, the MBO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
+Then, iteratively, the BO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
 The steps are repeated until a budget is exhausted.
@@
 
-The goal of this simulation study is to investigate the potential benefit of MBO for effectively finding a clinical trial design that maximizes a metric.
+The goal of this simulation study is to investigate the potential benefit of BO for effectively finding a clinical trial design that maximizes a metric.
 The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
@@
 
-In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand  MBO (Model-based optimization).
+In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand BO (Bayesian optimization).
 For different scenarios we compare the statistical power of the best found clinical trial design.
@@
 
-\paragraph{MBO (Model-based optimization)}
-For applying MBO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
+\paragraph{BO (Model-based optimization)}
+For applying BO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
 As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
@@
 
-As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:model_based_optimization}.
-We start MBO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
+As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
+We start BO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
 
@@
 
-The whole MBO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.
+The whole BO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.
 
-Additionally we investigate how \emph{MBO} performs when it is limited to return configurations that are within the grid.
-Therefore we introduce \textbf{MBO Grid}, which executes the optimization exactly as \emph{MBO}, but the best configuration is mapped to the closest value on the grid. 
-In other words, for \emph{MBO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{MBO}.
-The comparison of the results of \emph{MBO Gird} and \emph{Gird} will allow two possible conclusions:
-First, if \emph{MBO Grid} is worse then \emph{MBO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
-Second, if \emph{MBO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
+Additionally we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid.
+Therefore we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
+In other words, for \emph{BO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{BO}.
+The comparison of the results of \emph{BO Gird} and \emph{Gird} will allow two possible conclusions:
+First, if \emph{BO Grid} is worse then \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
+Second, if \emph{BO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
 Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.
@@
 The algorithms used in this study are implemented in R.
-For the model-based optimization, the R-package mlrMBO~\citep{bischl_mlrmbo_2017} is used.
-For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package asd~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
+For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
+For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
 \begin{table}[h]
@@
 Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $\tilde{f}$. 
-The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{MBO} are additional replicates.
+The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.
 
@@
 
-In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{MBO} and \emph{MBO Grid},
+In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{BO} and \emph{BO Grid},
 We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
@@
 
-The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
+The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
 Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
 Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
-In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
-Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension.
+In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+Across all scenarios, \emph{BO} and \emph{Grid} perform similar, even though \emph{BO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension.
 The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
-\emph{MBO} is approximately 20 times faster.
+\emph{BO} is approximately 20 times faster.
 \input{generated/tables/table_time.tex}
 
-For \emph{MBO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
-Comparing the results of \emph{MBO Grid} to those of \emph{MBO} we mainly see similar performances for both.
-This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{MBO}.
+For \emph{BO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
+Comparing the results of \emph{BO Grid} to those of \emph{BO} we mainly see similar performances for both.
+This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{BO}.
 
-Comparing \emph{MBO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{MBO Grid} achieves slightly better performance then \emph{Grid}.
+Comparing \emph{BO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{BO Grid} achieves slightly better performance then \emph{Grid}.
 The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e.
@@
 The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
-In these scenarios \emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
+In these scenarios \emph{BO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{BO}, as explained in Section~\ref{ssec:best_point}.
 Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
 
-If \emph{MBO} was not able to find a better solution then \emph{Grid}, then also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
-An increase in performance of \emph{MBO Grid} compared to \emph{MBO} is purely due to chance.
+If \emph{BO} was not able to find a better solution then \emph{Grid}, then also \emph{BO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{BO}.
+An increase in performance of \emph{BO Grid} compared to \emph{BO} is purely due to chance.
 
-To compare \emph{MBO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
-Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{MBO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
-Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
-Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
+To compare \emph{BO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
+Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
+Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{BO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
+Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
 In the first case, the absolute differences are negligible as all methods lead to a high power.
@@
 In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
-The advantage of \emph{MBO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect:paper2}, $n_{\text{treat}} = 500$.
+The advantage of \emph{BO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect:paper2}, $n_{\text{treat}} = 500$.
 
 Note that between the scenarios the absolute differences of the optimization results are of different practical relevance,
-i.e.\ for scenario (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$), \emph{Grid} is superior to \emph{MBO} but the difference is smaller than $0.25\%$.
+i.e.\ for scenario (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$), \emph{Grid} is superior to \emph{BO} but the difference is smaller than $0.25\%$.
 
-%After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{MBO} to find equally good configurations as the much more demanding exhaustive \emph{Grid}
+%After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{BO} to find equally good configurations as the much more demanding exhaustive \emph{Grid}
 We now investigate the results in more detail.
-In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{MBO} are shown, in order to analyze the differences in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
+In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{BO} are shown, in order to analyze the differences in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
 \begin{figure}[htb]
@@
   A line connects the mean outcomes for each configuration.
-  Additionally the best found configurations from \emph{MBO} are displayed.
+  Additionally the best found configurations from \emph{BO} are displayed.
   }
@@
 As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. 
-In all scenarios, \emph{MBO} identified the peak of all combined curves.
-Within the 20 stochastic repetitions per scenario, \emph{MBO} found different configurations, but those had similarly good outcomes.
+In all scenarios, \emph{BO} identified the peak of all combined curves.
+Within the 20 stochastic repetitions per scenario, \emph{BO} found different configurations, but those had similarly good outcomes.
 That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.
@@
 %In contrast to Figure~\ref{fig:plot_allbest},
-Figure~\ref{fig:plot_best_x} visualizes in detail which configurations were selected by \emph{MBO} and \emph{Grid} and the corresponding values for $y_{\text{valid}}$ (result of the independent validation).
+Figure~\ref{fig:plot_best_x} visualizes in detail which configurations were selected by \emph{BO} and \emph{Grid} and the corresponding values for $y_{\text{valid}}$ (result of the independent validation).
 \begin{figure}[htb]
@@
 \includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
-\caption{Best found $x$-values for \emph{Grid} and \emph{MBO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
+\caption{Best found $x$-values for \emph{Grid} and \emph{BO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
 \label{fig:plot_best_x}
@@
 In such cases the signal to noise ratio was high enough to reliably find the best configuration.
-These are also those scenarios where \emph{Grid} was superior the \emph{MBO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
+These are also those scenarios where \emph{Grid} was superior the \emph{BO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
 Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500)$ and \emph{effect: paper}, $n_{\text{treat}} = 500)$, small values of $r$ lead to an optimal outcome.
-Here, \emph{MBO} was not able to find these values close to the border of the search space.
+Here, \emph{BO} was not able to find these values close to the border of the search space.
 A common remedy is to log-transform this value.
 
-In other cases \emph{Grid} found varying configurations and \emph{MBO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+In other cases \emph{Grid} found varying configurations and \emph{BO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
 A special case is scenario \emph{effect: paper2}, $n_{\text{treat}} = 2000$:
@@
 For the exhaustive \emph{Grid}, we expect that the decreased noise leads to less selections of overly optimistic outcomes as result.
-Also \emph{MBO} should profit from a less noisy objective function as the surrogates predictions will be more precise.
+Also \emph{BO} should profit from a less noisy objective function as the surrogates predictions will be more precise.
 
@@
 So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
-Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{MBO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
-This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{MBO} for the low and the high value of $n_\text{sim}$.
+Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{BO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
+This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{BO} for the low and the high value of $n_\text{sim}$.
 \begin{figure}[htb]
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
-    \captionof{figure}{Optimization progress for \emph{MBO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
+    \captionof{figure}{Optimization progress for \emph{BO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
     \label{fig:plot_opt_path_5000}
@@
 But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
-We conclude with the observation, that \emph{MBO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.
+We conclude with the observation, that \emph{BO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.
 
@@
 We compared out approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-MBO was able to find competitive setups in a fraction of the time.
+BO was able to find competitive setups in a fraction of the time.
 
@@
 
-IDEA: Restricting MBO to only evaluate configurations on the grid to avoid evaluating identical / very close configurations.
+IDEA: Restricting BO to only evaluate configurations on the grid to avoid evaluating identical / very close configurations.
 IDEA: log transform parameters that have optimal settings close to the border.

### 2020-12-09 generator/main.R

@@ -76,7 +76,7 @@ select_labels_colors = #http://paletton.com/#uid=74Z140kqdu7ghF3lowyu1ppwGk7
     "random" = "#000000",
     "threshold rule" = "#C0E62A"
   )
-algorithm_labels = c("mbo" = "MBO", "eval" = "Grid", "mbo grid" = "MBO Grid", "grid" = "Grid", "grid7" = "Grid Small")
+algorithm_labels = c("mbo" = "BO", "eval" = "Grid", "mbo grid" = "BO Grid", "grid" = "Grid", "grid7" = "Grid Small")
 algorithm_labels_color = c("mbo" = "#F04A2B", "eval" = "#099438", "mbo grid" = "#690D86", "grid" = "#099438", "grid7" = "#44CA6D") #https://paletton.com/#uid=6060R0krPwadDMilcBuwcpxDti6
 
 ncases_labels = c(500,1000,2000)

### 2020-12-09 main.tex
reading 3

@@
 In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
-Then it becomes infeasible to evaluate the black-box for thousands of configurations.
+As then it becomes infeasible to evaluate the black-box for thousands of configurations.
 
@@
 where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
-To reduce the number of evaluations on $f$ the key idea of BO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
+To reduce the number of evaluations on $f$ the key idea of BO is to only evaluate values of $\theta$ that are estimated to lead to a high value of $f(\theta)$.
 The estimate is generated by a so called \emph{surrogate model}.
@@
 The proposal of a new configuration $\theta$ is obtained by maximizing a so called acquisition function.
-The acquisition function guides the search to promising new regions in the search space $\Theta$ by facilitating the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty prediction $\hat{s}^2(\theta)$ of the surrogate.
+The acquisition function guides the search to promising new regions in the search space $\Theta$ by facilitating the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty quantification $s^2(\theta)$ of the surrogate.
 It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation.
@@
 \begin{equation}
-\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2_{n}) %\text{ i.i.d.}
+\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2) %\text{ i.i.d.}
 \end{equation}
@@
 \begin{equation}
-  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot \hat{s}(\theta),
+  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot s(\theta),
 \end{equation}
@@
   \label{eq:aei}
-  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{\hat{s}(\theta)} \biggr) \\
-   + \hat{s}(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{\hat{s}(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma_{n}}{\sqrt{\sigma_{n}^2 + \hat{s}^2(\theta)}}}_{\text{correction}}\biggr),
+  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \\
+   + s(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma}{\sqrt{\sigma^2 + s^2(\theta)}}}_{\text{correction}}\biggr),
 \end{multline}
-where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma_{n}^2$ denotes the random error (nugget effect) in the Kriging model.
+where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma^2$ denotes the random error (nugget effect) in the Kriging model.
 The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
@@
 The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
-The correction term is necessary to avoid exploration in areas where the model uncertainty $\hat{s}^2(\theta)$ equals the random error $\sigma_{n}^2$, because in this case further evaluations will not decrease the model uncertainty.
+The correction term is necessary to avoid exploration in areas where the model uncertainty $s^2(\theta)$ equals the random error $\sigma^2$, because in this case further evaluations will not decrease the model uncertainty.
 
@@
 If we chose the configuration $\theta$ from all previously evaluated configurations according to the best outcome of $f$ as the optimization result, we would be overly optimistic.
-It is likely that a single best outcome can be partially attributed to the random error $\sigma_{n}^2$.
+It is likely that a single best outcome can be partially attributed to the random error $\sigma^2$.
 Instead, we are interested in the optimum of the true posterior mean.
@@
 In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
-The advantage of \emph{BO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect:paper2}, $n_{\text{treat}} = 500$.
+The advantage of \emph{BO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.
 
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
-    \captionof{figure}{Optimization progress for \emph{BO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
+    \captionof{figure}{Optimization progress for \emph{BO}: Best found outcome so far for each iteration in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
     \label{fig:plot_opt_path_5000}
@@
 Either because large computational resources are required for an exhaustive exploration of all different configurations, or because it involves a tedious manual manual try and error process.
-In this work we applied Model-based Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
+In this work we applied Bayesian Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
 We compared out approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.

### 2020-12-09 main.tex
besprechung 9.12.2020

@@
 Depending on the setting and the purpose of the simulation study, suitable metrics must be chosen to compare alternative designs or analysis strategies. 
-Especially relevant are neutral comparison studies, which are characterized as follows:
+Especially relevant are neutral comparison studies, which are characterized by~\citet{boulesteix_plea_2013} as follows:
 %FIXME Is there someting missing? (Jakob)
-The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way~\citep{boulesteix_plea_2013}.
+The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way.
 %While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios.
-A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation). ref %FIXME ref
+A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation)~\citep{friede_refinement_2010}.
 CSE was particularly developed to support the overall process of performing simulations of clinical trials, especially for complex designs and analysis strategies.
@@
 The proposal of a new configuration $\theta$ is obtained by maximizing a so called acquisition function.
-The acquisition function guides the search to promising new regions in the search space $\Theta$ by facilitating the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty quantification $s^2(\theta)$ of the surrogate.
+The acquisition function guides the search to promising new regions in the search space $\Theta$ by using the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty quantification $s^2(\theta)$ of the surrogate.
 It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation.

### 2020-12-10 main.tex
reformulate "5.2 caclulation of the statistical power"

@@
 Various frameworks and guidance proposals have been developed over the past years on how to plan, execute and interpret the results of such simulation studies.
-Method comparison studies are used to give clinicians and biostatisticians guidance when an approach improves currently used methods.  
+Method comparison studies are used to give clinicians and biostatisticians guidance when an approach improves currently used methods~\citep{hanneman_design_2008}.  
 Depending on the setting and the purpose of the simulation study, suitable metrics must be chosen to compare alternative designs or analysis strategies. 
@@
 %While method comparison studies aim to make statements when one approach might dominate others in terms of the chosen metrics, in designing an individual study or series of studies the objective is to identify a design that is optimal in terms of the metrics chosen in a certain scenario or, more importantly, across a set of scenarios.
-A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation)~\citep{friede_refinement_2010}.
+A comprehensive framework for such an objective assessment of competing strategies for clinical trial designs is CSE (Clinical Scenario Evaluation)~\citep{benda_aspects_2010,friede_refinement_2010}.
 CSE was particularly developed to support the overall process of performing simulations of clinical trials, especially for complex designs and analysis strategies.
@@
 %To obtain a final outcome $\nu^\ast = f(\theta^\ast)$ that is independent of the optimization process, $f(\theta^\ast)$ is calculated again.
-Using the stochastic outcome $\hat{\nu}^\ast := f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
+Using the stochastic outcome $f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
 Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimate of the true value of the best outcome.
@@
 
-The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
+The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
+The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
 The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control).
@@
 
-Let $n_{\text{treat}}$ be the total number of treatments in the overall trial, and let $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively.
+The simulation of the statistical power of the trail design can be formulated as a black-box function as follows:
+\begin{align}
+  \label{eq:bbox}
+  y & = f(n_{\text{stage1}}, n_{\text{stage2}}, \text{selection strategy}, \epsilon, \tau, \text{effect set}) \ ,
+\end{align}
+whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively and \emph{effect set} remains fixed as we are interested in the optimal trail design for a given \emph{effect set}.
+Our main optimization goal is to maximize the statistical power $y$ depending on the trial design.
+
+However, we are interested in keeping the total number of treatments $n_{\text{treat}}$ constant in the overall trail for a fair comparison of the power of different designs.
+Therefore we introduce the ratio of patients per treatment in stage 1, compared to patients per treatment in both stages
+\begin{align}
+  \label{eq:r}
+ r & = \frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}
+\end{align}
+as a new parameter that will replace $n_{\text{stage1}}$ and $n_{\text{stage2}}$ in~(\ref{eq:bbox}).
+
+In the following we will explain how we derive the values of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ in dependence of $r$ and $n_{\text{treat}}$ so that we can maximize
+\begin{equation}
+  y = \tilde{f}(r, \text{selection strategy}, \epsilon, \tau, \text{effect set}, n_{\text{treat}})
+\end{equation}
+instead of~(\ref{eq:bbox}), whereas $n_{\text{treat}}$ and \emph{effect set} remain fixed.
+
 Let $k_1$ and $k_2$ denote the number of treatments in stage 1 and stage 2, respectively. 
@@
   \label{eq:ntreat}
-  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}}.
+  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}} \ .
 \end{equation}
@@
 
-Our main optimization goal is to maximize the statistical power $y$ depending on the trial design. 
-The power can be formulated as a black-box as follows:
-\begin{align}
-  \label{eq:bbox}
-  y & = f(n_{\text{stage1}}, n_{\text{stage2}}, \text{effect set}, \text{selection strategy}, \epsilon, \tau).
-\end{align}
-
-The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
-We denote the proportion of patients per treatment in stage 1, compared to patients per treatment in both stages, with 
-\begin{align}
-  \label{eq:r}
- r & = \frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.
-\end{align}
 Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
@@
   \frac{r}{k_1\cdot
-  r+k_2\cdot (1-r)}, \nonumber \\
+  r+k_2\cdot (1-r)} \ \text{and} \nonumber \\
   n_{\text{stage2}} & =
@@
 This is only the case for selection strategies 1, 2, and 3.
+
 For the selection strategies \emph{eps} and \emph{thresh}, $k_2$ depends on the outcome of stage 1.
@@
 %The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
+The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
 
-The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
-Then the black-box from formula (\ref{eq:bbox}) can be reformulated as 
-\begin{equation}
-  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \epsilon, \tau).
-\end{equation}
-In our simulations, $n_{\text{treat}}$ is considered to be a constant.
-
-Summarizing, in the following we aim to find the clinical design setup, represented by the parameter values for $r$, \emph{effect set}, \emph{Selection strategy}, $\epsilon$ and $\tau$ that yield the highest statistical power, when evaluating $\tilde{f}$.
+Summarizing, in the following we aim to find the clinical design setup, represented by the parameter values for $r$, \emph{effect set}, \emph{Selection strategy}, $\epsilon$ and $\tau$ that yield the highest statistical power, when evaluating~$\tilde{f}$.
 

### 2020-12-10 main.tex
jakob ready

@@
 \usepackage{xcolor}
+\usepackage{nowidow}
+\usepackage{csquotes}
+\MakeOuterQuote{"}
 
@@
 \begin{equation}
+  \label{eq:bbox2}
   y = \tilde{f}(r, \text{selection strategy}, \epsilon, \tau, \text{effect set}, n_{\text{treat}})
@@
 
-Summarizing, in the following we aim to find the clinical design setup, represented by the parameter values for $r$, \emph{effect set}, \emph{Selection strategy}, $\epsilon$ and $\tau$ that yield the highest statistical power, when evaluating~$\tilde{f}$.
+The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and number of total treatments, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~\ref{eq:bbox2}.
 
@@
 
-In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand BO (Bayesian optimization).
+In the simulation study we compare the ability of two optimization approaches for finding an optimal parameter configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand BO (Bayesian optimization).
 For different scenarios we compare the statistical power of the best found clinical trial design.
@@
 
-The best point per evaluated grid is determined by selecting the configuration with the best outcome, i.e.\  the parameter set with the largest power.
+The best point per evaluated grid is determined by selecting the configuration with the best outcome, i.e.\  the concrete trail design with the largest power.
 The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the quality of the outcome.
@@
 \paragraph{BO (Model-based optimization)}
-For applying BO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
+For applying BO, we first define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter configurations) is generated.
 As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
@@
 Again, overoptimism must be avoided. 
-Thus, the best selected configuration (the best parameter set) is then independently evaluated $20$ times again.
+Thus, the best selected parameter configuration is then independently evaluated $20$ times again.
 
-The whole BO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.
+The whole BO procedure is repeated $20$ times to estimate the variance of the optimal found solution.
 
-Additionally we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid.
+Additionally we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid introduced previously.
 Therefore we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
@@
 The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
-Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $\tilde{f}$. 
+Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
 The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.
@@
 The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers of $n_{\text{treat}}$.
-Each optimizer is applied on each scenario.
+Each optimizer is applied on each scenario with 20 stochastic repetitions.
 
@@
 \caption{%
-  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$.
+  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$
   }
@@
 \end{figure}
-
-The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
+The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
 Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
@@
 Comparing \emph{BO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{BO Grid} achieves slightly better performance then \emph{Grid}.
-The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e.
+The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcomes and has no notion of the stochasticity of the problem, i.e.
 in this case the noise on the predicted power.
@@
 In these scenarios \emph{BO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{BO}, as explained in Section~\ref{ssec:best_point}.
-Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
+Further below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
 
@@
 
-To compare \emph{BO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
-Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
+To compare \emph{BO} with a grid search that uses approximately the same time (see Table~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
+Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO}, with only a small advantage for \emph{Grid Small}.
 Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{BO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
@@
 Note that between the scenarios the absolute differences of the optimization results are of different practical relevance,
-i.e.\ for scenario (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$), \emph{Grid} is superior to \emph{BO} but the difference is smaller than $0.25\%$.
+i.e.\ for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$, \emph{Grid} is superior to \emph{BO} but the difference is smaller than $0.25\%$.
 
@@
 We now investigate the results in more detail.
-In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{BO} are shown, in order to analyze the differences in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
+In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{BO} are shown, in order to analyze the differences in dependence of the \emph{selection strategy} and the number of treatments at stage 1 determined by the \emph{stage ratio} ($r$).
 \begin{figure}[htb]
@@
 \end{figure}
-Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments in the first stage.
-For the \emph{selection strategies} \emph{epsilon rule} and \emph{threshold rule} only curves representing the $\epsilon$ and $\tau$ values with the best outcome, respectively, are shown.
+For the selection strategies \emph{epsilon rule} and \emph{threshold rule} only curves representing the $\epsilon$ and $\tau$ values with the best outcome, respectively, are shown.
 Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are as expected nearly identical.
@@
 Within the 20 stochastic repetitions per scenario, \emph{BO} found different configurations, but those had similarly good outcomes.
-That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.
-different configurations yield similarly good outcomes.
+That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.\ different configurations yield similarly good outcomes.
 In cases in which multiple optimization runs result in different configurations it might be advisable to let a human expert choose the final configuration, also depending on the medical background.
@@
 
-%FIXME: Write more about the meaning of the results from a clinical point of view.
+%TODO: Write more about the meaning of the results from a clinical point of view?
 
-%In contrast to Figure~\ref{fig:plot_allbest},
 Figure~\ref{fig:plot_best_x} visualizes in detail which configurations were selected by \emph{BO} and \emph{Grid} and the corresponding values for $y_{\text{valid}}$ (result of the independent validation).
@@
 In other cases \emph{Grid} found varying configurations and \emph{BO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+Here we can assume, that \emph{BO} successfully accounted for the noise.
 A special case is scenario \emph{effect: paper2}, $n_{\text{treat}} = 2000$:
@@
 
-To investigate the effect of the function noise variance ($\sigma^2_n$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario \emph{effect: paper}, $n_{\text{treat}} = 2000$.
+To investigate the effect of the function noise variance ($\sigma^2$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario \emph{effect: paper}, $n_{\text{treat}} = 2000$.
 % This is expected to increase the runtime by a factor of 5. % obvious, also not of interest here
@@
 Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
-Furthermore, the average optimization curve indicates that less iterations can be sufficient to obtain an outcome that is comparable to \emph{Grid},
 It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
-So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
-Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{BO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
-This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{BO} for the low and the high value of $n_\text{sim}$.
+%So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
+%Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{BO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
+Despite the differences of the optimization paths, the median validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{BO} for the low and the high value of $n_\text{sim}$.
+Therefore we can assume that "bias correction" of \emph{BO} as explained in Section~\ref{ssec:best_point} seems to be effective.
 \begin{figure}[htb]
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
-    \captionof{figure}{Optimization progress for \emph{BO}: Best found outcome so far for each iteration in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
+    \captionof{figure}{Optimization progress for \emph{BO}: Best found outcome so far for each iteration in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$}
     \label{fig:plot_opt_path_5000}
@@
 But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
-We conclude with the observation, that \emph{BO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.
+We conclude with the observation, that \emph{BO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulations.
 
@@
 In this work we applied Bayesian Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
-We compared out approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-BO was able to find competitive setups in a fraction of the time.
+We compared our approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
+BO was able to find competitive setups in a fraction of the time and was able to cope well with the stochasticity of the simulation.
 
 % subsection Outlook
+One drawback of BO that we observed was, that it performed poorly if the optimal values of one parameter are close to the lower border of the search region.
+Therefore we would apply log transformation to such parameter in the future.
 
@@
 
-Another way of speeding up the optimization would be to adaptively increase the iterations of the design simulation. 
+Another way of speeding up the optimization would be to adaptively increase the simulation rounds of the trail design evaluation. 
 If few simulations already indicate a bad outcome it might not be promising to evaluate more simulation runs.
@@
 
-IDEA: Restricting BO to only evaluate configurations on the grid to avoid evaluating identical / very close configurations.
-IDEA: log transform parameters that have optimal settings close to the border.
+% IDEA: Restricting BO to only evaluate configurations on the grid to avoid evaluating identical / very close configurations.
 

### 2020-12-10 generator/main.R

@@ -93,7 +93,7 @@ tmp = lapply(tmp, function(x) cbind.data.frame(stage = rownames(x), x))
 tmp = rbindlist(tmp, idcol = TRUE)
 setnames(tmp, ".id", "Effect Set")
 setnames(tmp, "stage", "Stage")
-kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation") %>% 
+kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation in the first (early) and second (final) stage") %>% 
   collapse_rows(columns = 1:2, latex_hline = "major") %>% 
   add_header_above(c(" " = 2, "Treatment" = 5)) %>% 
   kable_styling(position = "center") %>% 
@@ -437,14 +437,14 @@ plot_wrapper(name = "plot_opt_path_5000", fig.height = 1.6 * FIG_HEIGHT * 0.35, 
   
   g = ggplot(df[dob > 0 & prop.type != "final_eval", ], aes(x = dob, y = cummax_y-best_y, color = algorithm))
   g = g + geom_line(alpha = 0.1, aes(group = paste0(repl)), size = 0.5)
-  g = g + geom_line(data = dfmean, aes(y = cummax_y_mean-best_y), color = colorspace::darken(algorithm_labels_color[["mbo"]], amount = 0.6))
+  g = g + geom_line(data = dfmean, aes(y = cummax_y_mean-best_y), color = colorspace::darken(algorithm_labels_color[["mbo"]], amount = 0.1))
   g = g + facet_grid(.~nsim, scales = "free")
-  g = g + geom_hline(yintercept = 0 , color = colorspace::darken(algorithm_labels_color[["grid"]], amount = 0.6))
+  g = g + geom_hline(yintercept = 0 , color = colorspace::darken(algorithm_labels_color[["grid"]], amount = 0.1))
   g = g + geom_label(data = df_best, aes(label = formatC(best_y, 4)), y = 0, x = 90, vjust = 1.5, show.legend = FALSE)
   g = g + scale_color_manual(labels = algorithm_labels, values = algorithm_labels_color)
   g = g + theme(legend.position = "bottom")
   g = g + labs(x = "iteration", y = expression(y[iter]*"*" - y[grid]*"*"), color = NULL)
-  g = g + guides(colour = guide_legend(override.aes = list(size=2)))
+  g = g + guides(colour = guide_legend(override.aes = list(size=1, alpha = 1)))
[…]

### 2020-12-10 main.tex
fine tuning

@@
 
-\title[Bayesian Optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian Optimization} %(please use sentence case, e.g.
+\title[Bayesian optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian optimization} %(please use sentence case, e.g.
 Score tests for exploring complex models)
@@
 Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
-In this work we propose to use Bayesian Optimization to quickly find a clinical trial setup with high power from a large number of setups options.
+In this work we propose to use Bayesian optimization to quickly find a clinical trial setup with high power from a large number of setups options.
 We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-We compare the Bayesian Optimization approach with an exhaustive evaluation of all candidate designs.
+We compare the Bayesian optimization approach with an exhaustive evaluation of all candidate designs.
 The optimization approach finds competitive setups in a fraction of the time.
@@
 
-Bayesian Optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
+Bayesian optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
 In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
@@
 Either because large computational resources are required for an exhaustive exploration of all different configurations, or because it involves a tedious manual manual try and error process.
-In this work we applied Bayesian Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
+In this work we applied Bayesian optimization to quickly find a clinical trial setup with high power from a large number of configuration options.
 We compared our approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.

### 2020-12-21 main.tex
Updates from Overleaf

@@
 
-\title[Bayesian optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian optimization} %(please use sentence case, e.g.
-Score tests for exploring complex models)
+\title[Bayesian optimization of Adaptive Seamless Designs]{Improving Adaptive Seamless Designs through Bayesian optimization} %(please use sentence case, e.g. Score tests for exploring complex models)
 %% Information for the first author.
@@
 \author[dd]{Tim Friede\inst{2}}
-\address[\inst{2}]{Institut für Medizinische Statistik, Universitätsmedizin Göttingen, 37073 Göttingen}
+\address[\inst{2}]{Institut für Medizinische Statistik, Universitätsmedizin Göttingen, and Deutsches Zentrum für Herz-Kreislauf-Forschung (DZHK), Standort Göttingen, 37073 Göttingen}
 %%%%    Information for the third author
@@
 \begin{abstract}
-In clinical trials, planning patient group sizes and test methods is a crucial task.
+In this paper we propose to use Bayesian optimization (BO) to improve the efficiency of the selection process of designs in clinical trials.
+BO is a method to optimize expensive black-box functions that uses a regression as a surrogate to guide the search.
+In clinical trials, planning test procedures and sample sizes is a crucial task.
 A common goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
@@
 Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
-In this work we propose to use Bayesian optimization to quickly find a clinical trial setup with high power from a large number of setups options.
+In this work we propose to use BO to quickly find a clinical trial setup with high power from a large number of setups options.
 We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-We compare the Bayesian optimization approach with an exhaustive evaluation of all candidate designs.
+We compare the BO approach with an exhaustive evaluation of all candidate designs.
 The optimization approach finds competitive setups in a fraction of the time.
@@
 % FIXME: Do we really take in this Assurance?
-Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
+% Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
 
@@
 We start by providing some more details on the study design considered (Sections \ref{ssec:two_stage_design_with_treatment_selection} and \ref{ssec:selection_rules}) and then introduce briefly a framework for optimizing such designs, namely clinical scenario evaluation (CSE) in Section \ref{ssec:clinical_scenario_evaluation_for_asd}.
-In Section \ref{ssec:simulating_asd} we make some brief comments on the simulation model.
+%In Section \ref{ssec:simulating_asd} we make some brief comments on the simulation model.
 
@@
 The intersection hypotheses are tested by combining stagewise $p$-values for the respective intersection hypotheses using a prespecified combination function.
-Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}:
+Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999} which is given by
 \begin{equation}
-  \frac{1}{\sqrt{K}} \sum_{k=1}^K \Phi^{-1}(1-p_k) \,
+   w_1 \Phi^{-1}(1-p_{\mathcal{K},1}) + w_2 \Phi^{-1}(1-p_{\mathcal{K},2}) \,
 \end{equation}
-where $\Phi^{-1}$ denotes the quantile function of the standard normal distribution.
+where $\Phi^{-1}(\cdot)$ denotes the quantile function of the standard normal distribution, $w_1$ and $w_2$ weights with $w_1^2+w_2^2 = 1$, and $p_{\mathcal{K},1}$ and $p_{\mathcal{K},2}$ stagewise $p$-values testing intersection hypothesis $H_{\mathcal{K}}$.
+The p-values for stages $1$ and $2$ are calculated on those patient recruited in stages $1$ and $2$, respectively.
+Note that one might not be able to calculate $p_{\mathcal{K},1}$ at the time of the interim analysis since some patients might still be under follow-up for the final endpoint.
 
@@
 \label{ssec:selection_rules}
-In practice, the totality of the data would be suggested by a data monitoring committee (DMC) to make a recommendation regarding the treatment selection.
-However, it is advisable for the sponsor to define a selection rule, although in comparison fairly simplistic, to provide some guidance for the DMC regarding sponsor's preferences.
+In practice, the totality of the data would be considered by a data monitoring committee (DMC) to make a recommendation regarding the treatment selection.
+However, it is advisable for the sponsor to define a selection rule, although in comparison fairly simplistic, to provide some guidance for the DMC regarding the sponsor's preferences.
 Furthermore, formal selection rules are indispensable to conduct simulation studies, e.g.
 to explore sample size and power.
+In the following we give some examples of selection rules which have previously been consider and which will feature below.
 
-\paragraph{k-Rule} %best k, evtl. anderer Name?
+The treatment selection might consider statistics $Z_{1,k}$ for the final (primary) outcome or statistics $Z_{1,k}^{(E)}$ calculated based on some early outcome.
+We present the selection rules using $Z_{1,k}^{(E)}$, but of course these would be replaced by $Z_{1,k}$ if the final outcome was used for treatment selection.
+Furthermore, larger statistics are considered better.
 
-\paragraph{Epsilon Rule}
+\paragraph{$\kappa$-best Rule} Let $Z_{1,(1)}^{(E)}, \dots, Z_{1,(K)}^{(E)}$ denote the ordered statistics with $Z_{1,(1)}^{(E)}$ being the largest and therefore best test statistic.
+The $\kappa$-best rule then selects the treatments associated with the $\kappa$ largest $Z_{1,k}^{(E)}$, i.e.$Z_{1,(1)}^{(E)}, \dots, Z_{1,(\kappa)}^{(E)}$.
 
-\paragraph{Threshold Rule}
+\paragraph{Epsilon Rule} To our knowledge this rule was first described by \citet{kelly_adaptive_2005} and has since then featured in a number of simulation studies including \citet{friede_comparison_2008, friede_adaptive_2020}.
+All treatments $k$ with $Z_{1,k}^{(E)} \ge \max_i Z_{1,i}^{(E)} - \epsilon$ with $\epsilon \ge 0$ are selected.
+With $\epsilon=0$ this rule is equivalent to the $\kappa$-best rule with $\kappa=1$.
 
-\subsection{Simulating ASD}
-\label{ssec:simulating_asd}
-The R-package \texttt{asd} by~\citet{parsons_software_2011} simulates the test statistics and not individual participant data.
+\paragraph{Threshold Rule} With this rule all treatments $k$ are selected with $Z_{1,k}^{(E)} \ge \tau$ where $\tau$ is a given threshold.
+
+We note that with the $\kappa$-best rule a fixed number of treatments is carried forward into the next stage whereas these are variable with the epsilon rule and the threshold rule.
 
@@
 
+The clinical scenario evaluation framework was first proposed by \citet{benda_aspects_2010} and subsequently developed further by \citet{friede_refinement_2010}. 
+Figure \ref{fig:friede_refinement_2010_fig2} provides a summary of this framework.
 \begin{figure}[ht]
@@
     \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
-      \item time point of interim analysis
-      \item treatment selection rules at interim Conventional phase 2/3
-      \item split of resources between phases
-      \item end oh phase futility criteria
+      \item Time point of interim analysis
+      \item Treatment selection rules at interim 
+      \item Conventional phase 2/3
+      \item Split of resources between phases
+      \item End oh phase futility criteria
     \end{itemize}
@@
 %Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
-\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis.}
+\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis (Figure adapted from \citet{friede_refinement_2010}).}
 \label{fig:friede_refinement_2010_fig2}
 \end{figure}
+A clinical scenario evaluation is defined by the following three elements: (i) disease-specific features; (ii) design options; and (iii) design performance measures.
+The design specific features describe distributional assumptions on the early and final outcomes including variances, correlations and treatment effect sizes.
+Here, those that can be estimated from data of previous clinical trials and those that are simply unknown.
+Considering adaptive seamless design, the design options include the time point of the interim analysis and the treatment selection rule, which can in principle both be chosen without any restrictions.
+In contrast, some other design options such as the number of treatments and the total sample size are constrained by the environment and infrastructure available.
+The performance measures (sometimes also referred to as metrics) are in the context with adaptive seamless designs e.g.
+power or sample size distributions.
+%\paragraph{Simulating ASD}
+%\label{ssec:simulating_asd}
+The evaluation of some performance measures requires Monte Carlo simulations, since no closed form solutions are available. 
+The R-package \texttt{asd} by~\citet{parsons_software_2011} can be used for such simulations.
+As it generates test statistics it is generally more efficient in comparison to simulating individual participant data.
+The simulation model was described in more detail by \citet{friede_adaptive_2020}. 
+It features selection at interim based on early as well as final outcomes.
+
+% Some corrections to the figure needed: e.g. design options (right panel)
 
@@
 
-\clearpage
-
 \section{Simulation Study}
@@
   \hline
-  \emph{all}     & Select all arms  \\
-  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
+  \emph{$\kappa$-best}    & Select $\kappa = 1, 2, 3$ or 4 (all) arms, respectively, according to maximal test statistic \\
   \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
@@
   \hline
-  Selection strategy               & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thresh} \}$ \\
+  Selection strategy               & \emph{\{1-best, 2-best, 3-best, all, eps, thresh\}} \\
   $r$ (ratio)                      & $(0,1)$ \\
@@
 
-\paragraph{Grid}
-We conduct an \emph{exhaustive grid search} with a resolution of 25 points per real-valued dimension and one point for each categorical dimension.
-Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of 25.
+\paragraph{Grid and Grid small}
+We conduct an \emph{exhaustive grid search} with a resolution of $k$ points per real-valued dimension and one point for each categorical dimension.
+Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of $k$.
 All 6 selection strategies have the numeric parameter $r$.
 The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively.
-This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
-We call this approach \textbf{Grid}.
-Further, we also consider a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \textbf{Grid Small}.
+Using $k=25$, we call this approach \textbf{Grid},
+with in a total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
+Using a subset of this grid with $k=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, we call the method \textbf{Grid Small}.
 Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.
@@
 
-\paragraph{BO (Model-based optimization)}
+\paragraph{BO (Model-based optimization) and BO Grid}
 For applying BO, we first define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter configurations) is generated.
@@
 
-The whole BO procedure is repeated $20$ times to estimate the variance of the optimal found solution.
+The whole BO procedure is repeated $20$ times to estimate the variance of the optimal solution found.
 
@@
 For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
+The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
+Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
+The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.
+
+% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
+
 \begin{table}[h]
@@
 \end{table}
-The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
-Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
-The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.
 
-% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
+\subsection{Main results for comparison of algorithms}
 
-\subsection{Main results}
-
-In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{BO} and \emph{BO Grid},
+In this chapter we present an overview of the results for the comparison of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{BO} and \emph{BO Grid}.
 We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
-The replicated evaluations give us an unbiased estimate of the performance ($y_{\text{valid}}$) of each optimizer.
+The evaluations provide unbiased estimates of the performance ($y_{\text{valid}}$) of all optimizers.
 
 For this study we consider 12 scenarios.
-The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers of $n_{\text{treat}}$.
+The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers for $n_{\text{treat}}$.
 Each optimizer is applied on each scenario with 20 stochastic repetitions.
@@
 Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
-\caption{%
-  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$
-  }
-\label{fig:plot_boxplot_valid_y}
-\end{figure}
 The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
@@
 In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
-Across all scenarios, \emph{BO} and \emph{Grid} perform similar, even though \emph{BO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension.
+Across all scenarios, \emph{BO} and \emph{Grid} perform similarly, although due to its evaluation budget, \emph{BO} is allowed only 116 evaluations as opposed to the 1350 evaluations for \emph{Grid}, when using a resolution of 25 points per dimension.
 The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
@@
 
+\begin{figure}[tbh]
+\centering
+\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
+\caption{%
+  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$ (500, 1000, 2000)
+  }
+\label{fig:plot_boxplot_valid_y}
+\end{figure}
+
 For \emph{BO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
 Comparing the results of \emph{BO Grid} to those of \emph{BO} we mainly see similar performances for both.
-This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{BO}.
+This indicates that the resolution of \emph{Grid} is fine enough to achieve similar performance to \emph{BO}.
 
-Comparing \emph{BO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{BO Grid} achieves slightly better performance then \emph{Grid}.
-The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcomes and has no notion of the stochasticity of the problem, i.e.
+Comparing \emph{BO Grid} to \emph{Grid}, we observe that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$, \emph{BO Grid} achieves slightly better performance than \emph{Grid}.
+This suggests that the \emph{Grid} optimization selects the best outcome "optimistically" from all observed outcomes and does not take into account the stochasticity of the problem, i.e.
 in this case the noise on the predicted power.
-Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
-The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
-In these scenarios \emph{BO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{BO}, as explained in Section~\ref{ssec:best_point}.
-Further below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.
+Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Grid} was misguided by too optimistic outcomes when selecting the final configuration as optimization result.
+In our evaluation, the independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
+In such scenarios \emph{BO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{BO}, as explained in Section~\ref{ssec:best_point}.
+Further below we will study on a single scenario if a reduced noise due to a higher number of simulations \texttt{nsim} decreases the effect of the overoptimistic selection of \emph{Grid}.
 
-If \emph{BO} was not able to find a better solution then \emph{Grid}, then also \emph{BO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{BO}.
-An increase in performance of \emph{BO Grid} compared to \emph{BO} is purely due to chance.
+In cases in which \emph{BO} was not able to find a better solution then \emph{Grid}, also \emph{BO Grid} cannot yield better results as it just selects the configuration on the Grid that is closest to the configuration found by \emph{BO}.
+Therefore any increase in performance of \emph{BO Grid} compared to \emph{BO} is purely due to chance.
 
-To compare \emph{BO} with a grid search that uses approximately the same time (see Table~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
+To compare \emph{BO} with a grid search that takes approximately the same time (see Table~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
 Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO}, with only a small advantage for \emph{Grid Small}.
-Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{BO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
+Despite the 10 more evaluations allowed, in the majority of cases \emph{BO} is superior to \emph{Gris Small}, see~Figure~\ref{fig:plot_boxplot_valid_y}.
+The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.
 Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
-In the first case, the absolute differences are negligible as all methods lead to a high power.
-For the second case we will later show, that the optimal configuration is close to the border of the search space.
-The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
-These points are not included in the coarse grid.
-In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
-The advantage of \emph{BO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.
+In the first case, the absolute differences are negligible and all methods lead to a high power.
+For the second case, we will later show that the optimal configuration is close to the border of the search space.
 
-Note that between the scenarios the absolute differences of the optimization results are of different practical relevance,
-i.e.\ for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$, \emph{Grid} is superior to \emph{BO} but the difference is smaller than $0.25\%$.
+As stated above, the high resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to stochasticity.
+These points are not included in the more coarse grid used by \emph{Grid Small}.
+In other words, if a configuration close to the optimum is included in the grid used by \emph{Grid Small}, then it is more likely to select it with less overoptimism, as the number of candidates in the region is smaller.
+
+Note that between the scenarios the absolute differences of the optimization results are of different practical relevance.
+For example, for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$ \emph{Grid} is superior to \emph{BO}, but the difference is smaller than $0.25\%$.
 
 %After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{BO} to find equally good configurations as the much more demanding exhaustive \emph{Grid}
+
+\subsection{Detailed analysis of identified configurations}
+
 We now investigate the results in more detail.
-In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{BO} are shown, in order to analyze the differences in dependence of the \emph{selection strategy} and the number of treatments at stage 1 determined by the \emph{stage ratio} ($r$).
+We analyze the differences depending on the \emph{selection strategy} and the number of treatments at stage 1, as determined by the \emph{stage ratio} ($r$).
+Figure~\ref{fig:plot_allbest} shows the observed outcomes for all evaluations on the grid and the final configurations found by \emph{BO}.
 \begin{figure}[htb]
@@
   All evaluated configurations and their outcomes on the \emph{Grid}.
-  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power ($y$) is displayed.
-  For each configuration in the grid 10 stochastic results are obtained.
+  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value with highest power ($y$) is displayed.
+  For each configuration in the grid 10 %stochastic
+  results are obtained.
   A line connects the mean outcomes for each configuration.
-  Additionally the best found configurations from \emph{BO} are displayed.
+  In addition, the best configurations identified from \emph{BO} are displayed.
   }
@@
 Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are as expected nearly identical.
-As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. 
+As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. corresponding to large values on the x-axis.
+
 In all scenarios, \emph{BO} identified the peak of all combined curves.
-Within the 20 stochastic repetitions per scenario, \emph{BO} found different configurations, but those had similarly good outcomes.
-That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.\ different configurations yield similarly good outcomes.
+Within the 20 stochastic repetitions per scenario, \emph{BO} found different configurations, but all had similar high power.
+The fact that no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e.\ different configurations yield similarly good outcomes.
 In cases in which multiple optimization runs result in different configurations it might be advisable to let a human expert choose the final configuration, also depending on the medical background.
+
 The curves also show the potential of optimization for the different scenarios.
-If the allowed treatment number is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less necessary.
-If the allowed treatment number is too small $n_{\text{treat}} = 500$, no configuration yields a practically usable design.
-Here, a fast optimization helps to determine if a desired treatment number can actually yield sufficiently powerful designs.
-Of special interest are the scenarios where the correct choice of the design configuration can make a difference between a design with a usable power or a design that is too weak, e.g.
+If the allowed treatment number is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less important.
+If the allowed treatment number is too small ($n_{\text{treat}} = 500$), then no configuration yields a practically usable design.
+In this situation, a fast optimization helps to determine if a desired treatment number can actually yield sufficiently powerful designs.
+Of special interest are the scenarios where the correct choice of the design configuration can make a difference between a design with acceptable power and a design with too low power, e.g.
 for the scenarios (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: linear}, $n_{\text{treat}} = 2000$, \emph{effect: paper}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
-Note, that the selection strategies \emph{thresh} and \emph{all} never gave powerful designs.
+Note, that the selection strategies \emph{thresh} and \emph{all} never selected powerful designs.
 
@@
 Figure~\ref{fig:plot_best_x} visualizes in detail which configurations were selected by \emph{BO} and \emph{Grid} and the corresponding values for $y_{\text{valid}}$ (result of the independent validation).
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
-\caption{Best found $x$-values for \emph{Grid} and \emph{BO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
-\label{fig:plot_best_x}
-\end{figure}
 The optimal values for $\epsilon$ and $\theta$ are not shown.
-For some scenarios most replications of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
-In such cases the signal to noise ratio was high enough to reliably find the best configuration.
-These are also those scenarios where \emph{Grid} was superior the \emph{BO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
+For some scenarios most runs of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
+In such cases the signal to noise ratio was sufficiently high to reliably find the best configuration.
+In some scenarios \emph{Grid} was superior to \emph{BO}, as could be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
 Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500)$ and \emph{effect: paper}, $n_{\text{treat}} = 500)$, small values of $r$ lead to an optimal outcome.
 Here, \emph{BO} was not able to find these values close to the border of the search space.
-A common remedy is to log-transform this value.
+A common remedy for this problem is to log-transform the values of~$r$.
 
-In other cases \emph{Grid} found varying configurations and \emph{BO} found more stable results that lead to better outcomes (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+In other cases \emph{Grid} found different configurations in different runs, while \emph{BO} lead to more stable results, also with higher power values (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
 Here we can assume, that \emph{BO} successfully accounted for the noise.
@@
 
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
+\caption{Best $x$-values found with \emph{Grid} and \emph{BO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
+\label{fig:plot_best_x}
+\end{figure}
+
 \subsection{Results for more simulation rounds}
 
-To investigate the effect of the function noise variance ($\sigma^2$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$ for the scenario \emph{effect: paper}, $n_{\text{treat}} = 2000$.
+To investigate the effect of the function noise variance ($\sigma^2$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$, restricting to the scenario \emph{effect: paper}, $n_{\text{treat}} = 2000$.
 % This is expected to increase the runtime by a factor of 5. % obvious, also not of interest here
-Additionally, we run 100 stochastic replicates instead of the previously 20 to obtain a more precise estimation of the independent validation.
+Additionally, we run the algorithms 100 times instead of the previously used 20, in order to obtain more precise estimates for the power values in the independent validation.
 
-We expect, that increasing $n_\text{sim}$ decreases the noise variance of the outcomes of the simulation that returns the statistical power of the design configuration.
-For the exhaustive \emph{Grid}, we expect that the decreased noise leads to less selections of overly optimistic outcomes as result.
-Also \emph{BO} should profit from a less noisy objective function as the surrogates predictions will be more precise.
+We expect that increasing $n_\text{sim}$ decreases the variance of the power values of the identified design configurations.
+For the exhaustive \emph{Grid}, we expect that the decreased noise leads to less selections of overly optimistic outcomes.
+\emph{BO} should profit as well from a less noisy objective function since the predictions of the surrogate function will be more precise.
 
-Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
-It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
+Figure~\ref{fig:plot_opt_path_5000} shows that indeed the lower noise level leads to more stable simulation runs, meaning that single optimization runs lead to results closer to the mean result.
+It has to be noted that this plot reports the best observed outcome so far and not the one that is chosen as final result.
 %So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
-    \captionof{figure}{Optimization progress for \emph{BO}: Best found outcome so far for each iteration in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$}
+    \captionof{figure}{Optimization progress for \emph{BO}: Best outcome found so far for each iteration, in comparison to the best outcome found in the grid search, for effect set \emph{paper} and $n_{\text{treat}} = 2000$}
     \label{fig:plot_opt_path_5000}
@@
 \end{figure}
-For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
-Also the absolute differences between the methods seem even more negligible.
-This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
-But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
-We conclude with the observation, that \emph{BO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulations.
+For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the solutions that are found vary less between the 100 stochastic replications.
+Also the absolute differences between the methods are even smaller now.
+This demonstrates that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
+However, an increased value of $n_\text{sim}$ also increases the runtime linearly, which conflicts the goal is to find the optimal configuration quickly.
+In summary, \emph{BO} can find configurations with a small number of required simulations (higher noise) that are nearly as good as the optimal solution which is identified with a large number of simulations.
 
@@
 % subsection Summary
-When planning clinical trials, determining patient group sizes and test methods is a crucial task.
-The main goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
+Bayesian optimization (BO) is a state-of-the art optimizer that can be used for parameter optimization in a wide range of applications.
+Here, we propose, for the first time, to optimize the power of adaptive designs using BO.
+In clinical trials, determining the test methods and sample sizes is a key step in the planning.
+One main goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
 If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
-Either because large computational resources are required for an exhaustive exploration of all different configurations, or because it involves a tedious manual manual try and error process.
-In this work we applied Bayesian optimization to quickly find a clinical trial setup with high power from a large number of configuration options.
+Either large computational resources are required for an exhaustive investigation of all the different configurations, or a tedious manual try-and-error process is required.
+
+In this paper we applied Bayesian optimization (BO) to quickly find a clinical trial setup with high power from a large number of configuration options.
 We compared our approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
 BO was able to find competitive setups in a fraction of the time and was able to cope well with the stochasticity of the simulation.
+% subsection Outlook
+Only in cases in which the optimal values of one parameter are close to the lower border, we sometimes observed poor performance of BO.
+Here, a possible improvement can be to apply a log transformation to such parameters in the future.
+% could be also applied for subgroup selection
+% https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8797
+Instead of keeping the number of total trials constant, we could apply multi-criteria optimization and include this number as an additional outcome.
+This multi-criteria approach would result in a set of points (Pareto front) that minimize the number of trials and at the same time maximize the statistical power of the trail design.
+The user could then see for which number of trials the power is above a certain threshold (e.g. $95\%$).
 
-% subsection Outlook
-One drawback of BO that we observed was, that it performed poorly if the optimal values of one parameter are close to the lower border of the search region.
-Therefore we would apply log transformation to such parameter in the future.
-
-Instead of keeping the number of total trials constant, we could apply multi-criteria optimization and include this number as an additional outcome.
-This multi-criteria approach would result in a set of points (Pareto front) that minimize the number of trails as well as maximize the statistical power of the trail design.
-The user could then see for which number of trails the power is above a certain threshold (e.g. $95\%$).
-
-Another way of speeding up the optimization would be to adaptively increase the simulation rounds of the trail design evaluation. 
-If few simulations already indicate a bad outcome it might not be promising to evaluate more simulation runs.
-Such procedure can be guided through statistical testing and is already common in machine learning under the terms racing or multi-fidelity optimization.
+Another way to speed up the optimization would be to iteratively increase the number of simulation rounds of the trial design evaluation. 
+If few simulations already indicate a bad outcome for a configuration then additional simulation rounds are not necessary.
+Such a procedure can be guided by statistical testing and is already common in machine learning under the terms \emph{racing} or \emph{multi-fidelity optimization}.
 
@@
 
-\noindent This is the body text. Only number equations which are referred to in the text body. If equations
-are numbered, these should be numbered continuously throughout the text. Not section wise! Please
-carefully follow the rules for mathematical expressions in the ``Instructions to Authors''.
+%\noindent This is the body text.
+Only number equations which are referred to in the text body.
+If equations are numbered, these should be numbered continuously throughout the text.
+Not section wise!
+Please carefully follow the rules for mathematical expressions in the ``Instructions to Authors''.
 

### 2020-12-22 main.tex
Updates from Overleaf

@@
 %{\small\it words)}\\[1pc]
-\noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not
-applicable)
+%\noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not applicable)
 }  %%% semicolon and fullpoint added here for keyword style
@@
 \begin{abstract}
-In this paper we propose to use Bayesian optimization (BO) to improve the efficiency of the selection process of designs in clinical trials.
-BO is a method to optimize expensive black-box functions that uses a regression as a surrogate to guide the search.
+We propose to use Bayesian optimization (BO) to improve the efficiency of the design selection process in clinical trials.
+BO is a method to optimize expensive black-box functions, by using a regression as a surrogate to guide the search.
 In clinical trials, planning test procedures and sample sizes is a crucial task.
-A common goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
-From a wide range of possible alternatives we want to select the best plan in a short period of time, to enable quick decisions.
-The standard approach to simulate the power for each single setup can become too time-consuming, when the number of possible setups becomes very large.
-Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
-In this work we propose to use BO to quickly find a clinical trial setup with high power from a large number of setups options.
+A common goal is to maximize the test power, given a set of treatments and corresponding effect sizes. % under the null hypothesis.
+From a wide range of possible designs we aim to select the best one in a short time to allow quick decisions.
+The standard approach to simulate the power for each single design can become too time-consuming.
+When the number of possible designs becomes very large, either large computational resources are required or an exhaustive exploration of all possible designs takes too long.
+Here, we propose to use BO to quickly find a clinical trial design with high power from a large number of candidate designs.
 We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-We compare the BO approach with an exhaustive evaluation of all candidate designs.
-The optimization approach finds competitive setups in a fraction of the time.
+Comparing BO with an exhaustive evaluation of all candidate designs shows that BO finds competitive designs in a fraction of the time.
 \end{abstract}
@@
 For instance, \citet{sheiner_learning_1997} describe the drug development process as two learning and confirming cycles:
-The first cycle includes learning about the tolerated dose (Phase I) and confirming the efficacy of the selected dose in a selected group of patients (Phase IIa), whereas the second cycle consists of learning about the optimal use in representative patients (Phase IIb) and confirming of an acceptable benefit or risk ratio (Phase III).
+The first cycle includes learning about the tolerated dose (Phase I) and confirming the efficacy of the selected dose in a selected group of patients (Phase IIa), whereas the second cycle consists of learning about the optimal use in representative patients (Phase IIb) and confirming an acceptable benefit or risk ratio (Phase III).
 
@@
 In general, study designs become more flexible, and more principled approaches to decision making using statistical modeling and integrated analyses across studies become popular. 
-A related principle on the analysis side is evidence synthesis, which is a way of combining information from several studies that have examined the same question to come to an overall understanding.
+A~related principle on the analysis side is evidence synthesis, which is a way of combining information from several studies that have examined the same question to come to an overall understanding.
 
@@
 Especially relevant are neutral comparison studies, which are characterized by~\citet{boulesteix_plea_2013} as follows:
-%FIXME Is there someting missing? (Jakob)
 The primary goal of the respective article is not to introduce a new promising method, the authors are reasonably neutral, and the evaluation metrics, methods, and data sets are chosen in a rational way.
@@
 Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
-Various implementations of BO methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
+Various implementations of BO methods are widely used, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
 For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
-This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
+This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
 Also, BO is well suited for multi-criteria optimization.
@@
 Many of them are implemented in \texttt{mlrMBO} as well.
-However, in this work we restrict ourselves to single-criteria optimization.
+However, in this work we restrict ourselves to single-criterion optimization.
 
@@
 The general idea of this paper is to apply BO for finding a trial design that is optimal with respect to a certain metric.
-As an example, we consider the maximization of the statistical power of a trail design, given a set of parameters to be chosen for the trial design.
+As an example, we consider the maximization of the statistical power of a trial design, given a set of parameters to be chosen for the trial design.
 We compare the computation time as well as the maximal obtained power between the results obtained with BO and with an exhaustive grid search over the space of adjustable parameters.
@@
 
-One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd}~\cite{parsons_software_2011} package and described in Section~\ref{sec:adaptive_seamless_designs}.
+One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd} package~\citep{parsons_software_2011}  and described in Section~\ref{sec:adaptive_seamless_designs}.
 %(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, )  bauer_combining_1999
@@
 In~\citet{friede_adaptive_2020} the authors present an application of \emph{adaptive seamless designs} for chronic obstructive pulmonary disease (COPD).
-We will use this COPD study to apply BO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
+We use this COPD study to apply BO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
 %within the given scenarios restrictions.
@@
 Indacaterol is a drug administered to COPD patients.
-The patients were randomized to four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}), active controls, and placebo control.
-As in \citet{friede_adaptive_2020}, here we ignore the active controls.
+The patients were randomized to four doses of indacaterol (\SI{75}{\micro\gram}, \SI{150}{\micro\gram}, \SI{300}{\micro\gram} and \SI{600}{\micro\gram}), to active controls, and to a placebo control.
+Here, as in \citet{friede_adaptive_2020}, we ignore the active controls.
 The primary outcome of the trial was the percentage of days of poor control over 26 weeks of the COPD patients.
@@
 
-In the analysis of \citet{friede_adaptive_2020}, the aim was to select two doses of indacaterol at the interim analysis that are passed to stage 2, resulting in a total sample size $5 \cdot 100 + 3 \cdot 300 = 1,400$ patients.
-In the simulations, a positive correlation between early and final outcomes of 0.4 was assumed.
+In the analysis of \citet{friede_adaptive_2020}, the aim was to select two doses of indacaterol at the interim analysis that are passed to stage 2, resulting in a total sample size of $5 \cdot 100 + 3 \cdot 300 = 1,400$ patients.
+In the corresponding simulations, a positive correlation between early and final outcomes of $0.4$ was assumed.
 Further, different settings with continuous early and continuous or binary final outcomes and different treatment selection strategies were compared.
@@
 
-% FIXME: Do we really take in this Assurance?
-% Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
-
 \section{Adaptive Seamless Designs}
@@
 % TODO: Tim
-Traditionally drug development is organized in four phases with individual clinical trials for the separate phases.
-Seamless designs combine elements from different development phases in a single trial thereby offering the promise of speeding up development times.
-For instance, seamless phase II/III designs combine the learning about dose-response relationship or subgroup heterogeneity typical for phase II with confirmation of treatment effects in phase III.
+Traditionally, drug development is organized in four phases with individual clinical trials for the separate phases.
+Seamless designs combine elements from different development phases in a single trial, thereby offering the promise of speeding the development process.
+For instance, seamless phase II/III designs combine the learning about dose-response relationship or subgroup heterogeneity typical for phase II with the confirmation of treatment effects in phase III.
 If the data of the phase II part are used for confirmatory purposes, then this needs to be accounted for in the testing strategy to maintain the familywise type I error rate in the strong sense.
-This type of designs are referred to as adaptive seamless designs.
+These types of designs are called adaptive seamless designs.
 A recent overview can be found in \citet{friede_adaptive_2020}.
 
-We start by providing some more details on the study design considered (Sections \ref{ssec:two_stage_design_with_treatment_selection} and \ref{ssec:selection_rules}) and then introduce briefly a framework for optimizing such designs, namely clinical scenario evaluation (CSE) in Section \ref{ssec:clinical_scenario_evaluation_for_asd}.
+We start by providing some more details on the study design considered (Sections \ref{ssec:two_stage_design_with_treatment_selection} and \ref{ssec:selection_rules}).
+Then, in Section \ref{ssec:clinical_scenario_evaluation_for_asd} we briefly introduce a framework for optimizing such designs, namely clinical scenario evaluation (CSE).
 %In Section \ref{ssec:simulating_asd} we make some brief comments on the simulation model.
@@
 \label{ssec:two_stage_design_with_treatment_selection}
-Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection.
+We consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection.
 The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control.
@@
 control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better).
-For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments.
+For the purpose of treatment selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered, with larger $Z_{1,k}$ indicating more beneficial treatments.
 These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis.
@@
 to explore sample size and power.
-In the following we give some examples of selection rules which have previously been consider and which will feature below.
+In the following we give some examples of selection rules which have previously been considered and which we will feature below.
 
@@
 We present the selection rules using $Z_{1,k}^{(E)}$, but of course these would be replaced by $Z_{1,k}$ if the final outcome was used for treatment selection.
-Furthermore, larger statistics are considered better.
+Again, larger statistics are considered better.
 
 \paragraph{$\kappa$-best Rule} Let $Z_{1,(1)}^{(E)}, \dots, Z_{1,(K)}^{(E)}$ denote the ordered statistics with $Z_{1,(1)}^{(E)}$ being the largest and therefore best test statistic.
-The $\kappa$-best rule then selects the treatments associated with the $\kappa$ largest $Z_{1,k}^{(E)}$, i.e.$Z_{1,(1)}^{(E)}, \dots, Z_{1,(\kappa)}^{(E)}$.
+The $\kappa$-best rule then selects the treatments associated with the $\kappa$ largest values of the test statistic, %$Z_{1,k}^{(E)}$,
+i.e. $Z_{1,(1)}^{(E)}, \dots, Z_{1,(\kappa)}^{(E)}$.
 
-\paragraph{Epsilon Rule} To our knowledge this rule was first described by \citet{kelly_adaptive_2005} and has since then featured in a number of simulation studies including \citet{friede_comparison_2008, friede_adaptive_2020}.
+\paragraph{Epsilon Rule} To our knowledge this rule was first described by \citet{kelly_adaptive_2005} and has since then featured in a number of simulation studies including \citet{friede_comparison_2008} and \citet{friede_adaptive_2020}.
 All treatments $k$ with $Z_{1,k}^{(E)} \ge \max_i Z_{1,i}^{(E)} - \epsilon$ with $\epsilon \ge 0$ are selected.
-With $\epsilon=0$ this rule is equivalent to the $\kappa$-best rule with $\kappa=1$.
+For $\epsilon=0$ this rule is equivalent to the $\kappa$-best rule with $\kappa=1$.
 
-\paragraph{Threshold Rule} With this rule all treatments $k$ are selected with $Z_{1,k}^{(E)} \ge \tau$ where $\tau$ is a given threshold.
+\paragraph{Threshold Rule} With this rule all treatments $k$ are selected with $Z_{1,k}^{(E)} \ge \tau$, where $\tau$ is a given threshold. \\
 
@@
 
-The clinical scenario evaluation framework was first proposed by \citet{benda_aspects_2010} and subsequently developed further by \citet{friede_refinement_2010}. 
+The clinical scenario evaluation framework was first proposed by \citet{benda_aspects_2010} and subsequently further developed by \citet{friede_refinement_2010}. 
 Figure \ref{fig:friede_refinement_2010_fig2} provides a summary of this framework.
@@
       \item Treatment selection rules at interim
-      \item Conventional phase 2/3
+    \end{itemize}
+    \vspace*{0.5ex}
+    Conventional phase 2/3
+    \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
       \item Split of resources between phases
-      \item End oh phase futility criteria
+      \item End of phase futility criteria
     \end{itemize}
@@
   };
-  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design Options}};
+  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design options}};
   \node[draw, below = 0.5cm of ClinicalScenarios] (DesignPerformance) {
     \textbf{Design performance} \\ e.g.\ statistical power};
-  \node[draw, right = of DesignPerformance, text width = 4cm, node distance = 2cm] (Evaluation) {
+  \node[draw, right = of DesignPerformance, text width = 5cm, node distance = 2cm] (Evaluation) {
     \textbf{Clinical Scenario Evaluation}\\
     Design performance measures evaluated across a wide range of clinical scenarios};
-  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
-  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
-  \draw[->] (CombinationPoints.223) -- (ClinicalScenarios);
-  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
-  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
-  \draw[->] (DesignPerformance) -- (Evaluation);
+  \draw[thick,->] (HistoricalDataPoints) -- (ClinicalScenarios);
+  \draw[thick,->] (UnknownPoints) -- (ClinicalScenarios);
+  \draw[thick,->] (CombinationPoints.223) -- (ClinicalScenarios);
+  \draw[thick,->] (ConstrainedPoints) -- (ClinicalScenarios);
+  \draw[thick,->] (ClinicalScenarios) -- (DesignPerformance);
+  \draw[thick,->] (DesignPerformance) -- (Evaluation);
 \end{tikzpicture}
 %Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
-\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis (Figure adapted from \citet{friede_refinement_2010}).}
+\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis (Figure adapted from Figure~2 in \citet{friede_refinement_2010}).}
 \label{fig:friede_refinement_2010_fig2}
@@
 A clinical scenario evaluation is defined by the following three elements: (i) disease-specific features; (ii) design options; and (iii) design performance measures.
-The design specific features describe distributional assumptions on the early and final outcomes including variances, correlations and treatment effect sizes.
-Here, those that can be estimated from data of previous clinical trials and those that are simply unknown.
-Considering adaptive seamless design, the design options include the time point of the interim analysis and the treatment selection rule, which can in principle both be chosen without any restrictions.
+The disease-specific features describe distributional assumptions on the early and final outcomes including variances, correlations and treatment effect sizes.
+Here, some can be estimated from data of previous clinical trials and some are simply unknown.
+Considering adaptive seamless designs, the design options include the time point of the interim analysis and the treatment selection rule, which can in principle both be chosen without any restrictions.
 In contrast, some other design options such as the number of treatments and the total sample size are constrained by the environment and infrastructure available.
-The performance measures (sometimes also referred to as metrics) are in the context with adaptive seamless designs e.g.
-power or sample size distributions.
+The performance measures (sometimes also referred to as metrics) in the context with adaptive seamless designs are, e.g., power or sample size distributions.
 %\paragraph{Simulating ASD}
 %\label{ssec:simulating_asd}
+
 The evaluation of some performance measures requires Monte Carlo simulations, since no closed form solutions are available. 
@@
 
-Bayesian optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
+Bayesian optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art technique for expensive black-box optimization problems~\citep{shahriari_taking_2016}.
 In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
-As then it becomes infeasible to evaluate the black-box for thousands of configurations.
+In this situation it becomes infeasible to evaluate the black-box for thousands of configurations.
 
@@
 
-The goal of this simulation study is to investigate the potential benefit of BO for effectively finding a clinical trial design that maximizes a metric.
+The goal of this simulation study is to investigate the potential benefit of BO for efficiently finding a clinical trial design that maximizes a given metric.
 The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
@@
 The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
-The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control).
+The $\kappa$-best strategies select the  1, 2, 3 or 4 best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control).
 The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
@@
 
-The simulation of the statistical power of the trail design can be formulated as a black-box function as follows:
+The simulation of the statistical power of the trial design can be formulated as a black-box function as follows:
 \begin{align}
@@
 \end{align}
-whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively and \emph{effect set} remains fixed as we are interested in the optimal trail design for a given \emph{effect set}.
+whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively and \emph{effect set} remains fixed as we are interested in the optimal trial design for a given \emph{effect set}.
 Our main optimization goal is to maximize the statistical power $y$ depending on the trial design.
 
-However, we are interested in keeping the total number of treatments $n_{\text{treat}}$ constant in the overall trail for a fair comparison of the power of different designs.
+However, we are interested in keeping the total number of treatments $n_{\text{treat}}$ constant in the overall trial for a fair comparison of the power of different designs.
 Therefore we introduce the ratio of patients per treatment in stage 1, compared to patients per treatment in both stages
@@
 
-The best point per evaluated grid is determined by selecting the configuration with the best outcome, i.e.\  the concrete trail design with the largest power.
+The best point per evaluated grid is determined by selecting the configuration with the best outcome, i.e.\  the concrete trial design with the largest power.
 The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the quality of the outcome.
@@
 For the selection strategies \emph{epsilon rule} and \emph{threshold rule} only curves representing the $\epsilon$ and $\tau$ values with the best outcome, respectively, are shown.
-Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are as expected nearly identical.
+Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
 As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. corresponding to large values on the x-axis.
@@
 Bayesian optimization (BO) is a state-of-the art optimizer that can be used for parameter optimization in a wide range of applications.
-Here, we propose, for the first time, to optimize the power of adaptive designs using BO.
-In clinical trials, determining the test methods and sample sizes is a key step in the planning.
-One main goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
+Here, we propose, for the first time, to optimize the power of adaptive seamless designs using BO.
+In clinical trials, determining the test methods and sample sizes is a key step in the planning process.
+One main goal is to maximize the test power, given a set of treatments with corresponding effect sizes. % under the null hypothesis.
 If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
@@
 
-In this paper we applied Bayesian optimization (BO) to quickly find a clinical trial setup with high power from a large number of configuration options.
-We compared our approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
-BO was able to find competitive setups in a fraction of the time and was able to cope well with the stochasticity of the simulation.
+%% FIXME: Do we really take in this Assurance?
+% Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
+% could be also applied for subgroup selection?
+% https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8797
+
+In this paper we analyzed if BO can quickly find an adaptive seamless design with high power, more efficiently than an exhaustive grid search.
+For various sets of treatment effect sizes, BO was able to find competitive designs in a fraction of the time.
+In contrast to grid search, BO was able to account for stochasticity in the metric that should be optimized.
+%In addition to results in other research fields, our results underline that BO is a powerful method for finding optimal configurations for problems that fit the black-box definition.
+Therefore, the application of BO to clinical trial design selection in general seems promising.
+The same concept could be applied for subgroup selection in adaptive enrichment designs, which make use of interim data in selecting the target population for the remainder of the trial~\citep{burnett_adaptive_2020}.
+
 % subsection Outlook
 Only in cases in which the optimal values of one parameter are close to the lower border, we sometimes observed poor performance of BO.
-Here, a possible improvement can be to apply a log transformation to such parameters in the future.
-% could be also applied for subgroup selection
-% https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8797
-Instead of keeping the number of total trials constant, we could apply multi-criteria optimization and include this number as an additional outcome.
-This multi-criteria approach would result in a set of points (Pareto front) that minimize the number of trials and at the same time maximize the statistical power of the trail design.
+Here, a possible improvement can be to apply a log transformation to such parameters. % in the future.
+Further, instead of keeping the number of treatments constant, we could apply multi-criteria optimization and include this number as an additional outcome.
+This multi-criteria approach would result in a set of points (a Pareto front) that minimize the number of treatments and at the same time maximize the statistical power of the trial design.
 The user could then see for which number of trials the power is above a certain threshold (e.g. $95\%$).
-
 Another way to speed up the optimization would be to iteratively increase the number of simulation rounds of the trial design evaluation. 
@@
 Such a procedure can be guided by statistical testing and is already common in machine learning under the terms \emph{racing} or \emph{multi-fidelity optimization}.
+
+In this work we assumed the effect sizes to be fixed. % under the null hypothesis.
+In~\citet{stallard_optimal_2009} the authors propose a Bayesian approach, where the effect sizes follow a prior, to calculate the expected trial performance.
+This requires a more expensive simulation making the optimization of a clinical trial design according to such metric an excellent candidate for BO.
 

### 2020-12-22 main.tex
languagetool

@@
 Adaptive designs are established, but are being further developed.
-In general, study designs become more flexible, and more principled approaches to decision making using statistical modeling and integrated analyses across studies become popular. 
+In general, study designs become more flexible, and more principled approaches to decision-making using statistical modeling and integrated analyses across studies become popular. 
 A~related principle on the analysis side is evidence synthesis, which is a way of combining information from several studies that have examined the same question to come to an overall understanding.
@@
 
-BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes in many different scenarios.
+BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes in many scenarios.
 Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
@@
 For instance, seamless phase II/III designs combine the learning about dose-response relationship or subgroup heterogeneity typical for phase II with the confirmation of treatment effects in phase III.
-If the data of the phase II part are used for confirmatory purposes, then this needs to be accounted for in the testing strategy to maintain the familywise type I error rate in the strong sense.
+If the data of the phase II part are used for confirmatory purposes, then this needs to be accounted for in the testing strategy to maintain the family-wise type I error rate in the strong sense.
 These types of designs are called adaptive seamless designs.
@@
 The R-package \texttt{asd} by~\citet{parsons_software_2011} can be used for such simulations.
-As it generates test statistics it is generally more efficient in comparison to simulating individual participant data.
+As it generates test statistics, it is generally more efficient in comparison to simulating individual participant data.
 The simulation model was described in more detail by \citet{friede_adaptive_2020}. 
@@
 Bayesian optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art technique for expensive black-box optimization problems~\citep{shahriari_taking_2016}.
-In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
+In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time-consuming.
 In this situation it becomes infeasible to evaluate the black-box for thousands of configurations.
@@
 
-The proposal of a new configuration $\theta$ is obtained by maximizing a so called acquisition function.
+The proposal of a new configuration $\theta$ is obtained by maximizing a so-called acquisition function.
 The acquisition function guides the search to promising new regions in the search space $\Theta$ by using the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty quantification $s^2(\theta)$ of the surrogate.
@@
 Using the stochastic outcome $f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
-Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimate of the true value of the best outcome.
+Therefore, an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimate of the true value of the best outcome.
 
@@
 However, we are interested in keeping the total number of treatments $n_{\text{treat}}$ constant in the overall trial for a fair comparison of the power of different designs.
-Therefore we introduce the ratio of patients per treatment in stage 1, compared to patients per treatment in both stages
+Therefore, we introduce the ratio of patients per treatment in stage 1, compared to patients per treatment in both stages
 \begin{align}
@@
 
-Additionally we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid introduced previously.
-Therefore we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
+Additionally, we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid introduced previously.
+Therefore, we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
 In other words, for \emph{BO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{BO}.
 The comparison of the results of \emph{BO Gird} and \emph{Gird} will allow two possible conclusions:
-First, if \emph{BO Grid} is worse then \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
-Second, if \emph{BO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
+First, if \emph{BO Grid} is worse than \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
+Second, if \emph{BO Grid} is better than \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
 Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.
@@
 In cases in which \emph{BO} was not able to find a better solution then \emph{Grid}, also \emph{BO Grid} cannot yield better results as it just selects the configuration on the Grid that is closest to the configuration found by \emph{BO}.
-Therefore any increase in performance of \emph{BO Grid} compared to \emph{BO} is purely due to chance.
+Therefore, any increase in performance of \emph{BO Grid} compared to \emph{BO} is purely due to chance.
 
@@
 Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
-As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. corresponding to large values on the x-axis.
+As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2.
 
@@
 For some scenarios most runs of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
-In such cases the signal to noise ratio was sufficiently high to reliably find the best configuration.
+In such cases the signal-to-noise ratio was sufficiently high to reliably find the best configuration.
 In some scenarios \emph{Grid} was superior to \emph{BO}, as could be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
@@
 We expect that increasing $n_\text{sim}$ decreases the variance of the power values of the identified design configurations.
-For the exhaustive \emph{Grid}, we expect that the decreased noise leads to less selections of overly optimistic outcomes.
+For the exhaustive \emph{Grid}, we expect that the decreased noise leads to fewer selections of overly optimistic outcomes.
 \emph{BO} should profit as well from a less noisy objective function since the predictions of the surrogate function will be more precise.
@@
 Despite the differences of the optimization paths, the median validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{BO} for the low and the high value of $n_\text{sim}$.
-Therefore we can assume that "bias correction" of \emph{BO} as explained in Section~\ref{ssec:best_point} seems to be effective.
+Therefore, we can assume that "bias correction" of \emph{BO} as explained in Section~\ref{ssec:best_point} seems to be effective.
 \begin{figure}[htb]
@@
 For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the solutions that are found vary less between the 100 stochastic replications.
-Also the absolute differences between the methods are even smaller now.
+Also, the absolute differences between the methods are even smaller now.
 This demonstrates that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
@@
 If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
-Either large computational resources are required for an exhaustive investigation of all the different configurations, or a tedious manual try-and-error process is required.
+Either large computational resources are required for an exhaustive investigation of all the different configurations, or a tedious manual trial-and-error process is required.
 

### 2021-03-15 response_to_the_reviewers.md
fixes and literature

# Response to the reviews for "Improving Adaptive Seamless Designs through Bayesian optimization"

We would like to thank all reviewers for their valuable and constructive comments on the paper.
We especially that Reviewer 1 for the thoughtful and in-depth feedback.

## Reviewer: 1

### Comments to the Author

> This is a well organised and written manuscript, describing a very thorough investigation of a reasonably interesting problem. Computational aspects of complex adaptive clinical trial designs have all too often been neglected. The speed gain proposed here is substantial, with no apparent loss in quality. The findings could help improve the uptake of adaptive seamless designs. I only have a few relatively minor suggestions that I hope the authors will find helpful.

### General

> 1) Power is used as a key performance metric, yet it lacks an explicit definition. The problem, of course, is that power is slightly more complicated in a multi-stage design with multiple arms (and therefore multiple hypotheses to be tested) than it is in a simple comparison of two groups. Given the scenario described in the motivating example, I would have assumed that power was defined as the probability of at least one (any) treatment group being demonstrated to be ‘better’ than control i.e. a ‘disjunctive’ or ‘minimal’ power definition (cf. Senn & Bretz 2007). However, on p. 10 it is stated that “the vector of treatment numbers for determining power counts rejections of 3 or 4 hypotheses”, implying a power definition more similar to ‘conjunctive’ or ‘maximal’ power.
>
> Reference: Senn S, Bretz F (2007) Power and sample size when multiple endpoints are considered. Pharmaceutical Statistics, 6(3), 161-170.

--> Tim Friede

>2) There is some inconsistent notation:
>
>a) It seems that n_treat is varyingly defined as the “total number of treatments” (p. 8, l. 29 and p. 9, l. 20), “total number of patients” (p. 8, l. 48 and p. 9, l. 14) and “allowed treatment number” (p. 12, l. 50 and p. 12, l. 51).

n_treat is the number of total treatments. 
If the patients are different in in both stages (i.e. each patient is only treated once), both numbers are the same.
However, we corrected the wording to avoid confusions.

> b) On p. 3 k is introduced as the running index for treatments (k = 1, …, K) but later on p. 9 k is the resolution of the search grid.

Thanks for catching that mistake. It is now fixed.

### Specific

>p. 2: For the benefit of readers less familiar with the term, the authors should explain at least briefly what they mean by “expensive black-box” optimisation problems.

We added a short explanation.

>p. 3 ff.: This may be implicitly clear but it should also be said explicitly that in the motivating example and throughout the remainder of the paper a normality assumption is made for all outcome data.

--> Tim? Gibt es diese assumption bei dir?
--> Mbo Teil habe ich übernommen.
We clarified the assumptions in chapter 4 and 5.

>p. 9: How were the values for the resolution of Grid and Grid Small (k=25 and k=7) selected?

We added an explanation and referenced previous studies, that used similar values.

> p. 10: Similarly, how was the number of randomly sampled points (16) for the initial design for BO chosen?

We added an explanation and a reference.

> p. 10: In section 5.4 ten additional stochastic simulations are mentions, which seems in conflict with the 20 additional stochastic simulations mentioned in other parts of the manuscript.

Thanks, for spotting that mistake. Indeed we use 20 stochastic repetitions.

> p. 10: It should be clarified whether the test level of 0.025 is one-sided.

--> Tim

> p. 11: Whilst I appreciate that the focus is on relative improvements in runtime, the authors should at least briefly describe the hardware setup with which these runtimes were achieved, to provide some context for the absolute runtimes.

We added the requested information.

> p. 12: It is stated that: “Only for the scenarios (effect: paper2, n_treat = 1000) and (effect: sigmoid, n_treat = 1000) Grid Small has a small advantage over BO and Grid.” I think it should be added that Grid Small also has an advantage over BO for (effect: paper, n_treat = 1000) and certainly for (effect: sigmoid, n_treat = 2000).

> p. 12: It says: “For example, for scenario effect: sigmoid, n_treat = 1000 Grid is superior to BO, but the difference is smaller than 0.25%.” This isn’t true for n_treat = 1000 where the difference between Grid and BO is around 2%, but rather for n_treat = 2000.

> p. 12: Apologies but I’m struggling to understand the following sentence: “As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2.” How can none of the treatments be used in stage 2 when on p. 8 it is specified that the number of treatments in stage 2 must always be between 2 and 5?

--> clarify

> p. 13: It is stated that: “Note, that the selection strategies thresh and all never selected powerful designs.” Why is that? Presumably, the former is true because the range of numerical values used for the threshold was chosen too narrow, whilst the latter shouldn’t be surprising at all, as it is the only non-adaptive design in the mix.

--> Maybe threshold is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization

> p. 14: What are the x-values referred to in the caption of Figure 4? No x is introduced in the notation, and Figure 4 itself plots y-values against r-values.

> p. 16: The citation of the technical report by Bischl et al. should include a URL if possible.

### Typos

> p. 9, l. 20: “in average”, which should read “on average”

Done.

> p. 9, l. 24: “in 8”, which should read “in (8)”

Done.

> p. 9, l. 53 and > p. 10, l. 14: “Mattern”, which should be spelt “Matérn”

> p. 10, l. 28: “Gird”

Done.

> p. 10, l. 39: “???”

Done.

> p. 12, l. 19: “Gris”

Done.

> p. 13, l. 48f.: remove closing bracket from “500)”

Done.

> p. 14, l. 55: “which conflicts the goal is to find the optimal” (grammar)

Done.

## Reviewer: 2

### Comments to the Author

> The authors propose to use Bayesian optimization (BO) to improve the efficiency of the design selection process in clinical trials. A set of parameters to be chose for the design optimization based on the power. The idea seem novel and could be a useful approach. However, there are several major issues and limitations of proposed methods: for example, power is often only part of measure of choosing design, other factors like study duration, and # of pts are also important. Also, the black-box function relay too much of the various parametric assumption of parameters, which need to be carefully decided to be more related to true clinical trial needs. Overall, I have the following comments for the authors.

--> In out method we keep the number of patients fixed
--> Tim Friede (assumption of parameters, true clinical trial needs)

### Major comments:

> In page 7 table 1, the effect size scenarios presented seem always assume 2nd stage effect size are higher than stage 1? The authors should also evaluate the vice visa situation, and consistent effect size case in simulation.

--> Argument: Falschrum
--> Often Realisitic scenario, also for the motivating example: First stage, second stage ...
--> Tim

> In page 8 table 2, the selection of “espilon” and “threshold” used in the arm selection should be based on the clinical meaningful cut, I am not sure how practical that they are part of parameters to be optimized? The authors should also evaluate with different setting.

--> Argument: Given a simulation setting we optimize the parameters eps and thr. This is independent of the application but of course the clinical application can give hints which simulation settings are adequate

> In section 5.2, the authors keep the total number of pts fixed, which seem not appropriate in practice, as overall power relying several factors, e.g. # of pts, the interim selection rule, stage 1 and stage 2 ratio. Often # of pts is also key goal of optimal design evaluation. I will strongly suggest # of pts should be a parameter in the black-box function, instead of fixing it to a constant value.

--> Argument: This will only lead to a result where more patients are selected. The idea is to run this optimization for different patient sizes, and then to decide which number of patients is required (also mention this in paper)

> In page 10 table 4, the number of simulation iteration is only 1000, which does not seem sufficient for complex optimization methods proposed by authors. I would suggest at least 5000 should be needed.

--> simulation iteration means that we evaluated 1000 times wheter the test detected a significant effect. this does not depend on the complexity of the optimization. As a matter of fact we use 100 optimization iterations which is sufficient as can be seen in plot ... where we reach convergence.

> Page 11 table 5, the authors present the average run time in hour, not sure how these information is helpful, as this depend on the what computing systems were used in the simulation. Please either add more information.

--> add computational information, relative values are still meaningfull

> Page 13, BO has issue to find optimal value when one parameters closer to the borders of search space. This is concerning and confirm the limitation of using too many assumptions for optimization, e.g. appropriate parameters range. I am not sure taking log-transform could solve this problem, could authors elaborate more details on this.  Additional simulation is necessary to confirm this.

--> log transform: good idea, outlook

> Given authors used COPD trial as motivating example, I would strongly suggest to add section of case study for the illustration of proposed methods.

--> in fact our effect set "paper" issues a case study where the effect sets are estimated on a true dataset. we do not propose a method to analyze data, but a method to select the best design given a known theoretical situation. (check that this is clear)

> Minor comments:

Page 8 line 29,  n_{treat} should be total # of pts, instead of “total number of treatment”.

### 2021-03-15 main.tex
Updates from Overleaf
fixes and literature

@@
 
-BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes in many scenarios.
+BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes (i.e.\ functions that take long to evaluate and are not given in closed-form) in many scenarios.
 Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
@@
 
+The expected improvement, as well as other acquisition functions, are derived under the assumption that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
+The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters \hat{\mu} and \sigma^2 under the assumption that $f$ is a realization of a Gaussian process.
+In practice, this assumption often cannot be verified.
+However, benchmarks show that the approach nevertheless works reliable for many practical problems~\citep{bischl_mlrmbo_2017,snoek_practical_2012}.
+
 \subsection{Acquisition function for noisy outcomes}
@@
 where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma^2$ denotes the random error (nugget effect) in the Kriging model.
-The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
-This assumption is also reflected by using a Gaussian process regression as a surrogate.
+The formula is derived under the assumption of a Gaussian posterior at each point $\theta$.
 The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
@@
 
-Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
+Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of treatments $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
 \begin{align}
@@
 We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed value for $n_{\text{stage1}}$.
-Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total patients is closest to $n_{\text{treat}}$.
+Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total treatments is closest to $n_{\text{treat}}$.
 More precisely, we minimize
@@
 %The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
-The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
-
-The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and number of total treatments, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~\ref{eq:bbox2}.
+The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (on average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
+
+The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and number of total treatments, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~(\ref{eq:bbox2}).
 
@@
 \paragraph{Grid and Grid small}
-We conduct an \emph{exhaustive grid search} with a resolution of $k$ points per real-valued dimension and one point for each categorical dimension.
-Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of $k$.
+We conduct an \emph{exhaustive grid search} with a resolution of $l$ points per real-valued dimension and one point for each categorical dimension.
+Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of $l$.
 All 6 selection strategies have the numeric parameter $r$.
 The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively.
-Using $k=25$, we call this approach \textbf{Grid},
-with in a total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
-Using a subset of this grid with $k=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, we call the method \textbf{Grid Small}.
+For $l=25$, we call this approach \textbf{Grid},
+with a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
+We chose $l=25$ generously to obtain an estimate of the optimum that should be close to the true optimum.
+Previous studies used half the resolution to assess the behavior of similar problems~\citep{friede_adaptive_2020}.
+We define \textbf{Grid Small} to be a subset of this grid with $l=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$.
+This gives a number of evaluations that can be easily carried out in practice.
+
 Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.
@@
 As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
-We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
+We configure Kriging to apply the Matérn$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
+Note, that by doing so, we implicitly assume that $f$ is a realization of a Gaussian process and that each single outcome is a realization of a Gaussian.
+However, as stated in Section~\ref{sec:bayesian_optimization}, we can expect the method to work well even if the assumption is slightly violated.
 
@@
 This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
-Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
+Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Matérn$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
 With the above steps we ensure that the input for the Kriging is purely numeric.
@@
 As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
-We start BO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
+We start BO with an initial design of $16$ randomly sampled points, following an established rule-of-thumb of $4$ points per dimension, which is also well in line with the recommendation of using around $10\%$ of your budget for the initial design~\cite{bossek_initial_2020}.
+We allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
 
@@
 In other words, for \emph{BO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{BO}.
-The comparison of the results of \emph{BO Gird} and \emph{Gird} will allow two possible conclusions:
+The comparison of the results of \emph{BO Grid} and \emph{Grid} will allow two possible conclusions:
 First, if \emph{BO Grid} is worse than \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
@@
 For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
-The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
+The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses. %???
 Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
-The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.
+The 20 additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.
 
 % (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
+%FIXME: Ist mir immer noch nicht klar / nicht mehr klar?
 
@@
   Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
-  Test level & \texttt{level} & $0.025$ \\
+  Test level & \texttt{level} & $0.025$ \\ %FIXME: @Tim: One sided?
   Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
@@
 Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
-The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
-Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
-Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
-In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
-Across all scenarios, \emph{BO} and \emph{Grid} perform similarly, although due to its evaluation budget, \emph{BO} is allowed only 116 evaluations as opposed to the 1350 evaluations for \emph{Grid}, when using a resolution of 25 points per dimension.
-The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
-\emph{BO} is approximately 20 times faster.
-\input{generated/tables/table_time.tex}
-
 \begin{figure}[tbh]
@@
 \end{figure}
+The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
+Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
+Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
+In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+Across all scenarios, \emph{BO} and \emph{Grid} perform similarly, although due to its evaluation budget, \emph{BO} is allowed only 116 evaluations as opposed to the 1350 evaluations for \emph{Grid}, when using a resolution of 25 points per dimension.
+The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
+Runtimes were measured on a single core of a compute node with 2 Intel Xeon E5-2697v2 (2.70 GHz) with 12 Cores each and 512 GB of RAM in an non-exclusive usage of the node.
+\emph{BO} is approximately 20 times faster.
+\input{generated/tables/table_time.tex}
 
@@
 Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO}, with only a small advantage for \emph{Grid Small}.
-Despite the 10 more evaluations allowed, in the majority of cases \emph{BO} is superior to \emph{Gris Small}, see~Figure~\ref{fig:plot_boxplot_valid_y}.
+Despite the 10 more evaluations allowed, in the majority of cases \emph{BO} is superior to \emph{Grid Small}, see~Figure~\ref{fig:plot_boxplot_valid_y}.
 The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.
@@
   For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value with highest power ($y$) is displayed.
-  For each configuration in the grid 10 %stochastic
+  For each configuration in the grid 20 %stochastic
   results are obtained.
@@
 The curves also show the potential of optimization for the different scenarios.
-If the allowed treatment number is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less important.
-If the allowed treatment number is too small ($n_{\text{treat}} = 500$), then no configuration yields a practically usable design.
+If the number of overall treatments is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less important.
+If the number of overall treatments is too small ($n_{\text{treat}} = 500$), then no configuration yields a practically usable design.
 In this situation, a fast optimization helps to determine if a desired treatment number can actually yield sufficiently powerful designs.
@@
 In some scenarios \emph{Grid} was superior to \emph{BO}, as could be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
-Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500)$ and \emph{effect: paper}, $n_{\text{treat}} = 500)$, small values of $r$ lead to an optimal outcome.
+Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 500$, small values of $r$ lead to an optimal outcome.
 Here, \emph{BO} was not able to find these values close to the border of the search space.
@@
 This demonstrates that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
-However, an increased value of $n_\text{sim}$ also increases the runtime linearly, which conflicts the goal is to find the optimal configuration quickly.
+However, an increased value of $n_\text{sim}$ also increases the runtime linearly, which conflicts with the goal to find the optimal configuration quickly.
 In summary, \emph{BO} can find configurations with a small number of required simulations (higher noise) that are nearly as good as the optimal solution which is identified with a large number of simulations.
@@
 
-\newpage
-\phantom{aaaa}
-%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
-\clearpage
-\begin{center}
-\textbf{\large Supplemental Materials: \newtitle}
-\end{center}
-\FloatBarrier
-%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
-%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
-\setcounter{equation}{0}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\setcounter{page}{1}
-\makeatletter
-\renewcommand{\theequation}{S\arabic{equation}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\bibnumfmt}[1]{[S#1]}
-%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
-
-\input{generated/tables/table_best.tex}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
-\caption{%
-  Optimization curves of each optimization run and the mean of all runs drawn as a black line.
-  }
-\label{fig:plot_opt_path} 
-\end{figure}
+% \newpage
+% \phantom{aaaa}
+
+% %%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
+% \clearpage
+% \begin{center}
+% \textbf{\large Supplemental Materials: \newtitle}
+% \end{center}
+% \FloatBarrier
+% %%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
+% %%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
+% \setcounter{equation}{0}
+% \setcounter{figure}{0}
+% \setcounter{table}{0}
+% \setcounter{page}{1}
+% \makeatletter
+% \renewcommand{\theequation}{S\arabic{equation}}
+% \renewcommand{\thefigure}{S\arabic{figure}}
+% \renewcommand{\thetable}{S\arabic{table}}
+% \renewcommand{\bibnumfmt}[1]{[S#1]}
+% %%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
+
+% \input{generated/tables/table_best.tex}
+
+% \begin{figure}[htb]
+% \centering
+% \includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
+% \caption{%
+%   Optimization curves of each optimization run and the mean of all runs drawn as a black line.
+%   }
+% \label{fig:plot_opt_path} 
+% \end{figure}
 

### 2021-03-18 response_to_the_reviewers.md
more review responses

@@
 If the patients are different in in both stages (i.e. each patient is only treated once), both numbers are the same.
-However, we corrected the wording to avoid confusions.
+However, we corrected the wording to avoid confusions. #FIXME
 
@@
 
+--> can I remove (effect: paper2, n_treat = 1000)?
+While we agree, that for (effect: sigmoid, n_treat = 2000) Grid small has an advantage over BO, and changed the text accordingly, we don't see that advantage for (effect: paper, n_treat = 1000).
+Here, the median of both methods lays within the box of the other method so we cannot say that one mehtod has an advantage over the other.
+
 > p.
 12: It says: “For example, for scenario effect: sigmoid, n_treat = 1000 Grid is superior to BO, but the difference is smaller than 0.25%.” This isn’t true for n_treat = 1000 where the difference between Grid and BO is around 2%, but rather for n_treat = 2000.
+
+Thank you for spotting that mistake.
 
@@
 
---> clarify
+If r=1, then all treatments are done in stage 1. The number of FIXME: How do we name treatments and "individual patients".
 
@@
 > p. 14: What are the x-values referred to in the caption of Figure 4? No x is introduced in the notation, and Figure 4 itself plots y-values against r-values.
+
+Thanks for spotting this mistake. It is supposed to say theta-values which r is a component of. We clearified the caption.
 
@@
 > p. 9, l. 53 and > p. 10, l. 14: “Mattern”, which should be spelt “Matérn”
+
+Done.
 
@@
 
---> In out method we keep the number of patients fixed
+We agree that the power is only one measure of choosing a design.
+Therefore, the idea behind our method is that you are able to quickly find out which power you can achieve given the other fixed factors like number of patients or number of treatments #FIXME.
+Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
 --> Tim Friede (assumption of parameters, true clinical trial needs)
@@
 
---> Argument: Falschrum
+In the table the effects for the second stage (final) are smaller then for the fist stage (early) which is in line with most realistic scenarios.
 --> Often Realisitic scenario, also for the motivating example: First stage, second stage ...
@@
 
---> Argument: This will only lead to a result where more patients are selected.
-The idea is to run this optimization for different patient sizes, and then to decide which number of patients is required (also mention this in paper)
+If we include the number of patients in the optimization, the optimizer will propose to choose the maximal number of patients because this gives the maximal performance.
+This can be seen as we tried different values of n_treat which equals the number of patients. 
+Here, a higher number always resulted in a higher power.
+As mentioned in the paper, the idea is to run this optimization for different patient sizes, and then to decide which number of patients is required.
+This brings us to the future goal that we mentioned in the discussion: We can include n_treat as another parameter in a multi-criteria setting, where the result would be a Pareto set.
 
@@
 
---> simulation iteration means that we evaluated 1000 times wheter the test detected a significant effect.
-this does not depend on the complexity of the optimization.
-As a matter of fact we use 100 optimization iterations which is sufficient as can be seen in plot ...
-where we reach convergence.
+The simulation iterations are a specific parameter of the black-box function we optimize.
+The higher this value, the more reliable is the estimator of the power of the design.
+Our optimization is run for 100 iterations which is sufficient to converge as Figure 5 shows.
+Therefore, we do not see the need to run more iterations.
 
@@
 
---> add computational information, relative values are still meaningfull
+We added the hardware setup that is used.
+However, we think that the relative speed up between the methods adds a meaningful information.
 
@@
 
---> log transform: good idea, outlook
+--> We actually mentioned the log transformation ourselfes.
+As we mentioned in the final Discussion section of the paper we see this as a matter for a future benchmark to further improve the performance of BO.
+The aim of this paper was benchmark the BO approach in a basic setting without the prior knowledge, that small values of r are of higher importance.
 
@@
 
---> in fact our effect set "paper" issues a case study where the effect sets are estimated on a true dataset.
-we do not propose a method to analyze data, but a method to select the best design given a known theoretical situation.
-(check that this is clear)
+In fact the effect set "paper" is based on a case study where the effect sets are estimated on a true dataset. 
+In this paper we do not propose a method to analyze data, but a method to select the best design given a known theoretical situation.
 
-> Minor comments:
+### Minor comments:
 
-Page 8 line 29,  n_{treat} should be total # of pts, instead of “total number of treatment”.
+> Page 8 line 29,  n_{treat} should be total # of pts, instead of “total number of treatment”.
+
+Thanks, we fixed that mistake.

### 2021-03-18 main.tex
more review responses

@@
 In clinical trials, planning test procedures and sample sizes is a crucial task.
-A common goal is to maximize the test power, given a set of treatments and corresponding effect sizes. % under the null hypothesis.
+A common goal is to maximize the test power, given a set of treatments and corresponding effect sizes. % under the null hypothesis. #FIXME and number of patients?
 From a wide range of possible designs we aim to select the best one in a short time to allow quick decisions.
@@
 In the following, we demonstrate that the so-called \emph{Bayesian optimization} (BO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
-As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
+As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of patients.
 Further, it is crucial that such a design is found in a short time.
@@
 The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.
-Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
+Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$), (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 2000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
 In the first case, the absolute differences are negligible and all methods lead to a high power.
@@
 Note that between the scenarios the absolute differences of the optimization results are of different practical relevance.
-For example, for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$ \emph{Grid} is superior to \emph{BO}, but the difference is smaller than $0.25\%$.
+For example, for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$ \emph{Grid} is superior to \emph{BO}, but the difference is smaller than $0.25\%$.
 
@@
 Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
-As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2.
+As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2 (i.e.\ $r=1$).
 
@@
 \includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
-\caption{Best $x$-values found with \emph{Grid} and \emph{BO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
+\caption{Best configurations found with \emph{Grid} and \emph{BO} runs, i.e.
+corresponding $y_{\text{valid}}$ values for the ratio $r$ for the selected configurations.
+For the epsilon rule, $\epsilon$ and for the threshold rule $\tau$ are not displayed.}
 \label{fig:plot_best_x}

### 2021-03-22 response_to_the_reviewers.md
tim call

@@
 
---> Tim Friede
+--> Tim Friede,
+--> Checken, was gemacht wurde, dann klarstellen
 
@@
 
---> Tim? Gibt es diese assumption bei dir?
---> Mbo Teil habe ich übernommen.
-We clarified the assumptions in chapter 4 and 5.
+There is no normality assumption for the single outcome data but for the test statistics Z_i,j a multivariate normal distribution is assumed which will hold approximately if the sample size is sufficiently large.
+We clarified the assumptions at the very end of chapter 3.
 
@@
 
---> Tim
+We use a one-sided test and added this information in the table.
 
@@
 Here, the median of both methods lays within the box of the other method so we cannot say that one mehtod has an advantage over the other.
+Thanks for bringing that up. We revised the list of cases where Grid Small is in advantage.
 
@@
 
-If r=1, then all treatments are done in stage 1. The number of FIXME: How do we name treatments and "individual patients".
+If r=1, then all treatments are done in stage 1. The number of FIXME: formulation total sample size
 
@@
 
---> Maybe threshold is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization
+Maybe threshold is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization.  
+In our analysis of the results we saw that the optimal thresh values are not close to the border.
+This indicates that the ranges are wide enough.
 
@@
 > p. 16: The citation of the technical report by Bischl et al. should include a URL if possible.
+
+FIXME
 
@@
 We agree that the power is only one measure of choosing a design.
-Therefore, the idea behind our method is that you are able to quickly find out which power you can achieve given the other fixed factors like number of patients or number of treatments #FIXME.
+Therefore, the idea behind our method is that you are able to quickly find out which power you can achieve given the other fixed factors like total sample size or number of treatments #FIXME.
 Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
 --> Tim Friede (assumption of parameters, true clinical trial needs)
+--> We added in the outlook:  (chater) length of the study
 
@@
 --> Tim
+Early outcome overoptimistic and therefore higher.
+This is a typical scenario.
 

### 2021-03-22 main.tex
tim call

@@
 In clinical trials, planning test procedures and sample sizes is a crucial task.
-A common goal is to maximize the test power, given a set of treatments and corresponding effect sizes. % under the null hypothesis. #FIXME and number of patients?
+A common goal is to maximize the test power, given a set of treatments and corresponding effect sizes. % under the null hypothesis. #FIXME and number of samples?
 From a wide range of possible designs we aim to select the best one in a short time to allow quick decisions.
@@
 In the following, we demonstrate that the so-called \emph{Bayesian optimization} (BO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
-As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of patients.
+As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and a fixed total sample size.
 Further, it is crucial that such a design is found in a short time.
@@
 In contrast, some other design options such as the number of treatments and the total sample size are constrained by the environment and infrastructure available.
-The performance measures (sometimes also referred to as metrics) in the context with adaptive seamless designs are, e.g., power or sample size distributions.
+The performance measures (sometimes also referred to as metrics) in the context with adaptive seamless designs are, e.g., power, sample size distributions or duration of the study.
 %\paragraph{Simulating ASD}
@@
 The R-package \texttt{asd} by~\citet{parsons_software_2011} can be used for such simulations.
-As it generates test statistics, it is generally more efficient in comparison to simulating individual participant data.
+As it generates test statistics, following multivariate normal distributions, it is applicable to a wide range of outcomes, at least approximately, and generally more efficient in comparison to simulating individual participant data.
 The simulation model was described in more detail by \citet{friede_adaptive_2020}. 
@@
 The expected improvement, as well as other acquisition functions, are derived under the assumption that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
-The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters \hat{\mu} and \sigma^2 under the assumption that $f$ is a realization of a Gaussian process.
+The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters $\hat{\mu}$ and $\sigma^2$ under the assumption that $f$ is a realization of a Gaussian process.
 In practice, this assumption often cannot be verified.
@@
 The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
-In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
+In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of the total samples (ratio $r$) allocated to the second stage are given.
 \begin{table}[h]
@@
 \end{align}
-whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively and \emph{effect set} remains fixed as we are interested in the optimal trial design for a given \emph{effect set}.
+whereas $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients (samples) allocated to each treatment in stage 1 and stage 2 of the trial, respectively and \emph{effect set} remains fixed as we are interested in the optimal trial design for a given \emph{effect set}.
 Our main optimization goal is to maximize the statistical power $y$ depending on the trial design.
 
-However, we are interested in keeping the total number of treatments $n_{\text{treat}}$ constant in the overall trial for a fair comparison of the power of different designs.
-Therefore, we introduce the ratio of patients per treatment in stage 1, compared to patients per treatment in both stages
+However, we are interested in keeping the total sample size $n_{\text{total}}$ constant in the overall trial for a fair comparison of the power of different designs.
+Therefore, we introduce the ratio of patients (samples) per treatment in stage 1, compared to patients per treatment in both stages
 \begin{align}
@@
 
-In the following we will explain how we derive the values of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ in dependence of $r$ and $n_{\text{treat}}$ so that we can maximize
+In the following we will explain how we derive the values of $n_{\text{stage1}}$ and $n_{\text{stage2}}$ in dependence of $r$ and $n_{\text{total}}$ so that we can maximize
 \begin{equation}
   \label{eq:bbox2}
-  y = \tilde{f}(r, \text{selection strategy}, \epsilon, \tau, \text{effect set}, n_{\text{treat}})
+  y = \tilde{f}(r, \text{selection strategy}, \epsilon, \tau, \text{effect set}, n_{\text{total}})
 \end{equation}
-instead of~(\ref{eq:bbox}), whereas $n_{\text{treat}}$ and \emph{effect set} remain fixed.
+instead of~(\ref{eq:bbox}), whereas $n_{\text{total}}$ and \emph{effect set} remain fixed.
 
@@
 Within one stage, the same number of patients are allocated to all different treatments. 
-Then $n_{\text{treat}}$ is given by
+Then $n_{\text{total}}$ is given by
 \begin{equation}
   \label{eq:ntreat}
-  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}} \ .
+  n_{\text{total}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}} \ .
 \end{equation}
@@
 
-Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of treatments $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
+Given the number of treatments $k_1$ and $k_2$ in the two stages and the total sample size $n_{\text{total}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
 \begin{align}
@@
   n_{\text{stage1}} & = 
-  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
-  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
-  & = n_{\text{treat}}\cdot
+  n_{\text{stage1}}\cdot \frac{n_{\text{total}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
+  n_{\text{total}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
+  & = n_{\text{total}}\cdot
   \frac{r}{k_1\cdot
@@
   n_{\text{stage2}} & =
-  n_{\text{treat}}\cdot
+  n_{\text{total}}\cdot
   \frac{1-r}{k_1\cdot
@@
 \end{align}
-%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
-These formulas give the numbers of patients per treatment arm in the two stages, if $k_2$ is known. 
+%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{total}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
+These formulas give the numbers of patients (samples) per treatment arm in the two stages, if $k_2$ is known. 
 This is only the case for selection strategies 1, 2, and 3.
@@
 In these cases, for given values of $n_{\text{stage1}}$ and $\epsilon$ or $\tau$, respectively, we can obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
-For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
+For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{total}}}{k_1} \rceil, \lceil \frac{n_{\text{total}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
 We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed value for $n_{\text{stage1}}$.
-Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total treatments is closest to $n_{\text{treat}}$.
+Then we select the value of $n_{\text{stage1}}$ such that the corresponding total sample size is closest to $n_{\text{total}}$.
 More precisely, we minimize
@@
   %}_{=n_{\text{stage2}}}
-  \Big) - n_{\text{treat}} \right)^2
+  \Big) - n_{\text{total}} \right)^2
 %  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
@@
 %The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
-The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (on average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
-
-The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and number of total treatments, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~(\ref{eq:bbox2}).
+The calibration step is necessary to guarantee that the total sample size $n_{\text{total}}$ is (on average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
+
+The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and a fixed total sample size, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~(\ref{eq:bbox2}).
 
@@
   Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
-  Test level & \texttt{level} & $0.025$ \\ %FIXME: @Tim: One sided?
-  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
+  One-sided significance level & \texttt{level} & $0.025$ \\ %FIXME: @Tim: One sided?
+  % Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
   \hline
@@
 For this study we consider 12 scenarios.
-The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers for $n_{\text{treat}}$.
+The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers for $n_{\text{total}}$.
 Each optimizer is applied on each scenario with 20 stochastic repetitions.
@@
 \caption{%
-  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$ (500, 1000, 2000)
+  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{total}}$ (500, 1000, 2000)
   }
@@
 The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
-Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
+Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{total}} = 2000$; \emph{effect: sigmoid}, $n_{\text{total}} = 1000$; \emph{effect: sigmoid}, $n_{\text{total}} = 2000$).
 Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
-In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{total}} = 2000$; \emph{effect: paper2}, $n_{\text{total}} = 500$; \emph{effect: sigmoid}, $n_{\text{total}} = 500$).
 Across all scenarios, \emph{BO} and \emph{Grid} perform similarly, although due to its evaluation budget, \emph{BO} is allowed only 116 evaluations as opposed to the 1350 evaluations for \emph{Grid}, when using a resolution of 25 points per dimension.
@@
 
-Comparing \emph{BO Grid} to \emph{Grid}, we observe that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$, \emph{BO Grid} achieves slightly better performance than \emph{Grid}.
+Comparing \emph{BO Grid} to \emph{Grid}, we observe that for the scenarios \emph{effect: paper}, $n_{\text{total}} = 500$ and \emph{effect: paper}, $n_{\text{total}} = 1000$, \emph{BO Grid} achieves slightly better performance than \emph{Grid}.
 This suggests that the \emph{Grid} optimization selects the best outcome "optimistically" from all observed outcomes and does not take into account the stochasticity of the problem, i.e.
@@
 Despite the 10 more evaluations allowed, in the majority of cases \emph{BO} is superior to \emph{Grid Small}, see~Figure~\ref{fig:plot_boxplot_valid_y}.
-The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.
-Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$), (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 2000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
+The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{total}} = 500$.
+Only for the scenarios (\emph{effect: paper2}, $n_{\text{total}} = 1000$), (\emph{effect: sigmoid}, $n_{\text{total}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{total}} = 2000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
 In the first case, the absolute differences are negligible and all methods lead to a high power.
@@
 Note that between the scenarios the absolute differences of the optimization results are of different practical relevance.
-For example, for scenario \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$ \emph{Grid} is superior to \emph{BO}, but the difference is smaller than $0.25\%$.
+For example, for scenario \emph{effect: sigmoid}, $n_{\text{total}} = 2000$ \emph{Grid} is superior to \emph{BO}, but the difference is smaller than $0.25\%$.
 
@@
 We now investigate the results in more detail.
-We analyze the differences depending on the \emph{selection strategy} and the number of treatments at stage 1, as determined by the \emph{stage ratio} ($r$).
+We analyze the differences depending on the \emph{selection strategy} and the number of samples at stage 1, as determined by the \emph{stage ratio} ($r$).
 Figure~\ref{fig:plot_allbest} shows the observed outcomes for all evaluations on the grid and the final configurations found by \emph{BO}.
@@
 Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are nearly identical.
-As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2 (i.e.\ $r=1$).
+As intuitively anticipated, in almost all scenarios the worst results were obtained if all available samples are used in stage 1 and none in stage 2 (i.e.\ $r=1$).
 
@@
 The curves also show the potential of optimization for the different scenarios.
-If the number of overall treatments is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less important.
-If the number of overall treatments is too small ($n_{\text{treat}} = 500$), then no configuration yields a practically usable design.
-In this situation, a fast optimization helps to determine if a desired treatment number can actually yield sufficiently powerful designs.
+If the total sample size is high ($n_{\text{total}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less important.
+If the total sample size is too small ($n_{\text{total}} = 500$), then no configuration yields a practically usable design.
+In this situation, a fast optimization helps to determine if a desired total sample size can actually yield sufficiently powerful designs.
 Of special interest are the scenarios where the correct choice of the design configuration can make a difference between a design with acceptable power and a design with too low power, e.g.
-for the scenarios (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: linear}, $n_{\text{treat}} = 2000$, \emph{effect: paper}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
+for the scenarios (\emph{effect: linear}, $n_{\text{total}} = 1000$; \emph{effect: linear}, $n_{\text{total}} = 2000$, \emph{effect: paper}, $n_{\text{total}} = 1000$; \emph{effect: sigmoid}, $n_{\text{total}} = 1000$).
 Note, that the selection strategies \emph{thresh} and \emph{all} never selected powerful designs.
@@
 The optimal values for $\epsilon$ and $\theta$ are not shown.
-For some scenarios most runs of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
+For some scenarios most runs of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{total}} = 1000$).
 In such cases the signal-to-noise ratio was sufficiently high to reliably find the best configuration.
 In some scenarios \emph{Grid} was superior to \emph{BO}, as could be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
-Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 500$, small values of $r$ lead to an optimal outcome.
+Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{total}} = 500$ and \emph{effect: paper}, $n_{\text{total}} = 500$, small values of $r$ lead to an optimal outcome.
 Here, \emph{BO} was not able to find these values close to the border of the search space.
@@
 
-In other cases \emph{Grid} found different configurations in different runs, while \emph{BO} lead to more stable results, also with higher power values (e.g.\ \emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
+In other cases \emph{Grid} found different configurations in different runs, while \emph{BO} lead to more stable results, also with higher power values (e.g.\ \emph{effect: linear}, $n_{\text{total}} = 1000$; \emph{effect: paper2}, $n_{\text{total}} = 500$; \emph{effect: sigmoid}, $n_{\text{total}} = 500$).
 Here we can assume, that \emph{BO} successfully accounted for the noise.
-A special case is scenario \emph{effect: paper2}, $n_{\text{treat}} = 2000$:
+A special case is scenario \emph{effect: paper2}, $n_{\text{total}} = 2000$:
 Here nearly all found configurations yield perfect outcomes.
@@
 
-To investigate the effect of the function noise variance ($\sigma^2$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$, restricting to the scenario \emph{effect: paper}, $n_{\text{treat}} = 2000$.
+To investigate the effect of the function noise variance ($\sigma^2$) on the optimization result we increase the simulation iterations from $n_\text{sim} = 1000$ to $n_\text{sim} = 5000$, restricting to the scenario \emph{effect: paper}, $n_{\text{total}} = 2000$.
 % This is expected to increase the runtime by a factor of 5. % obvious, also not of interest here
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
-    \captionof{figure}{Optimization progress for \emph{BO}: Best outcome found so far for each iteration, in comparison to the best outcome found in the grid search, for effect set \emph{paper} and $n_{\text{treat}} = 2000$}
+    \captionof{figure}{Optimization progress for \emph{BO}: Best outcome found so far for each iteration, in comparison to the best outcome found in the grid search, for effect set \emph{paper} and $n_{\text{total}} = 2000$}
     \label{fig:plot_opt_path_5000}
@@
     \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
-    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{treat}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
+    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{total}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
     \label{fig:plot_boxplot_valid_y_5000}
@@
 In clinical trials, determining the test methods and sample sizes is a key step in the planning process.
-One main goal is to maximize the test power, given a set of treatments with corresponding effect sizes. % under the null hypothesis.
+One main goal is to maximize the test power, given a set of treatments with corresponding effect sizes. % under the null hypothesis. %FIXME sample size
 If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
@@
 In this paper we analyzed if BO can quickly find an adaptive seamless design with high power, more efficiently than an exhaustive grid search.
-For various sets of treatment effect sizes, BO was able to find competitive designs in a fraction of the time.
+For various sets of treatment effect sizes, BO was able to find competitive designs in a fraction of the time. %FIXME and samples?
 In contrast to grid search, BO was able to account for stochasticity in the metric that should be optimized.
@@
 Here, a possible improvement can be to apply a log transformation to such parameters. % in the future.
-Further, instead of keeping the number of treatments constant, we could apply multi-criteria optimization and include this number as an additional outcome.
-This multi-criteria approach would result in a set of points (a Pareto front) that minimize the number of treatments and at the same time maximize the statistical power of the trial design.
+Further, instead of keeping the total sample size constant, we could apply multi-criteria optimization and include this number as an additional outcome.
+This multi-criteria approach would result in a set of points (a Pareto front) that minimizes the total sample size and at the same time maximize the statistical power of the trial design.
 The user could then see for which number of trials the power is above a certain threshold (e.g. $95\%$).

### 2021-03-22 response_to_the_reviewers.md
n total

@@
 
---> Tim Friede,
---> Checken, was gemacht wurde, dann klarstellen
+As in the original paper (Friede et. al. 2020) we count the rejections of any of the hypotheses 3 or 4 towards the power.
+We made this more clear throughout the paper and clarified that it is up to the user to determine the method how the power is calculated.
 
@@
 
-n_treat is the number of total treatments. 
-If the patients are different in in both stages (i.e. each patient is only treated once), both numbers are the same.
-However, we corrected the wording to avoid confusions. #FIXME
+We improved our wording, so that the number of treatment arms cannot longer be confused with the individual number of treatments of each patient.
+Therefore, n_total replaces n_treat and defines the total sample size (formerly “total number of treatments”).
+If the patients are different in in both stages (i.e. each patient is only treated once) the total number of patients is would be the same as the total sample size.
+However, we only us "total sample size" phrase to avoid confusions.
 

### 2021-03-22 generator/main.R

@@ -288,7 +288,7 @@ plot_wrapper(name = "plot_allbest", fig.height = 1.5 * FIG_HEIGHT, expr = {
   g = g + geom_point(aes(size = algorithm, shape = algorithm), alpha = 0.15, shape = 16)
   g = g + scale_color_manual(values = select_labels_colors)
   g = g + scale_size_manual(labels = algorithm_labels[c("mbo" ,"eval")], values = c("mbo" = 2, "eval" = 0.5))
-  g = g + facet_grid(effect~n_cases, scales = "free", labeller = label_bquote(cols = {n[treat]==.(n_cases)}))
+  g = g + facet_grid(effect~n_cases, scales = "free", labeller = label_bquote(cols = {n[total]==.(n_cases)}))
   g = g + labs(x = expression(k[1] %.% n[stage1]))
   g = g + theme(legend.position = "bottom")
   g
@@ -313,7 +313,7 @@ plot_wrapper(name = "plot_best_x", fig.height = 1.5 * FIG_HEIGHT, expr = {
   g = g + labs(x = expression(r), y = expression(y[valid]))
   g = g + theme(legend.position = "bottom", strip.background = element_blank(), strip.text.x = element_blank())
   #grid headlines
-  col_heads = lapply(unique(tmp$n_cases), function(x) bquote(n[treat]==.(x))) %>% do.call("expression",.) %>% lapply(textGrob, gp = gpar(fontsize = 10, color = "grey10"))
+  col_heads = lapply(unique(tmp$n_cases), function(x) bquote(n[total]==.(x))) %>% do.call("expression",.) %>% lapply(textGrob, gp = gpar(fontsize = 10, color = "grey10"))
   row_heads = levels(factor(tmp$effect)) %>% lapply(textGrob, gp = gpar(fontsize = 10, col = "grey10"), rot=90*3)
   layout_mat = matrix(c(
     1, 2, 3, NA,
@@ -357,7 +357,7 @@ plot_wrapper(name = "plot_opt_path", fig.height = 1.6 * FIG_HEIGHT, expr = {
   g = g + labs(x = "iteration", y = expression(y[iter]*"*" - y[grid]*"*"), color = NULL)
   g = g + guides(colour = guide_legend(override.aes = list(size=2)))
   #grid headlines
[…]

### 2021-03-22 main.tex

@@
 In clinical trials, planning test procedures and sample sizes is a crucial task.
-A common goal is to maximize the test power, given a set of treatments and corresponding effect sizes. % under the null hypothesis. #FIXME and number of samples?
+A common goal is to maximize the test power, given a set of treatments, corresponding effect sizes, and a total number of samples.
 From a wide range of possible designs we aim to select the best one in a short time to allow quick decisions.
@@
 The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.
+Depending on the application the power can be calculated as the rate of rightfully rejecting any, all, or a subset of the elementary hypotheses~\citet{senn_power_2007}.
 
@@
 For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
-The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses. %???
+%The vector of treatment numbers for determining power \texttt{ptest} counts rejections of treat3 or 4 hypotheses. %???
+The vector of treatment numbers \texttt{ptest} determines the hypotheses for which rejections are counted towards the power.
+Here, the rejection of one or both of the hypotheses that treatment 3 or 4 has no significant effect against the control determines the power, as in~\citet{friede_adaptive_2020}.
+
 Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
@@
 
-% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
-%FIXME: Ist mir immer noch nicht klar / nicht mehr klar?
+% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, genau da gucken wir drauf!
 
@@
   Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
-  One-sided significance level & \texttt{level} & $0.025$ \\ %FIXME: @Tim: One sided?
-  % Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
+  One-sided significance level & \texttt{level} & $0.025$ \\
+  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
   \hline
@@
 In clinical trials, determining the test methods and sample sizes is a key step in the planning process.
-One main goal is to maximize the test power, given a set of treatments with corresponding effect sizes. % under the null hypothesis. %FIXME sample size
+One main goal is to maximize the test power, given a set of treatments with corresponding effect sizes and a fixed number of samples.
 If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
@@
 In this paper we analyzed if BO can quickly find an adaptive seamless design with high power, more efficiently than an exhaustive grid search.
-For various sets of treatment effect sizes, BO was able to find competitive designs in a fraction of the time. %FIXME and samples?
+For various sets of treatment effect sizes and total sample sizes, BO was able to find competitive designs in a fraction of the time.
 In contrast to grid search, BO was able to account for stochasticity in the metric that should be optimized.

### 2021-03-22 response_to_the_reviewers.md
more responses

@@
 As in the original paper (Friede et. al. 2020) we count the rejections of any of the hypotheses 3 or 4 towards the power.
-We made this more clear throughout the paper and clarified that it is up to the user to determine the method how the power is calculated.
+We made this more clear throughout the paper and clarified that it is up to the user to determine the method of how the power is calculated.
 
@@
 We improved our wording, so that the number of treatment arms cannot longer be confused with the individual number of treatments of each patient.
-Therefore, n_total replaces n_treat and defines the total sample size (formerly “total number of treatments”).
+Therefore, the total sample size (formerly “total number of treatments”) is denoted as n_total (formerly n_treat)
 If the patients are different in in both stages (i.e. each patient is only treated once) the total number of patients is would be the same as the total sample size.
-However, we only us "total sample size" phrase to avoid confusions.
+However, we only use the phrase "total sample size" to avoid confusions.
 
@@
 
---> can I remove (effect: paper2, n_treat = 1000)?
-While we agree, that for (effect: sigmoid, n_treat = 2000) Grid small has an advantage over BO, and changed the text accordingly, we don't see that advantage for (effect: paper, n_treat = 1000).
-Here, the median of both methods lays within the box of the other method so we cannot say that one mehtod has an advantage over the other.
-Thanks for bringing that up. We revised the list of cases where Grid Small is in advantage.
+Thanks for bringing that up. We revised the list of cases where Grid Small shows a notable advantage.
+An advantage is not notable if the median of both methods lays within the box of the other method.
+Therefore, we don't see an advantage for (effect: paper, n_treat = 1000).
 
@@
 
-If r=1, then all treatments are done in stage 1. The number of FIXME: formulation total sample size
+If r=1, then all samples of the total number of allowed samples are done in stage 1. 
+Therefore no more treatments can be done in stage 2.
+We improved the wording to make this more clear.
 
@@
 
-Maybe threshold is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization.  
-In our analysis of the results we saw that the optimal thresh values are not close to the border.
+In the analysis of the results we saw that the optimal threshold values are not close to the border.
 This indicates that the ranges are wide enough.
+We assume that the threshold selection is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization.
 
@@
 
-FIXME
+We added the arxiv id.
 
@@
 
-> p. 9, l. 20: “in average”, which should read “on average”
-
-Done.
-
-> p. 9, l. 24: “in 8”, which should read “in (8)”
-
-Done.
-
-> p. 9, l. 53 and > p. 10, l. 14: “Mattern”, which should be spelt “Matérn”
-
-Done.
-
-> p. 10, l. 28: “Gird”
-
-Done.
-
-> p. 10, l. 39: “???”
-
-Done.
-
-> p. 12, l. 19: “Gris”
-
-Done.
-
-> p. 13, l. 48f.: remove closing bracket from “500)”
-
-Done.
-
-> p. 14, l. 55: “which conflicts the goal is to find the optimal” (grammar)
-
-Done.
+We corrected all spotted typos. Thank you.
 
@@
 We agree that the power is only one measure of choosing a design.
-Therefore, the idea behind our method is that you are able to quickly find out which power you can achieve given the other fixed factors like total sample size or number of treatments #FIXME.
+Therefore, the idea behind our method is that you are able to quickly find out which power you can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
 Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
---> Tim Friede (assumption of parameters, true clinical trial needs)
---> We added in the outlook:  (chater) length of the study
+We added study duration as a possible limitation in our conclusion and made sure that the limitations of our approach are mentioned more clearly.
+Assuming certain parameters under the null hypothesis is a standard approach.
+However, we see that a more Bayesian approach is certainly interesting, as we already stated in the last paragraph of the discussion.
 
@@
 In the table the effects for the second stage (final) are smaller then for the fist stage (early) which is in line with most realistic scenarios.
---> Often Realisitic scenario, also for the motivating example: First stage, second stage ...
---> Tim
-Early outcome overoptimistic and therefore higher.
-This is a typical scenario.
+Typically the early outcome is overoptimistic and therefore higher.
+This is a realistic scenario.
 
@@
 
---> Argument: Given a simulation setting we optimize the parameters eps and thr.
-This is independent of the application but of course the clinical application can give hints which simulation settings are adequate
+Given a simulation setting we optimize the parameters eps and thr. This is independent of the application but of course the clinical application can give hints which simulation settings are adequate.
 
@@
 
-If we include the number of patients in the optimization, the optimizer will propose to choose the maximal number of patients because this gives the maximal performance.
-This can be seen as we tried different values of n_treat which equals the number of patients. 
+If we included the number of patients in the optimization, the optimizer will propose to choose the maximal number of patients because this gives the maximal performance.
+This is also indicated in our results, as we tried different values of n_total which equals the number of patients. 
 Here, a higher number always resulted in a higher power.
 As mentioned in the paper, the idea is to run this optimization for different patient sizes, and then to decide which number of patients is required.
-This brings us to the future goal that we mentioned in the discussion: We can include n_treat as another parameter in a multi-criteria setting, where the result would be a Pareto set.
+This brings us to the future goal that we mentioned in the discussion: We can include n_total as another parameter and outcome in a multi-criteria optimisation setting.
 
@@
 
---> We actually mentioned the log transformation ourselfes.
-As we mentioned in the final Discussion section of the paper we see this as a matter for a future benchmark to further improve the performance of BO.
-The aim of this paper was benchmark the BO approach in a basic setting without the prior knowledge, that small values of r are of higher importance.
+As we mentioned in the final discussion section of the paper, we see this as a matter for a future benchmark to further improve the performance of BO.
+The aim of this paper was to benchmark BO in a basic setting without the prior knowledge, that small values of r are of higher importance.
 

### 2021-03-22 main.tex
Updates from Overleaf

@@
 The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{total}} = 500$.
-Only for the scenarios (\emph{effect: paper2}, $n_{\text{total}} = 1000$), (\emph{effect: sigmoid}, $n_{\text{total}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{total}} = 2000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
-In the first case, the absolute differences are negligible and all methods lead to a high power.
-For the second case, we will later show that the optimal configuration is close to the border of the search space.
+Only for the scenarios (\emph{effect: sigmoid}, $n_{\text{total}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{total}} = 2000$) \emph{Grid Small} has a notable advantage over \emph{BO} and \emph{Grid}.
+In the second case, the absolute differences are negligible and all methods lead to a high power.
+Also for this case, we will later show that the optimal configuration is close to the border of the search space.
 
@@
 Note that between the scenarios the absolute differences of the optimization results are of different practical relevance.
-For example, for scenario \emph{effect: sigmoid}, $n_{\text{total}} = 2000$ \emph{Grid} is superior to \emph{BO}, but the difference is smaller than $0.25\%$.
+For example, for scenario \emph{effect: sigmoid}, $n_{\text{total}} = 2000$ \emph{Grid} and \emph{Grid Small} are superior to \emph{BO}, but the difference is smaller than $0.5\%$.
 

### 2021-03-22 response_to_the_reviewers.md
typos + updated figs

@@
 
-Thanks for spotting this mistake. It is supposed to say theta-values which r is a component of. We clearified the caption.
+Thanks for spotting this mistake. It is supposed to say theta-values which r is a component of. We clarified the caption.
 
@@
 As mentioned in the paper, the idea is to run this optimization for different patient sizes, and then to decide which number of patients is required.
-This brings us to the future goal that we mentioned in the discussion: We can include n_total as another parameter and outcome in a multi-criteria optimisation setting.
+This brings us to the future goal that we mentioned in the discussion: We can include n_total as another parameter and outcome in a multi-criteria optimization setting.
 

### 2021-03-24 response_to_the_reviewers.md
review mit jörg

@@
 We would like to thank all reviewers for their valuable and constructive comments on the paper.
-We especially that Reviewer 1 for the thoughtful and in-depth feedback.
+We especially thank Reviewer 1 for the thoughtful feedback that helped us to clarify the definition of the total number of samples and the power.
 
@@
 
-As in the original paper (Friede et. al. 2020) we count the rejections of any of the hypotheses 3 or 4 towards the power.
-We made this more clear throughout the paper and clarified that it is up to the user to determine the method of how the power is calculated.
+As in the original paper (Friede et. al. 2020) in which our motivating example was introduced, we count the rejections of any of the hypotheses 3 or 4 for calculating the power.
+We now clearly state this in the paper and also clarify that there are different options and it is up to the user to determine the method of how the power is calculated.
 
@@
 Therefore, the total sample size (formerly “total number of treatments”) is denoted as n_total (formerly n_treat)
-If the patients are different in in both stages (i.e. each patient is only treated once) the total number of patients is would be the same as the total sample size.
+If the patients are different in both stages (i.e. each patient is treated only once) the total number of patients would be the same as the total sample size.
 However, we only use the phrase "total sample size" to avoid confusions.
@@
 
-Thanks for catching that mistake. It is now fixed.
+Thanks for catching that mistake. It is now fixed, we use l for the resolution of the search grid.
 
@@
 There is no normality assumption for the single outcome data but for the test statistics Z_i,j a multivariate normal distribution is assumed which will hold approximately if the sample size is sufficiently large.
-We clarified the assumptions at the very end of chapter 3.
+We clarified that the test statistics are multivariate normal distributed at least approximately at the very end of chapter 3.
 
@@
 
-We added an explanation and referenced previous studies, that used similar values.
+We added an explanation and referenced previous studies that used similar values.
 
@@
 
-We added an explanation and a reference.
+We added an explanation that often 10% of the budget is used for the initial design, together with a reference.
 
@@
 Thanks for bringing that up. We revised the list of cases where Grid Small shows a notable advantage.
-An advantage is not notable if the median of both methods lays within the box of the other method.
-Therefore, we don't see an advantage for (effect: paper, n_treat = 1000).
+An advantage is not notable if the median of both methods lies within the box of the other method.
+Therefore, we don't see an advantage for the configuration (effect: paper, n_treat = 1000).
 
@@
 
-Thank you for spotting that mistake.
+Thank you for spotting that mistake. This was changed.
 
@@
 
-If r=1, then all samples of the total number of allowed samples are done in stage 1. 
-Therefore no more treatments can be done in stage 2.
-We improved the wording to make this more clear.
+If r=1, then all samples of the total number of allowed samples are evaluated in stage 1. 
+Therefore no more treatments can be evaluated in stage 2.
+We improved the wording to make this clear.
 
@@
 
-Thanks for spotting this mistake. It is supposed to say theta-values which r is a component of. We clarified the caption.
+Thanks for spotting this mistake. We changed the caption, replacing "x" by "configuration".
 
@@
 
-We agree that the power is only one measure of choosing a design.
-Therefore, the idea behind our method is that you are able to quickly find out which power you can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
+We agree that the power is only one measure for choosing a design.
+Therefore, the idea behind our method is to quickly find out which power one can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
 Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
-We added study duration as a possible limitation in our conclusion and made sure that the limitations of our approach are mentioned more clearly.
+We added study duration as a possible limitation in our conclusion section and made sure that the limitations of our approach are mentioned more clearly.
 Assuming certain parameters under the null hypothesis is a standard approach.
@@
 
-In the table the effects for the second stage (final) are smaller then for the fist stage (early) which is in line with most realistic scenarios.
-Typically the early outcome is overoptimistic and therefore higher.
-This is a realistic scenario.
+In the table the effects for the second stage (final) are indeed smaller then for the fist stage (early).
+This is a realistic scenario, as often the early outcome is overoptimistic and therefore higher.
 
@@
 
-If we included the number of patients in the optimization, the optimizer will propose to choose the maximal number of patients because this gives the maximal performance.
+If we included the number of patients in the optimization, the optimizer would propose to choose the maximal number of patients, because this gives the maximal performance.
 This is also indicated in our results, as we tried different values of n_total which equals the number of patients. 
 Here, a higher number always resulted in a higher power.
-As mentioned in the paper, the idea is to run this optimization for different patient sizes, and then to decide which number of patients is required.
+As mentioned in the paper, the idea is to run this optimization for different sample sizes (denoted by n_total), and then to decide which sample size is required.
 This brings us to the future goal that we mentioned in the discussion: We can include n_total as another parameter and outcome in a multi-criteria optimization setting.
@@
 Therefore, we do not see the need to run more iterations.
+We added a corresponding statement in section 5.7.
 
@@
 
-We added the hardware setup that is used.
-However, we think that the relative speed up between the methods adds a meaningful information.
+Thank you. We added the hardware setup that was used.
 
@@
 
-As we mentioned in the final discussion section of the paper, we see this as a matter for a future benchmark to further improve the performance of BO.
-The aim of this paper was to benchmark BO in a basic setting without the prior knowledge, that small values of r are of higher importance.
+As we mentioned in the final discussion section of the paper, we see this as a topic for a future benchmark to further improve the performance of BO.
+The aim of this paper was to benchmark BO in a basic setting without the prior knowledge that small values of r are of higher importance.
 
@@
 
-In fact the effect set "paper" is based on a case study where the effect sets are estimated on a true dataset. 
-In this paper we do not propose a method to analyze data, but a method to select the best design given a known theoretical situation.
+In fact the effect set "paper" is based on a case study where the effect sets are estimated on a true dataset (Friede et al, 2020). 
+In this paper we do not propose a method for analyzing a real dataset, but a method to select the best design given a known theoretical situation.
 

### 2021-03-24 main.tex
review mit jörg

@@
 The expected improvement, as well as other acquisition functions, are derived under the assumption that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
-The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters $\hat{\mu}$ and $\sigma^2$ under the assumption that $f$ is a realization of a Gaussian process.
+The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters $\hat{\mu}$ and $s^2(\theta)$ under the assumption that $f$ is a realization of a Gaussian process.
 In practice, this assumption often cannot be verified.
@@
 As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
-We start BO with an initial design of $16$ randomly sampled points, following an established rule-of-thumb of $4$ points per dimension, which is also well in line with the recommendation of using around $10\%$ of your budget for the initial design~\cite{bossek_initial_2020}.
+We start BO with an initial design of $16$ randomly sampled points, following an established rule-of-thumb of $4$ points per dimension, which is also well in line with the recommendation of using around $10\%$ of the budget for the initial design~\cite{bossek_initial_2020}.
 We allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
@@
 Figure~\ref{fig:plot_opt_path_5000} shows that indeed the lower noise level leads to more stable simulation runs, meaning that single optimization runs lead to results closer to the mean result.
+Both curves indicate, that more iterations of \emph{BO} would not yield substantially better outcomes as the progress is stagnating after 80 iterations.
 It has to be noted that this plot reports the best observed outcome so far and not the one that is chosen as final result.

### 2021-03-25 response_to_the_reviewers.md
diff generation and review

@@
+---
+title: Test
+author: Author Name
+header-includes: |
+  \usepackage{framed}
+  \newlength{\leftbarwidth}
+  \setlength{\leftbarwidth}{3pt}
+  \newlength{\leftbarsep}
+  \setlength{\leftbarsep}{-10pt}
+  \newcommand*{\leftbarcolorcmd}{\color{leftbarcolor}}
+  \colorlet{leftbarcolor}{gray}
+  \renewenvironment{leftbar}{%
+      \def\FrameCommand{{\leftbarcolorcmd{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
+      \MakeFramed {\advance \hsize -\width \FrameRestore }%
+    \noindent
+  }{%
+      \endMakeFramed
+  }
+  \renewenvironment{quote}
+  {\vspace{5pt} \leftbar\list{}{\rightmargin\leftmargin}%
+  \item\vspace{-5pt}\relax}
+    {\endlist\endleftbar}
+---
+
 # Response to the reviews for "Improving Adaptive Seamless Designs through Bayesian optimization"
@@
 
-> This is a well organised and written manuscript, describing a very thorough investigation of a reasonably interesting problem.
+This is a well organised and written manuscript, describing a very thorough investigation of a reasonably interesting problem.
 Computational aspects of complex adaptive clinical trial designs have all too often been neglected.
@@
 
-> 1) Power is used as a key performance metric, yet it lacks an explicit definition.
+1) Power is used as a key performance metric, yet it lacks an explicit definition.
 The problem, of course, is that power is slightly more complicated in a multi-stage design with multiple arms (and therefore multiple hypotheses to be tested) than it is in a simple comparison of two groups.
@@
 10 it is stated that “the vector of treatment numbers for determining power counts rejections of 3 or 4 hypotheses”, implying a power definition more similar to ‘conjunctive’ or ‘maximal’ power.
->
-> Reference: Senn S, Bretz F (2007) Power and sample size when multiple endpoints are considered. Pharmaceutical Statistics, 6(3), 161-170.
 
-As in the original paper (Friede et. al. 2020) in which our motivating example was introduced, we count the rejections of any of the hypotheses 3 or 4 for calculating the power.
-We now clearly state this in the paper and also clarify that there are different options and it is up to the user to determine the method of how the power is calculated.
+Reference: Senn S, Bretz F (2007) Power and sample size when multiple endpoints are considered. Pharmaceutical Statistics, 6(3), 161-170.
 
->2) There is some inconsistent notation:
->
->a) It seems that n_treat is varyingly defined as the “total number of treatments” (p.
+> As in the original paper (Friede et. al. 2020) in which our motivating example was introduced, we count the rejections of any of the hypotheses 3 or 4 for calculating the power.
+> We now clearly state this in the paper and also clarify that there are different options and it is up to the user to determine the method of how the power is calculated.
+
+2) There is some inconsistent notation:
+
+a) It seems that n_treat is varyingly defined as the “total number of treatments” (p.
 8, l.
@@
 
-We improved our wording, so that the number of treatment arms cannot longer be confused with the individual number of treatments of each patient.
-Therefore, the total sample size (formerly “total number of treatments”) is denoted as n_total (formerly n_treat)
-If the patients are different in both stages (i.e. each patient is treated only once) the total number of patients would be the same as the total sample size.
-However, we only use the phrase "total sample size" to avoid confusions.
+> We improved our wording, so that the number of treatment arms cannot longer be confused with the individual number of treatments of each patient.
+> Therefore, the total sample size (formerly “total number of treatments”) is denoted as n_total (formerly n_treat)
+> If the patients are different in both stages (i.e. each patient is treated only once) the total number of patients would be the same as the total sample size.
+> However, we only use the phrase "total sample size" to avoid confusions.
 
-> b) On p. 3 k is introduced as the running index for treatments (k = 1, …, K) but later on p. 9 k is the resolution of the search grid.
+b) On p. 3 k is introduced as the running index for treatments (k = 1, …, K) but later on p. 9 k is the resolution of the search grid.
 
-Thanks for catching that mistake. It is now fixed, we use l for the resolution of the search grid.
+> Thanks for catching that mistake. It is now fixed, we use l for the resolution of the search grid.
 
@@
 
->p. 2: For the benefit of readers less familiar with the term, the authors should explain at least briefly what they mean by “expensive black-box” optimisation problems.
+p. 2: For the benefit of readers less familiar with the term, the authors should explain at least briefly what they mean by “expensive black-box” optimisation problems.
 
-We added a short explanation.
+> We added a short explanation.
 
->p.
+p.
 3 ff.: This may be implicitly clear but it should also be said explicitly that in the motivating example and throughout the remainder of the paper a normality assumption is made for all outcome data.
 
-There is no normality assumption for the single outcome data but for the test statistics Z_i,j a multivariate normal distribution is assumed which will hold approximately if the sample size is sufficiently large.
-We clarified that the test statistics are multivariate normal distributed at least approximately at the very end of chapter 3.
+> There is no normality assumption for the single outcome data but for the test statistics Z_i,j a multivariate normal distribution is assumed which will hold approximately if the sample size is sufficiently large.
+> We clarified that the test statistics are multivariate normal distributed at least approximately at the very end of chapter 3.
 
->p. 9: How were the values for the resolution of Grid and Grid Small (k=25 and k=7) selected?
+p. 9: How were the values for the resolution of Grid and Grid Small (k=25 and k=7) selected?
 
-We added an explanation and referenced previous studies that used similar values.
+> We added an explanation and referenced previous studies that used similar values.
 
-> p. 10: Similarly, how was the number of randomly sampled points (16) for the initial design for BO chosen?
+p. 10: Similarly, how was the number of randomly sampled points (16) for the initial design for BO chosen?
 
-We added an explanation that often 10% of the budget is used for the initial design, together with a reference.
+> We added an explanation that often 10% of the budget is used for the initial design, together with a reference.
 
-> p. 10: In section 5.4 ten additional stochastic simulations are mentions, which seems in conflict with the 20 additional stochastic simulations mentioned in other parts of the manuscript.
+p. 10: In section 5.4 ten additional stochastic simulations are mentions, which seems in conflict with the 20 additional stochastic simulations mentioned in other parts of the manuscript.
 
-Thanks, for spotting that mistake. Indeed we use 20 stochastic repetitions.
+> Thanks, for spotting that mistake. Indeed we use 20 stochastic repetitions.
 
-> p. 10: It should be clarified whether the test level of 0.025 is one-sided.
+p. 10: It should be clarified whether the test level of 0.025 is one-sided.
 
-We use a one-sided test and added this information in the table.
+> We use a one-sided test and added this information in the table.
 
-> p.
+p.
 11: Whilst I appreciate that the focus is on relative improvements in runtime, the authors should at least briefly describe the hardware setup with which these runtimes were achieved, to provide some context for the absolute runtimes.
 
-We added the requested information.
+> We added the requested information.
 
-> p.
+p.
 12: It is stated that: “Only for the scenarios (effect: paper2, n_treat = 1000) and (effect: sigmoid, n_treat = 1000) Grid Small has a small advantage over BO and Grid.” I think it should be added that Grid Small also has an advantage over BO for (effect: paper, n_treat = 1000) and certainly for (effect: sigmoid, n_treat = 2000).
 
-Thanks for bringing that up. We revised the list of cases where Grid Small shows a notable advantage.
-An advantage is not notable if the median of both methods lies within the box of the other method.
-Therefore, we don't see an advantage for the configuration (effect: paper, n_treat = 1000).
+> Thanks for bringing that up. We revised the list of cases where Grid Small shows a notable advantage.
+> An advantage is not notable if the median of both methods lies within the box of the other method.
+> Therefore, we don't see an advantage for the configuration (effect: paper, n_treat = 1000).
 
-> p.
+p.
 12: It says: “For example, for scenario effect: sigmoid, n_treat = 1000 Grid is superior to BO, but the difference is smaller than 0.25%.” This isn’t true for n_treat = 1000 where the difference between Grid and BO is around 2%, but rather for n_treat = 2000.
 
-Thank you for spotting that mistake. This was changed.
+> Thank you for spotting that mistake. This was changed.
 
-> p.
+p.
 12: Apologies but I’m struggling to understand the following sentence: “As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2.” How can none of the treatments be used in stage 2 when on p.
@@
 
-If r=1, then all samples of the total number of allowed samples are evaluated in stage 1. 
-Therefore no more treatments can be evaluated in stage 2.
-We improved the wording to make this clear.
+> If r=1, then all samples of the total number of allowed samples are evaluated in stage 1. 
+> Therefore no more treatments can be evaluated in stage 2.
+> We improved the wording to make this clear.
 
-> p.
+p.
 13: It is stated that: “Note, that the selection strategies thresh and all never selected powerful designs.” Why is that?
@@
 
-In the analysis of the results we saw that the optimal threshold values are not close to the border.
-This indicates that the ranges are wide enough.
-We assume that the threshold selection is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization.
+> In the analysis of the results we saw that the optimal threshold values are not close to the border.
+> This indicates that the ranges are wide enough.
+> We assume that the threshold selection is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization.
 
-> p. 14: What are the x-values referred to in the caption of Figure 4? No x is introduced in the notation, and Figure 4 itself plots y-values against r-values.
+p. 14: What are the x-values referred to in the caption of Figure 4? No x is introduced in the notation, and Figure 4 itself plots y-values against r-values.
 
-Thanks for spotting this mistake. We changed the caption, replacing "x" by "configuration".
+> Thanks for spotting this mistake. We changed the caption, replacing "x" by "configuration".
 
-> p. 16: The citation of the technical report by Bischl et al. should include a URL if possible.
+p. 16: The citation of the technical report by Bischl et al. should include a URL if possible.
 
-We added the arxiv id.
+> We added the arxiv id.
 
@@
 
-We corrected all spotted typos. Thank you.
+> We corrected all spotted typos. Thank you.
 
@@
 
-> The authors propose to use Bayesian optimization (BO) to improve the efficiency of the design selection process in clinical trials.
+The authors propose to use Bayesian optimization (BO) to improve the efficiency of the design selection process in clinical trials.
 A set of parameters to be chose for the design optimization based on the power.
@@
 
-We agree that the power is only one measure for choosing a design.
-Therefore, the idea behind our method is to quickly find out which power one can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
-Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
-We added study duration as a possible limitation in our conclusion section and made sure that the limitations of our approach are mentioned more clearly.
-Assuming certain parameters under the null hypothesis is a standard approach.
-However, we see that a more Bayesian approach is certainly interesting, as we already stated in the last paragraph of the discussion.
+> We agree that the power is only one measure for choosing a design.
+> Therefore, the idea behind our method is to quickly find out which power one can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
+> Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
+> We added study duration as a possible limitation in our conclusion section and made sure that the limitations of our approach are mentioned more clearly.
+> Assuming certain parameters under the null hypothesis is a standard approach.
+> However, we see that a more Bayesian approach is certainly interesting, as we already stated in the last paragraph of the discussion.
 
@@
 
-> In page 7 table 1, the effect size scenarios presented seem always assume 2nd stage effect size are higher than stage 1?
+In page 7 table 1, the effect size scenarios presented seem always assume 2nd stage effect size are higher than stage 1?
 The authors should also evaluate the vice visa situation, and consistent effect size case in simulation.
 
-In the table the effects for the second stage (final) are indeed smaller then for the fist stage (early).
-This is a realistic scenario, as often the early outcome is overoptimistic and therefore higher.
+> In the table the effects for the second stage (final) are indeed smaller then for the fist stage (early).
+> This is a realistic scenario, as often the early outcome is overoptimistic and therefore higher.
 
-> In page 8 table 2, the selection of “espilon” and “threshold” used in the arm selection should be based on the clinical meaningful cut, I am not sure how practical that they are part of parameters to be optimized?
+In page 8 table 2, the selection of “espilon” and “threshold” used in the arm selection should be based on the clinical meaningful cut, I am not sure how practical that they are part of parameters to be optimized?
 The authors should also evaluate with different setting.
 
-Given a simulation setting we optimize the parameters eps and thr. This is independent of the application but of course the clinical application can give hints which simulation settings are adequate.
+> Given a simulation setting we optimize the parameters epsilon (eps) and tau (thresh).
+This is independent of the application but of course the clinical application can give hints which simulation settings are adequate.
 
-> In section 5.2, the authors keep the total number of pts fixed, which seem not appropriate in practice, as overall power relying several factors, e.g.
+In section 5.2, the authors keep the total number of pts fixed, which seem not appropriate in practice, as overall power relying several factors, e.g.
 # of pts, the interim selection rule, stage 1 and stage 2 ratio.
@@
 
-If we included the number of patients in the optimization, the optimizer would propose to choose the maximal number of patients, because this gives the maximal performance.
-This is also indicated in our results, as we tried different values of n_total which equals the number of patients. 
-Here, a higher number always resulted in a higher power.
-As mentioned in the paper, the idea is to run this optimization for different sample sizes (denoted by n_total), and then to decide which sample size is required.
-This brings us to the future goal that we mentioned in the discussion: We can include n_total as another parameter and outcome in a multi-criteria optimization setting.
+> If we included the number of patients in the optimization, the optimizer would propose to choose the maximal number of patients, because this gives the maximal performance.
+> This is also indicated in our results, as we tried different values of n_total which equals the number of patients. 
+> Here, a higher number always resulted in a higher power.
+> As mentioned in the paper, the idea is to run this optimization for different sample sizes (denoted by n_total), and then to decide which sample size is required.
+> This brings us to the future goal that we mentioned in the discussion: We can include n_total as another parameter and outcome in a multi-criteria optimization setting.
 
-> In page 10 table 4, the number of simulation iteration is only 1000, which does not seem sufficient for complex optimization methods proposed by authors.
+In page 10 table 4, the number of simulation iteration is only 1000, which does not seem sufficient for complex optimization methods proposed by authors.
 I would suggest at least 5000 should be needed.
 
-The simulation iterations are a specific parameter of the black-box function we optimize.
-The higher this value, the more reliable is the estimator of the power of the design.
-Our optimization is run for 100 iterations which is sufficient to converge as Figure 5 shows.
-Therefore, we do not see the need to run more iterations.
-We added a corresponding statement in section 5.7.
+> The simulation iterations are a specific parameter of the black-box function we optimize.
+> The higher this value, the more reliable is the estimator of the power of the design.
+> Our optimization is run for 100 iterations which is sufficient to converge as Figure 5 shows.
+> Therefore, we do not see the need to run more iterations.
+> We added a corresponding statement in section 5.7.
 
-> Page 11 table 5, the authors present the average run time in hour, not sure how these information is helpful, as this depend on the what computing systems were used in the simulation.
+Page 11 table 5, the authors present the average run time in hour, not sure how these information is helpful, as this depend on the what computing systems were used in the simulation.
 Please either add more information.
 
-Thank you. We added the hardware setup that was used.
+> Thank you. We added the hardware setup that was used.
 
-> Page 13, BO has issue to find optimal value when one parameters closer to the borders of search space.
+Page 13, BO has issue to find optimal value when one parameters closer to the borders of search space.
 This is concerning and confirm the limitation of using too many assumptions for optimization, e.g.
@@
 
-As we mentioned in the final discussion section of the paper, we see this as a topic for a future benchmark to further improve the performance of BO.
-The aim of this paper was to benchmark BO in a basic setting without the prior knowledge that small values of r are of higher importance.
+> As we mentioned in the final discussion section of the paper, we see this as a topic for a future benchmark to further improve the performance of BO.
+> The aim of this paper was to benchmark BO in a basic setting without the prior knowledge that small values of r are of higher importance.
 
-> Given authors used COPD trial as motivating example, I would strongly suggest to add section of case study for the illustration of proposed methods.
+Given authors used COPD trial as motivating example, I would strongly suggest to add section of case study for the illustration of proposed methods.
 
-In fact the effect set "paper" is based on a case study where the effect sets are estimated on a true dataset (Friede et al, 2020). 
-In this paper we do not propose a method for analyzing a real dataset, but a method to select the best design given a known theoretical situation.
+> In fact the effect set "paper" is based on a case study where the effect sets are estimated on a true dataset (Friede et al, 2020). 
+> In this paper we do not propose a method for analyzing a real dataset, but a method to select the best design given a known theoretical situation.
 
@@
 
-> Page 8 line 29,  n_{treat} should be total # of pts, instead of “total number of treatment”.
+Page 8 line 29,  n_{treat} should be total # of pts, instead of “total number of treatment”.
 
-Thanks, we fixed that mistake.
+> Thanks, we fixed that mistake.

### 2021-03-25 generator/main.R

@@ -90,10 +90,11 @@ names(effect_labels) = effect_labels = c("linear", "paper", "paper2", "sigmoid")
 ## ----table_effect_names-------------------------------------------------------
 tmp = lapply(EFFECTS, do.call, what = rbind)
 tmp = lapply(tmp, function(x) cbind.data.frame(stage = rownames(x), x))
+tmp = lapply(tmp, function(x) {colnames(x) = c("stage", seq_len(ncol(x)-1)-1); x}) # colnames start at 0
 tmp = rbindlist(tmp, idcol = TRUE)
 setnames(tmp, ".id", "Effect Set")
 setnames(tmp, "stage", "Stage")
-kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation in the first (early) and second (final) stage") %>% 
+kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation in the first (early) and second (final) stage", label = "table_effect_names") %>% 
   collapse_rows(columns = 1:2, latex_hline = "major") %>% 
   add_header_above(c(" " = 2, "Treatment" = 5)) %>% 
   kable_styling(position = "center") %>% 
@@ -458,7 +459,7 @@ df = res_ave[,.SD[order(-mean_y)[1:3]], by = c("effect", "n_cases")][,.(effect, 
 res_ave_mbo = get_res_mbo()[, list(mean_y = mean(y)), by = algo.par.names.meta]
 df = rbind(df, res_ave_mbo[, .(effect, n_cases, mean_y, select = algorithm)], fill = TRUE)
 setkey(df, effect, n_cases, mean_y)
-knitr::kable(df, booktabs = TRUE, caption = "Best configurations per ncases and effects", longtable = FALSE, digits = 3) %>% # linesep = c(rep("",4), "\\addlinespace") 
+knitr::kable(df, booktabs = TRUE, caption = "Best configurations per ncases and effects", longtable = FALSE, digits = 3, label = "table_best") %>% # linesep = c(rep("",4), "\\addlinespace") 
   kable_styling(position = "center", font_size = FNT_SMALL) %>% 
   collapse_rows(1:2, latex_hline = "custom", custom_latex_hline = 1:2) %>% 
   kable_to_text("table_best")
@@ -478,7 +479,8 @@ kable(tmp2,
   booktabs = TRUE, 
   col.names = c("", as.character(tmp[algorithm == "mbo"]$n_cases), "\\emph{evals}"),
[…]

### 2021-04-08 response_to_the_reviewers.md
small fixes

@@
 ---
-title: Test
-author: Author Name
+title: Response to the Reviewers
+author: Jakob Richter
 header-includes: |
@@
 > As in the original paper (Friede et. al. 2020) in which our motivating example was introduced, we count the rejections of any of the hypotheses 3 or 4 for calculating the power.
-> We now clearly state this in the paper and also clarify that there are different options and it is up to the user to determine the method of how the power is calculated.
+> We now clearly state this in the paper and also clarify that there are different options, and it is up to the user to determine the method of how the power is calculated.
 
@@
 
-> Thanks, for spotting that mistake. Indeed we use 20 stochastic repetitions.
+> Thanks, for spotting that mistake. Indeed, we use 20 stochastic repetitions.
 
@@
 
-> In the table the effects for the second stage (final) are indeed smaller then for the fist stage (early).
+> In the table the effects for the second stage (final) are indeed smaller than for the fist stage (early).
 > This is a realistic scenario, as often the early outcome is overoptimistic and therefore higher.

### 2021-04-08 main.tex
diff generation and review
small fixes

@@
 This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
-Also, BO is well suited for multi-criteria optimization.
+Also, BO is well suited for multi-objective optimization.
 Different adaptions of BO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
 Many of them are implemented in \texttt{mlrMBO} as well.
-However, in this work we restrict ourselves to single-criterion optimization.
+However, in this work we restrict ourselves to single-objective optimization.
+
+Independently of this work, \citet{wilson_efficient_2021} successfully applied multi-objective BO to minimize the number of participants and the number of clusters in a cluster randomized controlled trial under the restriction that the power is above a given threshold.
 
@@
 In the sections~\ref{sec:adaptive_seamless_designs} and~\ref{sec:bayesian_optimization}, we explain the background on adaptive seamless designs (asd) and Bayesian optimization (BO), respectively.
-In section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of BO for optimizing trial designs is demonstrated.
+In Section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of BO for optimizing trial designs is demonstrated.
 In the last Section~\ref{sec:discussion}, we discuss promising extensions as well as limitations of the approach.
@@
 
-% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, genau da gucken wir drauf!
-
 \begin{table}[h]
@@
 The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
-Runtimes were measured on a single core of a compute node with 2 Intel Xeon E5-2697v2 (2.70 GHz) with 12 Cores each and 512 GB of RAM in an non-exclusive usage of the node.
+Runtimes were measured on a single core of a compute node with 2 Intel Xeon E5-2697v2 (2.70 GHz) with 12 Cores each and 512 GB of RAM in a non-exclusive usage of the node.
 \emph{BO} is approximately 20 times faster.
@@
 Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO}, with only a small advantage for \emph{Grid Small}.
-Despite the 10 more evaluations allowed, in the majority of cases \emph{BO} is superior to \emph{Grid Small}, see~Figure~\ref{fig:plot_boxplot_valid_y}.
+Despite the 10 more evaluations allowed, in the majority of cases \emph{BO} is superior to \emph{Grid Small}, see Figure~\ref{fig:plot_boxplot_valid_y}.
 The advantage is particularly apparent for all scenarios with effect set \emph{effect: linear} and for scenario \emph{effect: paper2}, $n_{\text{total}} = 500$.
@@
 In the second case, the absolute differences are negligible and all methods lead to a high power.
-Also for this case, we will later show that the optimal configuration is close to the border of the search space.
+Also, for this case we will later show that the optimal configuration is close to the border of the search space.
 
@@
 
-%% FIXME: Do we really take in this Assurance?
-% Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
-% could be also applied for subgroup selection?
-% https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8797
-
 In this paper we analyzed if BO can quickly find an adaptive seamless design with high power, more efficiently than an exhaustive grid search.
@@
 Here, a possible improvement can be to apply a log transformation to such parameters. % in the future.
-Further, instead of keeping the total sample size constant, we could apply multi-criteria optimization and include this number as an additional outcome.
-This multi-criteria approach would result in a set of points (a Pareto front) that minimizes the total sample size and at the same time maximize the statistical power of the trial design.
+Further, instead of keeping the total sample size constant, we could apply multi-objective optimization and include this number as an additional outcome.
+This multi-objective approach would result in a set of points (a Pareto front) that minimizes the total sample size and at the same time maximize the statistical power of the trial design.
 The user could then see for which number of trials the power is above a certain threshold (e.g. $95\%$).
@@
 Such a procedure can be guided by statistical testing and is already common in machine learning under the terms \emph{racing} or \emph{multi-fidelity optimization}.
+
+% Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln
+% could be also applied for subgroup selection?
+% https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8797
 

### 2021-04-12 response_to_the_reviewers.md
TF response

@@
 
-> As in the original paper (Friede et. al. 2020) in which our motivating example was introduced, we count the rejections of any of the hypotheses 3 or 4 for calculating the power.
+> As in Friede et. al. (2020), where our motivating example was introduced, we count the rejections of any of the hypotheses 3 or 4 for calculating the power.
 > We now clearly state this in the paper and also clarify that there are different options, and it is up to the user to determine the method of how the power is calculated.
@@
 > We improved our wording, so that the number of treatment arms cannot longer be confused with the individual number of treatments of each patient.
-> Therefore, the total sample size (formerly “total number of treatments”) is denoted as n_total (formerly n_treat)
-> If the patients are different in both stages (i.e. each patient is treated only once) the total number of patients would be the same as the total sample size.
+> Therefore, the total sample size (formerly “total number of treatments”) is denoted as n_total (formerly n_treat).
+> Since the patients are different in both stages (i.e. each patient is treated only once) the total number of patients is the same as the total sample size.
 > However, we only use the phrase "total sample size" to avoid confusions.
@@
 
-> There is no normality assumption for the single outcome data but for the test statistics Z_i,j a multivariate normal distribution is assumed which will hold approximately if the sample size is sufficiently large.
+> There is no normality assumption for the individual outcome data but for the test statistics Z_i,j a multivariate normal distribution is assumed which will hold approximately if the sample size is sufficiently large.
 > We clarified that the test statistics are multivariate normal distributed at least approximately at the very end of chapter 3.
@@
 > An advantage is not notable if the median of both methods lies within the box of the other method.
-> Therefore, we don't see an advantage for the configuration (effect: paper, n_treat = 1000).
+> Therefore, we do not see a notable advantage of Grid Small for the configuration (effect: paper, n_treat = 1000).
 
@@
 > Therefore no more treatments can be evaluated in stage 2.
-> We improved the wording to make this clear.
+> We revised the wording to make this clearer.
 
@@
 > We agree that the power is only one measure for choosing a design.
-> Therefore, the idea behind our method is to quickly find out which power one can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
+> Therefore, the idea behind our method is to quickly asses which power one can achieve given the other fixed factors such as total sample size (and therefore the number of patients).
 > Quickly obtaining the best possible performance for a specific design choice will also help decision makers to decide how to choose the design.
-> We added study duration as a possible limitation in our conclusion section and made sure that the limitations of our approach are mentioned more clearly.
+> Furthermore, we added study duration as a possible alternative performance measure. 
+> This is now discussed as a limitation in our conclusion section.
 > Assuming certain parameters under the null hypothesis is a standard approach.
@@
 > Our optimization is run for 100 iterations which is sufficient to converge as Figure 5 shows.
-> Therefore, we do not see the need to run more iterations.
+> Therefore, we do not see the need to run a larger number of iterations.
 > We added a corresponding statement in section 5.7.
@@
 > As we mentioned in the final discussion section of the paper, we see this as a topic for a future benchmark to further improve the performance of BO.
-> The aim of this paper was to benchmark BO in a basic setting without the prior knowledge that small values of r are of higher importance.
+> The aim of this paper was to benchmark BO in a basic setting without the prior knowledge that small values of **r** are of higher importance.
 
@@
 
-> In fact the effect set "paper" is based on a case study where the effect sets are estimated on a true dataset (Friede et al, 2020). 
+> In fact the effect set "paper" is based on a case study where the effect sets are estimated on a real dataset (Friede et al, 2020). 
 > In this paper we do not propose a method for analyzing a real dataset, but a method to select the best design given a known theoretical situation.

### 2021-04-12 main.tex
tf notes

@@
 However, in this work we restrict ourselves to single-objective optimization. 
-
-Independently of this work, \citet{wilson_efficient_2021} successfully applied multi-objective BO to minimize the number of participants and the number of clusters in a cluster randomized controlled trial under the restriction that the power is above a given threshold.
+Independently of this work, for instance, \citet{wilson_efficient_2021} successfully applied multi-objective BO to minimize the number of participants and the number of clusters in a cluster randomized controlled trial under the restriction that the power is above a given threshold.
 
@@
 
-One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd} package~\citep{parsons_software_2011} and described in Section~\ref{sec:adaptive_seamless_designs}.
+One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd} package~\citep{parsons_package_2012} and described in Section~\ref{sec:adaptive_seamless_designs}.
 %(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, )  bauer_combining_1999
@@
 The evaluation of some performance measures requires Monte Carlo simulations, since no closed form solutions are available. 
-The R-package \texttt{asd} by~\citet{parsons_software_2011} can be used for such simulations.
+The R-package \texttt{asd} by~\citet{parsons_package_2012} can be used for such simulations.
 As it generates test statistics, following multivariate normal distributions, it is applicable to a wide range of outcomes, at least approximately, and generally more efficient in comparison to simulating individual participant data.
@@
 
-The expected improvement, as well as other acquisition functions, are derived under the assumption that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
+The expected improvement as well as other acquisition functions are derived under the assumption that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
 The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters $\hat{\mu}$ and $s^2(\theta)$ under the assumption that $f$ is a realization of a Gaussian process.
@@
 The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.
-Depending on the application the power can be calculated as the rate of rightfully rejecting any, all, or a subset of the elementary hypotheses~\citet{senn_power_2007}.
+Depending on the application the power can be calculated as the proportion of correctly rejecting any, all, or a subset of the elementary hypotheses~\citep{senn_power_2007}.
 
@@
 We define \textbf{Grid Small} to be a subset of this grid with $l=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$.
-This gives a number of evaluations that can be easily carried out in practice.
+This results in a number of evaluations that can be easily carried out in practice.
 
@@
 As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
-We start BO with an initial design of $16$ randomly sampled points, following an established rule-of-thumb of $4$ points per dimension, which is also well in line with the recommendation of using around $10\%$ of the budget for the initial design~\cite{bossek_initial_2020}.
+We start BO with an initial design of $16$ randomly sampled points, following an established rule-of-thumb of $4$ points per dimension, which is also well in line with the recommendation of using around $10\%$ of the budget for the initial design~\citep{bossek_initial_2020}.
 We allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.
@@
 For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
-For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
+For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_package_2012} is used, with parameter values given in Table~\ref{tab:par_implement}.
 %The vector of treatment numbers for determining power \texttt{ptest} counts rejections of treat3 or 4 hypotheses. %???
 The vector of treatment numbers \texttt{ptest} determines the hypotheses for which rejections are counted towards the power.
-Here, the rejection of one or both of the hypotheses that treatment 3 or 4 has no significant effect against the control determines the power, as in~\citet{friede_adaptive_2020}.
+Here, the rejection of one or both of the hypotheses that treatment 3 or 4 has no effect against the control determines the power, as in~\citet{friede_adaptive_2020}.
 
@@
 Figure~\ref{fig:plot_opt_path_5000} shows that indeed the lower noise level leads to more stable simulation runs, meaning that single optimization runs lead to results closer to the mean result.
-Both curves indicate, that more iterations of \emph{BO} would not yield substantially better outcomes as the progress is stagnating after 80 iterations.
+Both curves indicate, that more iterations of \emph{BO} would not yield substantially better outcomes as no substantial improvements are observed after 80 iterations
 It has to be noted that this plot reports the best observed outcome so far and not the one that is chosen as final result.
@@
 In clinical trials, determining the test methods and sample sizes is a key step in the planning process.
-One main goal is to maximize the test power, given a set of treatments with corresponding effect sizes and a fixed number of samples.
+For our optimization we chose to maximize the test power, given a set of treatments with corresponding effect sizes and a fixed number of samples.
+%FIXME: TF: Sollten wir hier nochmal explizit auf alternative objectives wie study duration eingehen?
 If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
@@
 Here, a possible improvement can be to apply a log transformation to such parameters. % in the future.
-Further, instead of keeping the total sample size constant, we could apply multi-objective optimization and include this number as an additional outcome.
-This multi-objective approach would result in a set of points (a Pareto front) that minimizes the total sample size and at the same time maximize the statistical power of the trial design.
-The user could then see for which number of trials the power is above a certain threshold (e.g. $95\%$).
+We chose to maximize the power for a fixed number of samples.
+Other aspects like study duration or the cost of a trial in general could be of interest either as an additional constraint or an additional outcome.
+Also the number of samples can be included as an additional outcome instead of being fixed.
+Multiple outcomes that are subject to optimization give a (possibly restricted) multi-objective optimization problem.
+The result of which is a set of points (a Pareto front) that maximizes the statistical power and e.g. minimizes the total sample size at the same time.
+The user could then see for which sample size the power is above a certain threshold (e.g. $95\%$).
 Another way to speed up the optimization would be to iteratively increase the number of simulation rounds of the trial design evaluation. 

### 2021-05-19 arxiv/main.tex
arxiv init

--- 2021-04-12 main.tex
@@
-\documentclass[bimj,fleqn]{w-art}
-\usepackage{times}
-\usepackage{w-thm}
+\documentclass{article}
+
+\usepackage{arxiv}
+
 \usepackage[authoryear]{natbib}
-\setlength{\bibsep}{2pt}
-\setlength{\bibhang}{2em}
-\newcommand{\J}{J\"{o}reskog}
-\newcommand{\So}{S\"{o}rbom}
-\newcommand{\bcx}{{\bf X}}
-\newcommand{\bcy}{{\bf Y}}
-\newcommand{\bcz}{{\bf Z}}
-\newcommand{\bcu}{{\bf U}}
-\newcommand{\bcv}{{\bf V}}
-\newcommand{\bcw}{{\bf W}}
-\newcommand{\bci}{{\bf I}}
-\newcommand{\bch}{{\bf H}}
-\newcommand{\bcb}{{\bf B}}
-\newcommand{\bcr}{{\bf R}}
-\newcommand{\bcm}{{\bf M}}
-\newcommand{\bcf}{{\bf F}}
-\newcommand{\bcg}{{\bf G}}
-\newcommand{\bcs}{{\bf S}}
-\newcommand{\bca}{{\bf A}}
-\newcommand{\bcd}{{\bf D}}
-\newcommand{\bcc}{{\bf C}}
-\newcommand{\bce}{{\bf E}}
-\newcommand{\ba}{{\bf a}}
-\newcommand{\bb}{{\bf b}}
-\newcommand{\bc}{{\bf c}}
-\newcommand{\bd}{{\bf d}}
-\newcommand{\bx}{{\bf x}}
-\newcommand{\by}{{\bf y}}
-\newcommand{\bz}{{\bf z}}
-\newcommand{\bu}{{\bf u}}
-\newcommand{\bv}{{\bf v}}
-\newcommand{\bh}{{\bf h}}
-\newcommand{\bl}{{\bf l}}
-\newcommand{\be}{{\bf e}}
-\newcommand{\br}{{\bf r}}
-\newcommand{\bw}{{\bf w}}
-\newcommand{\de}{\stackrel{D}{=}}
-\newcommand{\bt}{\bigtriangleup}
-\newcommand{\bfequiv}{\mbox{\boldmath $\equiv$}}
-\newcommand{\bmu}{\mbox{\boldmath $\mu$}}
-\newcommand{\bnu}{\mbox{\boldmath $\nu$}}
-\newcommand{\bxi}{\mbox{\boldmath $\xi$}}
-\newcommand{\btau}{\mbox{\boldmath $\tau$}}
-\newcommand{\bgamma}{\mbox{\boldmath $\Gamma$}}
-\newcommand{\bphi}{\mbox{\boldmath $\Phi$}}
-\newcommand{\bfphi}{\mbox{\boldmath $\varphi$}}
-\newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
-\newcommand{\bpi}{\mbox{\boldmath $\Pi$}}
-\newcommand{\bequiv}{\mbox{\boldmath $\equiv$}}
-\newcommand{\bvarepsilon}{\mbox{\boldmath $\varepsilon$}}
-\newcommand{\btriangle}{\mbox{\boldmath $\triangle$}}
-\newcommand{\bdelta}{\mbox{\boldmath $\Delta$}}
-\newcommand{\beps}{\mbox{\boldmath $\epsilon$}}
-\newcommand{\btheta}{\mbox{\boldmath $\theta$}}
-\newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
-\newcommand{\bsphi}{\mbox{\boldmath $\varphi$}}
-\newcommand{\bsig}{\mbox{\boldmath $\sigma$}}
-\newcommand{\bfpsi}{\mbox{\boldmath $\psi$}}
-\newcommand{\bfdelta}{\mbox{\boldmath $\delta$}}
-\newcommand{\bsigma}{{\bf \Sigma}}
-\newcommand{\bzero}{{\bf 0}}
-\newcommand{\bpsi}{\mbox{\boldmath $\Psi$}}
-\newcommand{\bep}{\mbox{\boldmath $\epsilon$}}
-\newcommand{\bomega}{\mbox{\boldmath $\Omega$}}
-\newcommand{\bfomega}{\mbox{\boldmath $\omega$}}
-\newcommand{\blambda}{\mbox{\boldmath $\Lambda$}}
-\newcommand{\bflambda}{\mbox{\boldmath $\lambda$}}
-\newcommand{\bfsigma}{\mbox{\boldmath $\sigma$}}
-\newcommand{\bfpi}{{\mbox{\boldmath $\pi$}}}
-\newcommand{\bupsilon}{\mbox{\boldmath $\upsilon$}}
-\newcommand{\obs}{{\rm obs}}
-\newcommand{\mis}{{\rm mis}}
-\theoremstyle{plain}
-\newtheorem{criterion}{Criterion}
-\theoremstyle{definition}
-\newtheorem{condition}[theorem]{Condition}
-\usepackage[]{graphicx}
-\chardef\bslash=`\\ % p. 424, TeXbook
-\newcommand{\ntt}{\normalfont\ttfamily}
-\newcommand{\cn}[1]{{\protect\ntt\bslash#1}}
-\newcommand{\pkg}[1]{{\protect\ntt#1}}
-\let\fn\pkg
-\let\env\pkg
-\let\opt\pkg
-\hfuzz1pc % Don't bother to report overfull boxes if overage is < 1pc
-\newcommand{\envert}[1]{\left\lvert#1\right\rvert}
-\let\abs=\envert
+\usepackage{doi}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{caption}
 
@@
 
+% \keywords{Adaptive Seamless Designs; Bayesian Optimization; Clinical Trials; Treatment Selection\\\\
+
+\title{Improving Adaptive Seamless Designs through Bayesian optimization}
+\renewcommand{\shorttitle}{Bayesian optimization of Adaptive Seamless Designs}
+
+\hypersetup{
+pdftitle={Improving Adaptive Seamless Designs through Bayesian optimization},
+pdfauthor={Jakob Richter, Tim Friede, Jörg Rahnenführer},
+pdfkeywords={Adaptive Seamless Designs, Bayesian Optimization, Clinical Trials, Treatment Selection},
+}
+
+\author{
+  \href{https://orcid.org/0000-0003-4481-5554}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Jakob Richter}  \\
+    Fakultät Statistik \\
+    Technische Universität Dortmund \\
+    \texttt{richter@statistik.tu-dortmund.de}
+  \And
+  \href{https://orcid.org/0000-0001-5347-7441}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Tim Friede} \\
+    Institut für Medizinische Statistik \\
+    Universitätsmedizin Göttingen, and Deutsches Zentrum für Herz-Kreislauf-Forschung (DZHK), Standort Göttingen \\
+    \texttt{tim.friede@med.uni-goettingen.de}
+  \And
+  \href{https://orcid.org/0000-0002-8947-440X}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Jörg Rahnenführer} \\
+    Fakultät Statistik \\
+    Technische Universität Dortmund \\
+    \texttt{rahnenfuehrer@statistik.tu-dortmund.de}
+}
+
 \begin{document}
-
-%\DOIsuffix{bimj.DOIsuffix}
-\DOIsuffix{bimj.200100000}
-\Volume{52}
-\Issue{61}
-\Year{2020}
-\pagespan{1}{}
-\keywords{Adaptive Seamless Designs; Bayesian Optimization; Clinical Trials; Treatment Selection\\\\ %Up to five keywords are allowed and should be given in alphabetical order.
-Please capitalize the key
-%\noindent \hspace*{-4pc}\\
-%\hspace*{-4pc} 
-%{\small\it words)}\\[1pc]
-%\noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not applicable)
-} %%% semicolon and fullpoint added here for keyword style
-
-\title[Bayesian optimization of Adaptive Seamless Designs]{Improving Adaptive Seamless Designs through Bayesian optimization} %(please use sentence case, e.g. Score tests for exploring complex models)
-%% Information for the first author.
-\author[Jakob Richter {\it{et al.}}]{Jakob Richter\footnote{Corresponding author: {\sf{e-mail: richter@statistik.tu-dortmund.de}}}\inst{,1}} 
-\address[\inst{1}]{Fakultät Statistik, Technische Universität Dortmund, 44221 Dortmund}
-%%%%    Information for the second author
-\author[dd]{Tim Friede\inst{2}}
-\address[\inst{2}]{Institut für Medizinische Statistik, Universitätsmedizin Göttingen, and Deutsches Zentrum für Herz-Kreislauf-Forschung (DZHK), Standort Göttingen, 37073 Göttingen}
-%%%%    Information for the third author
-\author[]{Jörg Rahnenführer\inst{1}} %please provide full author names. Middle names should be indicated by initials only, i.e. Henry J. James
-%%%%    \dedicatory{This is a dedicatory.}
-\Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}
-
-\makeatletter
-\let\newtitle\@title
-\makeatother
+\maketitle
 
@@
 
-\begin{acknowledgement}
+\section*{Acknowledgments}
 This work was partly supported by Deutsche Forschungsgemeinschaft (DFG) within the Collaborative Research Center SFB 876, A3.
-\end{acknowledgement}
-\vspace*{1pc}
-
-\noindent {\bf{Conflict of Interest}}
-
-\noindent {\it{The authors have declared no conflict of interest.}} %(or please state any conflicts of interest)
 
@@
 
-\bibliographystyle{biometrical}
+\bibliographystyle{unsrtnat}
 \bibliography{literature,literature_zotero}

### 2021-06-02 generator/main.R
appendix

@@ -60,10 +60,12 @@ EFFECTS = list(
   # saturation = {x = c(0,5,7,7.8,8)/10; list(early = x, final = x/4)},
   sigmoid = {x = c(0,1,2,7,8)/10; list(early = x, final = x/4)}, #yes
   # paper_mod = list(early = c(0,0.68,0.82,0.93,0.93), final = c(0,0.13,0.17,0.23,0.20)), 
-  paper2 = list(early = c(0,0.68,0.82,0.95,0.91), final = 2*c(0,0.13,0.17,0.23,0.20)) #yes
+  paper2 = list(early = c(0,0.68,0.82,0.95,0.91), final = 2*c(0,0.13,0.17,0.23,0.20)), #yes
   # paper_mod2 = list(early = c(0,0.68,0.82,0.93,0.93), final = 2*c(0,0.13,0.17,0.23,0.20))
+  paper_rev = list(early = c(0,0.13,0.17,0.23,0.20), final = c(0,0.68,0.82,0.95,0.91))
 )
 effect_names = setNames(names(EFFECTS), names(EFFECTS)) #to be populated with better names
+main_effect_names = names(EFFECTS)[1:4]
 
 select_labels = c("0" = "all", "1" = "1 best", "2" = "2 best", "3" = "3 best", "4" = "epsilon rule", "5" = "random", "6" = "threshold rule")
 select_labels_colors = #http://paletton.com/#uid=74Z140kqdu7ghF3lowyu1ppwGk7
@@ -88,7 +90,7 @@ names(effect_labels) = effect_labels = c("linear", "paper", "paper2", "sigmoid")
 
 
 ## ----table_effect_names-------------------------------------------------------
-tmp = lapply(EFFECTS, do.call, what = rbind)
+tmp = lapply(EFFECTS[main_effect_names], do.call, what = rbind)
 tmp = lapply(tmp, function(x) cbind.data.frame(stage = rownames(x), x))
 tmp = lapply(tmp, function(x) {colnames(x) = c("stage", seq_len(ncol(x)-1)-1); x}) # colnames start at 0
 tmp = rbindlist(tmp, idcol = TRUE)
@@ -101,7 +103,7 @@ kable(tmp, booktabs = TRUE, caption = "Effect sizes used for simulation in the f
   kable_to_text("table_effect_names")
 
 ## ----read_data----------------------------------------------------------------
-read_results_uncached = function(select_labels, case = "default") {
+read_results_uncached = function(select_labels, case = "default", effect_names = main_effect_names) {
[…]

### 2021-06-07 generator/main.R
cleanup boxplots

@@ -468,15 +468,15 @@ plot_wrapper(name = "plot_boxplot_valid_y_5000", fig.height = 1.6 * FIG_HEIGHT *
 ## ----plot_appendix_boxplot_paper_rev-----------------------------------------------------
 plot_wrapper(name = "plot_appendix_boxplot_paper_rev", fig.height = FIG_HEIGHT * 0.7, fig.width = FIG_WIDTH, expr = {
   tmp = create_boxplot_df(case = "default", effect_names = c("paper", "paper_rev"))
-  g = ggplot(tmp, aes(x = as.factor(nsim), y = y_valid, color = algorithm, fill = algorithm))
+  g = ggplot(tmp, aes(x = algorithm, y = y_valid, color = algorithm, fill = algorithm))
   g = g + facet_wrap(effect~n_cases, scales = "free_y", ncol = 6, labeller = label_bquote({atop(.(effect), n[total]==.(n_cases))}))
   darker_colors = colorspace::darken(algorithm_labels_color, amount = 0.6)
   names(darker_colors) = names(algorithm_labels_color)
   g = g + scale_fill_manual(labels = algorithm_labels, values = algorithm_labels_color) + scale_color_manual(labels = algorithm_labels, values = darker_colors)
-  g = g + theme(legend.position = "bottom")
+  g = g + scale_x_discrete(labels = algorithm_labels)
   g = g + geom_boxplot()
-  g = g + labs(x = expression(n[sim]), y = expression(y[valid]), color = NULL, fill = NULL)
-  g = g + guides(fill=guide_legend(nrow=2,byrow=FALSE))
+  g = g + labs(x = NULL, y = expression(y[valid]), color = NULL, fill = NULL)
+  g = g + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
   g
 })
 
@@ -517,16 +517,19 @@ plot_wrapper(name = "plot_opt_path_5000", fig.height = 1.6 * FIG_HEIGHT * 0.35, 
 ## ----table_best---------------------------------------------------------------
 #best of grid
 #FIXME: calculate time of complete grid, devide through 10
-res_ave = get_res_eval_average()
[…]

### 2021-06-07 main.tex
appendix
cleanup boxplots
opt eps and thresh plots

@@
 
-% %%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
-% \clearpage
-% \begin{center}
-% \textbf{\large Supplemental Materials: \newtitle}
-% \end{center}
-% \FloatBarrier
-% %%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
-% %%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
-% \setcounter{equation}{0}
-% \setcounter{figure}{0}
-% \setcounter{table}{0}
-% \setcounter{page}{1}
-% \makeatletter
-% \renewcommand{\theequation}{S\arabic{equation}}
-% \renewcommand{\thefigure}{S\arabic{figure}}
-% \renewcommand{\thetable}{S\arabic{table}}
-% \renewcommand{\bibnumfmt}[1]{[S#1]}
-% %%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
-
-% \input{generated/tables/table_best.tex}
-
-% \begin{figure}[htb]
-% \centering
-% \includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
-% \caption{%
-%   Optimization curves of each optimization run and the mean of all runs drawn as a black line.
-%   }
-% \label{fig:plot_opt_path} 
-% \end{figure}
+%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
+\clearpage
+\begin{center}
+\textbf{\large Supplemental Materials: \newtitle}
+\end{center}
+\FloatBarrier
+%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
+%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
+\setcounter{equation}{0}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\setcounter{page}{1}
+\makeatletter
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\bibnumfmt}[1]{[S#1]}
+%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
+\caption{%
+  Optimization progress for \emph{BO}: Best outcome found so far for each iteration, in comparison to the best outcome found in the grid search, for all effect sets and all values of $n_{\text{total}}$.
+}
+\label{fig:plot_opt_path} 
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.5\linewidth]{generated/figures/plot_appendix_opt_epsilon.pdf}%
+\includegraphics[width=0.5\linewidth]{generated/figures/plot_appendix_opt_thresh.pdf}
+\caption{%
+  This figure shows the location of the optimal values of $\varepsilon$ and $\tau$ for their respective rules obtained through the \emph{Grid} search, averaged across the 20 stochastic replications.
+  It indicates that the search space for $r$ and $\varepsilon$ or $\tau$ is big enough but that each should be optimized on a $log$-scale.
+  The color indicates the value of $y$, with \emph{red} beeing good.
+  $y$-values are scaled for each single plot and cannot be compared across experiments.
+  The best points are indicated through a white dot.
+}
+\label{fig:plot_opt_eps_thresh} 
+\end{figure}
 

### 2021-06-07 generator/main.R

@@ -22,13 +22,15 @@ FIG_HEIGHT = 4.5
 FNT_SMALL = 6
 CACHE = cache_filesystem(path = "memoise")
 
-plot_wrapper_uncached = function(name, fig.height = FIG_HEIGHT, fig.width = FIG_WIDTH, expr) {
+plot_wrapper_uncached = function(name, fig.height = FIG_HEIGHT, fig.width = FIG_WIDTH, expr, crop = TRUE) {
   path = paste0("../generated/figures/", name, ".pdf")
   cairo_pdf(filename = path, width = fig.width, height = fig.height, onefile = TRUE)
   print(eval(expr))
   dev.off()
   #browser()
-  knitr::plot_crop(path)
+  if (crop) {
+    knitr::plot_crop(path)
+  }
 }
 plot_wrapper = memoise(plot_wrapper_uncached, cache = CACHE)
 
@@ -377,11 +379,11 @@ plot_wrapper(name = "plot_appendix_best_x_paper_rev", fig.height = FIG_HEIGHT, e
 })
 
 
-## ----plot_opt_path, fig.height = 1.6 * FIG_HEIGHT-----------------------------
+## ---- plot_opt_path -----------------------------
 
 # calculate y perf of mbo runs
 plot_wrapper(name = "plot_opt_path", fig.height = 1.6 * FIG_HEIGHT, expr = {
-  df = get_res_mbo()
+  df = get_res_mbo(effect_names = names(EFFECTS))
   common_names = intersect(colnames(df), colnames(df$opt.path[[1]]))
   data.table::setnames(df, common_names, paste0("opt.", common_names))
   df = tidyr::unnest(df, "opt.path")
@@ -389,7 +391,7 @@ plot_wrapper(name = "plot_opt_path", fig.height = 1.6 * FIG_HEIGHT, expr = {
   df = df[prop.type != "final_eval", cummax_y := cummax(y), by = c(algo.par.names.meta, "repl")]
   
   # calculate theoretical best y from grid
-  res_ave = get_res_eval_average()
+  res_ave = get_res_eval_average(effect_names = names(EFFECTS))
   df_best = res_ave[,.SD[order(-mean_y)[1]], by = c("effect", "n_cases", "algorithm")][,.(effect, n_cases, algorithm, best_y = mean_y)]
[…]

### 2021-07-11 main.tex
appendix

@@
 \centering
+\includegraphics[width=\linewidth]{generated/figures/plot_appendix_boxplot_paper_rev.pdf}
+\caption{%
+    Box plots that compares the performance (power) values on the effect set \emph{paper} and \emph{paper\_rev}.
+    \emph{paper\_rev} is an additonal scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
+  In contrast to \emph{paper}, the effect sizes are smaller for the \emph{early} stage.
+}
+\label{fig:plot_appendix_boxplot_paper_rev} 
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\linewidth]{generated/figures/plot_appendix_allbest_paper_rev.pdf}
+\caption{%
+  All evaluated configurations and their outcomes on the \emph{Grid} compared for the effect set \emph{paper} and \emph{paper\_rev}.
+  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value with highest power ($y$) is displayed.
+}
+\label{fig:plot_appendix_allbest_paper_rev} 
+\end{figure}
+
+\begin{figure}[htb]
+\centering
 \includegraphics[width=\linewidth]{generated/figures/plot_opt_path.pdf}
@@
 }
-\label{fig:plot_opt_eps_thresh} 
+\label{fig:plot_appendix_opt_episolon_thresh} 
 \end{figure}

### 2021-07-11 main.tex
lazy fixes

@@
 These sets reflect different realistic situations.
+A furthet effect set, where the numbers between the early and the final stage are flipped for \emph{paper} is presented in the Appendix.
 
 The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
+Choises for the boundaries of $\epsilon$ and $\tau$ are obtained from preliminary studies and based on the fact that these are thresholds to the absolute ammount of normally distributed random variables.
+Also Figure~\ref{fig:plot_appendix_opt_episolon_thresh} in the Appendix shows that the optimum is within the given boundaries.
 The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
@@
 %In addition to results in other research fields, our results underline that BO is a powerful method for finding optimal configurations for problems that fit the black-box definition.
-Therefore, the application of BO to clinical trial design selection in general seems promising. 
+For our optimization we chose wide intervals for the parameters of the selection rules (i.e. epsilon and threshold rule) to show that BO is generally applicable.
+In practical applications, however, these values might sometimes be based on clinical rather than statistical considerations or the range of these values might be restricted by such considerations.
+In general, the application of BO to clinical trial design selection in general seems promising.
+Additional restrictions on the search space decrease the its size and therefore promise even more reliable results.
 The same concept could be applied for subgroup selection in adaptive enrichment designs, which make use of interim data in selecting the target population for the remainder of the trial~\citep{burnett_adaptive_2020}.
@@
     \emph{paper\_rev} is an additonal scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
-  In contrast to \emph{paper}, the effect sizes are smaller for the \emph{early} stage.
+  In contrast to \emph{paper}, for \emph{paper\_rev} the effect sizes are smaller for the \emph{early} stage.
 }

### 2021-07-11 generator/main.R

@@ -597,7 +597,7 @@ plot_wrapper(name = "plot_appendix_opt_epsilon", fig.height = 1.5 * FIG_HEIGHT, 
   tmp = tmp[select == "epsilon rule",]
   tmp[, mean_y := BBmisc::normalize(x = mean_y, method = "range"), by = .(effect, n_cases)]
   best = tmp[, .SD[mean_y == max(mean_y)], by = .(effect, n_cases)]
-  ggplot(tmp, aes(x = stage_ratio, y = epsilon)) + geom_raster(aes(fill = mean_y)) + facet_grid(effect~n_cases) + geom_point(data = best, color = "white", size = 0.7) + scale_fill_continuous(guide = FALSE) + scale_fill_gradientn(colours = c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598", "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F", "#9E0142"), guide = FALSE) + labs(title = "Optimal ϵ values for epsilon rule", y = "ϵ") + theme(axis.text = element_text(size = 7)) + scale_x_continuous(labels = scales::label_number(drop0trailing = TRUE))
+  ggplot(tmp, aes(x = stage_ratio, y = epsilon)) + geom_raster(aes(fill = mean_y)) + facet_grid(effect~n_cases) + geom_point(data = best, color = "white", size = 0.7) + scale_fill_continuous(guide = FALSE) + scale_fill_gradientn(colours = c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598", "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F", "#9E0142"), guide = FALSE) + labs(title = "Optimal ϵ values for epsilon rule", x = "r", y = "ϵ") + theme(axis.text = element_text(size = 7)) + scale_x_continuous(labels = scales::label_number(drop0trailing = TRUE))
 })
 
 # --- optimal tau values ------

### 2021-07-11 main.tex
spelling -.-

@@
 These sets reflect different realistic situations.
-A furthet effect set, where the numbers between the early and the final stage are flipped for \emph{paper} is presented in the Appendix.
+A further effect set, where the numbers between the early and the final stage are flipped for \emph{paper} is presented in the Appendix.
 
 The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
-Choises for the boundaries of $\epsilon$ and $\tau$ are obtained from preliminary studies and based on the fact that these are thresholds to the absolute ammount of normally distributed random variables.
+Choices for the boundaries of $\epsilon$ and $\tau$ are obtained from preliminary studies and based on the fact that these are thresholds to the absolute amount of normally distributed random variables.
 Also Figure~\ref{fig:plot_appendix_opt_episolon_thresh} in the Appendix shows that the optimum is within the given boundaries.
@@
     Box plots that compares the performance (power) values on the effect set \emph{paper} and \emph{paper\_rev}.
-    \emph{paper\_rev} is an additonal scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
+    \emph{paper\_rev} is an additional scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
   In contrast to \emph{paper}, for \emph{paper\_rev} the effect sizes are smaller for the \emph{early} stage.

### 2021-08-28 response_to_the_reviewers2.md
jr+tf

---
title: Response to the Reviewers
author: Jakob Richter
header-includes: |
  \usepackage{framed}
  \newlength{\leftbarwidth}
  \setlength{\leftbarwidth}{3pt}
  \newlength{\leftbarsep}
  \setlength{\leftbarsep}{-10pt}
  \newcommand*{\leftbarcolorcmd}{\color{leftbarcolor}}
  \colorlet{leftbarcolor}{gray}
  \renewenvironment{leftbar}{%
      \def\FrameCommand{{\leftbarcolorcmd{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
      \MakeFramed {\advance \hsize -\width \FrameRestore }%
    \noindent
  }{%
      \endMakeFramed
  }
  \renewenvironment{quote}
  {\vspace{5pt} \leftbar\list{}{\rightmargin\leftmargin}%
  \item\vspace{-5pt}\relax}
    {\endlist\endleftbar}
---

# Second response to the reviews for "Improving Adaptive Seamless Designs through Bayesian optimization"

We would like to thank all reviewers again for their valuable and constructive comments on the paper.
We carried out a new simulation to address the comments by Reviewer 2 and included the results in the appendix.

## Reviewer: 1

### Comments to the Author

Thanks for addressing/responding to all my comments

## Reviewer: 2

### Comments to the Author
Thanks authors for the response to the comments. However, some of them are not addressed appropriately here. Please consider them carefully in the next round revision.
For page 7 table 1 effect size scenario, although it is common that early come is overoptimistic and higher, however, it is also often in practice that there are delayed treatment effect (e.g. in the immunotherapy clinical trial setting) where stage 1 effect are smaller than stage 2. I would strongly recommend authors to add difference simulation setting.

> We carried out new simulations to address this comment. 
> In the desired scenario, titled `paper_rev`, effect sizes for early and final are flipped compared to the standard scenario `paper`, so that the lower values occur in the early stage.
> The results are shown in the Appendix.

Similarly for the selection of “espilon” and “threshold” in page 8 table 2, there should be more practical evaluation where they are commonly based on the clinical meaningful cut, besides as optimization parameters. Suggest to add relevant discussion here to facilitate more practical usage of proposed methods.

> In this paper we consider an optimization problem maximizing the power given a total sample size.
> If, for example, the task was to select all treatments with clinically relevant effect sizes, then clearly the selection parameter needs to be informed by clinical reasoning.
> We added a respective statement to the discussion.

Page 8 table 3, authors should add more details or rationale on the how parameters range of search space is determined.

> We included a statement that “epsilon” and “threshold” are obtained through preliminary studies.
> Furthermore, we added a Figure S4 in the Appendix, which demonstrates that the ranges are chosen in a sensible way.

### Minor comments:
There is citation format typo in page 3: “as implemented in the asd package (?)”

> This occurred only in the diff-version, which will not be published.

### 2021-08-28 main.tex
jr+tf

@@
 The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
-In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
+In the first stage, five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
 The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.
@@
 For the effect set \emph{paper2} the effect sizes for the second stage are doubled.
-As further case we consider a \emph{linear} effect set with effect sizes linear increasing from 20\% to 80\% for the first stage and effect sizes divided by 4 for the second stage as well as a \emph{sigmoid} relationship, again with 4 times larger effect sizes for the first stage.
+As a further case, we consider a \emph{linear} effect set with effect sizes linear increasing from 20\% to 80\% for the first stage and effect sizes divided by 4 for the second stage as well as a \emph{sigmoid} relationship, again with 4 times larger effect sizes for the first stage.
 These sets reflect different realistic situations.
-A further effect set, where the numbers between the early and the final stage are flipped for \emph{paper} is presented in the Appendix.
+Sometimes, there are delayed treatment effects (e.g.\ in the immunotherapy clinical trial setting) where the early effects are smaller than the delayed final effects.
+To include this scenario, we added a further effect set, where the numbers between the early and the final stage of the scenario \emph{paper} are flipped (see Appendix).
 
@@
 Choices for the boundaries of $\epsilon$ and $\tau$ are obtained from preliminary studies and based on the fact that these are thresholds to the absolute amount of normally distributed random variables.
-Also Figure~\ref{fig:plot_appendix_opt_episolon_thresh} in the Appendix shows that the optimum is within the given boundaries.
+Also, Figure~\ref{fig:plot_appendix_opt_episolon_thresh} in the Appendix shows that the optimum is within the given boundaries.
 The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
@@
 In practical applications, however, these values might sometimes be based on clinical rather than statistical considerations or the range of these values might be restricted by such considerations.
+Here we consider an optimization problem maximizing the power given a total sample size.
+If, for example, the task was to select all treatments with clinically relevant effect sizes, then clearly the selection parameter needs to be informed by clinical reasoning.
 In general, the application of BO to clinical trial design selection in general seems promising.
-Additional restrictions on the search space decrease the its size and therefore promise even more reliable results.
+Additional restrictions on the search space decrease its size and therefore promise even more reliable results.
 The same concept could be applied for subgroup selection in adaptive enrichment designs, which make use of interim data in selecting the target population for the remainder of the trial~\citep{burnett_adaptive_2020}.
@@
 \caption{%
-    Box plots that compares the performance (power) values on the effect set \emph{paper} and \emph{paper\_rev}.
+    Box plots that compare the performance (power) values on the effect set \emph{paper} and \emph{paper\_rev}.
     \emph{paper\_rev} is an additional scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
